---
params:
  ld_block: null
  ld_blocks_file: "../../../pipeline_steps/temp_all_ld_blocks.txt"
  results_version: "1.0.0"
  block_list_name: "1_0_0"
output:
  html_document:
    self_contained: true
title: "Clustering mode comparison across LD blocks"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", warning = F)

library(igraph)
library(ggplot2)
library(dplyr)
library(tidyr)
library(vroom)

source("../../../pipeline_steps/constants.R")
source("clustering_comparison_helpers.R")

# LD block as ancestry/chr/start-stop (same string as pipeline / ld_block_data_dir layout)
ld_blocks_file <- params$ld_blocks_file
if (!file.exists(ld_blocks_file)) {
  stop("Missing LD block list: ", ld_blocks_file)
}
comparison_ld_blocks <- readLines(ld_blocks_file, warn = FALSE)
comparison_ld_blocks <- comparison_ld_blocks[nzchar(comparison_ld_blocks)]

ld_block <- params$ld_block
if (is.null(ld_block) || length(ld_block) != 1L || !nzchar(ld_block)) {
  ld_block <- comparison_ld_blocks[1L]
}
if (!(ld_block %in% comparison_ld_blocks)) {
  stop("params$ld_block is not present in ", ld_blocks_file)
}

block_data_dir <- file.path(ld_block_data_dir, ld_block)
report_results_dir <- file.path(results_dir, params$results_version)

h4_threshold <- posterior_prob_h4_threshold
subgraph_density_theshold <- 0.6

finemapped_file <- file.path(block_data_dir, "finemapped_studies.tsv")
coloc_file <- file.path(
  block_data_dir,
  glue::glue("coloc_pairwise_results_5e-6_no_bfdr_{params$block_list_name}.tsv.gz")
)
cluster_obj <- file.path(
  block_data_dir,
  glue::glue("igraph_clustered_results_5e-6_no_bfdr_{params$block_list_name}_mode_2.rds")
)
problemsnp_file <- file.path(block_data_dir, "problematic_finemapped_snps.tsv")

studies_file <- file.path(report_results_dir, "studies_processed.tsv.gz")
studies_db_file <- file.path(report_results_dir, "studies.db")
truthset_file <- file.path(report_results_dir, "analysis", "clustering", "tipairs_launched_truthset.tsv")
if (!file.exists(truthset_file)) {
  fallback_truthset <- "../data/tipairs_launched_truthset.tsv"
  if (file.exists(fallback_truthset)) {
    message("Using fallback truthset: ", fallback_truthset)
    truthset_file <- fallback_truthset
  }
}

mode_config <- data.frame(
  mode = 1:2,
  mode_label = c(
    "Mode 1: original",
    "Mode 2: original + calibrated isolated dyads"
  ),
  stringsAsFactors = FALSE
)

comparison_inputs <- tidyr::crossing(
  ld_block = comparison_ld_blocks,
  mode = mode_config$mode
) |>
  dplyr::left_join(mode_config, by = "mode") |>
  dplyr::mutate(
    block_data_dir = file.path(ld_block_data_dir, ld_block),
    cluster_file = file.path(
      block_data_dir,
      glue::glue("clustered_results_5e-6_no_bfdr_{params$block_list_name}_mode_{mode}.tsv.gz")
    ),
    rds_file = file.path(
      block_data_dir,
      glue::glue("igraph_clustered_results_5e-6_no_bfdr_{params$block_list_name}_mode_{mode}.rds")
    ),
    cluster_exists = file.exists(cluster_file),
    rds_exists = file.exists(rds_file)
  )

complete_blocks <- comparison_inputs |>
  dplyr::group_by(ld_block) |>
  dplyr::summarise(complete = all(cluster_exists & rds_exists), .groups = "drop") |>
  dplyr::filter(complete) |>
  dplyr::pull(ld_block)

message("Comparison LD blocks: ", length(comparison_ld_blocks))
message("Blocks with complete outputs for both modes: ", length(complete_blocks))
message("Detailed LD block: ", ld_block)
message("Block data dir: ", block_data_dir)

if (length(complete_blocks) == 0L) {
  stop("No LD block has complete clustered TSV and igraph RDS outputs for both modes.")
}

analysis_inputs <- comparison_inputs |>
  dplyr::filter(ld_block %in% complete_blocks) |>
  dplyr::arrange(match(ld_block, comparison_ld_blocks), mode)
```

## Clustering mode evaluation against the truth set

Modes are compared on `r length(complete_blocks)` LD blocks with complete outputs for both modes. The primary question is whether a clustering mode improves on **production GPMap cluster support** (`gpmap_support`) for truth-set target–indication pairs, without introducing many spurious same-cluster molecular–phenotype assignments.

**Gap targets (primary):** truth-set `ti_uid`s with `pairwisecoloc_support = TRUE` but `gpmap_support = FALSE` in the production truth set, and with at least one molecular and one phenotype study finemapped in the same analyzed LD block.

**Cluster support (test outcome):** the mode assigns at least one eligible molecular/phenotype extraction pair to the same final cluster component.

**Non-truth-set same-cluster pairs (false-positive burden):** H4-qualified molecular–phenotype coloc pairs placed in the same cluster by the mode but not matching any truth-set extraction pair in that block.

```{r prepare_truthset_comparison}
if (!file.exists(studies_file)) stop("Missing studies file: ", studies_file)
if (!file.exists(studies_db_file)) stop("Missing studies database: ", studies_db_file)
if (!file.exists(truthset_file)) stop("Missing truth-set file: ", truthset_file)
if (!requireNamespace("duckdb", quietly = TRUE)) {
  stop("The duckdb package is required to load gene pleiotropy scores.")
}

studies_processed_comparison <- vroom::vroom(studies_file, show_col_types = FALSE)
truthset_comparison <- vroom::vroom(truthset_file, show_col_types = FALSE)
truthset_ti_meta <- build_truthset_ti_meta(truthset_comparison)
study_types <- studies_processed_comparison |>
  dplyr::select(study_name, data_type) |>
  dplyr::distinct()

gene_study_lookup_comparison <- studies_processed_comparison |>
  dplyr::filter(data_type != "phenotype", !is.na(gene)) |>
  dplyr::select(study_name, gene, data_type) |>
  dplyr::distinct()

finemapped_by_block <- stats::setNames(
  lapply(complete_blocks, function(block) {
    path <- file.path(ld_block_data_dir, block, "finemapped_studies.tsv")
    if (!file.exists(path)) stop("Missing finemapped studies file: ", path)
    vroom::vroom(path, col_types = finemapped_column_types, show_col_types = FALSE) |>
      dplyr::filter(variant_type == "common") |>
      dplyr::select(study, unique_study_id, min_p) |>
      dplyr::distinct()
  }),
  complete_blocks
)

coloc_files_by_block <- stats::setNames(
  file.path(
    ld_block_data_dir,
    complete_blocks,
    glue::glue("coloc_pairwise_results_5e-6_no_bfdr_{params$block_list_name}.tsv.gz")
  ),
  complete_blocks
)

truthset_candidates <- truthset_comparison |>
  dplyr::filter(dplyr::coalesce(as.logical(pairwisecoloc_support), FALSE)) |>
  dplyr::inner_join(
    gene_study_lookup_comparison,
    by = c("target" = "gene"),
    relationship = "many-to-many"
  )

truthset_uid_mappings <- dplyr::bind_rows(lapply(complete_blocks, function(block) {
  finemapped <- finemapped_by_block[[block]]
  truthset_candidates |>
    dplyr::inner_join(
      finemapped |>
        dplyr::select(study, unique_study_id) |>
        dplyr::rename(
          study_name = study,
          molecular_unique_study_id = unique_study_id
        ),
      by = "study_name",
      relationship = "many-to-many"
    ) |>
    dplyr::inner_join(
      finemapped |>
        dplyr::select(study, unique_study_id) |>
        dplyr::rename(
          trait_gpmap = study,
          phenotype_unique_study_id = unique_study_id
        ),
      by = "trait_gpmap",
      relationship = "many-to-many"
    ) |>
    dplyr::mutate(ld_block = block, .before = 1L) |>
    dplyr::distinct(
      ld_block, ti_uid, target, indication_mesh_term, combined_max_phase,
      trait_gpmap, trait_name_gpmap, study_name, data_type,
      molecular_unique_study_id, phenotype_unique_study_id
    )
}))

if (nrow(truthset_uid_mappings) == 0L) {
  stop("No pairwise-supported truth-set indications map to finemapped studies in the complete LD blocks.")
}

evaluable_indications <- truthset_uid_mappings |>
  dplyr::distinct(
    ti_uid, target, indication_mesh_term, combined_max_phase
  ) |>
  dplyr::inner_join(truthset_ti_meta, by = "ti_uid")

gap_evaluable <- evaluable_indications |>
  dplyr::filter(gap_target)

studies_conn <- DBI::dbConnect(
  duckdb::duckdb(),
  studies_db_file,
  read_only = TRUE
)
gene_pleiotropy <- DBI::dbGetQuery(
  studies_conn,
  "
    SELECT
      ga.id AS gene_id,
      ga.gene AS target,
      gp.distinct_trait_categories
    FROM gene_annotations AS ga
    LEFT JOIN gene_pleiotropy AS gp
      ON ga.id = gp.gene_id
  "
)
DBI::dbDisconnect(studies_conn, shutdown = TRUE)

gene_pleiotropy <- gene_pleiotropy |>
  dplyr::group_by(target) |>
  dplyr::summarise(
    gene_annotation_found = TRUE,
    pleiotropy_score_found = any(!is.na(distinct_trait_categories)),
    distinct_trait_categories = if (all(is.na(distinct_trait_categories))) {
      0L
    } else {
      max(distinct_trait_categories, na.rm = TRUE)
    },
    .groups = "drop"
  )

evaluable_indications <- evaluable_indications |>
  dplyr::left_join(gene_pleiotropy, by = "target") |>
  dplyr::mutate(
    gene_annotation_found = dplyr::coalesce(gene_annotation_found, FALSE),
    pleiotropy_score_found = dplyr::coalesce(pleiotropy_score_found, FALSE),
    pleiotropy_bin = dplyr::case_when(
      !gene_annotation_found ~ "Unmapped",
      distinct_trait_categories <= 1 ~ "0-1",
      distinct_trait_categories <= 5 ~ "2-5",
      distinct_trait_categories <= 15 ~ "6-15",
      distinct_trait_categories > 15 ~ ">15",
      TRUE ~ "Unmapped"
    ),
    pleiotropy_bin = factor(
      pleiotropy_bin,
      levels = c("0-1", "2-5", "6-15", ">15", "Unmapped")
    )
  )

truthset_diagnostics <- data.frame(
  check = c(
    "Complete LD blocks analyzed",
    "Pairwise-supported ti_uids in truth set",
    "Evaluable ti_uids (finemapped in analyzed blocks)",
    "Gap targets evaluable (pairwise yes, production gpmap no)",
    "Evaluable gap targets / all gap targets in truth set",
    "Production GPMap-supported ti_uids among evaluable set"
  ),
  n = c(
    length(complete_blocks),
    sum(truthset_ti_meta$pairwisecoloc_support),
    nrow(evaluable_indications),
    nrow(gap_evaluable),
    NA_real_,
    sum(evaluable_indications$production_gpmap_support)
  ),
  stringsAsFactors = FALSE
)
truthset_diagnostics$n[5] <- paste0(
  nrow(gap_evaluable), " / ", sum(truthset_ti_meta$gap_target)
)

knitr::kable(
  truthset_diagnostics,
  caption = "Truth-set evaluation scope"
)
```

```{r evaluate_clustering_modes}
recovery_by_block <- dplyr::bind_rows(lapply(
  split(analysis_inputs, seq_len(nrow(analysis_inputs))),
  evaluate_mode_ti_recovery,
  truthset_uid_mappings = truthset_uid_mappings
))

coloc_burden_by_block <- dplyr::bind_rows(lapply(
  split(analysis_inputs, seq_len(nrow(analysis_inputs))),
  function(input_row) {
    evaluate_mode_coloc_burden(
      input_row = input_row,
      coloc_file = coloc_files_by_block[[input_row$ld_block]],
      h4_threshold = h4_threshold,
      study_types = study_types,
      truthset_uid_mappings = truthset_uid_mappings
    )
  }
))

mode_support_pooled <- recovery_by_block |>
  dplyr::group_by(mode, mode_label, ti_uid) |>
  dplyr::summarise(cluster_support = any(cluster_support), .groups = "drop") |>
  dplyr::inner_join(evaluable_indications, by = "ti_uid") |>
  dplyr::mutate(
    new_gpmap_support = cluster_support & !production_gpmap_support,
    maintained_gpmap_support = cluster_support & production_gpmap_support,
    lost_gpmap_support = !cluster_support & production_gpmap_support,
    still_missing_gpmap_support = !cluster_support & !production_gpmap_support
  )

gpmap_gap_summary <- mode_support_pooled |>
  dplyr::filter(gap_target) |>
  dplyr::group_by(mode, mode_label) |>
  dplyr::summarise(
    n_gap_evaluable = dplyr::n(),
    n_new_gpmap_support = sum(new_gpmap_support),
    n_still_missing = sum(still_missing_gpmap_support),
    gap_recovery_rate = n_new_gpmap_support / n_gap_evaluable,
    .groups = "drop"
  )

gpmap_concordance_summary <- mode_support_pooled |>
  dplyr::group_by(mode, mode_label) |>
  dplyr::summarise(
    n_evaluable = dplyr::n(),
    n_production_gpmap = sum(production_gpmap_support),
    n_mode_cluster_support = sum(cluster_support),
    n_agree = sum(cluster_support == production_gpmap_support),
    n_new_gpmap_support = sum(new_gpmap_support),
    n_lost_gpmap_support = sum(lost_gpmap_support),
    concordance_rate = n_agree / n_evaluable,
    .groups = "drop"
  )

false_positive_summary <- coloc_burden_by_block |>
  dplyr::group_by(mode, mode_label) |>
  dplyr::summarise(
    n_h4_mol_pheno_pairs = sum(n_h4_mol_pheno_pairs),
    n_same_cluster_h4_mol_pheno = sum(n_same_cluster_h4_mol_pheno),
    n_truthset_supported_same_cluster = sum(n_truthset_supported_same_cluster),
    n_non_truthset_same_cluster = sum(n_non_truthset_same_cluster),
    non_truthset_same_cluster_rate = dplyr::if_else(
      n_same_cluster_h4_mol_pheno > 0L,
      n_non_truthset_same_cluster / n_same_cluster_h4_mol_pheno,
      NA_real_
    ),
    n_pipeline_false_positive = sum(n_pipeline_false_positive),
    n_pipeline_false_negative = sum(n_pipeline_false_negative),
    .groups = "drop"
  )

knitr::kable(
  gpmap_gap_summary,
  digits = 4,
  caption = "Gap-target recovery: achieving cluster support where production GPMap support is absent"
)

knitr::kable(
  gpmap_concordance_summary,
  digits = 4,
  caption = "Concordance with production GPMap support labels across all evaluable pairwise-supported ti_uids"
)

knitr::kable(
  false_positive_summary,
  digits = 4,
  caption = "False-positive burden: same-cluster H4 molecular–phenotype pairs not in the truth set"
)
```

```{r gpmap_mode_gains_and_losses}
gap_support_wide <- mode_support_pooled |>
  dplyr::filter(gap_target) |>
  dplyr::select(ti_uid, mode, cluster_support) |>
  tidyr::pivot_wider(
    names_from = mode,
    values_from = cluster_support,
    names_prefix = "mode_",
    values_fill = FALSE
  )

mode_pairs <- utils::combn(mode_config$mode, 2L, simplify = FALSE)
gap_mode_gain_loss <- dplyr::bind_rows(lapply(mode_pairs, function(mode_pair) {
  mode_a <- paste0("mode_", mode_pair[1])
  mode_b <- paste0("mode_", mode_pair[2])
  supported_a <- gap_support_wide[[mode_a]]
  supported_b <- gap_support_wide[[mode_b]]
  data.frame(
    mode_a = mode_pair[1],
    mode_b = mode_pair[2],
    supported_by_both = sum(supported_a & supported_b),
    supported_only_by_a = sum(supported_a & !supported_b),
    supported_only_by_b = sum(!supported_a & supported_b),
    supported_by_neither = sum(!supported_a & !supported_b)
  )
}))

changed_gap_indications <- gap_support_wide |>
  dplyr::filter(
    rowSums(dplyr::across(dplyr::starts_with("mode_"))) > 0L,
    rowSums(dplyr::across(dplyr::starts_with("mode_"))) < nrow(mode_config)
  ) |>
  dplyr::left_join(evaluable_indications, by = "ti_uid") |>
  dplyr::arrange(pleiotropy_bin, target, ti_uid)

knitr::kable(
  gap_mode_gain_loss,
  caption = "Pairwise gains and losses on gap targets (pairwise yes, production gpmap no)"
)

if (nrow(changed_gap_indications) > 0L) {
  knitr::kable(
    changed_gap_indications,
    caption = "Gap-target ti_uids whose cluster support differs between clustering modes"
  )
} else {
  message("No gap-target ti_uids differ in cluster support between clustering modes.")
}
```

```{r gap_recovery_by_pleiotropy, fig.width=9, fig.height=5}
pleiotropy_gap_summary <- mode_support_pooled |>
  dplyr::filter(gap_target) |>
  dplyr::group_by(mode, mode_label, pleiotropy_bin, .drop = FALSE) |>
  dplyr::summarise(
    n_gap_evaluable = dplyr::n(),
    n_new_gpmap_support = sum(new_gpmap_support),
    gap_recovery_rate = if (n_gap_evaluable > 0L) {
      n_new_gpmap_support / n_gap_evaluable
    } else {
      NA_real_
    },
    .groups = "drop"
  )

knitr::kable(
  pleiotropy_gap_summary,
  digits = 4,
  caption = "Gap-target recovery by target-gene pleiotropy profile"
)

ggplot(
  dplyr::filter(pleiotropy_gap_summary, n_gap_evaluable > 0L),
  aes(x = pleiotropy_bin, y = gap_recovery_rate, fill = mode_label)
) +
  geom_col(position = position_dodge(width = 0.8), width = 0.72) +
  geom_text(
    aes(label = paste0(n_new_gpmap_support, "/", n_gap_evaluable)),
    position = position_dodge(width = 0.8),
    vjust = -0.25,
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x = "Target-gene distinct trait-category score",
    y = "Gap targets achieving cluster support",
    fill = "Clustering mode"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Graph structure by clustering mode

Both modes use the original p-value-derived edge weights and clustering procedure. Mode 2 additionally admits isolated two-study H4 components that pass a p-value-dependent H4 requirement. All graph metrics are therefore on the same weight scale; differences isolate the effect of calibrated dyad rescue.

```{r clustering_mode_graph_metrics}
graph_metrics <- dplyr::bind_rows(lapply(
  split(analysis_inputs, seq_len(nrow(analysis_inputs))),
  function(input_row) {
    clustered_result <- readRDS(input_row$rds_file)
    dplyr::bind_cols(
      input_row |>
        dplyr::select(ld_block, mode, mode_label),
      summarise_clustered_graphs(clustered_result)
    )
  }
))

graph_metrics_summary <- graph_metrics |>
  dplyr::group_by(mode, mode_label) |>
  dplyr::summarise(
    n_blocks = dplyr::n_distinct(ld_block),
    n_components_unpruned = sum(n_components_unpruned),
    n_components_pruned = sum(n_components_pruned),
    mean_weight_unpruned = mean(mean_weight_unpruned, na.rm = TRUE),
    mean_weight_pruned = mean(mean_weight_pruned, na.rm = TRUE),
    mean_transitivity_unpruned = mean(mean_transitivity_unpruned, na.rm = TRUE),
    mean_transitivity_pruned = mean(mean_transitivity_pruned, na.rm = TRUE),
    mean_conductance_pruned = mean(mean_conductance_pruned, na.rm = TRUE),
    median_size_unpruned = stats::median(median_size_unpruned, na.rm = TRUE),
    median_size_pruned = stats::median(median_size_pruned, na.rm = TRUE),
    .groups = "drop"
  )

mode_1_metrics <- graph_metrics |>
  dplyr::filter(mode == 1L) |>
  dplyr::select(
    ld_block,
    baseline_components_pruned = n_components_pruned,
    baseline_transitivity_pruned = mean_transitivity_pruned,
    baseline_conductance_pruned = mean_conductance_pruned,
    baseline_median_size_pruned = median_size_pruned
  )

graph_metric_deltas <- graph_metrics |>
  dplyr::left_join(mode_1_metrics, by = "ld_block") |>
  dplyr::transmute(
    ld_block,
    mode,
    mode_label,
    delta_components_pruned = n_components_pruned - baseline_components_pruned,
    delta_transitivity_pruned = mean_transitivity_pruned - baseline_transitivity_pruned,
    delta_conductance_pruned = mean_conductance_pruned - baseline_conductance_pruned,
    delta_median_size_pruned = median_size_pruned - baseline_median_size_pruned
  )

knitr::kable(
  graph_metrics,
  digits = 4,
  caption = "Graph structure metrics by LD block and clustering mode"
)

knitr::kable(
  graph_metrics_summary,
  digits = 4,
  caption = "Graph structure summary across complete LD blocks"
)

knitr::kable(
  graph_metric_deltas,
  digits = 4,
  caption = "Per-block changes in pruned graph metrics relative to mode 1"
)
```

## Detailed mode 2 exploration

The remainder of the report retains the original single-block diagnostic plots for `r ld_block`, using clustering mode 2.

```{r read_in_data}
required_detail_files <- c(finemapped_file, coloc_file, cluster_obj, studies_file)
missing_detail_files <- required_detail_files[!file.exists(required_detail_files)]
if (length(missing_detail_files) > 0L) {
  stop("Missing detailed-analysis input(s): ", paste(missing_detail_files, collapse = ", "))
}

coloc_results <- vroom::vroom(coloc_file, delim = '\t', show_col_types = F)
coloc_results_filt <- coloc_results |> dplyr::filter(ignore == FALSE, h4 >= h4_threshold)

finemapped_studies <- vroom::vroom(finemapped_file, col_types = finemapped_column_types, show_col_types = F) |>
  dplyr::filter(variant_type == "common") |>
  dplyr::arrange(unique_study_id)

clustered_results <- readRDS(cluster_obj)
clusters <- clustered_results$groups

studies_processed <- vroom::vroom(studies_file, show_col_types = F)

if (file.exists(problemsnp_file)) {
  problem_snps <- vroom::vroom(problemsnp_file, show_col_types = F)
} else {
  problem_snps <- data.frame()
}
```


```{r load_truthset}

truthset <- vroom::vroom(truthset_file, show_col_types = F)

# Build lookup: for non-phenotype studies, gene -> study_name
gene_study_lookup <- studies_processed |>
  dplyr::filter(data_type != "phenotype" & !is.na(gene)) |>
  dplyr::select(study_name, gene, data_type)

# Match truthset rows to finemapped studies in this block:
# - target (gene) matches gene_study_lookup$gene -> gives study_name for the molecular study
# - trait_gpmap matches finemapped_studies$study for the phenotype study
block_studies <- unique(finemapped_studies$study)

truthset_edges <- truthset |>
  dplyr::filter(pairwisecoloc_support == TRUE &
                  (gpmap_support == FALSE | preinfomap_support == FALSE)) |>
  dplyr::inner_join(gene_study_lookup, by = c("target" = "gene"), relationship = "many-to-many") |>
  dplyr::filter(study_name %in% block_studies & trait_gpmap %in% block_studies) |>
  dplyr::distinct(study_name, trait_gpmap, .keep_all = TRUE)

# Build set of unique_study_id pairs for edge highlighting
# A single study can have multiple unique_study_ids (credible sets) in this block
truthset_unique_pairs <- truthset_edges |>
  dplyr::inner_join(
    finemapped_studies |> dplyr::select(study, unique_study_id) |> dplyr::rename(uid_mol = unique_study_id),
    by = c("study_name" = "study"), relationship = "many-to-many"
  ) |>
  dplyr::inner_join(
    finemapped_studies |> dplyr::select(study, unique_study_id) |> dplyr::rename(uid_pheno = unique_study_id),
    by = c("trait_gpmap" = "study"), relationship = "many-to-many"
  )

message("Truthset edges to highlight in this block: ", nrow(truthset_edges))
message("Truthset unique_study_id pairs: ", nrow(truthset_unique_pairs))
```

### Coloc results:

```{r summarise_coloc_hits}

h4_cuts <- seq(0,1, by = 0.1)
coloc_results$h4_cuts <- cut(coloc_results$h4, breaks = h4_cuts)

message("No. studies in LD block: ", nrow(finemapped_studies))
message("No. Pairwise colocs in LD block: ", nrow(coloc_results))
message("Colocs at H4 range:")

tab <- table(coloc_results$h4_cuts)
names(tab) <- sub(",","-",sub("\\(","",sub("]","",names(tab))))
tab

clusters$study <- sub("_EUR.*","",clusters$unique_study_id)

check_clusters <- clusters |> dplyr::group_by(component, study) |> dplyr::summarise(n_regions = dplyr::n())
```

### Truthset pairs in this LD block

```{r truthset_table, results='asis'}

if (nrow(truthset_unique_pairs) > 0) {
  pair_info <- truthset_unique_pairs |>
    dplyr::left_join(
      coloc_results |> dplyr::select(unique_study_a, unique_study_b, h4, PP.H3.abf, PP.H4.abf, ignore),
      by = c("uid_mol" = "unique_study_a", "uid_pheno" = "unique_study_b")
    )

  # Also check reversed direction
  pair_info_rev <- truthset_unique_pairs |>
    dplyr::left_join(
      coloc_results |> dplyr::select(unique_study_a, unique_study_b, h4, PP.H3.abf, PP.H4.abf, ignore),
      by = c("uid_mol" = "unique_study_b", "uid_pheno" = "unique_study_a")
    )

  pair_info <- dplyr::bind_rows(pair_info, pair_info_rev) |>
    dplyr::filter(!is.na(h4)) |>
    dplyr::distinct()

  # Add min_p for each study
  min_p_lookup <- finemapped_studies |> dplyr::select(unique_study_id, min_p)

  pair_info <- pair_info |>
    dplyr::left_join(min_p_lookup |> dplyr::rename(min_p_mol = min_p), by = c("uid_mol" = "unique_study_id")) |>
    dplyr::left_join(min_p_lookup |> dplyr::rename(min_p_pheno = min_p), by = c("uid_pheno" = "unique_study_id")) |>
    dplyr::select(
      target, trait_gpmap, trait_name_gpmap, data_type,
      gpmap_support, preinfomap_support,
      uid_mol, uid_pheno,
      h4, PP.H3.abf, PP.H4.abf,
      min_p_mol, min_p_pheno
    ) |>
    dplyr::arrange(dplyr::desc(h4))

  knitr::kable(pair_info, digits = 4, caption = "Truthset pairs present in this LD block (pairwisecoloc_support=T & missing gpmap/preinfomap support)")
} else {
  message("No truthset pairs found in this LD block")
}
```

### Clustering studies:

#### Original graph:

```{r plot_clustering, fig.width = 6, fig.height = 6}

studies <- sub("_EUR.*", "", V(clustered_results$unpruned_graph)$name)

g <- clustered_results$unpruned_graph
g_comps <- igraph::components(g)

message("Number of graph verts: ", length(igraph::V(g)))
message("Number of graph edges: ", length(igraph::E(g)))

g_layout <- igraph::layout_with_kk(g)

plot(g,
     vertex.label = NA, 
     vertex.size = 3,
     vertex.color = "black",
     edge.width = 2,
     layout = g_layout,
     main = "Original graph")
```

#### Original subgraphs:

```{r subgraphs}

csize <- table(g_comps$csize) |> as.data.frame()
colnames(csize) <- c("Component size", "Freq")

message("Original cluster size and frequency:")
csize

comp_num <- as.numeric(which(table(g_comps$membership) > 1))

g_subgraphs <- lapply(comp_num, function(x){
  igraph::induced_subgraph(g, vids = igraph::V(g)[g_comps$membership == x])
})

g_density <- sapply(g_subgraphs, function(x){igraph::edge_density(x)})

# Build set of truthset edge pairs for highlighting
truthset_highlight_pairs <- NULL
if (nrow(truthset_unique_pairs) > 0) {
  truthset_highlight_pairs <- rbind(
    data.frame(V1 = truthset_unique_pairs$uid_mol, V2 = truthset_unique_pairs$uid_pheno, stringsAsFactors = FALSE),
    data.frame(V1 = truthset_unique_pairs$uid_pheno, V2 = truthset_unique_pairs$uid_mol, stringsAsFactors = FALSE)
  )
}

plot_subgraphs <- function(subgraph, colour_verts, highlight_pairs = NULL){
  g_density <- igraph::edge_density(subgraph)
  g_layout <- igraph::layout_with_kk(subgraph)
  
  if(colour_verts == FALSE){
    col_vert = "black"
    col_edge = "grey"
    size_vert = 3
  }else{
    col_vert = V(subgraph)$community
    communities <- unique(col_vert[col_vert > 0])
    v_colours <- c("white", rainbow(length(communities)))
    names(v_colours) <- c(0, communities)
    col_vert = as.vector(v_colours[as.character(col_vert)])
    
    col_edge = E(subgraph)$pruned
    col_edge = ifelse(col_edge == 0, "grey", "red")
    size_vert = 6
  }

  # Highlight truthset edges in gold
  edge_width <- rep(2, length(E(subgraph)))
  if (!is.null(highlight_pairs) && nrow(highlight_pairs) > 0) {
    el <- igraph::as_edgelist(subgraph)
    for (k in seq_len(nrow(el))) {
      v1 <- el[k, 1]
      v2 <- el[k, 2]
      match_fwd <- any(highlight_pairs$V1 == v1 & highlight_pairs$V2 == v2)
      match_rev <- any(highlight_pairs$V1 == v2 & highlight_pairs$V2 == v1)
      if (match_fwd || match_rev) {
        col_edge[k] <- "gold"
        edge_width[k] <- 4
      }
    }
  }
  
  plot(subgraph,
       vertex.label = NA,
       vertex.size = size_vert,
       edge.width = edge_width,
       vertex.color = col_vert,
       edge.color = col_edge,
       layout = g_layout)
}

plot_nrows <- ceiling(length(g_subgraphs)/10)
fig_height <- plot_nrows*2
```

### Clustering (with infomap):

```{r infomap_clustering, fig.width = 12, fig.height = 6}

par(mfrow = c(1,1))

clustered_graph <- clustered_results$pruned_graph
gp_comps <- igraph::components(clustered_graph)

comp_num_pruned <- as.numeric(which(table(gp_comps$membership) > 1))

gp_subgraphs <- lapply(comp_num_pruned, function(x){
  igraph::induced_subgraph(clustered_graph, vids = igraph::V(clustered_graph)[gp_comps$membership == x])
})

V(g)$community <- V(clustered_graph)$component[match(V(g)$name, V(clustered_graph)$name)]
V(g)$community[is.na(V(g)$community)] <- 0

pruned_edges <- clustered_results$pruned_edges
pruned_edge_ids <- get.edge.ids(g, t(as.matrix(pruned_edges)))

E(g)$pruned <- 0
E(g)$pruned[pruned_edge_ids] <- 1

plot(g,
     vertex.label = NA, 
     vertex.size = 3,
     vertex.color = V(g)$community,
     edge.color = "grey",
     edge.width = 2,
     layout = g_layout, main = "Original graph (coloured by identified infomap communities)")

plot_nrows_set <- max(1, ceiling(sum(g_density < subgraph_density_theshold) / 5))
fig_height_set <- max(4, plot_nrows_set * 4)
```

```{r plot_subgraphs_coloured, fig.width = 20, fig.height = fig_height_set}

lowdensity_subgraphs <- g_subgraphs[which(g_density < subgraph_density_theshold)]
n_plots <- length(lowdensity_subgraphs)

if (n_plots == 0) {
  message("No subgraphs below density threshold ", subgraph_density_theshold)
} else {
  plot_ncols <- 5
  plot_nrows_set <- max(1, ceiling(n_plots / plot_ncols))
  par(
    mfrow = c(plot_nrows_set, plot_ncols),
    mar = c(0.2, 0.2, 0.5, 0.2),
    oma = c(0, 0, 0, 0)
  )
  invisible(sapply(
    lowdensity_subgraphs,
    plot_subgraphs,
    colour_verts = TRUE,
    highlight_pairs = truthset_highlight_pairs
  ))
}
```

```{r select_split_subgraphs}

split_subgraph_info <- lapply(seq_along(lowdensity_subgraphs), function(i) {
  sg <- lowdensity_subgraphs[[i]]
  n_nodes <- length(V(sg))
  communities <- V(sg)$community
  communities <- communities[communities > 0]
  n_communities <- length(unique(communities))
  data.frame(
    index = i,
    n_nodes = n_nodes,
    n_communities = n_communities,
    density = igraph::edge_density(sg)
  )
}) |> dplyr::bind_rows()

candidates <- split_subgraph_info |>
  dplyr::filter(n_communities > 1 & n_nodes < 100) |>
  dplyr::arrange(dplyr::desc(n_communities), dplyr::desc(n_nodes))

n_to_show <- min(3, nrow(candidates))

if (n_to_show == 0) {
  message("No subgraphs found with multiple communities and < 100 nodes")
} else {
  selected <- candidates[1:n_to_show, ]
  message("Selected ", n_to_show, " subgraphs for exploration:")
  print(selected)
}
```

### Region overlay

```{r region_plot_functions}

plot_region <- function(subgraph){
  
  study_ids <- V(subgraph)$name
  finemapped <- finemapped_studies |> dplyr::filter(unique_study_id %in% study_ids)
  
  region_files <- finemapped$file
  regions <- lapply(region_files, function(x) {
    stopifnot(file.exists(x))
    data.table::fread(x)
  })
  names(regions) <- finemapped$study
  regions <- mapply(function(x,y){
    x <- x |> 
      dplyr::mutate(study_id = y,
                    scaled_LBF = LBF/max(LBF)) |>
      dplyr::filter(SE != 1)
    return(x)} , x = regions, y = finemapped$unique_study_id, SIMPLIFY = F)
  
  if(!("community" %in% vertex_attr_names(subgraph))){
    V(subgraph)$community <- V(subgraph)$component
  }
    
  component <- V(subgraph)$community
  names(component) <- V(subgraph)$name
  
  regions_combined <- data.table::rbindlist(regions, fill = TRUE) |>
    dplyr::mutate(component = component[study_id]) |>
    dplyr::filter(component != 0)
  
  tag_snp <- clusters |> 
    dplyr::filter(component %in% unique(V(subgraph)$community)) |> dplyr::select(component, snp) |> 
    unique()
  
  cluster_details <- regions_combined |> 
    dplyr::group_by(component) |> 
    dplyr::summarise(max(LBF)) |> 
    as.data.frame() |> 
    dplyr::arrange(desc(`max(LBF)`)) |>
    dplyr::left_join(tag_snp) |>
    dplyr::mutate(bp = as.numeric(sub(".*:(.*)_.*_.*","\\1",snp)))

  regions_combined <- regions_combined |> 
    dplyr::mutate(component = factor(component, levels = cluster_details$component)) |>
    dplyr::arrange(component, study_id)
  
  col_vert = V(subgraph)$community
  communities <- unique(col_vert[col_vert > 0])
  v_colours <- c("white", rainbow(length(communities)))
  names(v_colours) <- c(0, communities)
  
  p1 <- ggplot(regions_combined, 
         aes(x = BP/1000000, y = scaled_LBF, colour = as.factor(component))) +
    geom_point(size = 1) +
    geom_vline(xintercept = cluster_details$bp/1000000) +
    ggrepel::geom_label_repel(data = cluster_details, aes(x = bp/1000000, y = 1, label = snp), show.legend = F, max.overlaps = 30) +
    scale_colour_manual(name = "Cluster ID", values = v_colours) +
    theme_minimal() +
    theme(legend.position = "bottom") +
    labs(x = "Position (Mb)", y = "Scaled LBF")
  
  x_range <- c(min(cluster_details$bp) - 100000, max(cluster_details$bp) + 100000)
  
  p2 <- p1 + scale_x_continuous(limits = x_range / 1000000)

  gridExtra::arrangeGrob(p1, p2, ncol = 2, widths = c(0.55, 0.45))
}
```

#### Exploration of selected split subgraphs

For each selected subgraph, the network graph is shown first, followed by the overlayed manhattan plot.

```{r explore_selected_subgraphs, fig.width = 11, fig.height = 8, results='asis'}

if (n_to_show > 0) {
  for (j in seq_len(n_to_show)) {
    idx <- selected$index[j]
    sg <- lowdensity_subgraphs[[idx]]
    n_nodes <- selected$n_nodes[j]
    n_comms <- selected$n_communities[j]

    cat("\n\n##### Subgraph ", idx, ": ", n_nodes, " nodes, ", n_comms, " communities, density ",
        round(selected$density[j], 3), "\n\n", sep = "")

    par(mfrow = c(1, 1))
    plot_subgraphs(subgraph = sg, colour_verts = TRUE, highlight_pairs = truthset_highlight_pairs)

    tryCatch({
      gridExtra::grid.arrange(plot_region(subgraph = sg))
    }, error = function(e) {
      message("Overlay plot failed for subgraph ", idx, ": ", e$message)
    })
  }
}
```
