---
params:
  ld_block: null
  ld_blocks_file: "../../../pipeline_steps/temp_all_ld_blocks.txt"
  results_version: "1.0.0"
  n_cores: 50
output:
  html_document:
    self_contained: true
title: "Coloc Clustering Validation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5
)

library(igraph)
library(dplyr)
library(vroom)
library(scales)
library(parallel)

source("../../../pipeline_steps/constants.R")
source("clustering_comparison_helpers.R")

n_cores <- max(1L, as.integer(params$n_cores))

ld_blocks_file <- params$ld_blocks_file
if (!file.exists(ld_blocks_file)) {
  stop("Missing LD block list: ", ld_blocks_file)
}

ld_block_param <- params$ld_block
use_single_ld_block <- !(is.null(ld_block_param) ||
  (length(ld_block_param) == 0L) ||
  (is.character(ld_block_param) && length(ld_block_param) == 1L &&
    (is.na(ld_block_param) || !nzchar(trimws(ld_block_param)))))
if (is.character(ld_block_param) && length(ld_block_param) > 1L) {
  stop("params$ld_block must be null or a single character string.")
}

selected_ld_blocks <- if (use_single_ld_block) {
  trimws(ld_block_param)
} else {
  blocks <- readLines(ld_blocks_file, warn = FALSE)
  blocks[nzchar(blocks)]
}

comparison_files <- data.frame(
  method = c(
    "5e-6, no BFDR",
    "5e-6 + BFDR",
    "1e-5 + BFDR"
  ),
  pairwise_file = c(
    "coloc_pairwise_results_5e-6_no_bfdr_1_0_0.tsv.gz",
    "coloc_pairwise_results_5e-6_bfdr_1_0_0.tsv.gz",
    "coloc_pairwise_results_1e-5_bfdr_1_0_0.tsv.gz"
  ),
  cluster_file = c(
    "clustered_results_5e-6_no_bfdr_1_0_0.tsv.gz",
    "clustered_results_5e-6_bfdr_1_0_0.tsv.gz",
    "clustered_results_1e-5_bfdr_1_0_0.tsv.gz"
  ),
  rds_file = c(
    "igraph_clustered_results_5e-6_no_bfdr_1_0_0.rds",
    "igraph_clustered_results_5e-6_bfdr_1_0_0.rds",
    "igraph_clustered_results_1e-5_bfdr_1_0_0.rds"
  ),
  bfdr_file = c(
    "global_no_bfdr_threshold_p12_5e-6.tsv",
    "global_bfdr_threshold_p12_5e-6.tsv",
    "global_bfdr_threshold_p12_1e-5.tsv"
  ),
  stringsAsFactors = FALSE
)

studies_file <- file.path(results_dir, params$results_version, "studies_processed.tsv.gz")
if (!file.exists(studies_file)) {
  stop("Missing studies file: ", studies_file)
}
study_types <- vroom::vroom(studies_file, show_col_types = FALSE) |>
  dplyr::select(study_name, data_type, variant_type) |>
  dplyr::distinct()

study_extractions_file <- file.path(
  results_dir, params$results_version, "study_extractions.tsv.gz"
)
if (!file.exists(study_extractions_file)) {
  stop("Missing study extractions file: ", study_extractions_file)
}
study_extractions_meta <- vroom::vroom(
  study_extractions_file,
  col_select = c(unique_study_id, study, min_p),
  show_col_types = FALSE
)
study_min_p <- study_extractions_meta |>
  dplyr::select(unique_study_id, min_p)

# Denominator for coloc profiles: common phenotype extractions at genome-wide significance
phenotype_universe <- study_extractions_meta |>
  dplyr::filter(!is.na(min_p), min_p <= genome_wide_p_value_threshold) |>
  dplyr::left_join(study_types, by = c("study" = "study_name")) |>
  dplyr::filter(data_type == "phenotype", variant_type == "common") |>
  dplyr::pull(unique_study_id) |>
  unique()
```

## Overview

This report compares the three prior / BFDR clustering configurations across LD blocks.

**Single-block override:** set `params$ld_block` when rendering.

### Methods compared

| Method | Pairwise input used for clustering | Annotated pairwise output | Threshold file |
|--------|------------------------------------|---------------------------|----------------|
| 5e-6, no BFDR | `coloc_pairwise_results.tsv.gz` | `coloc_pairwise_results_5e-6_no_bfdr_1_0_0.tsv.gz` | `global_no_bfdr_threshold_p12_5e-6.tsv` |
| 5e-6 + BFDR | `coloc_pairwise_results.tsv.gz` | `coloc_pairwise_results_5e-6_bfdr_1_0_0.tsv.gz` | `global_bfdr_threshold_p12_5e-6.tsv` |
| 1e-5 + BFDR | `coloc_pairwise_results_p12_1e-5.tsv.gz` | `coloc_pairwise_results_1e-5_bfdr_1_0_0.tsv.gz` | `global_bfdr_threshold_p12_1e-5.tsv` |

```{r resolve_blocks}
ld_info <- data.frame(
  block = selected_ld_blocks,
  ld_block_data = file.path(ld_block_data_dir, selected_ld_blocks),
  stringsAsFactors = FALSE
)
ld_info <- ld_info[dir.exists(ld_info$ld_block_data), , drop = FALSE]

required_files <- c(
  comparison_files$pairwise_file,
  comparison_files$cluster_file,
  comparison_files$rds_file
)
complete_mask <- vapply(seq_len(nrow(ld_info)), function(i) {
  all(file.exists(file.path(ld_info$ld_block_data[i], required_files)))
}, logical(1))

incomplete_blocks <- ld_info$block[!complete_mask]
ld_info <- ld_info[complete_mask, , drop = FALSE]
blocks <- ld_info$block

if (length(blocks) == 0L) {
  stop("No LD blocks have complete outputs for all three clustering configurations.")
}
```

**Complete LD blocks analysed:** `r length(blocks)` of `r length(selected_ld_blocks)` requested. All tables below are **pooled across those complete blocks**.

### Graph structure metrics

Metrics are calculated per LD block from the unpruned and pruned graphs, then pooled across blocks.

```{r pruned_vs_unpruned_graph_structure_metrics}
per_block_graph <- dplyr::bind_rows(parallel::mclapply(
  seq_len(nrow(ld_info)),
  function(bi) {
    dplyr::bind_rows(lapply(seq_len(nrow(comparison_files)), function(method_i) {
      rds_path <- file.path(ld_info$ld_block_data[bi], comparison_files$rds_file[method_i])
      cr <- tryCatch(readRDS(rds_path), error = function(e) NULL)
      if (is.null(cr) || is.null(cr$unpruned_graph) || is.null(cr$pruned_graph)) {
        return(NULL)
      }
      u <- tryCatch(
        analyze_graph_structure_advanced(cr$unpruned_graph),
        error = function(e) NULL
      )
      p <- tryCatch(
        analyze_graph_structure_advanced(cr$pruned_graph, original_graph = cr$unpruned_graph),
        error = function(e) NULL
      )
      if (is.null(u) || is.null(p)) {
        return(NULL)
      }
      data.frame(
        ld_block = ld_info$block[bi],
        method = comparison_files$method[method_i],
        n_components_unpruned = u$n_components,
        n_components_pruned = p$n_components,
        mean_weight_unpruned = u$mean_weight,
        mean_weight_pruned = p$mean_weight,
        mean_transitivity_unpruned = u$mean_transitivity,
        mean_transitivity_pruned = p$mean_transitivity,
        mean_conductance_pruned = p$mean_conductance,
        median_size_unpruned = u$median_size,
        median_size_pruned = p$median_size,
        stringsAsFactors = FALSE
      )
    }))
  },
  mc.cores = n_cores
))

graph_structure_pooled <- per_block_graph |>
  dplyr::group_by(method) |>
  dplyr::summarise(
    n_components_unpruned = sum(n_components_unpruned),
    n_components_pruned = sum(n_components_pruned),
    mean_weight_unpruned = mean(mean_weight_unpruned, na.rm = TRUE),
    mean_weight_pruned = mean(mean_weight_pruned, na.rm = TRUE),
    mean_transitivity_unpruned = mean(mean_transitivity_unpruned, na.rm = TRUE),
    mean_transitivity_pruned = mean(mean_transitivity_pruned, na.rm = TRUE),
    mean_conductance_pruned = mean(mean_conductance_pruned, na.rm = TRUE),
    median_of_median_size_unpruned = stats::median(median_size_unpruned, na.rm = TRUE),
    median_of_median_size_pruned = stats::median(median_size_pruned, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  graph_structure_pooled,
  digits = 4,
  caption = "Graph structure metrics pooled across complete LD blocks"
)
```

### Stability across prior and BFDR settings

False-positive / false-negative flags come from the annotated pairwise outputs written by clustering:

- **False positive:** an H4-positive pair whose edge was removed during clustering
- **False negative:** an H4-negative pair whose studies nevertheless finish in the same cluster

FP/FN rates are restricted to **non-ignored** pairs where **both** study extractions have
`min_p < 5e-8`, and H4 significance uses `h4 > 0.8`. Both false-positive and false-negative
rates use H4-positive pairs as the denominator (`n_false_* / n_h4_positive`).

Method-to-method stability asks how much BFDR (or the 1e-5 prior) changes **pairwise H4
calls** versus **final Infomap communities**:

- **Pairwise change** — among the union of non-ignored pairs with H4 ≥ each method's clustering
  threshold, the fraction in the symmetric difference (`1 − Jaccard`)
- **Community-pair change** — same for co-clustered study pairs in the clustered results
- **Adjusted Rand Index (ARI)** — agreement of Infomap partitions on the shared study universe
  (studies present in only one clustering are treated as singletons in the other)

```{r compare_clustering_result_sets}
resolve_global_bfdr_path <- function(bfdr_file) {
  candidates <- c(
    file.path(pipeline_metadata_dir, bfdr_file),
    bfdr_file
  )
  existing <- candidates[file.exists(candidates)]
  if (length(existing) > 0L) existing[1L] else candidates[1L]
}

read_global_bfdr_threshold <- function(path) {
  if (!file.exists(path)) {
    warning("Missing global BFDR file; using the default H4 threshold: ", path)
    return(posterior_prob_h4_threshold)
  }
  global_bfdr <- vroom::vroom(path, show_col_types = FALSE)
  threshold_col <- intersect(c("threshold", "threhold"), names(global_bfdr))[1L]
  if (length(threshold_col) == 0L || is.na(threshold_col) || nrow(global_bfdr) == 0L) {
    warning("No threshold column/value in ", path, "; using the default H4 threshold.")
    return(posterior_prob_h4_threshold)
  }
  as.numeric(global_bfdr[[threshold_col]][1L])
}

method_thresholds <- stats::setNames(
  vapply(
    comparison_files$bfdr_file,
    function(f) read_global_bfdr_threshold(resolve_global_bfdr_path(f)),
    numeric(1)
  ),
  comparison_files$method
)

make_pair_id <- function(study_a, study_b) {
  paste(pmin(study_a, study_b), pmax(study_a, study_b), sep = "||")
}

# Hubert & Arabie adjusted Rand index for two labelings of the same objects.
adjusted_rand_index <- function(labels_a, labels_b) {
  stopifnot(length(labels_a) == length(labels_b))
  n <- length(labels_a)
  if (n < 2L) {
    return(NA_real_)
  }

  tab <- table(labels_a, labels_b)
  sum_comb <- function(x) sum(x * (x - 1) / 2)
  index <- sum_comb(as.vector(tab))
  row_comb <- sum_comb(rowSums(tab))
  col_comb <- sum_comb(colSums(tab))
  expected <- row_comb * col_comb / (n * (n - 1) / 2)
  max_index <- (row_comb + col_comb) / 2
  denom <- max_index - expected
  if (denom <= 0) {
    return(1)
  }
  (index - expected) / denom
}

cocluster_pair_ids <- function(clusters) {
  if (is.null(clusters) || nrow(clusters) == 0L) {
    return(character())
  }
  component_members <- split(clusters$unique_study_id, as.character(clusters$component))
  pairs <- lapply(component_members, function(studies) {
    studies <- unique(studies)
    if (length(studies) < 2L) {
      return(character())
    }
    apply(utils::combn(sort(studies), 2L), 2L, paste, collapse = "||")
  })
  unique(unlist(pairs, use.names = FALSE))
}

membership_from_clusters <- function(clusters) {
  if (is.null(clusters) || nrow(clusters) == 0L) {
    return(stats::setNames(character(), character()))
  }
  stats::setNames(as.character(clusters$component), clusters$unique_study_id)
}

align_memberships_for_ari <- function(membership_a, membership_b) {
  universe <- union(names(membership_a), names(membership_b))
  if (length(universe) == 0L) {
    return(list(a = character(), b = character()))
  }
  label_a <- membership_a[universe]
  label_b <- membership_b[universe]
  missing_a <- is.na(label_a)
  missing_b <- is.na(label_b)
  if (any(missing_a)) {
    label_a[missing_a] <- paste0("singleton_a_", universe[missing_a])
  }
  if (any(missing_b)) {
    label_b[missing_b] <- paste0("singleton_b_", universe[missing_b])
  }
  list(a = unname(label_a), b = unname(label_b), universe = universe)
}

compare_partition_pair <- function(
    method_a,
    method_b,
    h4_pairs_a,
    h4_pairs_b,
    clusters_a,
    clusters_b,
    ld_block
) {
  pair_union <- union(h4_pairs_a, h4_pairs_b)
  pair_shared <- intersect(h4_pairs_a, h4_pairs_b)
  n_pair_union <- length(pair_union)
  n_pair_shared <- length(pair_shared)

  cocluster_a <- cocluster_pair_ids(clusters_a)
  cocluster_b <- cocluster_pair_ids(clusters_b)
  community_union <- union(cocluster_a, cocluster_b)
  community_shared <- intersect(cocluster_a, cocluster_b)
  n_community_union <- length(community_union)
  n_community_shared <- length(community_shared)

  aligned <- align_memberships_for_ari(
    membership_from_clusters(clusters_a),
    membership_from_clusters(clusters_b)
  )
  ari <- if (length(aligned$universe) >= 2L) {
    adjusted_rand_index(aligned$a, aligned$b)
  } else {
    NA_real_
  }

  data.frame(
    ld_block = ld_block,
    method_a = method_a,
    method_b = method_b,
    n_studies = length(aligned$universe),
    n_h4_pairs_a = length(h4_pairs_a),
    n_h4_pairs_b = length(h4_pairs_b),
    n_h4_pairs_shared = n_pair_shared,
    n_h4_pairs_union = n_pair_union,
    pairwise_changed_frac = if (n_pair_union > 0L) {
      1 - n_pair_shared / n_pair_union
    } else {
      NA_real_
    },
    n_community_pairs_a = length(cocluster_a),
    n_community_pairs_b = length(cocluster_b),
    n_community_pairs_shared = n_community_shared,
    n_community_pairs_union = n_community_union,
    community_changed_frac = if (n_community_union > 0L) {
      1 - n_community_shared / n_community_union
    } else {
      NA_real_
    },
    adjusted_rand_index = ari,
    stringsAsFactors = FALSE
  )
}

read_comparison_set <- function(block_data_dir, method_i) {
  pairwise_path <- file.path(block_data_dir, comparison_files$pairwise_file[method_i])
  cluster_path <- file.path(block_data_dir, comparison_files$cluster_file[method_i])

  pairwise <- vroom::vroom(
    pairwise_path,
    delim = "\t",
    col_select = dplyr::any_of(c(
      "unique_study_a", "unique_study_b", "study_a", "study_b",
      "ignore", "h4", "PP.H4.abf", "false_positive", "false_negative"
    )),
    show_col_types = FALSE
  )
  clusters <- vroom::vroom(
    cluster_path,
    delim = "\t",
    col_select = dplyr::any_of(c("unique_study_id", "component")),
    show_col_types = FALSE
  )

  if (!"ignore" %in% names(pairwise)) pairwise$ignore <- FALSE
  if (!"h4" %in% names(pairwise) && "PP.H4.abf" %in% names(pairwise)) {
    pairwise$h4 <- pairwise$PP.H4.abf
  }
  for (flag in c("false_positive", "false_negative")) {
    if (!flag %in% names(pairwise)) {
      pairwise[[flag]] <- NA
    } else {
      pairwise[[flag]] <- as.logical(pairwise[[flag]])
    }
  }

  pairwise <- pairwise |>
    dplyr::mutate(ignore = dplyr::coalesce(as.logical(ignore), FALSE))

  list(
    method = comparison_files$method[method_i],
    h4_threshold = method_thresholds[[comparison_files$method[method_i]]],
    pairwise = pairwise,
    clusters = clusters
  )
}

h4_pair_ids <- function(pairwise, h4_threshold) {
  keep <- !pairwise$ignore & !is.na(pairwise$h4) & pairwise$h4 >= h4_threshold
  if (!any(keep)) {
    return(character())
  }
  unique(make_pair_id(pairwise$unique_study_a[keep], pairwise$unique_study_b[keep]))
}

pheno_ids_from_pairs <- function(annotated, keep_mask) {
  sub <- annotated[keep_mask, , drop = FALSE]
  if (nrow(sub) == 0L) {
    return(character())
  }
  unique(stats::na.omit(c(sub$unique_study_a, sub$unique_study_b)))
}

# Molecular QTL partners: qtl_data_types from constants.R (includes methylation;
# excludes phenotype and non_qtl_data_types such as cell_trait / metabolome / plasma_protein).
is_molecular_data_type <- function(data_type) {
  data_type %in% qtl_data_types
}

summarise_trait_molecular_links <- function(pairwise, h4_threshold) {
  empty <- list(
    summary = data.frame(
      n_h4_mol_pheno_pairs = 0,
      n_h4_no_fp_mol_pheno_pairs = 0,
      n_h4_no_fp_plus_fn_mol_pheno_pairs = 0
    ),
    trait_hits_h4_only = character(),
    trait_hits_h4_no_fp = character(),
    trait_hits_h4_no_fp_plus_fn = character(),
    trait_studies_h4_only = character(),
    trait_studies_h4_no_fp = character(),
    trait_studies_h4_no_fp_plus_fn = character(),
    pheno_pheno_hits_h4_only = character(),
    pheno_pheno_hits_h4_no_fp = character(),
    pheno_pheno_hits_h4_no_fp_plus_fn = character()
  )

  if (!all(c("study_a", "study_b") %in% names(pairwise))) {
    return(empty)
  }

  annotated <- pairwise |>
    dplyr::filter(!ignore) |>
    dplyr::left_join(study_types, by = c("study_a" = "study_name")) |>
    dplyr::rename(type_a = data_type) |>
    dplyr::left_join(study_types, by = c("study_b" = "study_name")) |>
    dplyr::rename(type_b = data_type) |>
    dplyr::left_join(study_min_p, by = c("unique_study_a" = "unique_study_id")) |>
    dplyr::rename(min_p_a = min_p) |>
    dplyr::left_join(study_min_p, by = c("unique_study_b" = "unique_study_id")) |>
    dplyr::rename(min_p_b = min_p) |>
    dplyr::mutate(
      type_a = dplyr::coalesce(type_a, "unknown"),
      type_b = dplyr::coalesce(type_b, "unknown"),
      both_gw_sig = !is.na(min_p_a) & !is.na(min_p_b) &
        min_p_a <= genome_wide_p_value_threshold &
        min_p_b <= genome_wide_p_value_threshold,
      is_mol_pheno = both_gw_sig & (
        (type_a == "phenotype" & is_molecular_data_type(type_b)) |
          (type_b == "phenotype" & is_molecular_data_type(type_a))
      ),
      is_pheno_pheno = both_gw_sig & type_a == "phenotype" & type_b == "phenotype",
      phenotype_unique_study_id = dplyr::case_when(
        type_a == "phenotype" & is_molecular_data_type(type_b) ~ unique_study_a,
        type_b == "phenotype" & is_molecular_data_type(type_a) ~ unique_study_b,
        TRUE ~ NA_character_
      ),
      phenotype_study = dplyr::case_when(
        type_a == "phenotype" & is_molecular_data_type(type_b) ~ study_a,
        type_b == "phenotype" & is_molecular_data_type(type_a) ~ study_b,
        TRUE ~ NA_character_
      ),
      false_positive = dplyr::coalesce(false_positive, FALSE),
      false_negative = dplyr::coalesce(false_negative, FALSE),
      h4_ok = !is.na(h4) & h4 >= h4_threshold,
      keep_pair_h4_only = h4_ok,
      keep_pair_h4_no_fp = h4_ok & !false_positive,
      keep_pair_h4_no_fp_plus_fn = (h4_ok & !false_positive) | false_negative,
      keep_h4_only = is_mol_pheno & keep_pair_h4_only,
      keep_h4_no_fp = is_mol_pheno & keep_pair_h4_no_fp,
      keep_h4_no_fp_plus_fn = is_mol_pheno & keep_pair_h4_no_fp_plus_fn
    )

  list(
    summary = data.frame(
      n_h4_mol_pheno_pairs = sum(as.numeric(annotated$keep_h4_only), na.rm = TRUE),
      n_h4_no_fp_mol_pheno_pairs = sum(as.numeric(annotated$keep_h4_no_fp), na.rm = TRUE),
      n_h4_no_fp_plus_fn_mol_pheno_pairs = sum(
        as.numeric(annotated$keep_h4_no_fp_plus_fn), na.rm = TRUE
      )
    ),
    trait_hits_h4_only = unique(stats::na.omit(
      annotated$phenotype_unique_study_id[annotated$keep_h4_only]
    )),
    trait_hits_h4_no_fp = unique(stats::na.omit(
      annotated$phenotype_unique_study_id[annotated$keep_h4_no_fp]
    )),
    trait_hits_h4_no_fp_plus_fn = unique(stats::na.omit(
      annotated$phenotype_unique_study_id[annotated$keep_h4_no_fp_plus_fn]
    )),
    trait_studies_h4_only = unique(stats::na.omit(
      annotated$phenotype_study[annotated$keep_h4_only]
    )),
    trait_studies_h4_no_fp = unique(stats::na.omit(
      annotated$phenotype_study[annotated$keep_h4_no_fp]
    )),
    trait_studies_h4_no_fp_plus_fn = unique(stats::na.omit(
      annotated$phenotype_study[annotated$keep_h4_no_fp_plus_fn]
    )),
    pheno_pheno_hits_h4_only = pheno_ids_from_pairs(
      annotated, annotated$is_pheno_pheno & annotated$keep_pair_h4_only
    ),
    pheno_pheno_hits_h4_no_fp = pheno_ids_from_pairs(
      annotated, annotated$is_pheno_pheno & annotated$keep_pair_h4_no_fp
    ),
    pheno_pheno_hits_h4_no_fp_plus_fn = pheno_ids_from_pairs(
      annotated, annotated$is_pheno_pheno & annotated$keep_pair_h4_no_fp_plus_fn
    )
  )
}

# Molecular if the coloc group contains any QTL data type (including methylation).
# Non-QTL types (cell_trait, metabolome, plasma_protein) do not count as molecular.
classify_phenotypes_from_clusters <- function(clusters) {
  empty <- list(
    molecular = character(),
    phenotype_only = character()
  )
  if (is.null(clusters) || nrow(clusters) == 0L || !"unique_study_id" %in% names(clusters)) {
    return(empty)
  }

  anno <- clusters |>
    dplyr::left_join(
      study_extractions_meta |> dplyr::select(unique_study_id, study),
      by = "unique_study_id"
    ) |>
    dplyr::left_join(study_types, by = c("study" = "study_name")) |>
    dplyr::mutate(
      data_type = dplyr::coalesce(data_type, "unknown"),
      component = as.character(component),
      is_molecular = is_molecular_data_type(data_type)
    )

  molecular <- anno |>
    dplyr::group_by(component) |>
    dplyr::filter(any(is_molecular)) |>
    dplyr::ungroup() |>
    dplyr::filter(data_type == "phenotype") |>
    dplyr::pull(unique_study_id) |>
    unique()

  phenotype_only <- anno |>
    dplyr::group_by(component) |>
    dplyr::filter(!any(is_molecular), any(data_type == "phenotype")) |>
    dplyr::ungroup() |>
    dplyr::filter(data_type == "phenotype") |>
    dplyr::pull(unique_study_id) |>
    unique()

  list(molecular = molecular, phenotype_only = phenotype_only)
}

compare_block <- function(ld_block, block_data_dir) {
  result_sets <- lapply(seq_len(nrow(comparison_files)), function(method_i) {
    read_comparison_set(block_data_dir, method_i)
  })
  names(result_sets) <- comparison_files$method

  method_names <- names(result_sets)
  h4_pairs_by_method <- lapply(result_sets, function(result_set) {
    h4_pair_ids(result_set$pairwise, result_set$h4_threshold)
  })

  partition_stability <- dplyr::bind_rows(lapply(seq_along(method_names), function(i) {
    dplyr::bind_rows(lapply(seq_along(method_names), function(j) {
      if (j <= i) {
        return(NULL)
      }
      compare_partition_pair(
        method_a = method_names[i],
        method_b = method_names[j],
        h4_pairs_a = h4_pairs_by_method[[i]],
        h4_pairs_b = h4_pairs_by_method[[j]],
        clusters_a = result_sets[[i]]$clusters,
        clusters_b = result_sets[[j]]$clusters,
        ld_block = ld_block
      )
    }))
  }))

  algorithm_error_summary <- dplyr::bind_rows(lapply(result_sets, function(result_set) {
    # Match pleiotropy_analysis.Rmd: GW-sig pairs on both sides, H4 > 0.8 for the rate.
    pairwise <- result_set$pairwise |>
      dplyr::left_join(study_min_p, by = c("unique_study_a" = "unique_study_id")) |>
      dplyr::rename(min_p_a = min_p) |>
      dplyr::left_join(study_min_p, by = c("unique_study_b" = "unique_study_id")) |>
      dplyr::rename(min_p_b = min_p) |>
      dplyr::filter(
        !ignore,
        !is.na(min_p_a),
        !is.na(min_p_b),
        min_p_a < genome_wide_p_value_threshold,
        min_p_b < genome_wide_p_value_threshold
      )

    h4_positive <- !is.na(pairwise$h4) & pairwise$h4 > posterior_prob_h4_threshold
    n_h4_positive <- sum(h4_positive)
    n_false_positive <- sum(h4_positive & pairwise$false_positive %in% TRUE)
    n_false_negative <- sum(pairwise$false_negative %in% TRUE)

    data.frame(
      ld_block = ld_block,
      method = result_set$method,
      clustering_h4_threshold = result_set$h4_threshold,
      n_h4_positive = n_h4_positive,
      n_false_positive = n_false_positive,
      n_false_negative = n_false_negative,
      false_positive_rate = if (n_h4_positive > 0L) {
        n_false_positive / n_h4_positive
      } else {
        NA_real_
      },
      # FN rate relative to H4-positive pairs (true-positive H4 discoveries)
      false_negative_rate = if (n_h4_positive > 0L) {
        n_false_negative / n_h4_positive
      } else {
        NA_real_
      }
    )
  }))

  trait_molecular_by_method <- lapply(result_sets, function(result_set) {
    trait_links <- summarise_trait_molecular_links(
      result_set$pairwise,
      result_set$h4_threshold
    )
    cluster_classes <- classify_phenotypes_from_clusters(result_set$clusters)
    list(
      summary = dplyr::bind_cols(
        data.frame(
          ld_block = ld_block,
          method = result_set$method,
          h4_threshold = result_set$h4_threshold,
          pairwise_file = comparison_files$pairwise_file[
            comparison_files$method == result_set$method
          ],
          stringsAsFactors = FALSE
        ),
        trait_links$summary
      ),
      ids = c(
        trait_links[c(
          "trait_hits_h4_only",
          "trait_hits_h4_no_fp",
          "trait_hits_h4_no_fp_plus_fn",
          "trait_studies_h4_only",
          "trait_studies_h4_no_fp",
          "trait_studies_h4_no_fp_plus_fn",
          "pheno_pheno_hits_h4_only",
          "pheno_pheno_hits_h4_no_fp",
          "pheno_pheno_hits_h4_no_fp_plus_fn"
        )],
        list(
          cluster_molecular = cluster_classes$molecular,
          cluster_phenotype_only = cluster_classes$phenotype_only
        )
      )
    )
  })

  list(
    partition_stability = partition_stability,
    algorithm_error_summary = algorithm_error_summary,
    trait_molecular_summary = dplyr::bind_rows(
      lapply(trait_molecular_by_method, `[[`, "summary")
    ),
    trait_molecular_ids = lapply(trait_molecular_by_method, `[[`, "ids")
  )
}

block_comparisons <- parallel::mclapply(
  seq_len(nrow(ld_info)),
  function(i) {
    compare_block(ld_info$block[i], ld_info$ld_block_data[i])
  },
  mc.cores = n_cores
)

partition_stability <- dplyr::bind_rows(
  lapply(block_comparisons, `[[`, "partition_stability")
)
algorithm_error_summary <- dplyr::bind_rows(
  lapply(block_comparisons, `[[`, "algorithm_error_summary")
)
trait_molecular_summary <- dplyr::bind_rows(
  lapply(block_comparisons, `[[`, "trait_molecular_summary")
)

partition_stability_pooled <- partition_stability |>
  dplyr::group_by(method_a, method_b) |>
  dplyr::summarise(
    n_blocks = dplyr::n(),
    # Compute weighted ARI before n_studies is overwritten by its sum.
    adjusted_rand_index = {
      ok <- !is.na(adjusted_rand_index) & n_studies > 0
      if (!any(ok)) {
        NA_real_
      } else {
        sum(adjusted_rand_index[ok] * n_studies[ok]) / sum(n_studies[ok])
      }
    },
    n_studies = sum(as.numeric(n_studies)),
    n_h4_pairs_a = sum(as.numeric(n_h4_pairs_a)),
    n_h4_pairs_b = sum(as.numeric(n_h4_pairs_b)),
    n_h4_pairs_shared = sum(as.numeric(n_h4_pairs_shared)),
    n_h4_pairs_union = sum(as.numeric(n_h4_pairs_union)),
    pairwise_changed_pct = if (n_h4_pairs_union > 0) {
      100 * (1 - n_h4_pairs_shared / n_h4_pairs_union)
    } else {
      NA_real_
    },
    n_community_pairs_a = sum(as.numeric(n_community_pairs_a)),
    n_community_pairs_b = sum(as.numeric(n_community_pairs_b)),
    n_community_pairs_shared = sum(as.numeric(n_community_pairs_shared)),
    n_community_pairs_union = sum(as.numeric(n_community_pairs_union)),
    community_changed_pct = if (n_community_pairs_union > 0) {
      100 * (1 - n_community_pairs_shared / n_community_pairs_union)
    } else {
      NA_real_
    },
    .groups = "drop"
  )

algorithm_error_pooled <- algorithm_error_summary |>
  dplyr::group_by(method) |>
  dplyr::summarise(
    clustering_h4_threshold = dplyr::first(clustering_h4_threshold),
    n_h4_positive = sum(as.numeric(n_h4_positive)),
    n_false_positive = sum(as.numeric(n_false_positive)),
    n_false_negative = sum(as.numeric(n_false_negative)),
    false_positive_rate = if (n_h4_positive > 0) {
      n_false_positive / n_h4_positive
    } else {
      NA_real_
    },
    false_negative_rate = if (n_h4_positive > 0) {
      n_false_negative / n_h4_positive
    } else {
      NA_real_
    },
    .groups = "drop"
  )

knitr::kable(
  algorithm_error_pooled,
  digits = 4,
  caption = paste(
    "False positives and false negatives marked by the clustering algorithm,",
    "pooled across LD blocks (non-ignored pairs with both min_p < 5e-8;",
    "FP and FN rates = count / n_h4_positive where h4 > 0.8)"
  )
)

knitr::kable(
  partition_stability_pooled |>
    dplyr::select(
      method_a, method_b,
      pairwise_changed_pct, community_changed_pct, adjusted_rand_index,
      n_h4_pairs_union, n_community_pairs_union, n_studies
    ),
  digits = 4,
  caption = paste(
    "Method-to-method stability pooled across LD blocks.",
    "pairwise_changed_pct = 100 × (1 − Jaccard) of H4≥threshold pairs;",
    "community_changed_pct = same for Infomap co-clustered study pairs;",
    "adjusted_rand_index = study-count–weighted mean ARI of Infomap partitions",
    "(studies present in only one clustering treated as singletons)."
  )
)

bfdr_contrast <- partition_stability_pooled |>
  dplyr::filter(
    method_a == "5e-6, no BFDR",
    method_b == "5e-6 + BFDR"
  )

if (nrow(bfdr_contrast) == 1L) {
  cat(
    sprintf(
      paste0(
        "\nApplying BFDR changed %.1f%% of pairwise colocalizations ",
        "but only %.1f%% of final graph community pairs ",
        "(Adjusted Rand Index = %.3f).\n"
      ),
      bfdr_contrast$pairwise_changed_pct,
      bfdr_contrast$community_changed_pct,
      bfdr_contrast$adjusted_rand_index
    )
  )
}
```

### Trait hits linked to a molecular phenotype

Using the annotated pairwise outputs (`coloc_pairwise_results_*`), count how many **phenotype extractions** (`unique_study_id`) and **phenotype studies** have at least one molecular QTL association under three definitions. Molecular partners are restricted to `qtl_data_types` from `constants.R` (eQTL / sQTL / tQTL / pQTL / methQTL), and both sides must have `min_p ≤ 5e-8` (matching `pipeline_summary.Rmd`).

1. **H4 only** — H4 ≥ method threshold and phenotype–QTL pair
2. **H4 minus false positives** — same as (1), excluding pairs marked `false_positive`
3. **H4 minus FP + false negatives** — pairs kept by (2), plus pairs marked `false_negative`

Then report **overall** colocalization profiles for genome-wide significant common phenotype loci (`min_p ≤ 5e-8`), for each of the **9** method × definition combinations:

- **H4 only** — pairwise H4 links to a GW-significant QTL partner
- **H4 minus FP** / **H4 minus FP + FN** — Infomap coloc-group membership from each method's clustered results (molecular if the group contains any `qtl_data_types` member, including methylation; phenotype-only if the group has phenotype but no QTL types). FN pairs are already co-clustered, so those two definitions share the same profile.

```{r trait_molecular_association}
trait_molecular_pair_counts <- trait_molecular_summary |>
  dplyr::group_by(method, pairwise_file, h4_threshold) |>
  dplyr::summarise(
    n_mol_pheno_pairs_h4_only = sum(as.numeric(n_h4_mol_pheno_pairs), na.rm = TRUE),
    n_mol_pheno_pairs_h4_no_fp = sum(as.numeric(n_h4_no_fp_mol_pheno_pairs), na.rm = TRUE),
    n_mol_pheno_pairs_h4_no_fp_plus_fn = sum(
      as.numeric(n_h4_no_fp_plus_fn_mol_pheno_pairs), na.rm = TRUE
    ),
    .groups = "drop"
  )

union_ids <- function(id_lists, field) {
  unique(unlist(lapply(id_lists, `[[`, field), use.names = FALSE))
}

trait_molecular_pooled <- dplyr::bind_rows(lapply(comparison_files$method, function(method) {
  id_lists <- lapply(block_comparisons, function(block_result) {
    block_result$trait_molecular_ids[[method]]
  })
  pair_row <- trait_molecular_pair_counts |>
    dplyr::filter(method == !!method)

  data.frame(
    method = method,
    pairwise_file = pair_row$pairwise_file[1],
    h4_threshold = pair_row$h4_threshold[1],
    n_mol_pheno_pairs_h4_only = pair_row$n_mol_pheno_pairs_h4_only[1],
    n_mol_pheno_pairs_h4_no_fp = pair_row$n_mol_pheno_pairs_h4_no_fp[1],
    n_mol_pheno_pairs_h4_no_fp_plus_fn = pair_row$n_mol_pheno_pairs_h4_no_fp_plus_fn[1],
    n_trait_hits_h4_only = length(union_ids(id_lists, "trait_hits_h4_only")),
    n_trait_hits_h4_no_fp = length(union_ids(id_lists, "trait_hits_h4_no_fp")),
    n_trait_hits_h4_no_fp_plus_fn = length(union_ids(id_lists, "trait_hits_h4_no_fp_plus_fn")),
    n_trait_studies_h4_only = length(union_ids(id_lists, "trait_studies_h4_only")),
    n_trait_studies_h4_no_fp = length(union_ids(id_lists, "trait_studies_h4_no_fp")),
    n_trait_studies_h4_no_fp_plus_fn = length(union_ids(id_lists, "trait_studies_h4_no_fp_plus_fn")),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  trait_molecular_pooled,
  digits = 4,
  caption = paste(
    "Trait hits linked to a GW-significant QTL (qtl_data_types) across annotated pairwise files.",
    "Both sides require min_p ≤ 5e-8. Trait hits = unique phenotype extractions;",
    "trait studies = unique phenotype study_name."
  )
)

n_phenotype_universe <- length(phenotype_universe)

coloc_profile_overall <- dplyr::bind_rows(lapply(comparison_files$method, function(method) {
  id_lists <- lapply(block_comparisons, function(block_result) {
    block_result$trait_molecular_ids[[method]]
  })

  cluster_mol <- intersect(union_ids(id_lists, "cluster_molecular"), phenotype_universe)
  cluster_pheno_only <- intersect(
    union_ids(id_lists, "cluster_phenotype_only"),
    phenotype_universe
  )
  # A locus in a molecular group should not also count as phenotype-only
  cluster_pheno_only <- setdiff(cluster_pheno_only, cluster_mol)

  pairwise_states <- list(
    "H4 only" = list(
      mol = intersect(union_ids(id_lists, "trait_hits_h4_only"), phenotype_universe),
      pheno_only = setdiff(
        intersect(union_ids(id_lists, "pheno_pheno_hits_h4_only"), phenotype_universe),
        intersect(union_ids(id_lists, "trait_hits_h4_only"), phenotype_universe)
      )
    ),
    "H4 minus FP" = list(
      mol = cluster_mol,
      pheno_only = cluster_pheno_only
    ),
    "H4 minus FP + FN" = list(
      mol = cluster_mol,
      pheno_only = cluster_pheno_only
    )
  )

  dplyr::bind_rows(lapply(names(pairwise_states), function(definition) {
    mol_ids <- pairwise_states[[definition]]$mol
    pheno_only_ids <- pairwise_states[[definition]]$pheno_only
    n_mol <- length(mol_ids)
    n_pheno_only <- length(pheno_only_ids)
    n_none <- n_phenotype_universe - n_mol - n_pheno_only

    data.frame(
      method = method,
      definition = definition,
      n_phenotype_loci = n_phenotype_universe,
      n_molecular = n_mol,
      n_phenotype_only = n_pheno_only,
      n_no_coloc = n_none,
      percent_molecular = n_mol / n_phenotype_universe,
      percent_phenotype_only = n_pheno_only / n_phenotype_universe,
      percent_no_coloc = n_none / n_phenotype_universe,
      stringsAsFactors = FALSE
    )
  }))
}))

coloc_profile_overall <- coloc_profile_overall |>
  dplyr::mutate(
    method = factor(method, levels = comparison_files$method),
    definition = factor(
      definition,
      levels = c("H4 only", "H4 minus FP", "H4 minus FP + FN")
    )
  ) |>
  dplyr::arrange(method, definition)

knitr::kable(
  coloc_profile_overall |>
    dplyr::mutate(
      n_phenotype_loci = scales::comma(n_phenotype_loci),
      n_molecular = scales::comma(n_molecular),
      n_phenotype_only = scales::comma(n_phenotype_only),
      n_no_coloc = scales::comma(n_no_coloc),
      percent_molecular = scales::percent(percent_molecular, accuracy = 0.1),
      percent_phenotype_only = scales::percent(percent_phenotype_only, accuracy = 0.1),
      percent_no_coloc = scales::percent(percent_no_coloc, accuracy = 0.1)
    ),
  caption = paste(
    "Overall colocalization profiles of fine-mapped common phenotype loci (min_p ≤ 5e-8).",
    "H4 only = pairwise H4 to a GW-significant QTL partner (qtl_data_types, including methQTL);",
    "H4 minus FP / H4 minus FP + FN = coloc-group membership with any QTL member.",
    "non_qtl_data_types (cell_trait, metabolome, plasma_protein) do not count as molecular."
  )
)
```
