---
title: "Genomic regions"
author: Gibran Hemani
date: "2024-04-18"
categories: []
cache: true
---

## Background

What does it look like to try to cluster traits based on GWAS hits, on a per-region basis?

## Example

Extract all the tophits from UKB GWAS

```{r}
library(dplyr)
library(ieugwasr)
library(ggplot2)
library(tidyr)

gwasinfo <- gwasinfo()
ukb_th <- tophits(grep("ukb-b", gwasinfo$id, value = TRUE))
```

```{r}
ggplot(ukb_th, aes(y = -log10(p), x = position, color = chr)) +
  geom_point() +
  facet_grid(. ~ as.numeric(chr), space="free_x", scale="free_x") +
  theme(axis.text.x=element_blank(), legend.position = "none")
```

Get pre-calculated LD regions

```{r}
load(url("https://github.com/explodecomputer/simulateGP/raw/master/data/ldetect.rdata"))

ldetect <- subset(ldetect, pop == "EUR") %>%
    mutate(chr = gsub("chr", "", chr), code = paste0(chr, ":", start, "-", stop))
ldetect %>% head
```

Group UKB tophits by LD region

```{r}
ukb_th <- group_by(ldetect, code) %>%
    do({
        subset(ukb_th, chr == .$chr & position >= min(.$start) & position <= max(.$stop))
    })
```

Number of GWAS hits per LD region

```{r}
ukb_th %>% group_by(code) %>% summarise(n = n()) %>% {hist(.$n, breaks=100)}
```

## Example region

Take one region that has hits

```{r}
a <- lapply(subset(ukb_th, code == unique(ukb_th$code)[3])$id, \(id) {
    message(id)
    associations(unique(ukb_th$code)[3], id)
}) %>% bind_rows()
str(a)
```

These are the traits:

```{r}
unique(a$trait)
```

We could also search for traits that might have associations in this region that are not quite significant, but let's leave that for another time..

Look at regional plots for some of these:

```{r, fig.height=15}
ggplot(subset(a, id %in% unique(a$id)[c(1, 3, 6, 7, 8, 13, 17, 29)]), aes(y=-log10(p), x=position)) +
geom_point(aes(colour=p < 5e-4)) +
facet_wrap(~ trait, scales = "free_y", ncol=1)

```

So the objective here

-   Fine map each trait to get the marginal associations for each GWAS hit

-   Colocalise marginal regions to find groups of traits that share at least one GWAS hit

-   Output: A list of traits colocalising for every finemapped hit

## Clustering

Try clustering the traits using principal components?

```{r}
aw <- tidyr::pivot_wider(a %>% select(position, beta, id) %>% filter(!duplicated(paste(position, id))), names_from = id, values_from = beta)

awm <- as.matrix(aw[,-1])
awm[is.na(awm)] <- 0

awp <- prcomp(t(awm))
```

PCs 1-2

```{r}
plot(awp$x[,1], awp$x[,2], col=as.factor(unique(a$id)))
text(awp$x[,1], awp$x[,2], labels=as.factor(unique(a$trait)), cex=0.4, col="grey")
```

PCs 3-4

```{r}
plot(awp$x[,3], awp$x[,4], col=as.factor(unique(a$id)))
text(awp$x[,3], awp$x[,4], labels=as.factor(unique(a$trait)), cex=0.4, col="grey")
```

PCs 5-6

```{r}
plot(awp$x[,5], awp$x[,6], col=as.factor(unique(a$id)))
text(awp$x[,5], awp$x[,6], labels=as.factor(unique(a$trait)), cex=0.4, col="grey")
```

------------------------------------------------------------------------

```{r}
sessionInfo()
```
