---
title: "Basic region simulation"
format: html
editor: visual
---

## Background

Simulate GWAS summary data with the following characteristics

-   A single region with realistic LD patterns

-   `t` traits

-   `m` SNPs tested for each trait

-   `n` samples used to generate the summary statistics

-   A given `m x m` LD matrix and allele frequencies for all `m` SNPs

-   A causal matrix of `p` causal variants influencing the `t` traits. For example

    -   SNP 10 influences traits 1 and 2

    -   SNP 20 influences traits 1 and 3

    -   SNP 30 influences trait 4

From the marginal summary statistics for each trait, recapitulate the causal matrix

## Simulation function

```{r}
# remotes::install_github("jean997/GWASBrewer")
library(GWASBrewer)
library(dplyr)
library(ggplot2)
data("ld_mat_list")
data("AF")
```

```{r}
#' Simulate summary stats for traits in one region 
#' 
#' @param ld LD Matrix
#' @param af Vector of allele frequencies
#' @param pi Matrix of which SNPs have causal effects on which traits. First column is the SNP, each subsequent column is the trait, where 0=no effect and 1=effect
#' @param n Sample size for each trait
#' @param h2 Heritability of each trait
sim <- function(ld, af, pi, n, h2) {
  
  mat <- matrix(0, length(af), ncol(pi)-1)
  for(i in 1:nrow(pi)) {
    mat[pi[i,1],] <- pi[i,-1]
  }

  o <- sim_mv(G = ncol(mat),
    J = length(af), 
    N = n, 
    h2 = h2, 
    pi = mat, 
    R_LD = list(ld), 
    af = af)

  out <- lapply(1:ncol(mat), \(i) {
    tibble(
      trait = i,
      SNP = 1:length(af),
      af = af,
      beta = o$direct_SNP_effects_joint[,i],
      bhat = o$beta_hat[,i],
      se = o$se_beta_hat[,i],
      n = n,
      pval = pnorm(abs(bhat)/se, low=F)
    )
  }) %>% bind_rows()
  return(out)
}
```

## Example run

```{r}
# Causal matrix
pi <- matrix(c(
  10, 0,0,0,0,1,1,
  50, 0,0,0,0,1,1,
  80, 1,0,0,0,1,0,
  100,0,1,1,1,0,0
), 4, byrow=TRUE)

# Run the simulation
set.seed(100)
s <- sim(
  ld = ld_mat_list[[1]], 
  af = AF[1:140], 
  pi = pi, 
  n = 100000, 
  h2 = c(0.1, 0.1, 0.05, 0.05, 0.05, 0.08)
)
str(s)
```

Show marginal associations

```{r}
ggplot(s, aes(x=SNP, y=abs(bhat)/se)) +
  geom_point() +
  facet_grid(trait ~ .)
```
