---
title: "Current State of GP Map Data"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_fragment
params:
  compare_results: F
---


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r load-data, echo=F}
source('constants.R')
ld_blocks <- vroom::vroom('data/ld_blocks.tsv', show_col_types = F)
ld_info <- construct_ld_block(ld_blocks$ancestry, ld_blocks$chr, ld_blocks$start, ld_blocks$stop) |>
  dplyr::filter(dir.exists(ld_block_data))

pad_data_frames <- function(df1, df2) {
  df1 <- as.data.frame(df1)
  df2 <- as.data.frame(df2)
  nrow1 <- nrow(df1)
  ncol1 <- ncol(df1)
  nrow2 <- nrow(df2)
  ncol2 <- ncol(df2)
  
  # Create column names for consistency
  all_columns <- union(names(df1), names(df2))
  
  # Add missing columns filled with 0
  df1 <- cbind(df1, setNames(as.data.frame(matrix(0, nrow = nrow1, ncol = length(setdiff(all_columns, names(df1))))), 
                             setdiff(all_columns, names(df1))))
  df2 <- cbind(df2, setNames(as.data.frame(matrix(0, nrow = nrow2, ncol = length(setdiff(all_columns, names(df2))))), 
                             setdiff(all_columns, names(df2))))
  
  # Reorder columns to match
  df1 <- df1[, all_columns]
  df2 <- df2[, all_columns]
  
  # Add missing rows filled with 0
  if (nrow1 < nrow2) {
    df1 <- rbind(df1, as.data.frame(matrix(0, nrow = nrow2 - nrow1, ncol = ncol(df1))))
  } else if (nrow2 < nrow1) {
    df2 <- rbind(df2, as.data.frame(matrix(0, nrow = nrow1 - nrow2, ncol = ncol(df2))))
  }
  
  return(list(df1 = df1, df2 = df2))
}


subtract_data_frames <- function(df1, df2) {
  numeric_columns <- sapply(df1, is.numeric)
  result <- df1[, numeric_columns] - df2[, numeric_columns]
  final_result <- cbind(df1[, !numeric_columns, drop = FALSE], result)
  return(final_result)
}


pad_data_frames <- function(df1, df2) {
  df1 <- as.data.frame(df1)
  df2 <- as.data.frame(df2)
  nrow1 <- nrow(df1)
  ncol1 <- ncol(df1)
  nrow2 <- nrow(df2)
  ncol2 <- ncol(df2)
  
  # Create column names for consistency
  all_columns <- union(names(df1), names(df2))
  
  # Add missing columns filled with 0
  df1 <- cbind(df1, setNames(as.data.frame(matrix(0, nrow = nrow1, ncol = length(setdiff(all_columns, names(df1))))), 
                             setdiff(all_columns, names(df1))))
  df2 <- cbind(df2, setNames(as.data.frame(matrix(0, nrow = nrow2, ncol = length(setdiff(all_columns, names(df2))))), 
                             setdiff(all_columns, names(df2))))
  
  # Reorder columns to match
  df1 <- df1[, all_columns]
  df2 <- df2[, all_columns]
  
  # Add missing rows filled with 0
  if (nrow1 < nrow2) {
    df1 <- rbind(df1, as.data.frame(matrix(0, nrow = nrow2 - nrow1, ncol = ncol(df1))))
  } else if (nrow2 < nrow1) {
    df2 <- rbind(df2, as.data.frame(matrix(0, nrow = nrow1 - nrow2, ncol = ncol(df2))))
  }
  
  return(list(df1 = df1, df2 = df2))
}


subtract_data_frames <- function(df1, df2) {
  numeric_columns <- sapply(df1, is.numeric)
  result <- df1[, numeric_columns] - df2[, numeric_columns]
  final_result <- cbind(df1[, !numeric_columns, drop = FALSE], result)
  return(final_result)
}

load_results_data <- function(specific_results_dir) {
  studies_conn <- duckdb::dbConnect(duckdb::duckdb(), file.path(specific_results_dir, 'studies.db'))
  associations_conn <- duckdb::dbConnect(duckdb::duckdb(), file.path(specific_results_dir, 'associations.db'))
  ld_conn <- duckdb::dbConnect(duckdb::duckdb(), file.path(specific_results_dir, 'ld.db'))

  # imputed studies are not stored in the database, so we need to load them from the file
  return(list(
    studies_conn = studies_conn,
    associations_conn = associations_conn,
    ld_conn = ld_conn
  ))
}

last_two_pipeline_runs <- head(sort(list.files(results_dir, pattern = "^20", include.dirs=T), decreasing=T), 2)
latest_results <- load_results_data(latest_results_dir)
# previous_results <- load_results_data(last_two_pipeline_runs[2])
```

### What has been ingested?

Broken down by data source:
```{r data-source, echo=F}

latest_sources <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT study_sources.name, COUNT(studies.*) as num_studies
  FROM studies
  JOIN study_sources ON studies.source_id = study_sources.id
  GROUP BY study_sources.name"
)
latest_sources$num_studies <- scales::comma(latest_sources$num_studies)
knitr::kable(latest_sources)
```

Broken down by data type:
```{r data-type, echo=F}
latest_data_types <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT data_type, COUNT(*) as num_studies
  FROM studies GROUP BY data_type"
)
latest_data_types$num_studies <- scales::comma(latest_data_types$num_studies)
knitr::kable(latest_data_types)
```

### Study Breakdown 
```{r study-extractions, echo=F}
latest_study_extractions <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT 
  study_sources.name as source_name,
  studies.id as study_id,
  studies.data_type as data_type,
  study_sources.id as study_sources_id,
  study_extractions.id as study_extraction_id,
  study_extractions.min_p as min_p,
  study_extractions.cis_trans,
  study_extractions.unique_study_id as unique_study_id,
  FROM studies
  JOIN study_sources ON studies.source_id = study_sources.id
  JOIN study_extractions ON studies.id = study_extractions.study_id"
)

unique_data_types <- unique(latest_study_extractions$data_type)

study_extractions_summary <- lapply(unique_data_types, function(this_data_type) {
  study_extractions <- dplyr::filter(latest_study_extractions, data_type == this_data_type)
  num_studies <- length(unique(study_extractions$study_id))

  num_study_extractions <- nrow(study_extractions)
  num_extracted_regions <- nrow(dplyr::filter(study_extractions, grepl('_1$', unique_study_id)))

  return(data.frame(
    data_type = this_data_type,
    num_studies = scales::comma(num_studies),
    num_extracted_regions = scales::comma(num_extracted_regions),
    extracted_region_ratio = scales::number(num_extracted_regions / num_studies, accuracy = 0.01),
    num_study_extractions = scales::comma(num_study_extractions),
    finemap_ratio = scales::number(num_study_extractions / num_extracted_regions, accuracy = 0.01),
    num_cis = scales::comma(nrow(dplyr::filter(study_extractions, cis_trans == 'cis'))),
    num_trans = scales::comma(nrow(dplyr::filter(study_extractions, cis_trans == 'trans'))),
    p_between_1e4_and_1e5 = scales::comma(nrow(dplyr::filter(study_extractions, min_p >=1e-5))),
    p_between_1e5_and_1e8 = scales::comma(nrow(dplyr::filter(study_extractions, min_p < 1e-5 & min_p >= 1e-8))),
    p_under_1e8 = scales::comma(nrow(dplyr::filter(study_extractions, min_p < 1e-8)))
  ))
}) |> dplyr::bind_rows()

knitr::kable(dplyr::select(study_extractions_summary, data_type, num_studies, num_extracted_regions, extracted_region_ratio, num_study_extractions, finemap_ratio, num_cis, num_trans))
knitr::kable(dplyr::select(study_extractions_summary, data_type, p_between_1e4_and_1e5, p_between_1e5_and_1e8, p_under_1e8))
```

### Number of colocs per data type

```{r colocs-summary-by-data-type, echo=F, eval=F}
latest_coloc_results <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT
  colocalisations.*, studies.data_type, study_sources.name as source_name, study_extractions.min_p as min_p
  FROM colocalisations 
  JOIN studies ON colocalisations.study_id = studies.id
  JOIN study_sources ON studies.source_id = study_sources.id
  JOIN study_extractions ON studies.id = study_extractions.study_id"
  )

#percent of study_extractions that have colocs
unique_data_types <- unique(latest_coloc_results$data_type)

coloc_summary <- lapply(unique_data_types, function(this_data_type) {
  coloc_results <- dplyr::filter(latest_coloc_results, data_type == this_data_type)
  high_coloc_results <- dplyr::filter(coloc_results, posterior_prob > 0.8)
  mid_coloc_results <- dplyr::filter(coloc_results, posterior_prob > 0.6 & posterior_prob <= 0.8)
  num_study_extractions <- nrow(latest_study_extractions)
  num_colocs <- nrow(coloc_results)
  percent_colocs <- scales::percent(num_colocs / num_study_extractions)

  return(data.frame(
    data_type = this_data_type,
    num_study_extractions = scales::comma(num_study_extractions),
    num_colocs = scales::comma(num_colocs),
    percent_colocs = percent_colocs
  ))
}) |> dplyr::bind_rows()

knitr::kable(coloc_summary)
```

### Direction concordance?


## Breakdown by dataset:

```{r echo=F, eval=params$compare_results}
breakdown_by_dataset <- function(results) {
  unique_data_types <- unique(results$studies$data_type)

  summary_data <- lapply(unique_data_types, function(used_data_type) {
    results$studies <- dplyr::filter(results$studies, data_type == used_data_type)
    num_studies <- length(unique(results$studies$study))
    finemapped_usable <- dplyr::filter(results$studies, min_p < lowest_p_value_threshold)

    more_1e5 <- nrow(dplyr::filter(finemapped_usable, p_val_bucket == '> 1e-5'))
    more_1e6 <- nrow(dplyr::filter(finemapped_usable, p_val_bucket == '1e-5 - 1e-8'))
    less_1e8 <- nrow(dplyr::filter(finemapped_usable, p_val_bucket == '< 1e-8'))
    cis_markers <- nrow(dplyr::filter(finemapped_usable, cis_trans == 'cis'))
    trans_markers <- nrow(dplyr::filter(finemapped_usable, cis_trans == 'trans'))
    
    return(data.frame(
      data_type = used_data_type,
      studies = num_studies,
      cis_markers = cis_markers,
      trans_markers = trans_markers,
      finemapped = nrow(finemapped_usable),
      `> 1e-5` = more_1e5,
      `1e-5 - 1e-8` = more_1e6,
      `< 1e-8` = less_1e8,
      check.names = F
    ))
  }) |> dplyr::bind_rows()

  return(summary_data)
}

latest_summary_data <- breakdown_by_dataset(latest_results)
knitr::kable(latest_summary_data)
```

Compared to last pipeline run:
```{r echo=F, eval=params$compare_results}
previous_summary_data <- breakdown_by_dataset(previous_results)
knitr::kable(previous_summary_data)
# summary_data <- pad_data_frames(latest_summary_data, previous_summary_data)
# knitr::kable(subtract_data_frames(summary_data$df1, summary_data$df2))
```

# Imputed vs. Finemapped Results per dataset: 

```{r echo=F, eval=params$compare_results}
unique_data_types <- unique(latest_results$all_finemapped_results$data_type)
dont_print <- lapply(unique_data_types, function(used_data_type) {
  studies_by_data_type <- dplyr::filter(latest_results$all_finemapped_results, data_type == used_data_type)
  # TODO: add this back when ld_block is included in the results
  # dedup_studies <- studies_by_data_type |> dplyr::distinct(study, ld_block, .keep_all = TRUE)
  studies_by_data_type <- dplyr::mutate(studies_by_data_type, study_id=sub('_\\d$', '', unique_study_id))
  dedup_studies <- studies_by_data_type |> dplyr::distinct(study_id, .keep_all = TRUE)
  dedup_studies$first_finemap_num_results[is.na(dedup_studies$first_finemap_num_results)] <- 1
  hist_size <- max(dedup_studies$first_finemap_num_results)
  hist(dedup_studies$first_finemap_num_results,
    xlim = c(0,hist_size),
    breaks=seq(0,hist_size),
    main=glue::glue('{used_data_type} (first pass) finemapped results per extracted study region'),
    cex.main = 0.7
  )
})
```


```{r echo=F, eval=params$compare_results}
finemapping_breakdown <- function(results) {
  unique_data_types <- unique(results$studies$data_type)

  summary_data <- lapply(unique_data_types, function(used_data_type) {
    results$studies <- dplyr::filter(results$studies, data_type == used_data_type)
    num_studies <- length(unique(results$studies$study))
    imputed_studies <- dplyr::filter(results$studies, grepl('_1$', unique_study_id))
    finemapped_usable <- dplyr::filter(results$studies, min_p < lowest_p_value_threshold)
    
    return(data.frame(
      data_type = used_data_type,
      studies = num_studies,
      imputed = nrow(imputed_studies),
      finemapped = nrow(finemapped_usable),
      finemap_ratio = nrow(finemapped_usable) / nrow(imputed_studies),
      check.names = F
    ))
  }) |> dplyr::bind_rows()

  return(summary_data)
}
latest_finemap_breakdown <- finemapping_breakdown(latest_results)
knitr::kable(latest_finemap_breakdown)


```

Compared to last pipeline run:
```{r echo=F, eval=params$compare_results}
previous_finemap_breakdown <- finemapping_breakdown(previous_results)
knitr::kable(previous_finemap_breakdown)
# summary_data <- pad_data_frames(latest_summary_data, previous_summary_data)
# knitr::kable(subtract_data_frames(summary_data$df1, summary_data$df2))
```

# Number of hyprcoloc clusters by posterior_prob (across all studies)
```{r echo=F, eval=params$compare_results}

hyprcoloc_clusters <- function(results) {
  more_than_50 <- dplyr::filter(results$raw_coloc_results, posterior_prob > 0.5)
  more_than_60 <- dplyr::filter(results$raw_coloc_results, posterior_prob > 0.6)
  more_than_70 <- dplyr::filter(results$raw_coloc_results, posterior_prob > 0.7)
  more_than_80 <- dplyr::filter(results$raw_coloc_results, posterior_prob > 0.8)
  more_than_90 <- dplyr::filter(results$raw_coloc_results, posterior_prob > 0.9)

  clusters <- tibble::tribble(~coloc_threshold, ~num_results, ~mean_studies, ~median_studies, ~max_studies,
                ' > 0.5', nrow(more_than_50), mean(more_than_50$num_traits), median(more_than_50$num_traits), max(more_than_50$num_traits),
                ' > 0.6', nrow(more_than_60), mean(more_than_60$num_traits), median(more_than_60$num_traits), max(more_than_60$num_traits),
                ' > 0.7', nrow(more_than_70), mean(more_than_70$num_traits), median(more_than_70$num_traits), max(more_than_70$num_traits),
                ' > 0.8', nrow(more_than_80), mean(more_than_80$num_traits), median(more_than_80$num_traits), max(more_than_80$num_traits),
                ' > 0.9', nrow(more_than_90), mean(more_than_90$num_traits), median(more_than_90$num_traits), max(more_than_90$num_traits)
  )
  return(clusters)
}

latest_results$raw_coloc_results$num_traits <- stringr::str_count(latest_results$raw_coloc_results$traits, ',') + 1
max_traits <- max(latest_results$raw_coloc_results$num_traits)
hist(latest_results$raw_coloc_results$num_traits, xlim = c(1,25), breaks=seq(1,max_traits), main='Frequency of # of studies in hyprcoloc result')

latest_clusters <- hyprcoloc_clusters(latest_results)
knitr::kable(latest_clusters)
```

Compared to last pipeline run:
```{r echo=F, eval=params$compare_results}
previous_clusters <- hyprcoloc_clusters(previous_results)
knitr::kable(previous_clusters)
# summary_data <- pad_data_frames(latest_clusters, previous_clusters)
# knitr::kable(subtract_data_frames(summary_data$df1, summary_data$df2))
```

# Coloc Results Broken Down p-val threshold and data source:
```{r echo=F, eval=params$compare_results}

additional_coloc_columns <- function(results) {
  study_a_match <- match(results$coloc_results$unique_study_a, results$studies$unique_study_id)
  study_b_match <- match(results$coloc_results$unique_study_b, results$studies$unique_study_id)

  results$coloc_results$source_a <- results$studies$source[study_a_match]
  results$coloc_results$source_b <- results$studies$source[study_b_match]
  results$coloc_results$data_type_a <- results$studies$data_type[study_a_match]
  results$coloc_results$data_type_b <- results$studies$data_type[study_b_match]
  results$coloc_results$cis_trans_a <- results$studies$cis_trans[study_a_match]
  results$coloc_results$cis_trans_b <- results$studies$cis_trans[study_b_match]
  results$coloc_results$tissue_a <- results$studies$tissue[study_a_match]
  results$coloc_results$tissue_b <- results$studies$tissue[study_b_match]
  results$coloc_results$p_val_bucket_a <- results$studies$p_val_bucket[study_a_match]
  results$coloc_results$p_val_bucket_b <- results$studies$p_val_bucket[study_b_match]
  results$coloc_results$gene_a <- results$all_study_results$known_gene[match(results$coloc_results$unique_study_a, results$all_study_results$unique_study_id)]
  results$coloc_results$gene_b <- results$all_study_results$known_gene[match(results$coloc_results$unique_study_b, results$all_study_results$unique_study_id)]

  return(results$coloc_results)

}

latest_coloc_results <- additional_coloc_columns(latest_results)
previous_coloc_results <- additional_coloc_columns(previous_results)
```

```{r echo=F, eval=params$compare_results}
coloc_by_p_val_and_type <- function(coloc_results) {
  coloc_split_p_val <- split(coloc_results,
    interaction(coloc_results$p_val_bucket_a, coloc_results$p_val_bucket_b)
  )

  coloc_split_p_val_results <- lapply(names(coloc_split_p_val), function(bucket_name) {
    bucket <- coloc_split_p_val[[bucket_name]]
    pheno_pheno <- dplyr::filter(bucket, data_type_a == 'phenotype' & data_type_b == 'phenotype')
    pheno_ge <- dplyr::filter(bucket, data_type_a == 'gene_expression' & data_type_b == 'phenotype')
    pheno_protein <- dplyr::filter(bucket, data_type_a == 'protein' & data_type_b == 'phenotype')
    ge_ge <- dplyr::filter(bucket, data_type_a == 'gene_expression' & data_type_b == 'gene_expression')
    protein_ge <- dplyr::filter(bucket, data_type_a == 'gene_expression' & data_type_b == 'protein')
    protein_protein <- dplyr::filter(bucket, data_type_a == 'protein' & data_type_b == 'protein')

    return(data.frame(
      p_val = bucket_name,
      pheno_pheno = nrow(pheno_pheno),
      pheno_gene_exp = nrow(pheno_ge),
      pheno_protein = nrow(pheno_protein),
      gene_exp_gene_exp = nrow(ge_ge),
      protein_ge = nrow(protein_ge),
      protein_protein = nrow(protein_protein)
    ))
  }) |> dplyr::bind_rows()

  return(coloc_split_p_val_results)
}

latest_coloc_split_results <- coloc_by_p_val_and_type(latest_coloc_results)
knitr::kable(latest_coloc_split_results)
```

Compared to last pipeline run:
```{r echo=F, eval=params$compare_results}
previous_coloc_split_results <- coloc_by_p_val_and_type(previous_coloc_results)
knitr::kable(previous_coloc_split_results)
# summary_data <- pad_data_frames(latest_coloc_split_results, previous_coloc_split_results)
# knitr::kable(subtract_data_frames(summary_data$df1, summary_data$df2))
```

# Coloc results broken down by cis, trans, or phenotype
```{r echo=F, eval=params$compare_results}

coloc_cis_trans <- function(coloc_results) {
  coloc_for_cis_trans <- coloc_results
  coloc_for_cis_trans$cis_trans_a[is.na(coloc_for_cis_trans$cis_trans_a)] <- 'phenotype'
  coloc_for_cis_trans$cis_trans_b[is.na(coloc_for_cis_trans$cis_trans_b)] <- 'phenotype'
  coloc_for_cis_trans$cis_trans_a_order <- pmin(coloc_for_cis_trans$cis_trans_a, coloc_for_cis_trans$cis_trans_b)  # Smaller value first
  coloc_for_cis_trans$cis_trans_b_order <- pmax(coloc_for_cis_trans$cis_trans_a, coloc_for_cis_trans$cis_trans_b)  # Smaller value first
  coloc_split_cis_trans <- split(coloc_for_cis_trans,
    interaction(coloc_for_cis_trans$cis_trans_a_order, coloc_for_cis_trans$cis_trans_b_order)
  )

  coloc_split_cis_trans_results <- lapply(names(coloc_split_cis_trans), function(bucket_name) {
    bucket <- coloc_split_cis_trans[[bucket_name]]

    return(data.frame(
      cis_trans_pair = bucket_name,
      num_pairs= nrow(bucket)
    ))
  }) |> dplyr::bind_rows()

  return(dplyr::filter(coloc_split_cis_trans_results, num_pairs > 0) |>
    dplyr::arrange(num_pairs, decreasing=T)
  )
}
latest_coloc_cis_trans <- coloc_cis_trans(latest_coloc_results)
knitr::kable(latest_coloc_cis_trans)
```

Compared to last pipeline run:
```{r echo=F, eval=params$compare_results}
previous_coloc_cis_trans <- coloc_cis_trans(previous_coloc_results)
knitr::kable(previous_coloc_cis_trans)
# summary_data <- pad_data_frames(latest_coloc_cis_trans, previous_coloc_cis_trans)
# knitr::kable(subtract_data_frames(summary_data$df1, summary_data$df2))
```

# Pipeline Metadata

```{r echo=F, eval=params$compare_results}
metadata <- latest_results$results_metadata |> dplyr::select(number_extracted, number_standardised, number_imputed, number_finemapped, num_finemap_failed)
```

sum of results across all LD blocks
```{r echo=F, eval=params$compare_results}
knitr::kable(colSums(metadata))
```
mean of results per LD block
```{r echo=F, eval=params$compare_results}
knitr::kable(colMeans(metadata))
hist_size <- max(metadata$number_finemapped)
hist(metadata$number_finemapped, xlim = c(1,hist_size), breaks=seq(1,hist_size), main='Finemapped results per LD Region')
```

Regions worst finemapping rate
```{r echo=F, eval=params$compare_results}
results_subset <- dplyr::select(latest_results$results_metadata, ld_block, number_finemapped, finemapped_per_imputed, num_finemap_failed) |>
  dplyr::arrange(dplyr::desc(num_finemap_failed / number_finemapped))

knitr::kable(head(results_subset, n=10))
```
