---
title: "Current State of GP Map Data"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_fragment
---


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6, dpi = 300) 
library(ggplot2)
library(scales)
library(tidyr)
cb_blue <- "#0072B2"
cb_light_blue <- "#80B9D9"
cb_orange <- "#E69F00"
cb_light_orange <- "#F3CF80"
```

```{r load-data, echo=F}
source('constants.R')
ld_blocks <- vroom::vroom('data/ld_blocks.tsv', show_col_types = F)
ld_info <- construct_ld_block(ld_blocks$ancestry, ld_blocks$chr, ld_blocks$start, ld_blocks$stop) |>
  dplyr::filter(dir.exists(ld_block_data))

pad_data_frames <- function(df1, df2) {
  df1 <- as.data.frame(df1)
  df2 <- as.data.frame(df2)
  nrow1 <- nrow(df1)
  ncol1 <- ncol(df1)
  nrow2 <- nrow(df2)
  ncol2 <- ncol(df2)
  
  # Create column names for consistency
  all_columns <- union(names(df1), names(df2))
  
  # Add missing columns filled with 0
  df1 <- cbind(df1, setNames(as.data.frame(matrix(0, nrow = nrow1, ncol = length(setdiff(all_columns, names(df1))))), 
                             setdiff(all_columns, names(df1))))
  df2 <- cbind(df2, setNames(as.data.frame(matrix(0, nrow = nrow2, ncol = length(setdiff(all_columns, names(df2))))), 
                             setdiff(all_columns, names(df2))))
  
  # Reorder columns to match
  df1 <- df1[, all_columns]
  df2 <- df2[, all_columns]
  
  # Add missing rows filled with 0
  if (nrow1 < nrow2) {
    df1 <- rbind(df1, as.data.frame(matrix(0, nrow = nrow2 - nrow1, ncol = ncol(df1))))
  } else if (nrow2 < nrow1) {
    df2 <- rbind(df2, as.data.frame(matrix(0, nrow = nrow1 - nrow2, ncol = ncol(df2))))
  }
  
  return(list(df1 = df1, df2 = df2))
}


subtract_data_frames <- function(df1, df2) {
  numeric_columns <- sapply(df1, is.numeric)
  result <- df1[, numeric_columns] - df2[, numeric_columns]
  final_result <- cbind(df1[, !numeric_columns, drop = FALSE], result)
  return(final_result)
}


pad_data_frames <- function(df1, df2) {
  df1 <- as.data.frame(df1)
  df2 <- as.data.frame(df2)
  nrow1 <- nrow(df1)
  ncol1 <- ncol(df1)
  nrow2 <- nrow(df2)
  ncol2 <- ncol(df2)
  
  # Create column names for consistency
  all_columns <- union(names(df1), names(df2))
  
  # Add missing columns filled with 0
  df1 <- cbind(df1, setNames(as.data.frame(matrix(0, nrow = nrow1, ncol = length(setdiff(all_columns, names(df1))))), 
                             setdiff(all_columns, names(df1))))
  df2 <- cbind(df2, setNames(as.data.frame(matrix(0, nrow = nrow2, ncol = length(setdiff(all_columns, names(df2))))), 
                             setdiff(all_columns, names(df2))))
  
  # Reorder columns to match
  df1 <- df1[, all_columns]
  df2 <- df2[, all_columns]
  
  # Add missing rows filled with 0
  if (nrow1 < nrow2) {
    df1 <- rbind(df1, as.data.frame(matrix(0, nrow = nrow2 - nrow1, ncol = ncol(df1))))
  } else if (nrow2 < nrow1) {
    df2 <- rbind(df2, as.data.frame(matrix(0, nrow = nrow1 - nrow2, ncol = ncol(df2))))
  }
  
  return(list(df1 = df1, df2 = df2))
}


subtract_data_frames <- function(df1, df2) {
  numeric_columns <- sapply(df1, is.numeric)
  result <- df1[, numeric_columns] - df2[, numeric_columns]
  final_result <- cbind(df1[, !numeric_columns, drop = FALSE], result)
  return(final_result)
}

load_results_data <- function(specific_results_dir) {
  studies_db_file <- file.path(specific_results_dir, 'studies.db')

  studies_conn <- duckdb::dbConnect(duckdb::duckdb(), studies_db_file, read_only = TRUE)
  coloc_pairs_conn <- duckdb::dbConnect(duckdb::duckdb(), file.path(specific_results_dir, 'coloc_pairs_full.db'), read_only = TRUE)
  coloc_pairs_significant_conn <- duckdb::dbConnect(duckdb::duckdb(), file.path(specific_results_dir, 'coloc_pairs.db'), read_only = TRUE)
  associations_conn <- duckdb::dbConnect(duckdb::duckdb(), file.path(specific_results_dir, 'associations.db'), read_only = TRUE)
  ld_conn <- duckdb::dbConnect(duckdb::duckdb(), file.path(specific_results_dir, 'ld.db'), read_only = TRUE)

  dont_print <- DBI::dbExecute(coloc_pairs_conn, glue::glue("ATTACH '{studies_db_file}' AS studies"))

  # imputed studies are not stored in the database, so we need to load them from the file
  return(list(
    studies_conn = studies_conn,
    coloc_pairs_conn = coloc_pairs_conn,
    coloc_pairs_significant_conn = coloc_pairs_significant_conn,
    associations_conn = associations_conn,
    ld_conn = ld_conn
  ))
}

latest_results <- load_results_data(current_results_dir)
# last_two_pipeline_runs <- head(sort(list.files(results_dir, pattern = "^20", include.dirs=T), decreasing=T), 2)
# previous_results <- load_results_data(last_two_pipeline_runs[2])
```

### What has been ingested?

```{r high-level-metrics, echo=F, results='asis'}
# Select specific columns to avoid duplicate names from JOIN
num_loci <- DBI::dbGetQuery(latest_results$studies_conn, "
  SELECT study_extractions.*, 
         studies.variant_type, 
         studies.data_type
  FROM study_extractions 
  JOIN studies ON study_extractions.study_id = studies.id
")
num_common <- nrow(dplyr::filter(num_loci, min_p <= 5e-8 & variant_type == 'common'))
all_num_rare <- nrow(dplyr::filter(num_loci, min_p <= 5e-8 & variant_type != 'common'))
sig_num_rare <- nrow(dplyr::filter(num_loci, min_p <= 5e-8 & variant_type != 'common'))
num_complex <- nrow(dplyr::filter(num_loci, min_p <= 5e-8 & (data_type == 'phenotype' | data_type == 'cell_trait' | data_type == 'plasma_protein')))
num_qtl <- nrow(dplyr::filter(num_loci, min_p <= 5e-8 & (data_type == 'protein' | data_type == 'methylation' | data_type == 'splice_variant' | data_type == 'gene_expression' | data_type == 'transcript')))


coloc_pairs <- DBI::dbGetQuery(latest_results$coloc_pairs_significant_conn, "SELECT * FROM coloc_pairs")
num_coloc_pairs <- nrow(dplyr::filter(coloc_pairs, h4 >= 0.8))

coloc_groups <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT * FROM coloc_groups")
num_coloc_groups <- max(coloc_groups$coloc_group_id)
num_snps_in_coloc_groups <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT COUNT(DISTINCT snp_id) as num_snps FROM coloc_groups")
num_coloc_groups_with_same_snp <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT COUNT(*) AS repeated_snp_count
  FROM (
      SELECT snp_id
      FROM coloc_groups
      GROUP BY snp_id
      HAVING COUNT(DISTINCT coloc_group_id) > 1
  ) as repeated_snps;")

average_h4_connectedness <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT coloc_group_id, h4_connectedness FROM coloc_groups") |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::summarise(h4_connectedness = mean(h4_connectedness), .groups = "drop")
average_h4_connectedness <- mean(average_h4_connectedness$h4_connectedness)

rare_results <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT * FROM rare_results")
num_rare_results <- max(rare_results$rare_result_group_id)
```


#### Study Data

- Number distinct loci: `r scales::comma(nrow(num_loci))`
- Number distinct loci with P ≤ 5e-8: `r scales::comma(nrow(dplyr::filter(num_loci, min_p <= 5e-8)))`
- Number distinct loci with P ≤ 5e-8 and variant type common: `r scales::comma(num_common)`
- Number distinct loci with variant type rare: `r scales::comma(all_num_rare)`
- Number distinct loci with P ≤ 5e-8 and variant type rare: `r scales::comma(sig_num_rare)`

#### Coloc Data

- Number of coloc pairs: `r scales::comma(num_coloc_pairs)`
- Number of coloc groups: `r scales::comma(num_coloc_groups)`
- Average H4 connectedness: `r round(average_h4_connectedness, 3)`
- Sentinal variants from coloc group: `r scales::comma(num_snps_in_coloc_groups$num_snps)`
- SNPs that show up in multiple coloc groups: `r scales::comma(num_coloc_groups_with_same_snp$repeated_snp_count)`

```{r exit-early, echo=F, eval=TRUE}
knitr::knit_exit()
```

Broken down by data source:
```{r data-source, echo=F, fig.cap="Number of studies by data source"}

latest_sources <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT study_sources.name, COUNT(studies.*) as num_studies
  FROM studies
  JOIN study_sources ON studies.source_id = study_sources.id
  GROUP BY study_sources.name"
)

latest_sources <- latest_sources |>
  dplyr::mutate(
    ordered_name = forcats::fct_reorder(name, num_studies),
    study_label = scales::comma(num_studies)
  )

latest_sources_table <- latest_sources |>
  dplyr::select(name, num_studies)
latest_sources_table$num_studies <- scales::comma(latest_sources_table$num_studies)
knitr::kable(latest_sources_table)

# Create bar chart
p1_label_threshold <- max(latest_sources$num_studies) * 0.12

p1 <- ggplot(latest_sources, aes(x = ordered_name, y = num_studies)) +
  geom_bar(stat = "identity", fill = cb_blue, alpha = 0.7) +
  geom_text(
    aes(
      label = study_label,
      hjust = ifelse(num_studies >= p1_label_threshold, 1.05, -0.1),
      color = ifelse(num_studies >= p1_label_threshold, "inside", "outside")
    ),
    size = 3.2,
    show.legend = FALSE
  ) +
  coord_flip() +
  labs(x = "Data Source", y = "Number of Studies", title = "(A) Number of Studies by Data Source") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  scale_color_manual(values = c("inside" = "white", "outside" = "black"))

ggsave(file.path(results_analysis_dir, 'summary_a_data_source.png'), p1, width = 10, height = 6, dpi = 300)

```

### Colocalization Clustering Information:

```{r colocalization-clustering-information, echo=F}
# false positive rate and false discovery rate
false_positive_rate <- DBI::dbGetQuery(latest_results$coloc_pairs_conn, "SELECT count(*) FROM coloc_pairs where false_positive = TRUE ")
total_positive_pairs <- DBI::dbGetQuery(latest_results$coloc_pairs_conn, "SELECT count(*) FROM coloc_pairs WHERE h4 >= 0.8")
false_negative_rate <- DBI::dbGetQuery(latest_results$coloc_pairs_conn, "SELECT count(*) FROM coloc_pairs where false_negative = TRUE ")
total_negative_pairs <- DBI::dbGetQuery(latest_results$coloc_pairs_conn, "SELECT count(*) FROM coloc_pairs WHERE h4 < 0.8")

average_connectedness_score <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT coloc_group_id, h4_connectedness FROM coloc_groups") |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::summarise(h4_connectedness = mean(h4_connectedness), .groups = "drop")

# false_positive_rate_table <- false_positive_rate |>
#   dplyr::mutate(false_positive_rate = scales::percent(false_positive_rate))
# false_negative_rate_table <- false_negative_rate |>
#   dplyr::mutate(false_negative_rate = scales::percent(false_negative_rate))
# knitr::kable(false_positive_rate_table)
# knitr::kable(false_negative_rate_table)
```

### Study Breakdown by data type:
```{r data-type, echo=F}
latest_data_types <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT data_type, COUNT(*) as num_studies
  FROM studies GROUP BY data_type"
)

extractions_by_data_type <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT studies.data_type, COUNT(study_extractions.id) as num_extractions
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  GROUP BY studies.data_type"
)

common_count_by_data_type <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT studies.data_type, COUNT(study_extractions.id) as common_extractions
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  WHERE studies.variant_type = 'common'
  GROUP BY studies.data_type, studies.variant_type"
)
rare_count_by_data_type <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT studies.data_type, COUNT(study_extractions.id) as rare_extractions
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  WHERE studies.variant_type != 'common'
  GROUP BY studies.data_type, studies.variant_type"
)

latest_data_types <- latest_data_types |>
  dplyr::left_join(extractions_by_data_type, by = "data_type") |>
  dplyr::left_join(common_count_by_data_type, by = "data_type") |>
  dplyr::left_join(rare_count_by_data_type, by = "data_type") |>
  dplyr::mutate(
    common_extractions = ifelse(is.na(common_extractions), 0, common_extractions),
    rare_extractions = ifelse(is.na(rare_extractions), 0, rare_extractions)
  )

latest_data_types_table <- latest_data_types |>
  dplyr::mutate(
    num_studies = scales::comma(num_studies),
    common_extractions = scales::comma(common_extractions),
    rare_extractions = scales::comma(rare_extractions),
    num_extractions = scales::comma(num_extractions)
  )

knitr::kable(latest_data_types_table)

# Prepare data for visualization
data_types_viz <- latest_data_types |>
  dplyr::select(data_type, num_studies, num_extractions, common_extractions, rare_extractions) |>
  tidyr::pivot_longer(cols = c(num_studies, num_extractions), names_to = "metric", values_to = "count") |>
  dplyr::mutate(metric = dplyr::case_when(
    metric == "num_studies" ~ "Studies",
    metric == "num_extractions" ~ "Extractions"
  ))

# Create grouped bar chart
p2 <- ggplot(data_types_viz, aes(x = reorder(data_type, count), y = count, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  geom_text(aes(label = scales::comma(count)), 
            position = position_dodge(width = 0.9), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(x = "Data Type", y = "Count", fill = "Metric",
       title = "Study Breakdown by Data Type") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom",
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2")

ggsave(file.path(results_analysis_dir, 'data_type.png'), p2, width = 10, height = 6, dpi = 300)

# Create stacked bar chart for variant types
variant_types_viz <- latest_data_types |>
  dplyr::select(data_type, common_extractions, rare_extractions) |>
  tidyr::pivot_longer(cols = c(common_extractions, rare_extractions), 
                      names_to = "variant_type", values_to = "count") |>
  dplyr::mutate(variant_type = dplyr::case_when(
    variant_type == "common_extractions" ~ "Common",
    variant_type == "rare_extractions" ~ "Rare"
  ))

p2b <- ggplot(variant_types_viz, aes(x = reorder(data_type, count), y = count, fill = variant_type)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.8) +
  geom_text(aes(label = ifelse(count > 0, scales::comma(count), "")), 
            position = position_stack(vjust = 0.5), 
            size = 3, color = "white", fontface = "bold") +
  coord_flip() +
  labs(x = "Data Type", y = "Number of Extractions", fill = "Variant Type",
       title = "Extractions by Variant Type") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom",
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_brewer(palette = "Dark2")

ggsave(file.path(results_analysis_dir, 'data_type_variant_type.png'), p2b, width = 10, height = 6, dpi = 300)
```

```{r study-extractions-by-tissue, echo=F}
all_coloc_groups <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT coloc_groups.*, studies.data_type, study_extractions.min_p,
  snp_annotations.feature_type, snp_annotations.consequence, snp_annotations.impact, snp_annotations.cdna_position,
  snp_annotations.cds_position, snp_annotations.protein_position, snp_annotations.amino_acids, snp_annotations.codons,
  snp_annotations.gene as snp_gene, traits.trait_category as trait_category, traits.trait_name as trait_name
  FROM coloc_groups
  JOIN snp_annotations ON coloc_groups.snp_id = snp_annotations.id
  JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  WHERE study_extractions.min_p <= 5e-8
  AND coloc_groups.coloc_group_id IN (
    SELECT coloc_group_id 
    FROM coloc_groups AS cg
    JOIN study_extractions as se ON cg.study_extraction_id = se.id
    WHERE se.min_p <= 5e-8
    GROUP BY coloc_group_id 
    HAVING COUNT(*) > 1
  )
")

study_extractions_by_tissue <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT 
  study_extractions.id AS study_extraction_id,
  study_extractions.min_p,
  study_extractions.study_id,
  studies.tissue,
  studies.data_type
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  WHERE studies.tissue IS NOT NULL AND POSITION('_' IN studies.tissue) = 0 AND studies.variant_type = 'common'
")

study_extractions_by_tissue <- study_extractions_by_tissue |>
  dplyr::mutate(
    pval_bucket = dplyr::if_else(min_p <= 5e-8, "P ≤ 5e-8", "P > 5e-8"),
    tissue = dplyr::case_when(
      tissue == 'Whole blood' ~ 'Whole Blood',
      tissue == 'Blood' ~ 'Whole Blood',
      TRUE ~ tissue
    )
  )

tissue_counts <- study_extractions_by_tissue |>
  dplyr::group_by(tissue, pval_bucket) |>
  dplyr::summarise(count = dplyr::n(), .groups = "drop")

tissue_totals <- tissue_counts |>
  dplyr::group_by(tissue) |>
  dplyr::summarise(total = sum(count), .groups = "drop") |>
  dplyr::arrange(dplyr::desc(total))

tissue_order <- tissue_totals$tissue

tissue_counts <- tissue_counts |>
  dplyr::left_join(tissue_totals, by = "tissue") |>
  dplyr::mutate(
    tissue = factor(tissue, levels = rev(tissue_order)),
    pval_bucket = factor(pval_bucket, levels = c("P ≤ 5e-8", "P > 5e-8"))
  )

# Add tissue information to all_coloc_groups
all_coloc_groups_with_tissue <- all_coloc_groups |>
  dplyr::left_join(
    study_extractions_by_tissue |> dplyr::select(study_extraction_id, tissue),
    by = "study_extraction_id"
  )

coloc_group_has_phenotype <- all_coloc_groups |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::summarise(has_phenotype = any(data_type == "phenotype"), .groups = "drop")

# For coloc groups without phenotype, determine if all studies are from the same tissue
coloc_group_tissue_info <- all_coloc_groups_with_tissue |>
  dplyr::left_join(coloc_group_has_phenotype, by = "coloc_group_id") |>
  dplyr::filter(!has_phenotype) |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::summarise(
    unique_tissues = dplyr::n_distinct(tissue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    same_tissue_only = dplyr::case_when(
      unique_tissues == 0 ~ NA,  # All tissues are NA, can't determine
      unique_tissues == 1 ~ TRUE,  # All studies have the same tissue
      TRUE ~ FALSE  # Multiple different tissues
    )
  ) |>
  dplyr::select(coloc_group_id, same_tissue_only)

coloc_status_by_extraction <- all_coloc_groups |>
  dplyr::left_join(coloc_group_has_phenotype, by = "coloc_group_id") |>
  dplyr::left_join(coloc_group_tissue_info, by = "coloc_group_id") |>
  dplyr::select(study_extraction_id, has_phenotype, same_tissue_only) |>
  dplyr::distinct() |>
  dplyr::mutate(
    coloc_status = dplyr::case_when(
      has_phenotype ~ "Coloc with Phenotype",
      same_tissue_only ~ "QTL only Coloc (Same Tissue)",
      !is.na(same_tissue_only) & !same_tissue_only ~ "QTL only Coloc (Multiple Tissues)",
      TRUE ~ "QTL only Coloc (Multiple Tissues)"
    )
  ) |>
  dplyr::select(study_extraction_id, coloc_status)

tissue_coloc_status <- study_extractions_by_tissue |>
  dplyr::filter(min_p <= 5e-8) |>
  dplyr::left_join(coloc_status_by_extraction, by = "study_extraction_id") |>
  dplyr::mutate(
    tissue = factor(tissue, levels = tissue_order),
    coloc_status = dplyr::coalesce(coloc_status, "No Coloc"),
    coloc_status = factor(coloc_status, levels = c(
      "Coloc with Phenotype", 
      "QTL only Coloc (Same Tissue)", 
      "QTL only Coloc (Multiple Tissues)", 
      "No Coloc"
    ))
  )

# Calculate totals for filtered data (min_p <= 5e-8 only) for proper percentage calculation
tissue_coloc_totals <- tissue_coloc_status |>
  dplyr::group_by(tissue) |>
  dplyr::summarise(total = dplyr::n(), .groups = "drop")

# Ensure all tissues from graph (F) are included, even if they have no min_p <= 5e-8 extractions
all_tissue_coloc_combinations <- tidyr::expand_grid(
  tissue = tissue_order,
  coloc_status = c(
    "Coloc with Phenotype", 
    "QTL only Coloc (Same Tissue)", 
    "QTL only Coloc (Multiple Tissues)", 
    "No Coloc"
  )
)

tissue_coloc_summary <- tissue_coloc_status |>
  dplyr::group_by(tissue, coloc_status) |>
  dplyr::summarise(count = dplyr::n(), .groups = "drop") |>
  dplyr::right_join(all_tissue_coloc_combinations, by = c("tissue", "coloc_status")) |>
  dplyr::mutate(count = tidyr::replace_na(count, 0)) |>
  dplyr::left_join(tissue_coloc_totals, by = "tissue") |>
  dplyr::mutate(
    total = tidyr::replace_na(total, 0),
    percentage = dplyr::if_else(total > 0, count / total, 0)
  ) |>
  dplyr::mutate(tissue = factor(tissue, levels = rev(tissue_order)))

p_tissue_counts <- ggplot(tissue_counts, aes(x = reorder(tissue, total), y = count, fill = pval_bucket)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE), alpha = 0.85) +
  coord_flip() +
  labs(
    x = "Tissue",
    y = "Count",
    fill = "",
    title = "(F) Number of Ingested Loci by Tissue"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    axis.text = element_text(color = "black"),
    axis.text.y = element_text(size = 7),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(
    values = c(
      "P ≤ 5e-8" = cb_blue,
      "P > 5e-8" = cb_orange 
    ),
    breaks = c("P ≤ 5e-8", "P > 5e-8")
  )

p_tissue_coloc <- ggplot(tissue_coloc_summary, aes(x = tissue, y = percentage, fill = coloc_status)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE), alpha = 0.85) +
  coord_flip() +
  labs(
    x = "Tissue",
    y = "Percentage",
    fill = "",
    title = "(G) Colocalization Membership by Tissue"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(hjust = 0.5, size = 10),
    legend.text = element_text(hjust = 0.5, size = 8),
    axis.text = element_text(color = "black"),
    axis.text.y = element_text(size = 7),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.25)) +
  scale_fill_manual(values = c(
    "Coloc with Phenotype" = cb_blue,
    "QTL only Coloc (Same Tissue)" = cb_light_orange,
    "QTL only Coloc (Multiple Tissues)" = cb_orange,
    "No Coloc" = cb_light_blue
  ))

ggsave(file.path(results_analysis_dir, 'summary_f_study_extractions_by_tissue_pvalue.png'), p_tissue_counts, width = 10, height = 6, dpi = 300)
ggsave(file.path(results_analysis_dir, 'summary_g_study_extractions_by_tissue_coloc.png'), p_tissue_coloc, width = 10, height = 6, dpi = 300)
```

```{r study-extractions, echo=F}
latest_study_extractions <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT 
  study_sources.name as source_name,
  studies.id as study_id,
  studies.data_type as data_type,
  studies.variant_type as variant_type,
  study_sources.id as study_sources_id,
  study_extractions.id as study_extraction_id,
  study_extractions.min_p as min_p,
  study_extractions.cis_trans,
  study_extractions.unique_study_id as unique_study_id
  FROM studies
  JOIN study_sources ON studies.source_id = study_sources.id
  JOIN study_extractions ON studies.id = study_extractions.study_id"
)

unique_data_types <- unique(latest_study_extractions$data_type)

study_extractions_summary <- lapply(unique_data_types, function(this_data_type) {
  study_extractions <- dplyr::filter(latest_study_extractions, data_type == this_data_type)
  num_studies <- length(unique(study_extractions$study_id))

  num_study_extractions <- nrow(study_extractions)
  num_extracted_regions <- nrow(dplyr::filter(study_extractions, grepl('_1$', unique_study_id)))

  return(data.frame(
    data_type = this_data_type,
    num_extracted_regions = num_extracted_regions,
    num_study_extractions = num_study_extractions,
    finemap_ratio = if (num_extracted_regions > 0) num_study_extractions / num_extracted_regions else NA_real_,
    num_cis = as.integer(nrow(dplyr::filter(study_extractions, cis_trans == 'cis'))),
    num_trans = as.integer(nrow(dplyr::filter(study_extractions, cis_trans == 'trans'))),
    p_over_5e8= as.integer(nrow(dplyr::filter(study_extractions, min_p > 5e-8))),
    p_under_5e8 = as.integer(nrow(dplyr::filter(study_extractions, min_p <= 5e-8)))
  ))
}) |> dplyr::bind_rows()

totals_row <- study_extractions_summary |>
  dplyr::summarise(
    dplyr::across(where(is.integer), ~ as.integer(sum(.x, na.rm = TRUE)))
  ) |>
  dplyr::mutate(
    data_type = "Total",
    finemap_ratio = ifelse(num_extracted_regions > 0,
                           num_study_extractions / num_extracted_regions,
                           NA_real_)
  ) |>
  dplyr::select(any_of(names(study_extractions_summary)))

study_extractions_summary <- dplyr::bind_rows(study_extractions_summary, totals_row)

study_extractions_summary_table <- study_extractions_summary |>
  dplyr::mutate(
    finemap_ratio = scales::number(finemap_ratio, accuracy = 0.01),
    dplyr::across(where(is.integer), scales::comma)
  )

knitr::kable(dplyr::select(study_extractions_summary_table, data_type, finemap_ratio, num_cis, num_trans))
knitr::kable(dplyr::select(study_extractions_summary_table, data_type, p_over_5e8, p_under_5e8))

# Prepare data for cis/trans visualization
cis_trans_viz <- study_extractions_summary |>
  dplyr::select(data_type, num_cis, num_trans) |>
  tidyr::pivot_longer(cols = c(num_cis, num_trans), names_to = "type", values_to = "count") |>
  dplyr::mutate(type = dplyr::case_when(
    type == "num_cis" ~ "Cis",
    type == "num_trans" ~ "Trans"
  ))

# Create cis/trans bar chart
p3a <- ggplot(cis_trans_viz, aes(x = reorder(data_type, count), y = count, fill = type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  geom_text(aes(label = scales::comma(count)), 
            position = position_dodge(width = 0.9), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(x = "Data Type", y = "Count", fill = "Type",
       title = "Study Extractions by Cis/Trans Classification") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom",
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Accent")

ggsave(file.path(results_analysis_dir, 'study_extractions_cis_trans.png'), p3a, width = 10, height = 6, dpi = 300)

# Prepare data for p-value threshold visualization
friendly_data_type_names <- unlist(data_type_names)

pval_variant_levels <- c(
  "Rare: 1.5e-4 < P < 5e-8",
  "Rare: P ≤ 5e-8",
  "Common: 1.5e-4 < P < 5e-8",
  "Common: P ≤ 5e-8"
)

pval_variant_viz <- latest_study_extractions |>
  dplyr::mutate(
    variant_group = dplyr::if_else(variant_type == "common", "Common", "Rare"),
    pval_bucket = dplyr::case_when(
      min_p <= 5e-8 ~ "P ≤ 5e-8",
      min_p > 5e-8 & min_p < 1.5e-4 ~ "1.5e-4 < P < 5e-8",
      TRUE ~ NA_character_
    )
  ) |>
  dplyr::filter(!is.na(pval_bucket)) |>
  dplyr::mutate(
    pval_variant_category = paste(variant_group, pval_bucket, sep = ": "),
    data_type_label = dplyr::recode(data_type, !!!friendly_data_type_names, .default = data_type)
  ) |>
  dplyr::group_by(data_type_label, pval_variant_category) |>
  dplyr::summarise(count = dplyr::n(), .groups = "drop")

pval_variant_totals <- pval_variant_viz |>
  dplyr::group_by(data_type_label) |>
  dplyr::summarise(total = sum(count), .groups = "drop")

pval_variant_viz <- pval_variant_viz |>
  dplyr::left_join(pval_variant_totals, by = "data_type_label") |>
  dplyr::mutate(
    pval_variant_category = factor(
      pval_variant_category,
      levels = pval_variant_levels
    )
  )

# Create p-value threshold bar chart
p3b <- ggplot(pval_variant_viz, aes(x = reorder(data_type_label, total), y = count, fill = pval_variant_category)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.85) +
  coord_flip() +
  labs(x = "Data Type", y = "Count", fill = "",
       title = "(B) Number of Ingested Loci by Data Type") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom",
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(
    values = c(
      "Common: P ≤ 5e-8" = cb_blue,
      "Common: 1.5e-4 < P < 5e-8" = cb_light_blue,
      "Rare: P ≤ 5e-8" = cb_orange,
      "Rare: 1.5e-4 < P < 5e-8" = cb_light_orange
    ),
    breaks = pval_variant_levels,
    guide = guide_legend(reverse = TRUE)
  )

ggsave(file.path(results_analysis_dir, 'summary_b_study_extractions_p_value_threshold.png'), p3b, width = 10, height = 6, dpi = 300)

# Prepare data for p-value threshold visualization by trait category (phenotype only)
phenotype_pval_variant_viz <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT 
  study_extractions.min_p as min_p,
  studies.variant_type as variant_type,
  traits.trait_category as trait_category
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  WHERE studies.data_type = 'phenotype'
") |>
  dplyr::mutate(
    variant_group = dplyr::if_else(variant_type == "common", "Common", "Rare"),
    pval_bucket = dplyr::case_when(
      min_p <= 5e-8 ~ "P ≤ 5e-8",
      min_p > 5e-8 & min_p < 1.5e-4 ~ "1.5e-4 < P < 5e-8",
      TRUE ~ NA_character_
    )
  ) |>
  dplyr::filter(!is.na(pval_bucket) & !is.na(trait_category)) |>
  dplyr::mutate(
    pval_variant_category = paste(variant_group, pval_bucket, sep = ": ")
  ) |>
  dplyr::group_by(trait_category, pval_variant_category) |>
  dplyr::summarise(count = dplyr::n(), .groups = "drop")

phenotype_pval_variant_totals <- phenotype_pval_variant_viz |>
  dplyr::group_by(trait_category) |>
  dplyr::summarise(total = sum(count), .groups = "drop")

# Get count for "Common: P ≤ 5e-8" for ordering
phenotype_common_5e8_counts <- phenotype_pval_variant_viz |>
  dplyr::filter(pval_variant_category == "Common: P ≤ 5e-8") |>
  dplyr::select(trait_category, common_5e8_count = count)

# Note: trait category ordering will be set later based on molecular colocalization percentage
# This is done in the phenotype_coloc_rates_by_trait_category section to reuse the logic

# Finemap ratio visualization
finemap_viz <- study_extractions_summary |>
  dplyr::filter(!is.na(finemap_ratio) & data_type != "summary") |>
  dplyr::select(data_type, finemap_ratio)

if (nrow(finemap_viz) > 0) {
  p3c <- ggplot(finemap_viz, aes(x = reorder(data_type, finemap_ratio), y = finemap_ratio)) +
    geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
    geom_text(aes(label = scales::number(finemap_ratio, accuracy = 0.01)), 
              hjust = -0.1, size = 3.5) +
    coord_flip() +
    labs(x = "Data Type", y = "Finemap Ratio", 
         title = "Finemap Ratio by Data Type") +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          axis.text = element_text(color = "black"),
          panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank()) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
  
  ggsave(file.path(results_analysis_dir, 'study_extractions_finemap_ratio.png'), p3c, width = 10, height = 6, dpi = 300)
}
```

### Number of coloc pairs with H4 > 0.8 per data type

```{r coloc_pairs_summary_by_data_type, echo=F}
latest_coloc_pair_results <- DBI::dbGetQuery(latest_results$coloc_pairs_conn, 'SELECT
  coloc_pairs.*,
  -- Study extraction A details
  study_extraction_a.unique_study_id AS study_extraction_a_unique_study_id,
  study_extraction_a.min_p AS min_p_a,
  study_extraction_a.cis_trans AS cis_trans_a,
  -- Study A details
  study_a.study_name AS study_name_a,
  study_a.id AS study_id_a,
  study_a.gene AS gene_a,
  study_a.tissue AS tissue_a,
  study_a.data_type AS data_type_a,
  trait_a.trait_category AS trait_category_a,
  -- Study extraction B details
  study_extraction_b.unique_study_id AS study_extraction_b_unique_study_id,
  study_extraction_b.min_p AS min_p_b,
  study_extraction_b.cis_trans AS cis_trans_b,
  -- Study B details
  study_b.study_name AS study_name_b,
  study_b.id AS study_id_b,
  study_b.gene AS gene_b,
  study_b.tissue AS tissue_b,
  study_b.data_type AS data_type_b,
  trait_b.trait_category AS trait_category_b
  FROM coloc_pairs
  -- Join for study extraction A
  JOIN studies.study_extractions AS study_extraction_a ON coloc_pairs.study_extraction_a_id = study_extraction_a.id
  JOIN studies.studies AS study_a ON study_extraction_a.study_id = study_a.id
  JOIN studies.traits AS trait_a ON study_a.trait_id = trait_a.id
  -- Join for study extraction B
  JOIN studies.study_extractions AS study_extraction_b ON coloc_pairs.study_extraction_b_id = study_extraction_b.id
  JOIN studies.studies AS study_b ON study_extraction_b.study_id = study_b.id
  JOIN studies.traits AS trait_b ON study_b.trait_id = trait_b.id
  WHERE coloc_pairs.h4 > 0.6 OR coloc_pairs.h3 > 0.8
')

all_coloc_groups <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT coloc_groups.*, studies.data_type, study_extractions.min_p,
  snp_annotations.feature_type, snp_annotations.consequence, snp_annotations.impact, snp_annotations.cdna_position,
  snp_annotations.cds_position, snp_annotations.protein_position, snp_annotations.amino_acids, snp_annotations.codons,
  snp_annotations.gene as snp_gene, traits.trait_category as trait_category, traits.trait_name as trait_name
  FROM coloc_groups
  JOIN snp_annotations ON coloc_groups.snp_id = snp_annotations.id
  JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  WHERE study_extractions.min_p <= 5e-8
  AND coloc_groups.coloc_group_id IN (
    SELECT coloc_group_id 
    FROM coloc_groups AS cg
    JOIN study_extractions as se ON cg.study_extraction_id = se.id
    WHERE se.min_p <= 5e-8
    GROUP BY coloc_group_id 
    HAVING COUNT(*) > 1
  )
")

coloc_pair_summary_by_p_val <- function(p_val_threshold) {
  coloc_summary <- lapply(unique_data_types, function(this_data_type) {
    data_type_study_extractions <- dplyr::filter(latest_study_extractions, data_type == this_data_type & variant_type == 'common' & min_p < p_val_threshold)
    coloc_results <- dplyr::filter(latest_coloc_pair_results, h4 > 0.8 & (data_type_a == this_data_type | data_type_b == this_data_type))

    unique_coloc_study_extractions <- unique(c(coloc_results$study_extraction_a_id, coloc_results$study_extraction_b_id))
    unique_coloc_group_study_extractions <- unique(all_coloc_groups$study_extraction_id)

    num_coloc_pairs_not_in_coloc_group <- length(setdiff(unique_coloc_study_extractions, unique_coloc_group_study_extractions))
    num_study_extractions_not_in_coloc <- length(setdiff(data_type_study_extractions$study_extraction_id, unique_coloc_study_extractions))

    num_study_extractions <- nrow(data_type_study_extractions)

    unique_studies <- unique(data_type_study_extractions$study_id)

    unique_coloc_studies <- unique(c(coloc_results$study_id_a, coloc_results$study_id_b))
    num_studies_not_in_coloc <- length(setdiff(data_type_study_extractions$study_id, unique_coloc_studies))
    num_studies <- length(unique_studies)

    return(data.frame(
      data_type = this_data_type,
      study_extractions = num_study_extractions,
      study_extractions_not_in_coloc = num_study_extractions_not_in_coloc,
      study_extractions_in_coloc = num_study_extractions - num_study_extractions_not_in_coloc,
      percent_extractions_not_in_coloc_pairs = if (num_study_extractions > 0) num_study_extractions_not_in_coloc / num_study_extractions else NA_real_,
      coloc_pairs_not_in_coloc_group = num_coloc_pairs_not_in_coloc_group,
      unique_coloc_study_extractions = length(unique_coloc_study_extractions),
      percent_coloc_pairs_not_in_coloc_group = num_coloc_pairs_not_in_coloc_group / length(unique_coloc_study_extractions),
      studies = num_studies,
      studies_not_in_colocs = num_studies_not_in_coloc,
      studies_in_colocs = num_studies - num_studies_not_in_coloc,
      percent_studies_not_in_colocs = if (num_studies > 0) num_studies_not_in_coloc / num_studies else NA_real_,
      coloc_pairs = nrow(coloc_results)
    ))
  }) |> dplyr::bind_rows()

  totals_row <- coloc_summary |>
    dplyr::summarise(
      study_extractions = sum(study_extractions, na.rm = TRUE),
      study_extractions_not_in_coloc = sum(study_extractions_not_in_coloc, na.rm = TRUE),
      study_extractions_in_coloc = sum(study_extractions_in_coloc, na.rm = TRUE),
      unique_coloc_study_extractions = sum(unique_coloc_study_extractions, na.rm = TRUE),
      coloc_pairs_not_in_coloc_group = sum(coloc_pairs_not_in_coloc_group, na.rm = TRUE),
      percent_coloc_pairs_not_in_coloc_group = sum(coloc_pairs_not_in_coloc_group, na.rm = TRUE) / sum(unique_coloc_study_extractions),
      studies = sum(studies, na.rm = TRUE),
      studies_not_in_colocs = sum(studies_not_in_colocs, na.rm = TRUE),
      studies_in_colocs = sum(studies_in_colocs, na.rm = TRUE),
      coloc_pairs = sum(coloc_pairs, na.rm = TRUE)
    ) |>
    dplyr::mutate(
      data_type = "summary",
      percent_extractions_not_in_coloc_pairs = if (study_extractions > 0) study_extractions_not_in_coloc / study_extractions else NA_real_,
      percent_studies_not_in_colocs = if (studies > 0) studies_not_in_colocs / studies else NA_real_,
      percent_coloc_pairs_not_in_coloc_group = sum(coloc_pairs_not_in_coloc_group, na.rm = TRUE) / sum(unique_coloc_study_extractions)
    ) |>
    dplyr::select(any_of(names(coloc_summary)))

  coloc_summary <- dplyr::bind_rows(coloc_summary, totals_row) |>
    dplyr::mutate(
      study_extractions = scales::comma(study_extractions),
      study_extractions_not_in_coloc = scales::comma(study_extractions_not_in_coloc),
      study_extractions_in_coloc = scales::comma(study_extractions_in_coloc),
      percent_extractions_not_in_coloc_pairs = scales::percent(percent_extractions_not_in_coloc_pairs),
      studies = scales::comma(studies),
      studies_not_in_colocs = scales::comma(studies_not_in_colocs),
      studies_in_colocs = scales::comma(studies_in_colocs),
      percent_studies_not_in_colocs = scales::percent(percent_studies_not_in_colocs),
      percent_coloc_pairs_not_in_coloc_group = scales::percent(percent_coloc_pairs_not_in_coloc_group),
      coloc_pairs = scales::comma(coloc_pairs)
    )

  return(coloc_summary)
}

significant_coloc_summary <- coloc_pair_summary_by_p_val(5e-8)
coloc_summary_important_rows <- dplyr::select(significant_coloc_summary, data_type, study_extractions, coloc_pairs, study_extractions_not_in_coloc, study_extractions_in_coloc, percent_extractions_not_in_coloc_pairs, percent_coloc_pairs_not_in_coloc_group)

knitr::kable(coloc_summary_important_rows, caption = "Breakdown for results for 5e-8 p-value threshold")

# Get numeric version before formatting for visualization
significant_coloc_summary_numeric <- coloc_pair_summary_by_p_val(5e-8)
# Recalculate without formatting
coloc_summary_viz <- lapply(unique_data_types, function(this_data_type) {
  data_type_study_extractions <- dplyr::filter(latest_study_extractions, data_type == this_data_type & variant_type == 'common' & min_p < 5e-8)
  coloc_results <- dplyr::filter(latest_coloc_pair_results, h4 > 0.8 & (data_type_a == this_data_type | data_type_b == this_data_type))
  
  unique_coloc_study_extractions <- unique(c(coloc_results$study_extraction_a_id, coloc_results$study_extraction_b_id))
  unique_coloc_group_study_extractions <- unique(all_coloc_groups$study_extraction_id)
  
  num_coloc_pairs_not_in_coloc_group <- length(setdiff(unique_coloc_study_extractions, unique_coloc_group_study_extractions))
  num_study_extractions_not_in_coloc <- length(setdiff(data_type_study_extractions$study_extraction_id, unique_coloc_study_extractions))
  
  num_study_extractions <- nrow(data_type_study_extractions)
  
  return(data.frame(
    data_type = this_data_type,
    study_extractions = num_study_extractions,
    study_extractions_not_in_coloc = num_study_extractions_not_in_coloc,
    study_extractions_in_coloc = num_study_extractions - num_study_extractions_not_in_coloc,
    coloc_pairs = nrow(coloc_results),
    coloc_pairs_not_in_coloc_group = num_coloc_pairs_not_in_coloc_group
  ))
}) |> dplyr::bind_rows()

# Create visualization for study extractions in/not in coloc
extractions_coloc_viz <- coloc_summary_viz |>
  dplyr::select(data_type, study_extractions_not_in_coloc, study_extractions_in_coloc) |>
  tidyr::pivot_longer(cols = c(study_extractions_not_in_coloc, study_extractions_in_coloc), 
                      names_to = "coloc_status", values_to = "count") |>
  dplyr::mutate(coloc_status = dplyr::case_when(
    coloc_status == "study_extractions_not_in_coloc" ~ "Not in Coloc",
    coloc_status == "study_extractions_in_coloc" ~ "In Coloc"
  ))

p4a <- ggplot(extractions_coloc_viz, aes(x = reorder(data_type, count), y = count, fill = coloc_status)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.8) +
  geom_text(aes(label = ifelse(count > 0, scales::comma(count), "")), 
            position = position_stack(vjust = 0.5), 
            size = 3, color = "white", fontface = "bold") +
  coord_flip() +
  labs(x = "Data Type", y = "Number of Study Extractions", fill = "Coloc Status",
       title = "Study Extractions: In Coloc vs Not in Coloc (H4 > 0.8, P ≤ 5×10⁻⁸)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom",
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("In Coloc" = cb_blue, "Not in Coloc" = cb_orange))

ggsave(file.path(results_analysis_dir, 'study_extractions_in_coloc_vs_not_in_coloc.png'), p4a, width = 10, height = 6, dpi = 300)

# Create visualization for coloc pairs
p4b <- ggplot(coloc_summary_viz, aes(x = reorder(data_type, coloc_pairs), y = coloc_pairs)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = scales::comma(coloc_pairs)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(x = "Data Type", y = "Number of Coloc Pairs (H4 > 0.8)", 
       title = "Coloc Pairs by Data Type") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1)))

ggsave(file.path(results_analysis_dir, 'coloc_pairs_by_data_type.png'), p4b, width = 10, height = 6, dpi = 300)
```

### Breakdown of phenotype coloc rates by trait category
```{r phenotype_coloc_rates_by_trait_category, echo=F}
all_phenotype_study_extractions <- DBI::dbGetQuery(latest_results$studies_conn, "SELECT study_extractions.*,
  snp_annotations.feature_type, snp_annotations.consequence, snp_annotations.impact, snp_annotations.cdna_position,
  snp_annotations.cds_position, snp_annotations.protein_position, snp_annotations.amino_acids, snp_annotations.codons,
  snp_annotations.gene as snp_gene, traits.trait_name as trait_name, traits.trait_category as trait_category
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  JOIN snp_annotations ON study_extractions.snp_id = snp_annotations.id
  WHERE studies.data_type = 'phenotype' and studies.variant_type = 'common' and study_extractions.min_p <= 5e-8
")


trait_categories_list <- unique(all_coloc_groups$trait_category[all_coloc_groups$data_type == "phenotype"]) |>
  stats::na.omit()

# Get numeric version for visualization (before percentage formatting)
gwas_hits_numeric <- lapply(trait_categories_list, function(tc) {
  all_study_extractions_in_coloc_groups <- unique(all_coloc_groups$study_extraction_id)
  all_phenotype_study_extractions_for_tc <- dplyr::filter(all_phenotype_study_extractions, trait_category == tc)
  
  all_phenotype_study_extractions_without_coloc <- all_phenotype_study_extractions_for_tc |>
    dplyr::filter(!id %in% all_study_extractions_in_coloc_groups)
  
  num_studies <- length(unique(all_phenotype_study_extractions_for_tc$study_id))
  
  coloc_groups_with_tc <- all_coloc_groups |>
    dplyr::filter(trait_category == tc) |>
    dplyr::pull(coloc_group_id) |>
    unique()
  
  study_extractions_with_phenotype_only_coloc <- all_coloc_groups |>
    dplyr::filter(coloc_group_id %in% coloc_groups_with_tc) |>
    dplyr::group_by(coloc_group_id) |>
    dplyr::filter(all(data_type == "phenotype")) |>
    dplyr::ungroup() |>
    dplyr::filter(trait_category == tc & data_type == "phenotype")
  num_study_extractions_with_phenotype_only_coloc <- length(unique(study_extractions_with_phenotype_only_coloc$study_extraction_id))
  
  study_extractions_molecular_coloc <- all_coloc_groups |>
    dplyr::filter(coloc_group_id %in% coloc_groups_with_tc) |>
    dplyr::group_by(coloc_group_id) |>
    dplyr::filter(any(data_type != "phenotype")) |>
    dplyr::ungroup() |>
    dplyr::filter(data_type == "phenotype" & trait_category == tc)
  num_study_extractions_molecular_coloc <- length(unique(study_extractions_molecular_coloc$study_extraction_id))

  all_significant_non_molecular_coloc <- num_study_extractions_with_phenotype_only_coloc
  total_extractions <- nrow(all_phenotype_study_extractions_for_tc)
  
  return(data.frame(
    trait_category = tc,
    num_study_extractions = total_extractions,
    num_study_extractions_without_molecular_coloc = all_significant_non_molecular_coloc,
    num_study_extractions_molecular_coloc = num_study_extractions_molecular_coloc,
    num_study_extractions_no_coloc = nrow(all_phenotype_study_extractions_without_coloc),
    num_study_extractions_total = total_extractions,
    percent_without_molecular_coloc = all_significant_non_molecular_coloc / total_extractions,
    percent_with_molecular_coloc = num_study_extractions_molecular_coloc / total_extractions,
    percent_with_no_coloc = nrow(all_phenotype_study_extractions_without_coloc) / total_extractions
  ))
}) |> dplyr::bind_rows()

gwas_totals_row <- gwas_hits_numeric |>
  dplyr::summarise(
    trait_category = "Total",
    num_study_extractions = sum(num_study_extractions),
    num_study_extractions_without_molecular_coloc = sum(num_study_extractions_without_molecular_coloc),
    num_study_extractions_molecular_coloc = sum(num_study_extractions_molecular_coloc),
    num_study_extractions_no_coloc = sum(num_study_extractions_no_coloc),
    num_study_extractions_total = sum(num_study_extractions_total),
    percent_without_molecular_coloc = num_study_extractions_without_molecular_coloc / num_study_extractions_total,
    percent_with_molecular_coloc = num_study_extractions_molecular_coloc / num_study_extractions_total,
    percent_with_no_coloc = num_study_extractions_no_coloc / num_study_extractions_total
  )

gwas_hits_table <- gwas_hits_numeric |>
  dplyr::mutate(
    num_study_extractions = scales::comma(num_study_extractions),
    num_study_extractions_without_molecular_coloc = scales::comma(num_study_extractions_without_molecular_coloc),
    num_study_extractions_molecular_coloc = scales::comma(num_study_extractions_molecular_coloc),
    num_study_extractions_no_coloc = scales::comma(num_study_extractions_no_coloc),
    num_study_extractions_total = scales::comma(num_study_extractions_total),
    percent_without_molecular_coloc = scales::percent(percent_without_molecular_coloc, accuracy = 0.1),
    percent_with_molecular_coloc = scales::percent(percent_with_molecular_coloc, accuracy = 0.1),
    percent_with_no_coloc = scales::percent(percent_with_no_coloc, accuracy = 0.1)
  )

gwas_total_table <- gwas_totals_row |>
  dplyr::mutate(
    num_study_extractions = scales::comma(num_study_extractions),
    num_study_extractions_without_molecular_coloc = scales::comma(num_study_extractions_without_molecular_coloc),
    num_study_extractions_molecular_coloc = scales::comma(num_study_extractions_molecular_coloc),
    num_study_extractions_no_coloc = scales::comma(num_study_extractions_no_coloc),
    num_study_extractions_total = scales::comma(num_study_extractions_total),
    percent_without_molecular_coloc = scales::percent(percent_without_molecular_coloc, accuracy = 0.1),
    percent_with_molecular_coloc = scales::percent(percent_with_molecular_coloc, accuracy = 0.1),
    percent_with_no_coloc = scales::percent(percent_with_no_coloc, accuracy = 0.1)
  )

gwas_hits_table <- dplyr::bind_rows(gwas_hits_table, gwas_total_table) |>
  dplyr::select(trait_category, num_study_extractions, percent_without_molecular_coloc, percent_with_molecular_coloc, percent_with_no_coloc)

knitr::kable(gwas_hits_table, caption = "Breakdown of phenotype info by trait category")

# Create shared trait category ordering based on highest percentage with molecular colocalization
# This ordering will be used across all trait category graphs (C), (D), and (E) for consistency
shared_trait_category_order <- gwas_hits_numeric |>
  dplyr::arrange(dplyr::desc(percent_with_molecular_coloc)) |>
  dplyr::pull(trait_category) |>
  unique()

# Create visualization
gwas_viz <- gwas_hits_numeric |>
  dplyr::select(trait_category, percent_with_no_coloc, percent_without_molecular_coloc, percent_with_molecular_coloc) |>
  tidyr::pivot_longer(cols = c(percent_with_no_coloc, percent_without_molecular_coloc, percent_with_molecular_coloc), 
                      names_to = "coloc_type", values_to = "percentage") |>
  dplyr::mutate(
    coloc_type = dplyr::case_when(
      coloc_type == "percent_with_no_coloc" ~ "No Coloc",
      coloc_type == "percent_with_molecular_coloc" ~ "With Molecular Coloc",
      coloc_type == "percent_without_molecular_coloc" ~ "Without Molecular Coloc"
    ),
    coloc_type = factor(coloc_type, levels = c("No Coloc", "Without Molecular Coloc", "With Molecular Coloc")),
    trait_category = factor(trait_category, levels = rev(shared_trait_category_order))
  )

p5 <- ggplot(gwas_viz, aes(x = trait_category, y = percentage, fill = coloc_type)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.8) +
  geom_text(aes(label = scales::percent(percentage, accuracy = 0.1)),
            position = position_stack(vjust = 0.5),
            size = 3, color = "white") +
  coord_flip() +
  labs(x = "", y = "Percentage", fill = "Colocalization Type",
       title = "(E) Clustered Molecular Colocalization Rates") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.justification = "center",
    legend.box.just = "center",
    legend.box.margin = margin(t = 5, r = 0, b = 5, l = 0),
    legend.title = element_text(hjust = 0.5, size = 10),
    legend.text = element_text(hjust = 0.5, size = 8),
    axis.text = element_text(color = "black"),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot",
    plot.margin = margin(t = 15, r = 30, b = 30, l = 15)
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.25)) +
  scale_fill_manual(values = c(
    "With Molecular Coloc" = cb_blue,
    "Without Molecular Coloc" = cb_light_blue,
    "No Coloc" = cb_orange
  )) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5))


ggsave(file.path(results_analysis_dir, 'summary_e_phenotype_coloc_rates_by_trait_category.png'), p5, width = 10, height = 6, dpi = 300)
```

```{r phenotype_study_extractions_by_trait_category_c, echo=F}
# Create p-value threshold bar chart by trait category
# Use shared trait category ordering based on molecular colocalization percentage from section (E)
phenotype_pval_variant_viz_ordered <- phenotype_pval_variant_viz |>
  dplyr::mutate(
    pval_variant_category = factor(
      pval_variant_category,
      levels = pval_variant_levels
    ),
    trait_category = factor(trait_category, levels = rev(shared_trait_category_order))
  )

p3b_trait <- ggplot(phenotype_pval_variant_viz_ordered, aes(x = trait_category, y = count, fill = pval_variant_category)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.85) +
  coord_flip() +
  labs(x = "Trait Category", y = "Count", fill = "",
       title = "(C) Number of Ingested Loci by Trait Category") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom",
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(
    values = c(
      "Common: P ≤ 5e-8" = cb_blue,
      "Common: 1.5e-4 < P < 5e-8" = cb_light_blue,
      "Rare: P ≤ 5e-8" = cb_orange,
      "Rare: 1.5e-4 < P < 5e-8" = cb_light_orange
    ),
    breaks = pval_variant_levels,
    guide = guide_legend(reverse = TRUE)
  )

ggsave(file.path(results_analysis_dir, 'summary_c_phenotype_study_extractions_p_value_threshold_by_trait_category.png'), p3b_trait, width = 10, height = 6, dpi = 300)
```

### Genome-wide phenotype hits with strong H3/H4 support

```{r phenotype_h3_h4_rates_by_trait_category, echo=F}
phenotype_molecular_coloc_pairs_a <- latest_coloc_pair_results |>
  dplyr::filter(
    data_type_a == "phenotype" & min_p_a <= 5e-8 & data_type_b != "phenotype" & min_p_b <= 5e-8
  ) |>
  dplyr::mutate(
    phenotype_study_extraction_id = study_extraction_a_id,
    phenotype_trait_category = trait_category_a
  )

full_phenotype_molecular_coloc_pairs_a <- latest_coloc_pair_results |>
  dplyr::filter(
    data_type_a == "phenotype" & min_p_a <= 5e-8 & data_type_b != "phenotype"
  ) |>
  dplyr::mutate(
    phenotype_study_extraction_id = study_extraction_a_id,
    phenotype_trait_category = trait_category_a
  )

phenotype_molecular_coloc_pairs_b <- latest_coloc_pair_results |>
  dplyr::filter(
    data_type_b == "phenotype" & min_p_b <= 5e-8 & data_type_a != "phenotype" & min_p_a <= 5e-8
  ) |>
  dplyr::mutate(
    phenotype_study_extraction_id = study_extraction_b_id,
    phenotype_trait_category = trait_category_b
  )

full_phenotype_molecular_coloc_pairs_b <- latest_coloc_pair_results |>
  dplyr::filter(
    data_type_b == "phenotype" & min_p_b <= 5e-8 & data_type_a != "phenotype"
  ) |>
  dplyr::mutate(
    phenotype_study_extraction_id = study_extraction_b_id,
    phenotype_trait_category = trait_category_b
  )


full_h3_h4_by_hit <- dplyr::bind_rows(full_phenotype_molecular_coloc_pairs_a, full_phenotype_molecular_coloc_pairs_b) |>
  dplyr::group_by(phenotype_study_extraction_id, phenotype_trait_category) |>
  dplyr::summarise(
    has_h3 = any(h3 > 0.8, na.rm = TRUE),
    has_h4 = any(h4 > 0.8, na.rm = TRUE),
    .groups = "drop"
  )

h3_h4_by_hit <- dplyr::bind_rows(phenotype_molecular_coloc_pairs_a, phenotype_molecular_coloc_pairs_b) |>
  dplyr::group_by(phenotype_study_extraction_id, phenotype_trait_category) |>
  dplyr::summarise(
    has_h3 = any(h3 > 0.8, na.rm = TRUE),
    has_h4 = any(h4 > 0.8, na.rm = TRUE),
    .groups = "drop"
  )

phenotype_hits_with_probs <- all_phenotype_study_extractions |>
  dplyr::select(id, trait_category, study_id, trait_name) |>
  dplyr::left_join(h3_h4_by_hit, by = c("id" = "phenotype_study_extraction_id", "trait_category" = "phenotype_trait_category")) |>
  dplyr::mutate(
    has_h3 = tidyr::replace_na(has_h3, FALSE),
    has_h4 = tidyr::replace_na(has_h4, FALSE),
    h4_support= dplyr::case_when(
      has_h4 ~ "H4 > 0.8",
      TRUE ~ "No H4 > 0.8"
    ),
    h3_support= dplyr::case_when(
      has_h3 ~ "H3 > 0.8",
      TRUE ~ "No H3 > 0.8"
    )
  ) |>
  dplyr::filter(!is.na(trait_category))

h3_h4_support_numeric <- phenotype_hits_with_probs |>
  dplyr::group_by(trait_category) |>
  dplyr::summarise(
    total_hits = dplyr::n(),
    hits_h4 = sum(has_h4),
    hits_h3 = sum(has_h3),
    .groups = "drop"
  )

h3_h4_table_main <- h3_h4_support_numeric |>
  dplyr::arrange(dplyr::desc(hits_h4)) |>
  dplyr::mutate(
    pct_h4_support = scales::percent(hits_h4 / total_hits, accuracy = 0.1),
    pct_h3_support = scales::percent(hits_h3 / total_hits, accuracy = 0.1),
    total_hits = scales::comma(total_hits),
    hits_h4 = scales::comma(hits_h4),
    hits_h3 = scales::comma(hits_h3)
  )

h3_h4_totals_row <- h3_h4_support_numeric |>
  dplyr::summarise(
    trait_category = "Total",
    total_hits = sum(total_hits),
    hits_h4 = sum(hits_h4),
    hits_h3 = sum(hits_h3),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pct_h4_support = scales::percent(hits_h4 / total_hits, accuracy = 0.1),
    pct_h3_support = scales::percent(hits_h3 / total_hits, accuracy = 0.1),
    total_hits = scales::comma(total_hits),
    hits_h4 = scales::comma(hits_h4),
    hits_h3 = scales::comma(hits_h3)
  )

h3_h4_table <- dplyr::bind_rows(h3_h4_table_main, h3_h4_totals_row)

knitr::kable(h3_h4_table, caption = "Genome-wide phenotype hits with strong colocalization support by trait category")

h3_h4_full_hits <- all_phenotype_study_extractions |>
  dplyr::select(id, trait_category, study_id, trait_name) |>
  dplyr::left_join(full_h3_h4_by_hit, by = c("id" = "phenotype_study_extraction_id", "trait_category" = "phenotype_trait_category")) |>
  dplyr::mutate(
    has_h3 = tidyr::replace_na(has_h3, FALSE),
    has_h4 = tidyr::replace_na(has_h4, FALSE)
  ) |>
  dplyr::filter(!is.na(trait_category))

h3_h4_full_support_numeric <- h3_h4_full_hits |>
  dplyr::group_by(trait_category) |>
  dplyr::summarise(
    total_hits = dplyr::n(),
    hits_h4 = sum(has_h4),
    hits_h3 = sum(has_h3),
    .groups = "drop"
  )

h3_h4_full_table_main <- h3_h4_full_support_numeric |>
  dplyr::arrange(dplyr::desc(hits_h4)) |>
  dplyr::mutate(
    pct_h4_support = scales::percent(hits_h4 / total_hits, accuracy = 0.1),
    pct_h3_support = scales::percent(hits_h3 / total_hits, accuracy = 0.1),
    total_hits = scales::comma(total_hits),
    hits_h4 = scales::comma(hits_h4),
    hits_h3 = scales::comma(hits_h3)
  )

h3_h4_full_totals_row <- h3_h4_full_support_numeric |>
  dplyr::summarise(
    trait_category = "Total",
    total_hits = sum(total_hits),
    hits_h4 = sum(hits_h4),
    hits_h3 = sum(hits_h3),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pct_h4_support = scales::percent(hits_h4 / total_hits, accuracy = 0.1),
    pct_h3_support = scales::percent(hits_h3 / total_hits, accuracy = 0.1),
    total_hits = scales::comma(total_hits),
    hits_h4 = scales::comma(hits_h4),
    hits_h3 = scales::comma(hits_h3)
  )

h3_h4_full_table <- dplyr::bind_rows(h3_h4_full_table_main, h3_h4_full_totals_row)

knitr::kable(h3_h4_full_table, caption = "Genome-wide phenotype hits with any H3/H4 support (full threshold) by trait category")

h3_h4_plot_data <- h3_h4_support_numeric |>
  dplyr::mutate(
    pct_h4 = hits_h4 / total_hits,
    pct_h3 = hits_h3 / total_hits
  ) |>
  dplyr::filter(!is.na(trait_category))

# Use shared trait category ordering for consistency across all graphs (C), (D), and (E)
# This ordering is based on molecular colocalization percentage from section (E)
shared_trait_order <- shared_trait_category_order[shared_trait_category_order %in% h3_h4_plot_data$trait_category]
if (length(shared_trait_order) == 0) {
  shared_trait_order <- sort(unique(h3_h4_plot_data$trait_category))
}

h3_h4_viz <- h3_h4_plot_data |>
  dplyr::mutate(trait_category = factor(trait_category, levels = rev(shared_trait_order))) |>
  tidyr::pivot_longer(
    cols = c(pct_h4, pct_h3),
    names_to = "support_type",
    values_to = "percentage"
  ) |>
  dplyr::mutate(
    support_type = dplyr::recode(
      support_type,
      pct_h4 = "H4 > 0.8",
      pct_h3 = "H3 > 0.8"
    ),
    support_type = factor(support_type, levels = c("H4 > 0.8", "H3 > 0.8"))
  )

p_h3h4 <- ggplot(h3_h4_viz, aes(x = trait_category, y = percentage, fill = support_type)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7, alpha = 0.8) +
  geom_text(aes(label = scales::percent(percentage, accuracy = 0.1)),
            position = position_dodge(width = 0.75),
            hjust = -0.1, size = 2, color = "black") +
  coord_flip() +
  labs(
    x = "",
    y = "Percentage",
    fill = "Colocalization Type",
    title = "(D) Pairwise Molecular Colocalization Rates"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.justification = "center",
    legend.box.just = "center",
    legend.box.margin = margin(t = 5, r = 0, b = 5, l = 0),
    legend.title = element_text(hjust = 0.5, size = 10),
    legend.text = element_text(hjust = 0.5, size = 8),
    axis.text = element_text(color = "black"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot",
    plot.margin = margin(t = 15, r = 30, b = 30, l = 15)
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.25)) +
  scale_fill_manual(values = c(
    "H4 > 0.8" = cb_orange,
    "H3 > 0.8" = cb_blue
  )) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5))

ggsave(file.path(results_analysis_dir, 'summary_d_phenotype_h3_h4_rates_by_trait_category.png'), p_h3h4, width = 10, height = 6, dpi = 300)
```

## Matrix of coloc results by data type and by cis-trans:

```{r matrix_of_coloc_results, echo=F}
coloc_matrix <- dplyr::filter(latest_coloc_pair_results, h4 > 0.8)

coloc_matrix_by_type <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(coloc_matrix_by_type) <- unique_data_types
colnames(coloc_matrix_by_type) <- unique_data_types

coloc_matrix_by_cis_trans <- matrix(0, nrow = 3, ncol = 3)
rownames(coloc_matrix_by_cis_trans) <- c('cis', 'trans', 'phenotype')
colnames(coloc_matrix_by_cis_trans) <- c('cis', 'trans', 'phenotype')

for (i in 1:nrow(coloc_matrix)) {
  type_a <- coloc_matrix$data_type_a[i]
  type_b <- coloc_matrix$data_type_b[i]

  coloc_matrix_by_type[type_a, type_b] <- coloc_matrix_by_type[type_a, type_b] + 1
  if (type_a != type_b) {
    coloc_matrix_by_type[type_b, type_a] <- coloc_matrix_by_type[type_b, type_a] + 1
  }

  cis_trans_a <- coloc_matrix$cis_trans_a[i]
  if (is.na(cis_trans_a)) cis_trans_a <- 'phenotype'

  cis_trans_b <- coloc_matrix$cis_trans_b[i]
  if (is.na(cis_trans_b)) cis_trans_b <- 'phenotype'
  
  coloc_matrix_by_cis_trans[cis_trans_a, cis_trans_b] <- coloc_matrix_by_cis_trans[cis_trans_a, cis_trans_b] + 1
  if (cis_trans_a != cis_trans_b) {
    coloc_matrix_by_cis_trans[cis_trans_b, cis_trans_a] <- coloc_matrix_by_cis_trans[cis_trans_b, cis_trans_a] + 1
  }
}

coloc_matrix_by_type[lower.tri(coloc_matrix_by_type)] <- NA
coloc_matrix_by_type <- as.data.frame(coloc_matrix_by_type)
coloc_matrix_by_type[, unique_data_types] <- lapply(coloc_matrix_by_type[, unique_data_types], scales::comma)

knitr::kable(coloc_matrix_by_type, caption = "Number of colocalizations (H4 > 0.8) between data types")

coloc_matrix_by_cis_trans[lower.tri(coloc_matrix_by_cis_trans)] <- NA
coloc_matrix_by_cis_trans <- as.data.frame(coloc_matrix_by_cis_trans)
coloc_matrix_by_cis_trans[, c('cis', 'trans', 'phenotype')] <- lapply(coloc_matrix_by_cis_trans[, c('cis', 'trans', 'phenotype')], scales::comma)

knitr::kable(as.data.frame(coloc_matrix_by_cis_trans), caption = "Number of colocalizations (H4 > 0.8) between cis-trans")
```

## Number of coloc pairs by h4 threshold (across all studies)
```{r coloc_pairs_by_h4_threshold, echo=F}

coloc_pairs <- function(results) {
  more_than_60 <- dplyr::filter(results, h4 > 0.60)
  more_than_65 <- dplyr::filter(results, h4 > 0.65)
  more_than_75 <- dplyr::filter(results, h4 > 0.75)
  more_than_80 <- dplyr::filter(results, h4 > 0.80)
  more_than_85 <- dplyr::filter(results, h4 > 0.85)
  more_than_90 <- dplyr::filter(results, h4 > 0.90)
  more_than_95 <- dplyr::filter(results, h4 > 0.95)
  more_than_99 <- dplyr::filter(results, h4 > 0.99)

  clusters <- tibble::tribble(~coloc_threshold, ~num_results, 
    ' > 0.99', nrow(more_than_99),
    ' > 0.95', nrow(more_than_95),
    ' > 0.90', nrow(more_than_90),
    ' > 0.85', nrow(more_than_85),
    ' > 0.80', nrow(more_than_80),
    ' > 0.75', nrow(more_than_75),
    ' > 0.65', nrow(more_than_65),
    ' > 0.60', nrow(more_than_60)
  )
  return(clusters)
}

latest_coloc_pairs <- coloc_pairs(latest_coloc_pair_results)
latest_coloc_pairs_table <- latest_coloc_pairs
latest_coloc_pairs_table$num_results <- scales::comma(latest_coloc_pairs_table$num_results)
knitr::kable(latest_coloc_pairs_table)

# Create bar chart
p7 <- ggplot(latest_coloc_pairs, aes(x = reorder(coloc_threshold, num_results), y = num_results)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = scales::comma(num_results)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(x = "H4 Threshold", y = "Number of Coloc Pairs", 
       title = "Number of Coloc Pairs by H4 Threshold (Across All Studies)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text = element_text(color = "black"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1)))

ggsave(file.path(results_analysis_dir, 'coloc_pairs_by_h4_threshold.png'), p7, width = 10, height = 6, dpi = 300)
```
