---
title: "BMI tissue-partitioned MR, example"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
library(data.table)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(UpSetR)
#pak::pak("MRCIEU/gpmapr") # update 
library(gpmapr)
library(TwoSampleMR)
library(MVMR)
library(ieugwasr)

```


```{r, include=FALSE}

out.dir <- "/Users/rj18633/OneDrive - University of Bristol/postdoc/gp_map/results/vignettes/repo_rerun/"

```


Search BMI....

```{r, include=TRUE} 
search <- search_options_api()
search <- search$search_terms
bmi_deets <- search[which(search$type_id == 1992),]

print(bmi_deets)
```

Query BMI... 

```{r search_bmi, include=TRUE}
bmi <-  trait(1992, include_associations = TRUE, include_coloc_pairs = TRUE)  


## study extractions
bmi_se <- bmi$study_extractions

## coloc groups 
bmi_cg <- bmi$coloc_groups

## pairwise coloc H3/H4 
bmi_cp <- bmi$coloc_pairs

## bmi_assocs  
bmi_effs <- bmi_cg[which(bmi_cg$study_id == 1992),]  



```


Annotate the coloc_pairs results with trait names...

coloc_pairs shows the pairwise coloc results for BMI with other traits mapped by study_extraction id. Map these to trait names using the coloc_groups dataframe to make a more intuitive summary of the paired coloc results. 

- Identify the study_extraction_ids for BMI
- Filter coloc_pairs to just the pairwise analyses with BMI (mapped with study_extraction ids)
- Complete mapping the remaining study_extraction IDs in the pairwise results 

```{r bmi_se_ids, include=TRUE}

ids <- bmi_effs$study_extraction_id

print(paste0("Number of BMI loci which underwent coloc analysis: ", NROW(unique(ids))))


bmi_coloc_res <- bmi_cp %>%
  dplyr::filter(
    study_extraction_a_id %in% ids |
      study_extraction_b_id %in% ids
  )


## annotate the coloc results with trait names 
annotate_coloc <- function(coloc_res, coloc_groups) {
  # join study_extraction_a_id
  coloc_res <- coloc_res %>%
    left_join(
      coloc_groups %>%
        select(study_extraction_id, trait_name, source_name, tissue) %>%
        rename(trait_a = trait_name,
              source_a = source_name),
      by = c("study_extraction_a_id" = "study_extraction_id")
    )
  
  # join study_extraction_b_id
  coloc_res <- coloc_res %>%
    left_join(
      coloc_groups %>%
        select(study_extraction_id, trait_name, source_name, tissue) %>%
        rename(trait_b = trait_name,
              source_b = source_name),
      by = c("study_extraction_b_id" = "study_extraction_id")
    )
  
  return(coloc_res)
}

bmi_coloc_annotated <- annotate_coloc(bmi_coloc_res, bmi_cg)

colnames(bmi_coloc_annotated)[c(11,14)] <- c("tissue_a", "tissue_b")


```


Check out the tissue eQTL profile of all BMI coloc results...  

```{r bmi_tissues, include=TRUE}

## grab gtex, eqtlGen, brain emeta and brain Meta 
srcs <- c("GTEx Atlas", "Brain eMeta","Brain Meta", "eQTLGen")

tissue_res <- bmi_coloc_annotated[which(bmi_coloc_annotated$source_a %in% srcs | 
                                       bmi_coloc_annotated$source_b %in% srcs), ]



## filter sqtls 
tissue_res <- tissue_res %>%
  filter(!apply(across(everything(), as.character), 1,
                function(row) any(grepl("sQTL", row))))


### tidy
## harmonise trait a/b 
dat1 <- tissue_res %>%
  mutate(swap = trait_b == "Body mass index") %>% # update trait name
  mutate(
    study_extraction_a_id_new = if_else(swap, study_extraction_b_id, study_extraction_a_id),
    study_extraction_b_id_new = if_else(swap, study_extraction_a_id, study_extraction_b_id),
    trait_a_new = if_else(swap, trait_b, trait_a),
    trait_b_new = if_else(swap, trait_a, trait_b),
    source_a_new = if_else(swap, source_b, source_a),
    source_b_new = if_else(swap, source_a, source_b),
    tissue_a_new  = if_else(swap, tissue_b, tissue_a),
    tissue_b_new  = if_else(swap, tissue_a, tissue_b)
  ) %>%
  select(-swap, -study_extraction_a_id, -study_extraction_b_id, -trait_a, -trait_b, -source_a, -source_b, -tissue_a, -tissue_b) %>%
  rename(
    study_extraction_a_id = study_extraction_a_id_new,
    study_extraction_b_id = study_extraction_b_id_new,
    trait_a = trait_a_new,
    trait_b = trait_b_new,
    source_a = source_a_new,
    source_b = source_b_new,
     tissue_a = tissue_a_new,
    tissue_b = tissue_b_new
  )


dat1 <- dat1[,c("snp_id","trait_a", "study_extraction_a_id", "trait_b", "study_extraction_b_id", "source_b", "tissue_b", "h3", "h4")]

dat1 <- dat1 %>%
  mutate(tissue_b = if_else(tissue_b == "Blood", "Whole Blood", tissue_b))


## merge with trait effect estimates 
effs <- bmi_effs[,c("snp_id", "study_extraction_id", "min_p", "display_snp","rsid", "beta", "se", "p", "eaf", "imputed")]


## rm rows missing summ stats 
effs <- effs %>% 
  filter(!is.na(beta))


dat2 <- merge(dat1, effs, by = c("snp_id"))

# filter cell line dat 
cells <- c("Cells EBV transformed lymphocytes", "Cells Cultured fibroblasts")
dat2 <- dat2[!(dat2$tissue_b %in% cells),] 


print(paste0("Number of BMI loci which with evidence for colocalization with a tissue eQTL: ", NROW(unique(dat2$snp_id)))) 

```


Identify tissues which contribute most.. 


```{r instruments_per_tissue, fig.width=12, fig.height=14}
# how many snps per tissue 
tbl <- dat2 %>%
  group_by(tissue_b) %>%
  summarise(n_unique_snps = n_distinct(snp_id)) %>%
  arrange(desc(n_unique_snps))

print(tbl)


## subset dat2 
tp_snps <- dat2[,c("tissue_b", "trait_b", "source_b", "snp_id", "display_snp","rsid", "study_extraction_a_id", "beta", "se", "p", "eaf", "h4")]
colnames(tp_snps) <- c("tissue", "gene", "source", "snp_id", "display_snp","rsid", "study_extraction_id_trait", "beta", "se", "p", "eaf", "h4")

```

Upset plot.. 

- illustrate to what extent SNPs/instruments are unique to a tissue.

```{r}

# data source combined 
snp_matrix <- tp_snps %>%
  select(snp_id, tissue) %>%
  distinct() %>%  # avoid duplicates
  mutate(present = 1) %>%
  pivot_wider(
    names_from = tissue,
    values_from = present,
    values_fill = list(present = 0)
  )

upset_data <- as.data.frame(snp_matrix)
rownames(upset_data) <- upset_data$snp_id
upset_data$snp_id <- NULL

# plot
upp1 <- upset(
   upset_data,
   nsets = 18,      
   nintersects = NA, 
   order.by = "freq"
 )


upp1

## summarise the subset data 
set_sizes <- colSums(upset_data)
top_sets <- names(sort(set_sizes, decreasing = TRUE))[1:18]

upset_top <- upset_data[, top_sets]
upset_top_nonzero <- upset_top[rowSums(upset_top) > 0, ]
n_distinct_intersections <- nrow(
  unique(upset_top_nonzero)
)


intersection_table <- upset_top_nonzero %>%
  as.data.frame() %>%
  count(across(everything()), name = "n_snps") %>%
  arrange(desc(n_snps))

head(intersection_table, 10)

tail_distribution <- intersection_table %>%
  count(n_snps, name = "n_intersections") %>%
  arrange(n_snps)

tail_distribution

n_intersections <- nrow(unique(upset_top_nonzero))
 
n_intersections 


#write.table(tail_distribution, paste0(out.dir, "summary_of_upset_dat.txt"), col.names = T, row.names = FALSE, quote = FALSE, sep = "\t")

# 
# tiff(
#   filename = file.path(out.dir, "bmi_upset.tiff"),
#   width = 13,
#   height = 6,
#   units = "in",
#   res = 300
# )

print(
  upset(
    upset_data,
    nsets = 18,
    #nintersects = 70,
    nintersects = n_intersections,
    order.by = "freq"
  )
)


#dev.off() 



```

Characterize the trait composition of lead tissues... 

```{r}


```

Dig in to the complex trait profile of putative tissue partitioned exposure vars.... 
- PCA analysis 

```{r}
top_tissues <- c(tbl$tissue_b[c(1:6,9)])
top_tissues

df <- tp_snps[which(tp_snps$tissue %in% top_tissues),]
expsres <- unique(df$tissue)

print(paste0("Number of BMI loci incorporated into TPMR across tissues: ", NROW(unique(df$rsid))))

snps <- unique(df$snp_id)

## get the BMI/complex trait profile for these SNPs 
pheno_dat <- bmi_coloc_annotated[which(bmi_coloc_annotated$snp_id %in% snps),]
pheno_dat <- pheno_dat %>% 
  filter(is.na(tissue_a))
srcs1 <- c("GoDMC", "eQTLGen", "UK Biobank Proteomics", "OneK1K")
pheno_dat <- pheno_dat[!(pheno_dat$source_b %in% srcs1),]
pheno_dat1 <- pheno_dat[,c(1,2,3,6,9,10,12,13)]

## harmonise traits so bmi always a: 
idx <- pheno_dat1$trait_b == "Body mass index" & pheno_dat1$trait_a != "Body mass index"
cols <- c("study_extraction", "source", "tissue", "trait")

for (x in cols) {
  a <- grep(paste0("^", x, "_a"), colnames(pheno_dat1), value = TRUE)
  b <- grep(paste0("^", x, "_b"), colnames(pheno_dat1), value = TRUE)

  tmp <- pheno_dat1[idx, a]
  pheno_dat1[idx, a] <- pheno_dat1[idx, b]
  pheno_dat1[idx, b] <- tmp
}

pheno_dat11 <- na.omit(pheno_dat1)

## rm repeat bmi traits  
rmbmi <- c("Body mass index (UKB data field 21001)", "Body mass index (BMI)")
pheno_dat11 <- pheno_dat11[!(pheno_dat11$trait_b %in% rmbmi),]

## how many unique complex traits 
NROW(unique(pheno_dat11$trait_b))  

## complex trait profile per tissue  
df2 <- df1[,c(1,2,3,4,5,6)]

## tissue by trait matrix 
df2_sub <- df2 %>%
  select(snp_id, tissue) %>%
  distinct()


tissue_trait_counts <- df2_sub %>%
  inner_join(pheno_sub, by = "snp_id") %>%
  count(tissue, trait_b) %>%
  pivot_wider(
    names_from = trait_b,
    values_from = n,
    values_fill = 0
  )


mat <- tissue_trait_counts %>%
  column_to_rownames("tissue") %>%
  as.matrix()

##  
colnames(mat)[apply(mat, 2, sd) == 0] 
mat_nz <- mat[, apply(mat, 2, sd) != 0]

## snp counts 
snp_counts <- df2_sub %>%
   distinct(tissue, snp_id) %>%
   count(tissue, name = "n_snps")

pca_res <- prcomp(mat_nz, scale. = TRUE) # tissue*trait snp count matrix scaled by tissue

loadings <- pca_res$rotation  # columns are PCs, rows are traits
head(loadings)  
contrib_PC1 <- sort(abs(loadings[,1]), decreasing = TRUE)  # top contributors to PC1
contrib_PC2 <- sort(abs(loadings[,2]), decreasing = TRUE) 
var_explained <- summary(pca_res)$importance[2, 1:2] * 100
var_explained


## plot 
pca_scores <- as.data.frame(pca_res$x) # extract tissue scores 
pca_scores$tissue <- rownames(pca_scores)

pca_scores <- pca_scores %>%
  left_join(snp_counts, by = "tissue") # add snp counts 


pc1_var2 <- round(100 * summary(pca_res)$importance[2, 1], 1)
pc2_var2 <- round(100 * summary(pca_res)$importance[2, 2], 1)


## colour palette
cb_blue <- "#0072B2"
cb_light_blue <- "#80B9D9"

cb_orange <- "#E69F00"
cb_light_orange <- "#F3CF80"


#### plot

pca_scores <- pca_scores %>%
  mutate(
    quadrant = case_when(
      PC1 >= 0 & PC2 >= 0 ~ "Q1",
      PC1 <  0 & PC2 >= 0 ~ "Q2",
      PC1 <  0 & PC2 <  0 ~ "Q3",
      PC1 >= 0 & PC2 <  0 ~ "Q4"
    )
  )

quad_cols <- c(
  Q1 = cb_blue,
  Q2 = cb_light_blue,
  Q3 = cb_orange,
  Q4 = cb_light_orange
)


pcap1 <- ggplot(
  pca_scores,
  aes(PC1, PC2, label = tissue, size = n_snps, colour = quadrant)
) +
  geom_point(alpha = 0.85) +
  geom_text_repel(
    size = 4,
    seed = 123,
    show.legend = FALSE,
    box.padding = 0.6,
    point.padding = 0.5,
    force = 1.5,
    min.segment.length = 0
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey70") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  scale_colour_manual(values = quad_cols, guide = "none") +
  scale_size_continuous(range = c(3, 8)) +
  guides(
    size = guide_legend(
      override.aes = list(
        colour = cb_blue,
        alpha = 1
      )
    )
  ) +
  theme_minimal(base_size = 14) +
  labs(
    title = "PCA of the trait profiles of tissue-stratified BMI loci",
    x = paste0("PC1 (", pc1_var2, "%)"),
    y = paste0("PC2 (", pc2_var2, "%)"),
    size = "nSNPs"
  )


pcap1

# ggsave(
#   filename = paste0(out.dir, "bmi_tpmr_trait_profile_pca_quadrants.tiff"),
#   plot = pcap1,
#   width = 9,
#   height = 6,
#   units = "in",
#   dpi = 300       
# )

```


```{r}
## tissue pleiotropy table  
snps_per_tissue <- df2_sub %>%
  distinct(tissue, snp_id)


pheno_links <- pheno_dat %>%
  select(snp_id, trait_b) %>%
  distinct()


summary <- snps_per_tissue %>%
  left_join(pheno_links, by = "snp_id") %>%
  group_by(tissue) %>%
  summarise(
    n_snps   = n_distinct(snp_id),
    n_traits = n_distinct(trait_b, na.rm = TRUE),
    .groups = "drop"
  )

summary <- summary[order(-summary$n_traits), ]
print(summary)

# write.table(summary_tbl_like, paste0(out.dir, "bmi_tpmr_pleiotropy_summary1.txt"), col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE) 

```


Trait enrichment in tissues... 
- hierarchical clustering 

```{r}
mat1 <- mat_nz

#d <- dist(mat1, method = "euclidean", diag = FALSE, upper = FALSE, p = 2)
#hclust(d, method = "complete", members = NULL)


#### 
mat_prop <- mat1 / rowSums(mat1)


trait_z <- scale(mat_prop)
p2 <- pheatmap(
  t(trait_z),
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  show_colnames = TRUE,
  show_rownames = FALSE
)

p2

## save plot 
# pheatmap(
#   t(trait_z),
#   cluster_cols = TRUE,
#   cluster_rows = TRUE,
#   show_colnames = TRUE,
#   show_rownames = FALSE,
#   filename = paste0(out.dir, "bmi_tissue_traits_z_heatmap.png"),
#   width = 8,
#   height = 10
# )
# 


#### enriched traits per tissue 
wb_z <- trait_z["Whole Blood", ]
wb_top <- sort(wb_z, decreasing = TRUE)

head(wb_top, 30)

#top_n <- 20
z_thresh <- 2

### all traits above z threshold 
top_traits_tbl <- lapply(rownames(trait_z), function(tissue) {
  z <- trait_z[tissue, ]
  
  tibble(
    tissue  = tissue,
    trait   = names(z),
    z_score = as.numeric(z)
  ) %>%
    filter(z_score > z_thresh) %>%     # filter on z score 
    arrange(desc(z_score)) #%>%
  #  slice_head(n = top_n)
}) %>%
  bind_rows()

top_traits_tbl


#write.table(top_traits_tbl, paste0(out.dir, "bmi_tissue_traits_z_rep_traits.txt"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t") 


```


#### MVMR/TPMR... 
- Select top tissues (brain and subcutaneous adipose) to evluate in TPMR 
- Format exposure dfs... 

```{r format_tissue_expsrs}

top_tissues
top_tissues <- c("Brain (Meta)", "Adipose Subcutaneous")

df <- tp_snps[which(tp_snps$tissue %in% top_tissues),]


# update effect allele coding - effect allele is second 
df$other_allele <- sub(".+ ([A-Z]+)/([A-Z]+)$", "\\1", df$display_snp)
df$effect_allele  <- sub(".+ ([A-Z]+)/([A-Z]+)$", "\\2", df$display_snp)


brain1 <-  df[which(df$tissue == top_tissues[1]),]
brain1 <- brain1[,c("tissue", "source", "snp_id", "rsid", "h4", "beta", "se", "p", "eaf", "effect_allele", "other_allele")]

## drop duplicated snps  - preferentially keep rows from Meta brain because higher power 
brain1 <- brain1 %>%
  arrange(factor(source, levels = c("Brain Meta", "Brain eMeta"))) %>% 
  distinct(snp_id, .keep_all = TRUE)



```


Format all tp-BMI exposures... 


```{r}

# reference tissue 
brain1 <- brain1 %>%
  select(tissue, snp_id, rsid, h4, beta, se, p, eaf, effect_allele, other_allele)

tissue_name <- top_tissues[2]

tissue_df <- df %>%
    filter(tissue == tissue_name) %>%
    select(tissue, snp_id, rsid, h4, beta, se, p, eaf, effect_allele, other_allele) %>%
    distinct(snp_id, .keep_all = TRUE)
  
  # harmonise: add missing snps from one to the other
  add_b1 <- tissue_df %>%
    filter(!snp_id %in% brain1$snp_id) %>%
    mutate(
      tissue = unique(brain1$tissue)[1],
      h4 = 0.001
    )
  
  brain_tmp <- bind_rows(brain1, add_b1)
  
  add_t1 <- brain_tmp %>%
    filter(!snp_id %in% tissue_df$snp_id) %>%
    mutate(
      tissue = unique(tissue_df$tissue)[1],
      h4 = 0.001
    )
  
  tissue_tmp <- bind_rows(tissue_df, add_t1)
  
  # combine & calculate weighted beta
  expdf <- bind_rows(brain_tmp, tissue_tmp) %>%
    mutate(weighted_beta = beta * h4) %>%
    select(
      Phenotype = tissue,
      SNP = rsid,
      beta = weighted_beta,
      se, pval = p, eaf, effect_allele, other_allele
    )
  


```



Run example tpmr.... 


``` {r mr_all}

  
  exp_name <- top_tissues[2]
  exp_dat <- format_data(expdf, type = 'exposure')
  
  # traits to test against
  traits <- c('ieu-a-7', 'ebi-a-GCST006414', 'ebi-a-GCST009541', 'ebi-a-GCST006906', 'finn-b-I9_PAD')
  
  
  results_file <- paste0(out.dir, "tpmr_", exp_name, "_results.txt")
  fstats_file  <- paste0(out.dir, "tpmr_", exp_name, "_fstats.txt")
  

  # loop over traits 
  for (i in traits)  {
    
    #i <- traits[1]
    
    ## MVMR
    outcome_dat <- extract_outcome_data(exp_dat$SNP, i)
    
    mvdat <- mv_harmonise_data(exp_dat, outcome_dat, harmonise_strictness = 1)
    
    ### Conditional F-stats 
    F.data <- format_mvmr(
      BXGs = mvdat$exposure_beta[, c(1, 2)],
      BYG = mvdat$outcome_beta,
      seBXGs = mvdat$exposure_se[, c(1, 2)],
      seBYG = mvdat$outcome_se,
      RSID = row.names(mvdat$exposure_beta)
    )
    
    pheno_mat <- cor(mvdat$exposure_beta[, c(1, 2)])
    seBXGs <- mvdat$exposure_se[, c(1, 2)]
    pos <- phenocov_mvmr(pheno_mat, seBXGs) 
    fstats <- strength_mvmr(r_input = F.data, gencov = pos)[1, ]
    
    res <- mv_multiple(mvdat)
    res <- res$result
    res$OR <- exp(res$b)
    res$lower_CI <- exp(res$b - (1.96 * res$se))
    res$upper_CI <- exp(res$b + (1.96 * res$se)) 
    
    #
    write.table(
      res,
      file = results_file,
      col.names = !file.exists(results_file),
      row.names = FALSE,
      quote = FALSE,
      sep = "\t",
      append = TRUE
    )
    
    write.table(
      fstats,
      file = fstats_file,
      col.names = !file.exists(fstats_file),
      row.names = FALSE,
      quote = FALSE,
      sep = "\t",
      append = TRUE
    )
  }
#}


```


Add separate analysis for T2D 

```{r mr}

#exp_dat <- format_data(expdf, type='exposure')
tmp <- load("/Volumes/MRC-IEU-research/projects/ieu2/p1/131/working/results/inclusion_criteria/outcome_gwas/T2D_Mahajan2018.RData")



# 

  outcome_dat <- out_dat[which(out_dat$SNP %in% exp_dat$SNP),]
  
  
  mvdat <- mv_harmonise_data(exp_dat, outcome_dat, harmonise_strictness = 1)
  
  ### Conditional F-stats 
  F.data<-format_mvmr(BXGs=mvdat$exposure_beta[,c(1,2)],
                      BYG=mvdat$outcome_beta,
                      seBXGs=mvdat$exposure_se[,c(1,2)],
                      seBYG=mvdat$outcome_se,
                      RSID=row.names(mvdat$exposure_beta))
  
  pheno_mat <- cor(mvdat$exposure_beta[,c(1,2)])
  seBXGs=mvdat$exposure_se[,c(1,2)]
  pos <- phenocov_mvmr(pheno_mat, seBXGs) 
  fstats <- strength_mvmr(r_input=F.data,gencov=pos)[1,]
  res <- mv_multiple(mvdat)
  
  res <- res$result
  res$OR <- exp(res$b)
  res$lower_CI <- exp(res$b -(1.96*res$se))
  res$upper_CI <- exp(res$b +(1.96*res$se)) 



 # append res
    write.table(
      res,
      file = results_file,
      col.names = !file.exists(results_file),
      row.names = FALSE,
      quote = FALSE,
      sep = "\t",
      append = TRUE
    )

    message("Appended results to: ", results_file)
        
    write.table(
      fstats,
      file = fstats_file,
      col.names = !file.exists(fstats_file),
      row.names = FALSE,
      quote = FALSE,
      sep = "\t",
      append = TRUE
    )
  


```
