---
params:
  # null or empty: run every LD block that has `dir.exists(ld_block_data)`.
  # A string, e.g. "EUR/1/93071870-95684841": run that block only (must match `ld_info$block`).
  ld_block: null
output:
  html_document:
    self_contained: true
title: "Coloc Clustering Validation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)

library(igraph)
library(ggplot2)
library(dplyr)
library(vroom)

source("../../pipeline_steps/constants.R")

# Whether to scan all on-disk LD block dirs or a single `params$ld_block`
ld_block_param <- params$ld_block
run_all_ld_blocks <- is.null(ld_block_param) ||
  (length(ld_block_param) == 0L) ||
  (is.character(ld_block_param) && length(ld_block_param) == 1L &&
    (is.na(ld_block_param) || !nzchar(trimws(ld_block_param))))
if (is.character(ld_block_param) && length(ld_block_param) > 1L) {
  stop("params$ld_block must be null (all blocks) or a single character string.")
}
```

## Overview

This report is used to validate the clustering method.  If we decide to change the clustering method, we can use this report to compare the new method to the old one.

**This run:** `r if (run_all_ld_blocks) "All LD blocks with a data directory." else paste0("Single block: ", ld_block_param, ".")`

To render only one block: `rmarkdown::render("coloc_clustering_validation.Rmd", params = list(ld_block = "EUR/1/93071870-95684841"))`.  To render all blocks, use `params = list(ld_block = NULL)` or omit `ld_block`.

Part 1: It calculates a series of graph structure metrics for the unpruned and pruned graphs.
Part 2: It compares the results based on real world target-indication pairs.

### Graph Structure Metrics
The metrics we calculate are:
1. Number of components
2. Mean weight
3. Mean transitivity
4. Mean conductance
5. Median size

We then compare the metrics for the unpruned and pruned graphs.

```{r pruned_vs_unpruned_graph_structure_metrics}
# `apply` only accepts matrices — use lapply/seq_along. Paths use ld_info$ld_block_data.
ld_blocks_tsv_path <- file.path("../../pipeline_steps/data/ld_blocks.tsv")
ld_rows <- vroom::vroom(ld_blocks_tsv_path, show_col_types = FALSE)
ld_info <- with(ld_rows, construct_ld_block(ancestry, chr, start, stop))
ld_info <- ld_info[base::dir.exists(ld_info$ld_block_data), , drop = FALSE]

if (!run_all_ld_blocks) {
  key <- trimws(ld_block_param)
  ld_info <- ld_info[ld_info$block == key, , drop = FALSE]
  if (nrow(ld_info) == 0L) {
    stop(
      "params$ld_block not found among blocks with an existing data directory: ",
      key
    )
  }
}

blocks <- ld_info$block

analyze_graph_structure_advanced <- function(g, original_graph = NULL) {
  if (is.null(g) || igraph::vcount(g) < 1L) {
    return(list(
      n_components = 0L,
      mean_weight = NA_real_,
      mean_transitivity = NA_real_,
      mean_conductance = NA_real_,
      median_size = NA_real_
    ))
  }
  comp <- components(g)
  if (comp$no < 1L) {
    return(list(
      n_components = 0L,
      mean_weight = NA_real_,
      mean_transitivity = NA_real_,
      mean_conductance = NA_real_,
      median_size = NA_real_
    ))
  }

  comp_metrics <- lapply(seq_len(comp$no), function(i) {
    v_indices <- which(comp$membership == i)
    subgraph <- induced_subgraph(g, v_indices)

    m_weight <- if (ecount(subgraph) > 0) mean(E(subgraph)$weight) else 0
    trans <- transitivity(subgraph, type = "global")
    if (is.na(trans)) trans <- 0

    cond <- NA
    if (!is.null(original_graph)) {
      node_names <- V(g)[v_indices]$name
      internal_weight <- sum(E(subgraph)$weight)
      inc_edges_list <- incident_edges(original_graph, v = node_names, mode = "all")
      all_edge_ids <- unique(unlist(inc_edges_list))
      total_original_weight <- sum(E(original_graph)[all_edge_ids]$weight)

      if (total_original_weight > 0) {
        cond <- (total_original_weight - internal_weight) / total_original_weight
      } else {
        cond <- 0
      }
    }

    data.frame(
      weight = m_weight,
      transitivity = trans,
      conductance = cond,
      size = vcount(subgraph)
    )
  })

  results_df <- do.call(rbind, comp_metrics)
  if (is.null(results_df) || nrow(results_df) == 0) {
    return(list(
      n_components = as.integer(comp$no),
      mean_weight = NA_real_,
      mean_transitivity = NA_real_,
      mean_conductance = NA_real_,
      median_size = NA_real_
    ))
  }

  list(
    n_components = comp$no,
    mean_weight = mean(results_df$weight, na.rm = TRUE),
    mean_transitivity = mean(results_df$transitivity, na.rm = TRUE),
    mean_conductance = mean(results_df$conductance, na.rm = TRUE),
    median_size = stats::median(results_df$size)
  )
}

per_block <- lapply(seq_along(blocks), function(bi) {
  block <- blocks[bi]
  rds_path <- file.path(ld_info$ld_block_data[bi], "igraph_clustered_results.rds")
  if (!file.exists(rds_path)) {
    return(NULL)
  }
  cr <- tryCatch(readRDS(rds_path), error = function(e) NULL)
  if (is.null(cr) || is.null(cr$unpruned_graph) || is.null(cr$pruned_graph)) {
    return(NULL)
  }
  u <- tryCatch(
    analyze_graph_structure_advanced(cr$unpruned_graph),
    error = function(e) NULL
  )
  p <- tryCatch(
    analyze_graph_structure_advanced(cr$pruned_graph, original_graph = cr$unpruned_graph),
    error = function(e) NULL
  )
  if (is.null(u) || is.null(p)) {
    return(NULL)
  }
  data.frame(
    ld_block = block,
    n_components_unpruned = u$n_components,
    n_components_pruned = p$n_components,
    mean_weight_unpruned = u$mean_weight,
    mean_weight_pruned = p$mean_weight,
    mean_transitivity_unpruned = u$mean_transitivity,
    mean_transitivity_pruned = p$mean_transitivity,
    mean_conductance_pruned = p$mean_conductance,
    median_size_unpruned = u$median_size,
    median_size_pruned = p$median_size,
    stringsAsFactors = FALSE
  )
})
blocks_graph_struct <- dplyr::bind_rows(Filter(Negate(is.null), per_block))

if (nrow(blocks_graph_struct) == 0) {
  knitr::kable(
    data.frame(message = "No valid igraph_clustered_results.rds found for any block."),
    caption = "Per-LD-block graph structure (empty)"
  )
} else {
  knitr::kable(
    blocks_graph_struct,
    digits = 4,
    caption = if (run_all_ld_blocks) {
      "Graph structure metrics for each LD block"
    } else {
      paste0("Graph structure metrics — ", ld_block_param)
    }
  )
}

# Totals and means across all blocks that contributed a row
if (nrow(blocks_graph_struct) > 0) {
  tot_tbl <- data.frame(
    Metric = c(
      "Unpruned components (sum)", "Pruned components (sum)",
      "Mean weight unpruned", "Mean weight pruned",
      "Mean transitivity unpruned", "Mean transitivity pruned",
      "Mean conductance (pruned clusters only, vs unpruned ref.)",
      "Median of per-block median cluster size — unpruned", "— pruned"
    ),
    Value = c(
      sum(blocks_graph_struct$n_components_unpruned, na.rm = TRUE),
      sum(blocks_graph_struct$n_components_pruned, na.rm = TRUE),
      mean(blocks_graph_struct$mean_weight_unpruned, na.rm = TRUE),
      mean(blocks_graph_struct$mean_weight_pruned, na.rm = TRUE),
      mean(blocks_graph_struct$mean_transitivity_unpruned, na.rm = TRUE),
      mean(blocks_graph_struct$mean_transitivity_pruned, na.rm = TRUE),
      mean(blocks_graph_struct$mean_conductance_pruned, na.rm = TRUE),
      stats::median(blocks_graph_struct$median_size_unpruned, na.rm = TRUE),
      stats::median(blocks_graph_struct$median_size_pruned, na.rm = TRUE)
    )
  )
  knitr::kable(
    tot_tbl,
    digits = 4,
    caption = if (run_all_ld_blocks) {
      "Aggregated results over all LD blocks in the per-block table"
    } else {
      paste0("Summary for single LD block: ", ld_block_param)
    }
  )
}

```