---
params:
  ld_block: null
  ld_blocks:
    - "EUR/1/153358890-156424168"
    - "EUR/1/158933969-163173938"
    - "EUR/3/10633999-13113221"
    - "EUR/22/44582535-46439661"
output:
  html_document:
    self_contained: true
title: "Coloc Clustering Validation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)

library(igraph)
library(ggplot2)
library(dplyr)
library(vroom)

source("../../pipeline_steps/constants.R")

# A single `ld_block` supplied at render time overrides the four default blocks.
ld_block_param <- params$ld_block
use_single_ld_block <- !(is.null(ld_block_param) ||
  (length(ld_block_param) == 0L) ||
  (is.character(ld_block_param) && length(ld_block_param) == 1L &&
    (is.na(ld_block_param) || !nzchar(trimws(ld_block_param)))))
if (is.character(ld_block_param) && length(ld_block_param) > 1L) {
  stop("params$ld_block must be null or a single character string.")
}

selected_ld_blocks <- if (use_single_ld_block) {
  trimws(ld_block_param)
} else {
  unname(unlist(params$ld_blocks, use.names = FALSE))
}

comparison_files <- data.frame(
  method = c("Original", "BFDR corrected", "Prior 1e-5 + BFDR corrected"),
  pairwise_file = c(
    "coloc_pairwise_results_5e-6_no_bfdr_1_0_0.tsv.gz",
    "coloc_pairwise_results_5e-6_bfdr_1_0_0.tsv.gz",
    "coloc_pairwise_results_1e-5_bfdr_1_0_0.tsv.gz"
  ),
  cluster_file = c(
    "clustered_results_5e-6_no_bfdr_1_0_0.tsv.gz",
    "clustered_results_5e-6_bfdr_1_0_0.tsv.gz",
    "clustered_results_1e-5_bfdr_1_0_0.tsv.gz"
  ),
  rds_file = c(
    "igraph_clustered_results_5e-6_no_bfdr_1_0_0.rds",
    "igraph_clustered_results_5e-6_bfdr_1_0_0.rds",
    "igraph_clustered_results_1e-5_bfdr_1_0_0.rds"
  ),
  bfdr_file = c(
    "global_no_bfdr_threshold_p12_5e-6.tsv",
    "global_bfdr_threshold_p12_5e-6.tsv",
    "global_bfdr_threshold_p12_1e-5.tsv"
  ),
  stringsAsFactors = FALSE
)
```

## Overview

This report is used to validate the clustering method.  If we decide to change the clustering method, we can use this report to compare the new method to the old one.

**LD blocks in this run:** `r paste(selected_ld_blocks, collapse = ", ")`

To override the default four blocks and render only one, use `rmarkdown::render("coloc_clustering_validation.Rmd", params = list(ld_block = "EUR/1/93071870-95684841"))`.

Part 1 calculates graph structure metrics for the unpruned and pruned graphs. Part 2 compares the pairwise coloc and clustering results produced by the three prior/BFDR settings.

### Graph Structure Metrics
The metrics we calculate are:
1. Number of components
2. Mean weight
3. Mean transitivity
4. Mean conductance
5. Median size

We then compare the metrics for the unpruned and pruned graphs.

```{r pruned_vs_unpruned_graph_structure_metrics}
ld_blocks_tsv_path <- file.path("../../pipeline_steps/data/ld_blocks.tsv")
ld_rows <- vroom::vroom(ld_blocks_tsv_path, show_col_types = FALSE)
ld_info <- with(ld_rows, construct_ld_block(ancestry, chr, start, stop))
ld_info <- ld_info[base::dir.exists(ld_info$ld_block_data), , drop = FALSE]
ld_info <- ld_info[ld_info$block %in% selected_ld_blocks, , drop = FALSE]

missing_ld_blocks <- setdiff(selected_ld_blocks, ld_info$block)
if (length(missing_ld_blocks) > 0L) {
  stop(
    "Selected LD block(s) not found among blocks with an existing data directory: ",
    paste(missing_ld_blocks, collapse = ", ")
  )
}

# Preserve the requested display order.
ld_info <- ld_info[match(selected_ld_blocks, ld_info$block), , drop = FALSE]
blocks <- ld_info$block

analyze_graph_structure_advanced <- function(g, original_graph = NULL) {
  if (is.null(g) || igraph::vcount(g) < 1L) {
    return(list(
      n_components = 0L,
      mean_weight = NA_real_,
      mean_transitivity = NA_real_,
      mean_conductance = NA_real_,
      median_size = NA_real_
    ))
  }
  comp <- components(g)
  if (comp$no < 1L) {
    return(list(
      n_components = 0L,
      mean_weight = NA_real_,
      mean_transitivity = NA_real_,
      mean_conductance = NA_real_,
      median_size = NA_real_
    ))
  }

  comp_metrics <- lapply(seq_len(comp$no), function(i) {
    v_indices <- which(comp$membership == i)
    subgraph <- induced_subgraph(g, v_indices)

    m_weight <- if (ecount(subgraph) > 0) mean(E(subgraph)$weight) else 0
    trans <- transitivity(subgraph, type = "global")
    if (is.na(trans)) trans <- 0

    cond <- NA
    if (!is.null(original_graph)) {
      node_names <- V(g)[v_indices]$name
      internal_weight <- sum(E(subgraph)$weight)
      inc_edges_list <- incident_edges(original_graph, v = node_names, mode = "all")
      all_edge_ids <- unique(unlist(inc_edges_list))
      total_original_weight <- sum(E(original_graph)[all_edge_ids]$weight)

      if (total_original_weight > 0) {
        cond <- (total_original_weight - internal_weight) / total_original_weight
      } else {
        cond <- 0
      }
    }

    data.frame(
      weight = m_weight,
      transitivity = trans,
      conductance = cond,
      size = vcount(subgraph)
    )
  })

  results_df <- do.call(rbind, comp_metrics)
  if (is.null(results_df) || nrow(results_df) == 0) {
    return(list(
      n_components = as.integer(comp$no),
      mean_weight = NA_real_,
      mean_transitivity = NA_real_,
      mean_conductance = NA_real_,
      median_size = NA_real_
    ))
  }

  list(
    n_components = comp$no,
    mean_weight = mean(results_df$weight, na.rm = TRUE),
    mean_transitivity = mean(results_df$transitivity, na.rm = TRUE),
    mean_conductance = mean(results_df$conductance, na.rm = TRUE),
    median_size = stats::median(results_df$size)
  )
}

per_block <- lapply(seq_along(blocks), function(bi) {
  block <- blocks[bi]

  dplyr::bind_rows(lapply(seq_len(nrow(comparison_files)), function(method_i) {
    method <- comparison_files$method[method_i]
    rds_path <- file.path(ld_info$ld_block_data[bi], comparison_files$rds_file[method_i])

    if (!file.exists(rds_path)) {
      return(NULL)
    }
    cr <- tryCatch(readRDS(rds_path), error = function(e) NULL)
    if (is.null(cr) || is.null(cr$unpruned_graph) || is.null(cr$pruned_graph)) {
      return(NULL)
    }
    u <- tryCatch(
      analyze_graph_structure_advanced(cr$unpruned_graph),
      error = function(e) NULL
    )
    p <- tryCatch(
      analyze_graph_structure_advanced(cr$pruned_graph, original_graph = cr$unpruned_graph),
      error = function(e) NULL
    )
    if (is.null(u) || is.null(p)) {
      return(NULL)
    }
    data.frame(
      ld_block = block,
      method = method,
      n_components_unpruned = u$n_components,
      n_components_pruned = p$n_components,
      mean_weight_unpruned = u$mean_weight,
      mean_weight_pruned = p$mean_weight,
      mean_transitivity_unpruned = u$mean_transitivity,
      mean_transitivity_pruned = p$mean_transitivity,
      mean_conductance_pruned = p$mean_conductance,
      median_size_unpruned = u$median_size,
      median_size_pruned = p$median_size,
      stringsAsFactors = FALSE
    )
  }))
})
blocks_graph_struct <- dplyr::bind_rows(Filter(Negate(is.null), per_block))

if (nrow(blocks_graph_struct) == 0) {
  knitr::kable(
    data.frame(message = "No valid igraph clustered results found for any selected block or method."),
    caption = "Per-LD-block graph structure (empty)"
  )
} else {
  knitr::kable(
    blocks_graph_struct,
    digits = 4,
    caption = "Graph structure metrics for each LD block and clustering analysis"
  )
}
```

### Stability across prior and BFDR settings

These comparisons treat each analysis as the reference in turn. They are intended to show whether changing the coloc prior and applying global BFDR materially changes the retained coloc pairs or final clustering.

```{r compare_clustering_result_sets}
evaluate_clustering <- function(true_membership, predicted_membership) {
  if (length(true_membership) != length(predicted_membership)) {
    stop("True and predicted memberships must be the same length")
  }

  n <- length(true_membership)
  if (n < 2L) {
    return(list(TPR = NA_real_, TDR = NA_real_, TP = 0L, FP = 0L, FN = 0L))
  }

  TP <- FP <- FN <- 0L
  for (i in seq_len(n - 1L)) {
    for (j in (i + 1L):n) {
      same_true <- true_membership[i] == true_membership[j]
      same_pred <- predicted_membership[i] == predicted_membership[j]

      if (same_true && same_pred) {
        TP <- TP + 1L
      } else if (!same_true && same_pred) {
        FP <- FP + 1L
      } else if (same_true && !same_pred) {
        FN <- FN + 1L
      }
    }
  }

  list(
    TPR = if ((TP + FN) > 0L) TP / (TP + FN) else NA_real_,
    TDR = if ((TP + FP) > 0L) TP / (TP + FP) else NA_real_,
    TP = TP,
    FP = FP,
    FN = FN
  )
}

resolve_global_bfdr_path <- function(bfdr_file, block_data_dir) {
  candidates <- c(
    file.path(pipeline_metadata_dir, bfdr_file),
    file.path(block_data_dir, bfdr_file),
    bfdr_file
  )
  existing <- candidates[file.exists(candidates)]
  if (length(existing) > 0L) existing[1L] else candidates[1L]
}

read_global_bfdr_threshold <- function(path) {
  if (!file.exists(path)) {
    warning("Missing global BFDR file; using the default H4 threshold: ", path)
    return(posterior_prob_h4_threshold)
  }

  global_bfdr <- vroom::vroom(path, show_col_types = FALSE)
  threshold_col <- intersect(c("threshold", "threhold"), names(global_bfdr))[1L]
  if (length(threshold_col) == 0L || is.na(threshold_col) || nrow(global_bfdr) == 0L) {
    warning("No threshold column/value in ", path, "; using the default H4 threshold.")
    return(posterior_prob_h4_threshold)
  }

  as.numeric(global_bfdr[[threshold_col]][1L])
}

make_pair_id <- function(study_a, study_b) {
  paste(pmin(study_a, study_b), pmax(study_a, study_b), sep = "||")
}

read_comparison_set <- function(block_data_dir, method_i) {
  pairwise_path <- file.path(block_data_dir, comparison_files$pairwise_file[method_i])
  cluster_path <- file.path(block_data_dir, comparison_files$cluster_file[method_i])
  bfdr_path <- resolve_global_bfdr_path(
    comparison_files$bfdr_file[method_i],
    block_data_dir
  )

  if (!file.exists(pairwise_path)) stop("Missing pairwise file: ", pairwise_path)
  if (!file.exists(cluster_path)) stop("Missing cluster file: ", cluster_path)

  pairwise <- vroom::vroom(
    pairwise_path,
    delim = "\t",
    col_select = dplyr::any_of(c(
      "unique_study_a", "unique_study_b", "ignore", "h4", "PP.H4.abf"
    )),
    show_col_types = FALSE
  )
  clusters <- vroom::vroom(
    cluster_path,
    delim = "\t",
    col_select = dplyr::any_of(c("unique_study_id", "component")),
    show_col_types = FALSE
  )

  if (!"ignore" %in% names(pairwise)) pairwise$ignore <- FALSE
  if (!"h4" %in% names(pairwise) && "PP.H4.abf" %in% names(pairwise)) {
    pairwise$h4 <- pairwise$PP.H4.abf
  }
  required_pairwise_columns <- c("unique_study_a", "unique_study_b", "h4")
  if (!all(required_pairwise_columns %in% names(pairwise))) {
    stop("Missing required columns in ", pairwise_path)
  }
  if (!all(c("unique_study_id", "component") %in% names(clusters))) {
    stop("Missing required columns in ", cluster_path)
  }

  pairwise <- pairwise |>
    dplyr::mutate(
      ignore = dplyr::coalesce(as.logical(ignore), FALSE),
      pair_id = make_pair_id(unique_study_a, unique_study_b)
    )

  list(
    method = comparison_files$method[method_i],
    h4_threshold = read_global_bfdr_threshold(bfdr_path),
    pairwise = pairwise,
    clusters = clusters
  )
}

compare_sets <- function(named_sets) {
  methods <- names(named_sets)
  dplyr::bind_rows(lapply(seq_along(methods), function(i) {
    dplyr::bind_rows(lapply(seq_along(methods), function(j) {
      if (i >= j) return(NULL)
      left <- named_sets[[i]]
      right <- named_sets[[j]]
      shared <- intersect(left, right)
      union_set <- union(left, right)
      data.frame(
        method_a = methods[i],
        method_b = methods[j],
        n_a = length(left),
        n_b = length(right),
        n_shared = length(shared),
        n_union = length(union_set),
        jaccard = if (length(union_set) > 0L) {
          length(shared) / length(union_set)
        } else {
          NA_real_
        }
      )
    }))
  }))
}

cocluster_pairs <- function(clusters) {
  if (nrow(clusters) == 0L) return(character())

  component_members <- split(clusters$unique_study_id, clusters$component)
  pairs <- lapply(component_members, function(studies) {
    studies <- unique(studies)
    if (length(studies) < 2L) return(character())
    apply(utils::combn(sort(studies), 2L), 2L, paste, collapse = "||")
  })
  unique(unlist(pairs, use.names = FALSE))
}

complete_membership <- function(clusters, universe, method) {
  membership <- paste(method, "singleton", seq_along(universe), sep = "_")
  names(membership) <- universe

  if (nrow(clusters) > 0L) {
    cluster_membership <- paste(method, clusters$component, sep = "_component_")
    names(cluster_membership) <- clusters$unique_study_id
    membership[names(cluster_membership)] <- cluster_membership
  }
  membership
}

compare_block <- function(ld_block, block_data_dir) {
  result_sets <- lapply(seq_len(nrow(comparison_files)), function(method_i) {
    read_comparison_set(block_data_dir, method_i)
  })
  names(result_sets) <- comparison_files$method

  pair_sets <- lapply(result_sets, function(result_set) {
    pairwise <- result_set$pairwise
    unique(pairwise$pair_id[
      !pairwise$ignore & pairwise$h4 >= result_set$h4_threshold
    ])
  })
  pair_overlap <- dplyr::mutate(
    compare_sets(pair_sets),
    ld_block = ld_block,
    .before = 1L
  )

  cocluster_sets <- lapply(result_sets, function(result_set) {
    cocluster_pairs(result_set$clusters)
  })
  cocluster_overlap <- dplyr::mutate(
    compare_sets(cocluster_sets),
    ld_block = ld_block,
    .before = 1L
  )

  cluster_universe <- unique(unlist(lapply(result_sets, function(result_set) {
    result_set$clusters$unique_study_id
  }), use.names = FALSE))
  memberships <- lapply(result_sets, function(result_set) {
    complete_membership(result_set$clusters, cluster_universe, result_set$method)
  })

  method_names <- names(memberships)
  clustering_eval <- dplyr::bind_rows(lapply(seq_along(method_names), function(i) {
    dplyr::bind_rows(lapply(seq_along(method_names), function(j) {
      if (i == j) return(NULL)
      eval_result <- evaluate_clustering(memberships[[i]], memberships[[j]])
      data.frame(
        ld_block = ld_block,
        true_method = method_names[i],
        predicted_method = method_names[j],
        TPR = eval_result$TPR,
        TDR = eval_result$TDR,
        TP = eval_result$TP,
        FP = eval_result$FP,
        FN = eval_result$FN
      )
    }))
  }))

  list(
    pair_overlap = pair_overlap,
    cocluster_overlap = cocluster_overlap,
    clustering_eval = clustering_eval
  )
}

block_comparisons <- lapply(seq_len(nrow(ld_info)), function(i) {
  compare_block(ld_info$block[i], ld_info$ld_block_data[i])
})

pair_overlap <- dplyr::bind_rows(lapply(block_comparisons, `[[`, "pair_overlap"))
cocluster_overlap <- dplyr::bind_rows(lapply(block_comparisons, `[[`, "cocluster_overlap"))
clustering_eval <- dplyr::bind_rows(lapply(block_comparisons, `[[`, "clustering_eval"))

knitr::kable(
  pair_overlap,
  digits = 4,
  caption = "Overlap of non-ignored pairwise coloc pairs at each method's H4 threshold, by LD block"
)

knitr::kable(
  cocluster_overlap,
  digits = 4,
  caption = "Overlap of co-clustered study pairs, by LD block"
)

knitr::kable(
  clustering_eval,
  digits = 4,
  caption = "Pairwise evaluation of cluster memberships by LD block (each method used as the reference in turn)"
)
```
