---
title: "Pleiotropy Investigation"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
params:
  # Path to a pipeline results directory (contains studies.db, etc.). Empty string uses current_results_dir from constants.R.
  results_dir: "/local-scratch/projects/genotype-phenotype-map/results/1.0.0"
---

```{r setup, include=FALSE, echo=F} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(ggplot2)
library(dplyr)
library(tidyr)
library(data.table)
library(purrr)
source("../../pipeline_steps/constants.R")

report_results_dir <- params$results_dir
report_analysis_dir <- file.path(report_results_dir, "analysis")
dir.create(report_analysis_dir, recursive = TRUE, showWarnings = FALSE)
message("Using results_dir: ", report_results_dir)
message("Analysis figures: ", report_analysis_dir)

studies_db_file <- file.path(report_results_dir, "studies.db")
associations_db_file <- file.path(report_results_dir, "associations.db")
coloc_pairs_full_db_file <- file.path(report_results_dir, "coloc_pairs_full.db")
sig_coloc_pairs_db_file <- file.path(report_results_dir, "coloc_pairs.db")
ld_db_file <- file.path(report_results_dir, "ld.db")

studies_conn <- duckdb::dbConnect(duckdb::duckdb(), studies_db_file, read_only = TRUE)
associations_conn <- duckdb::dbConnect(duckdb::duckdb(), associations_db_file, read_only = TRUE)
coloc_pairs_conn <- duckdb::dbConnect(duckdb::duckdb(), coloc_pairs_full_db_file, read_only = TRUE)
sig_coloc_pairs_conn <- duckdb::dbConnect(duckdb::duckdb(), sig_coloc_pairs_db_file, read_only = TRUE)
ld_conn <- duckdb::dbConnect(duckdb::duckdb(), ld_db_file, read_only = TRUE)

dont_print <- DBI::dbExecute(coloc_pairs_conn, glue::glue("ATTACH '{studies_db_file}' AS studies"))
dont_print <- DBI::dbExecute(sig_coloc_pairs_conn, glue::glue("ATTACH '{studies_db_file}' AS studies"))

cb_blue <- "#0072B2"
cb_light_blue <- "#80B9D9"
cb_orange <- "#E69F00"
cb_light_orange <- "#F3CF80"
```


### Questions to attempt to answer:

1. Association vs. colocalization: how often do we see a genetic association with a trait that doesn't have a coloc result?
   * Prevelance of H3 > 0.8 vs. H4 > 0.8
2. How many colocalization groups have only qtls?  QTL and phenotype, QTL and multiple phenotypes
3. How many of these chosen causal SNPs are in LD with each other?
4. Modularity of pleiotropy
   * How can we quantify modularity?
   * How many genes are associated with a trait on average?
   * Does it change when including trans-QTLs?
   * What about gene-gene modularity? Could we have a graph where we count the number of associations between genes tagged?
   * Genes that colocalise with the most traits (or trait categories?).  
5. What fraction of known GWAS hits are no known molecular trait associated with it?
6. Is there consistency between of molecule-disease associations between common cis, rare cis and trans instruments?
7. Tissue specificity: can we break down colocalization groups by tissue?
8. Can we calculate a 'pleiotropy score' for each SNP?  Would it be useful?


```{r load-data, echo=F}
joined_coloc_results <- DBI::dbGetQuery(coloc_pairs_conn, 'SELECT
  coloc_pairs.*,
  -- Study extraction A details
  study_extraction_a.unique_study_id AS study_extraction_a_unique_study_id,
  study_extraction_a.min_p AS min_p_a,
  study_extraction_a.cis_trans AS cis_trans_a,
  study_extraction_a.bp AS bp_a,
  study_extraction_a.chr AS chr_a,
  -- Study A details
  study_a.study_name AS study_name_a,
  study_a.gene AS gene_a,
  study_a.tissue AS tissue_a,
  study_a.sample_size AS sample_size_a,
  study_a.data_type AS data_type_a,
  trait_a.trait_category AS trait_category_a,
  trait_a.trait_name AS trait_name_a,
  -- Study extraction B details
  study_extraction_b.unique_study_id AS study_extraction_b_unique_study_id,
  study_extraction_b.min_p AS min_p_b,
  study_extraction_b.cis_trans AS cis_trans_b,
  study_extraction_b.bp AS bp_b,
  study_extraction_b.chr AS chr_b,
  -- Study B details
  study_b.study_name AS study_name_b,
  study_b.gene AS gene_b,
  study_b.tissue AS tissue_b,
  study_b.sample_size AS sample_size_b,
  study_b.data_type AS data_type_b,
  trait_b.trait_category AS trait_category_b,
  trait_b.trait_name AS trait_name_b
  FROM coloc_pairs
  -- Join for study extraction A
  JOIN studies.study_extractions AS study_extraction_a ON coloc_pairs.study_extraction_a_id = study_extraction_a.id
  JOIN studies.studies AS study_a ON study_extraction_a.study_id = study_a.id
  JOIN studies.traits AS trait_a ON study_a.trait_id = trait_a.id
  -- Join for study extraction B
  JOIN studies.study_extractions AS study_extraction_b ON coloc_pairs.study_extraction_b_id = study_extraction_b.id
  JOIN studies.studies AS study_b ON study_extraction_b.study_id = study_b.id
  JOIN studies.traits AS trait_b ON study_b.trait_id = trait_b.id
  WHERE coloc_pairs.h4 > 0.6 OR coloc_pairs.h3 > 0.6
')
unique_data_types <- DBI::dbGetQuery(studies_conn, "SELECT DISTINCT data_type FROM studies")$data_type
```

## Gene pleiotropy scores

```{r pleiotropy_scores, echo=F}
gene_pleiotropy_scores <- DBI::dbGetQuery(studies_conn, "SELECT * FROM gene_pleiotropy")

# Fit linear model for gene pleiotropy
lm_gene <- lm(distinct_protein_coding_genes ~ distinct_trait_categories, data = gene_pleiotropy_scores)
lm_gene_summary <- summary(lm_gene)
gene_r2 <- lm_gene_summary$r.squared
gene_p <- pf(lm_gene_summary$fstatistic[1], lm_gene_summary$fstatistic[2], lm_gene_summary$fstatistic[3], lower.tail = FALSE)
gene_slope <- coef(lm_gene)[2]

# Create annotation text
gene_stats_text <- paste0(
  "Slope = ", round(gene_slope, 3), "\n",
  "R² = ", round(gene_r2, 3), "\n",
  "p = ", formatC(gene_p, format = "e", digits = 2)
)

p_gene_pleiotropy <- ggplot(gene_pleiotropy_scores, aes(x = distinct_trait_categories, y = distinct_protein_coding_genes)) +
  geom_point(alpha = 0.6, size = 2, color = cb_blue) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  annotate("label", x = -Inf, y = Inf, label = gene_stats_text, 
           hjust = -0.1, vjust = 1.5, size = 3.5, family = "mono",
           fill = "white", alpha = 0.8) +
  labs(
    title = "Relationship between gene pleiotropy metrics",
    x = "Number of trait categories colocalizing with the anchor gene",
    y = "Number of protein-coding genes colocalizing with the anchor gene"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(p_gene_pleiotropy, filename = file.path(report_analysis_dir, "pleiotropy_e_gene_pleiotropy_plot.png"), width = 8, height = 6)
print(p_gene_pleiotropy)

# Create histogram of distinct_protein_coding_genes distribution
gene_pleiotropy_binned <- gene_pleiotropy_scores %>%
  mutate(
    protein_coding_genes_bin = case_when(
      distinct_protein_coding_genes == 0 ~ "0",
      distinct_protein_coding_genes <= 20 ~ as.character(distinct_protein_coding_genes),
      TRUE ~ ">20"
    )
  )

# Convert to factor with proper ordering
gene_pleiotropy_binned$protein_coding_genes_bin <- factor(
  gene_pleiotropy_binned$protein_coding_genes_bin,
  levels = c(as.character(0:20), ">20")
)

# Count by bins
gene_pleiotropy_bin_counts <- gene_pleiotropy_binned %>%
  group_by(protein_coding_genes_bin) %>%
  summarise(count = n(), .groups = 'drop') %>%
  filter(!is.na(protein_coding_genes_bin))

# Filter out 0 gene groups for summary statistics
gene_pleiotropy_filtered <- gene_pleiotropy_binned %>%
  filter(distinct_protein_coding_genes > 0)

p_gene_pleiotropy_dist <- ggplot(gene_pleiotropy_bin_counts, aes(x = protein_coding_genes_bin, y = count)) +
  geom_bar(stat = "identity", fill = cb_blue, color = "black", alpha = 0.7) +
  geom_text(aes(label = count), vjust = -0.5, size = 3) +
  scale_x_discrete(limits = c(as.character(1:20), ">20")) +
  labs(
    title = "Distribution gene pleiotropy scores",
    x = "Number of protein-coding genes colocalizing with the anchor gene",
    y = "Count of genes",
    caption = paste("Total genes (with colocalizing genes):", nrow(gene_pleiotropy_filtered),
      "| Mean:", round(mean(gene_pleiotropy_filtered$distinct_protein_coding_genes), 2),
      "| Median:", round(median(gene_pleiotropy_filtered$distinct_protein_coding_genes), 2)
    )
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

ggsave(p_gene_pleiotropy_dist, filename = file.path(report_analysis_dir, "pleiotropy_gene_protein_coding_genes_distribution.png"), width = 8, height = 6)
print(p_gene_pleiotropy_dist)

variant_pleiotropy_scores <- DBI::dbGetQuery(studies_conn, "SELECT * FROM variant_pleiotropy")

# Fit linear model for SNP pleiotropy
lm_snp <- lm(distinct_protein_coding_genes ~ distinct_trait_categories, data = variant_pleiotropy_scores)
lm_snp_summary <- summary(lm_snp)
snp_r2 <- lm_snp_summary$r.squared
snp_p <- pf(lm_snp_summary$fstatistic[1], lm_snp_summary$fstatistic[2], lm_snp_summary$fstatistic[3], lower.tail = FALSE)
snp_slope <- coef(lm_snp)[2]

# Create annotation text
snp_stats_text <- paste0(
  "Slope = ", round(snp_slope, 3), "\n",
  "R² = ", round(snp_r2, 3), "\n",
  "p = ", formatC(snp_p, format = "e", digits = 2)
)

p_variant_pleiotropy <- ggplot(variant_pleiotropy_scores, aes(x = distinct_trait_categories, y = distinct_protein_coding_genes)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = cb_blue) +
  annotate("label", x = Inf, y = Inf, label = snp_stats_text, 
           hjust = 1.1, vjust = 1.5, size = 3.5, family = "mono",
           fill = "white", alpha = 0.8) +
  labs(
    title = "Relationship between SNP Pleiotropy Metrics",
    subtitle = "Trait categories vs. number of protein-coding genes colocalizing with the anchor SNP",
    x = "Number of trait categories",
    y = "Number of protein-coding genes colocalizing with the anchor SNP"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(p_variant_pleiotropy, filename = file.path(report_analysis_dir, "variant_pleiotropy_plot.png"), width = 8, height = 6)
print(p_variant_pleiotropy)

plei <- DBI::dbGetQuery(studies_conn, "select * from variant_pleiotropy order by variant_id;")

plei_summary <- plei |>
  mutate(genecat = cut(distinct_protein_coding_genes, breaks = c(0, 1, 2, 5, 10, 1000))) |>
  filter(!is.na(genecat)) |>
  mutate(genecat = case_when(
    genecat == "(0,1]" ~ "1",
    genecat == "(1,2]" ~ "2",
    genecat == "(2,5]" ~ "3-5",
    genecat == "(5,10]" ~ "6-10",
    genecat == "(10,1e+03]" ~ "10+"
  )) |>
  mutate(genecat = factor(genecat, levels = c("1", "2", "3-5", "6-10", "10+"))) |>
  group_by(genecat) |>
  summarise(m = mean(distinct_trait_categories), s = sd(distinct_trait_categories),
            n = n(), se = s / sqrt(n), lci = m - 1.96 * se, uci = m + 1.96 * se,
            .groups = 'drop') |>
  mutate(n_label = paste0("n=", scales::comma(n)))

p_variant_pleiotropy_gene_distribution <- ggplot(plei_summary, aes(x = genecat, y = m)) +
  geom_bar(stat = "identity", fill = cb_blue, color = "black", alpha = 0.7) +
  geom_errorbar(aes(ymin = lci, ymax = uci), width = 0.2) +
  geom_text(aes(y = 0, label = n_label), vjust = -0.5, size = 8 / .pt, color = "black") +
  labs(
    title = "Mean number of trait categories by number of colocalizing genes (per variant)",
    x = "Number of distinct protein-coding genes colocalizing with the variant (binned)",
    y = "Mean distinct trait categories per variant",
    caption = "Bars: mean number of distinct trait categories among variants in each gene-count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 11.5)
  )
ggsave(p_variant_pleiotropy_gene_distribution, filename = file.path(report_analysis_dir, "variant_pleiotropy_gene_distribution.png"), width = 8, height = 6)
print(p_variant_pleiotropy_gene_distribution)

coloc_group_members <- DBI::dbGetQuery(sig_coloc_pairs_conn, "
  SELECT cg.coloc_group_id, cg.study_extraction_id, cg.ld_block_id
  FROM studies.coloc_groups AS cg
  JOIN studies.study_extractions AS se ON cg.study_extraction_id = se.id
  WHERE se.min_p < 5e-8
")

# Total maximum possible edges across all coloc groups: sum of n*(n-1)/2 per group
pairs_in_groups <- coloc_group_members %>%
  count(coloc_group_id) %>%
  mutate(max_edges = n * (n - 1) / 2) %>%
  summarise(total = sum(max_edges)) %>%
  pull(total)

# Calculate H4 false positive / false negative numbers and rates
h4_fp_fn_stats <- joined_coloc_results %>%
  filter(min_p_a < 5e-8 & min_p_b < 5e-8) %>%
  mutate(
    is_h4_sig = h4 > 0.8
  ) %>%
  summarise(
    total_pairs = n(),
    total_h4_sig = sum(is_h4_sig, na.rm = TRUE),
    h4_false_positive_n = sum(is_h4_sig & false_positive, na.rm = TRUE),
    h4_false_positive_rate = h4_false_positive_n / total_h4_sig,
    pairs_in_groups = pairs_in_groups,
    h4_false_negative_n = sum(false_negative, na.rm = TRUE),
    h4_false_negative_rate = h4_false_negative_n / pairs_in_groups
  )

knitr::kable(
  h4_fp_fn_stats %>%
    mutate(
      h4_false_positive_rate = scales::percent(h4_false_positive_rate, accuracy = 0.01),
      h4_false_negative_rate = scales::percent(h4_false_negative_rate, accuracy = 0.01)
    ),
  caption = "H4 false positive / false negative counts and rates"
)

# False negative rate from stored h4_connectedness in coloc_groups table
# h4_connectedness = fraction of within-group pairs with h4 >= 0.8
# so false_negative_rate = 1 - h4_connectedness
coloc_group_connectedness <- DBI::dbGetQuery(sig_coloc_pairs_conn, "
  SELECT cg.coloc_group_id, cg.h4_connectedness, COUNT(*) AS group_size
  FROM studies.coloc_groups AS cg
  JOIN studies.study_extractions AS se ON cg.study_extraction_id = se.id
  GROUP BY cg.coloc_group_id, cg.h4_connectedness
") %>%
  mutate(expected_pairs = group_size * (group_size - 1) / 2)

fn_from_connectedness <- tibble::tibble(
  method = c("Unweighted (mean of per-group scores)", "Weighted by group size"),
  mean_h4_connectedness = c(
    mean(coloc_group_connectedness$h4_connectedness, na.rm = TRUE),
    weighted.mean(coloc_group_connectedness$h4_connectedness,
                  coloc_group_connectedness$group_size, na.rm = TRUE)
  ),
  implied_fn_rate = 1 - mean_h4_connectedness
)

knitr::kable(
  fn_from_connectedness %>%
    mutate(
      mean_h4_connectedness = scales::percent(mean_h4_connectedness, accuracy = 0.01),
      implied_fn_rate = scales::percent(implied_fn_rate, accuracy = 0.01)
    ),
  caption = "False negative rate derived from stored h4_connectedness in coloc_groups"
)

sig_coloc_pairs <- DBI::dbGetQuery(sig_coloc_pairs_conn,
  "SELECT cp.*, sea.min_p AS min_p_a, seb.min_p AS min_p_b FROM coloc_pairs AS cp
    JOIN studies.study_extractions AS sea ON cp.study_extraction_a_id = sea.id
    JOIN studies.study_extractions AS seb ON cp.study_extraction_b_id = seb.id
    WHERE cp.h4 >= 0.8 AND min_p_a < 5e-8 AND min_p_b < 5e-8"
)

unused_sig_coloc_pairs <- sig_coloc_pairs %>%
  filter(is.na(variant_id))

# Recreate h4_connectedness per coloc group from pairwise data
# 1. Get coloc group membership (GW-sig study extractions only)
coloc_group_members <- DBI::dbGetQuery(sig_coloc_pairs_conn, "
  SELECT cg.coloc_group_id, cg.study_extraction_id, cg.ld_block_id
  FROM studies.coloc_groups AS cg
  JOIN studies.study_extractions AS se ON cg.study_extraction_id = se.id
  WHERE se.min_p < 5e-8
")

# 2. Get ALL coloc pairs from coloc_pairs.db with both studies GW-sig
#    (coloc_pairs.db stores pairs with h4 > 0.5, so includes non-colocalizing pairs)
all_gw_sig_pairs <- DBI::dbGetQuery(sig_coloc_pairs_conn, "
  SELECT cp.study_extraction_a_id, cp.study_extraction_b_id,
         cp.ld_block_id, cp.h3, cp.h4, cp.false_positive, cp.false_negative
  FROM coloc_pairs AS cp
  JOIN studies.study_extractions AS sea ON cp.study_extraction_a_id = sea.id
  JOIN studies.study_extractions AS seb ON cp.study_extraction_b_id = seb.id
  WHERE sea.min_p < 5e-8 AND seb.min_p < 5e-8 AND cp.h4 >= 0.8
")

# 3. Map each coloc pair to coloc_group membership
#    A pair is "within group" if both study extractions belong to the same coloc_group_id
pair_groups <- all_gw_sig_pairs %>%
  inner_join(coloc_group_members, by = c("study_extraction_a_id" = "study_extraction_id",
                                          "ld_block_id" = "ld_block_id")) %>%
  rename(coloc_group_a = coloc_group_id) %>%
  inner_join(coloc_group_members, by = c("study_extraction_b_id" = "study_extraction_id",
                                          "ld_block_id" = "ld_block_id")) %>%
  rename(coloc_group_b = coloc_group_id) %>%
  filter(coloc_group_a == coloc_group_b) %>%
  rename(coloc_group_id = coloc_group_a) %>%
  select(-coloc_group_b)

# 4. Get the expected number of pairs per group (n choose 2) to know coverage
group_sizes <- coloc_group_members %>%
  count(coloc_group_id, name = "n_members") %>%
  mutate(expected_pairs = n_members * (n_members - 1) / 2)

# 5. Calculate h4_connectedness per coloc group
#    Since all_gw_sig_pairs is filtered to h4 >= 0.8, observed_pairs = pairs with h4 >= 0.8.
#    h4_connectedness = observed h4>=0.8 pairs / expected total pairs (n choose 2)
h4_connectedness_scores <- pair_groups %>%
  group_by(coloc_group_id) %>%
  summarise(
    observed_h4_pairs = n(),
    fp_n = sum(false_positive, na.rm = TRUE),
    fn_n = sum(false_negative, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  right_join(group_sizes, by = "coloc_group_id") %>%
  mutate(
    observed_h4_pairs = tidyr::replace_na(observed_h4_pairs, 0),
    fp_n = tidyr::replace_na(fp_n, 0),
    fn_n = tidyr::replace_na(fn_n, 0),
    h4_connectedness = observed_h4_pairs / expected_pairs,
    fp_rate = fp_n / expected_pairs,
    fn_rate = (expected_pairs - observed_h4_pairs) / expected_pairs
  )

# H4 connectedness distribution
p_h4_connectedness <- ggplot(h4_connectedness_scores, aes(x = h4_connectedness)) +
  geom_histogram(bins = 20, fill = cb_blue, color = "black", alpha = 0.7) +
  labs(
    title = "H4 Connectedness per Colocalization Group (GW-sig pairs)",
    subtitle = "Fraction of within-group GW-sig pairs with H4 >= 0.8",
    x = "H4 Connectedness",
    y = "Number of Colocalization Groups",
    caption = paste("N =", nrow(h4_connectedness_scores), "colocalization groups")
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

ggsave(p_h4_connectedness, filename = file.path(report_analysis_dir, "h4_connectedness_distribution.png"), width = 8, height = 6)
print(p_h4_connectedness)

```


```{r l_shaped_pleiotropy, echo=F}
num_coloc_groups <- DBI::dbGetQuery(studies_conn, "SELECT COUNT(DISTINCT coloc_group_id) FROM coloc_groups")
coloc_group_data <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*,
  studies.gene, studies.sample_size, studies.data_type, studies.tissue, studies.study_name,
  gene_annotations.chr, gene_annotations.start, gene_annotations.stop, gene_annotations.gene_biotype,
  study_extractions.cis_trans, traits.trait_category
  FROM coloc_groups
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN studies.study_extractions ON coloc_groups.study_extraction_id = studies.study_extractions.id
  JOIN traits ON studies.trait_id = traits.id
  LEFT JOIN gene_annotations ON studies.gene_id = gene_annotations.id
  WHERE study_extractions.min_p < 5e-8
")
multiple_genes_in_coloc_group <- function(only_protein_coding = FALSE, include_trans = TRUE) {
  # Start with all coloc groups, but apply filters for counting
  coloc_group_analysis <- coloc_group_data %>%
    # Apply gene type filter if specified
    {if (only_protein_coding) {
      filter(., gene_biotype == "protein_coding")
    } else {
      .
    }} %>%
    # Apply cis/trans filter if specified
    {if (!include_trans) {
      filter(., cis_trans != "trans")
    } else {
      .
    }} %>%
    filter(data_type != "methylation") %>%
    group_by(coloc_group_id) %>%
    summarise(
      num_genes = n(),
      unique_genes = n_distinct(gene, na.rm = TRUE),
      genes = paste(unique(gene[!is.na(gene)]), collapse = ", "),
      .groups = 'drop'
    ) %>%
    mutate(
      genes = ifelse(genes == "", NA, genes),
      unique_genes = ifelse(genes == "" | is.na(genes), 0, unique_genes)
    )
  
  # Now we need to ensure we include ALL coloc groups, even those that might have been filtered out
  # Get all unique coloc_group_ids from the original data
  all_coloc_groups <- coloc_group_data %>%
    select(coloc_group_id) %>%
    distinct()
  
  # Left join to ensure all coloc groups are represented
  coloc_group_analysis <- all_coloc_groups %>%
    left_join(coloc_group_analysis, by = "coloc_group_id") %>%
    mutate(
      num_genes = ifelse(is.na(num_genes), 0, num_genes),
      unique_genes = ifelse(is.na(unique_genes), 0, unique_genes),
      genes = ifelse(is.na(genes), NA, genes)
    )

  coloc_group_analysis$unique_gene_bin <- case_when(
    coloc_group_analysis$unique_genes == 0 ~ "0",
    coloc_group_analysis$unique_genes <= 20 ~ as.character(coloc_group_analysis$unique_genes),
    TRUE ~ paste0(">20")
  )
  
  # Convert to factor with proper ordering
  coloc_group_analysis$unique_gene_bin <- factor(coloc_group_analysis$unique_gene_bin, 
                                                levels = c(as.character(0:20), ">20"))
  
  # Count by bins
  unique_gene_bin_counts <- coloc_group_analysis %>%
    group_by(unique_gene_bin) %>%
    summarise(count = n(), .groups = 'drop') %>%
    filter(!is.na(unique_gene_bin))
  
  # Filter out 0 gene groups for summary statistics to match the plot
  coloc_group_analysis_filtered <- coloc_group_analysis %>%
    filter(unique_genes > 0)
  
  p_gene_dist <- ggplot(unique_gene_bin_counts, aes(x = unique_gene_bin, y = count)) +
    geom_bar(stat = "identity", fill = cb_blue, color = "black", alpha = 0.7) +
    geom_text(aes(label = count), vjust = -0.5, size = 3) +
    scale_x_discrete(limits = c(as.character(1:20), ">20")) +
    labs(
      title = glue::glue("Distribution of {if (only_protein_coding) 'protein-coding' else 'all'} gene counts across colocalization groups {if (include_trans) ', including trans' else ''}"),
      x = "Number of unique protein-coding genes per colocalization group",
      y = "Frequency",
      caption = paste("Total groups (with genes):", nrow(coloc_group_analysis_filtered),
        "| Mean:", round(mean(coloc_group_analysis_filtered$unique_genes), 2),
        "| Median:", round(median(coloc_group_analysis_filtered$unique_genes), 2)
      )
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
  p_gene_dist_file <- file.path(report_analysis_dir, glue::glue("pleiotropy_b_num_genes_per_coloc_group_{if (only_protein_coding) 'protein_coding' else 'all'}_{if (include_trans) 'including_trans' else 'excluding_trans'}.png"))
  ggsave(p_gene_dist, filename = p_gene_dist_file, width = 8, height = 6)

  

  return(p_gene_dist)
}

multiple_genes_in_coloc_group(only_protein_coding = TRUE, include_trans = FALSE)
```

### SNP-to-Gene Pleiotropy Amalgamation
Does combining the trait categories from all SNPs associated with a gene significantly increase the number of trait categories, or do the same categories tend to repeat across different SNPs for that gene?

```{r snp_gene_pleiotropy_amalgamation, echo=F}
# 1. Get raw data from coloc_groups_wide
# We want the distinct list of trait categories per SNP and per Gene
# We allow gene_id and trait_category to be NULL initially to capture all relevant data
raw_coloc_data <- DBI::dbGetQuery(studies_conn, "
  SELECT variant_id, gene_id, trait_category 
  FROM coloc_groups_wide 
  WHERE data_type != 'methylation'
")

# 2. Calculate per-SNP trait category lists
# We filter for non-null trait_category here to count actual pleiotropy
snp_trait_lists <- raw_coloc_data %>%
  filter(!is.na(trait_category)) %>%
  group_by(variant_id) %>%
  summarise(
    snp_trait_categories = list(unique(trait_category)),
    num_snp_traits = n_distinct(trait_category),
    .groups = 'drop'
  )

# 3. Amalgamate by Gene
# We need a mapping of SNPs to Genes. A SNP can be associated with multiple genes.
snp_to_gene_map <- raw_coloc_data %>%
  filter(!is.na(gene_id)) %>%
  select(variant_id, gene_id) %>%
  distinct()

gene_amalgamation <- snp_to_gene_map %>%
  inner_join(snp_trait_lists, by = "variant_id") %>%
  group_by(gene_id) %>%
  summarise(
    num_snps = n(),
    mean_traits_per_snp = mean(num_snp_traits),
    max_traits_per_snp = max(num_snp_traits),
    # Amalgamated list (all unique categories across all SNPs for this gene)
    amalgamated_trait_categories = list(unique(unlist(snp_trait_categories))),
    num_amalgamated_traits = n_distinct(unlist(snp_trait_categories)),
    .groups = 'drop'
  ) %>%
  mutate(
    trait_increase = num_amalgamated_traits - max_traits_per_snp,
    percent_increase = ifelse(max_traits_per_snp > 0, (trait_increase / max_traits_per_snp) * 100, 0)
  )

# 4. Summary Table
amalgamation_summary <- gene_amalgamation %>%
  summarise(
    total_genes = n(),
    avg_snps_per_gene = mean(num_snps),
    avg_traits_per_snp = mean(mean_traits_per_snp),
    avg_amalgamated_traits = mean(num_amalgamated_traits),
    avg_trait_increase = mean(trait_increase),
    genes_with_any_increase = sum(trait_increase > 0),
    percent_genes_with_increase = (sum(trait_increase > 0) / n()) * 100
  )

knitr::kable(amalgamation_summary, caption = "Summary of SNP-to-Gene Trait Amalgamation", digits = 2)

# 5. Visualization: Amalgamated vs. Max SNP Traits
p_amalgamation <- ggplot(gene_amalgamation, aes(x = max_traits_per_snp, y = num_amalgamated_traits)) +
  geom_jitter(alpha = 0.4, color = cb_blue, width = 0.2, height = 0.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Trait Category Amalgamation: Gene vs. Best SNP",
    subtitle = "Red line represents no new categories added by additional SNPs",
    x = "Max Trait Categories for a the most pleiotropic SNP of a gene",
    y = "Total Amalgamated Trait Categories for Gene"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

ggsave(p_amalgamation, filename = file.path(report_analysis_dir, "pleiotropy_snp_gene_amalgamation.png"), width = 8, height = 6)
print(p_amalgamation)

# 6. Statistical Investigation: Added categories vs. Number of SNPs
# We want to see if having more SNPs actually leads to discovering more categories
# or if it's just redundant.

# Bin the number of SNPs to make the visualization clearer
gene_amalgamation_stats <- gene_amalgamation %>%
  mutate(
    snp_bin = case_when(
      num_snps == 1 ~ "1",
      num_snps == 2 ~ "2",
      num_snps <= 5 ~ "3-5",
      num_snps <= 10 ~ "6-10",
      TRUE ~ ">10"
    ),
    snp_bin = factor(snp_bin, levels = c("1", "2", "3-5", "6-10", ">10"))
  )

# Plot 1: Distribution of "Added" traits by SNP count bin
p_added_traits <- ggplot(gene_amalgamation_stats, aes(x = snp_bin, y = trait_increase)) +
  geom_boxplot(fill = cb_light_blue, outlier.alpha = 0.3) +
  labs(
    title = "Additional trait categories discovered vs. variants associated with gene",
    subtitle = "Increase in unique categories beyond the most pleiotropic single SNP",
    x = "Number of variants associated with gene",
    y = "Number of 'new' trait categories added"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

ggsave(p_added_traits, filename = file.path(report_analysis_dir, "pleiotropy_added_traits_by_snp_count.png"), width = 8, height = 6)
print(p_added_traits)

# Statistical test: Does SNP count significantly predict trait increase?
# Using Poisson regression since trait_increase is a count
fit_increase <- glm(trait_increase ~ num_snps, family = poisson(), data = gene_amalgamation)
summary_fit <- summary(fit_increase)

cat("\n**Statistical Test: Does the number of SNPs significantly predict the discovery of new trait categories?**\n")
print(summary_fit$coefficients)
```

### Comparison of Gene-centric vs. Coloc-group-centric Pleiotropy
Does the number of genes colocalizing with a specific gene match the distribution of genes found within individual colocalization groups?

```{r distribution_comparison, echo=F}
# Recalculate Distribution 2 (Protein-coding, excluding trans, non-methylation)
# matching the logic of pleiotropy_b_num_genes_per_coloc_group_protein_coding_excluding_trans
dist2_data <- coloc_group_data %>%
  filter(gene_biotype == "protein_coding") %>%
  filter(cis_trans != "trans" | is.na(cis_trans)) %>% 
  filter(data_type != "methylation") %>%
  group_by(coloc_group_id) %>%
  summarise(unique_genes = n_distinct(gene, na.rm = TRUE), .groups = 'drop') %>%
  filter(unique_genes > 0)

dist1_vals <- gene_pleiotropy_filtered$distinct_protein_coding_genes
dist2_vals <- dist2_data$unique_genes

# Calculate summary stats
stats_comparison <- data.frame(
  Metric = c("Mean", "Median", "SD", "Max", "N"),
  Gene_centric = c(mean(dist1_vals), median(dist1_vals), sd(dist1_vals), max(dist1_vals), length(dist1_vals)),
  Coloc_group_centric = c(mean(dist2_vals), median(dist2_vals), sd(dist2_vals), max(dist2_vals), length(dist2_vals))
)

# KS test
ks_res <- ks.test(dist1_vals, dist2_vals)

# Print results
knitr::kable(stats_comparison, caption = "Comparison of Pleiotropy Distributions", digits = 3)

cat("\n**Kolmogorov-Smirnov test for difference in distributions:**\n")
cat("D =", round(ks_res$statistic, 4), ", p-value =", formatC(ks_res$p.value, format = "e", digits = 3), "\n")
```

### Gene colocalization vs regional gene density (within ±1Mb)
```{r gene_colocalization_vs_regional_gene_density, echo=F}
# Use protein-coding genes only to match graph (B) context
protein_coding_genes <- DBI::dbGetQuery(studies_conn, "
  SELECT gene, chr, start, stop, gene_biotype
  FROM gene_annotations
  WHERE gene_biotype = 'protein_coding'
")

genes_dt <- as.data.table(protein_coding_genes)
# Define ±1Mb window around each gene; ensure ordering for foverlaps
genes_dt[, region_start := start - 1e6]
genes_dt[, region_end   := stop  + 1e6]
setkey(genes_dt, chr, start, stop)

# Count neighbouring protein-coding genes within ±1Mb (exclude self)
overlaps <- foverlaps(
  genes_dt[, .(gene, chr, region_start, region_end)],
  genes_dt,
  by.x = c("chr", "region_start", "region_end"),
  by.y = c("chr", "start", "stop"),
  type = "any",
  nomatch = 0
)

gene_density <- overlaps[
  ,
  .(neighborhood_genes = .N - 1L),  # subtract self
  by = gene
]

# For each gene, count distinct protein-coding genes it colocalises with (cis, min_p < 5e-8)
coloc_protein <- coloc_group_data |>
  filter(gene_biotype == "protein_coding", !is.na(gene)) |>
  group_by(coloc_group_id) |>
  summarise(genes = list(unique(gene)), .groups = "drop") |>
  filter(lengths(genes) > 0)

gene_partner_pairs <- coloc_protein |>
  mutate(pairs = map(genes, function(g) {
    if (length(g) < 2) {
      return(tibble::tibble(gene = character(), partner = character()))
    }
    cmb <- combn(g, 2)
    tibble::tibble(
      gene    = c(cmb[1, ], cmb[2, ]),
      partner = c(cmb[2, ], cmb[1, ])
    )
  })) |>
  select(pairs) |>
  tidyr::unnest(pairs)

gene_coloc_partners <- gene_partner_pairs |>
  group_by(gene) |>
  summarise(num_coloc_partner_genes = n_distinct(partner), .groups = "drop")

gene_coloc_density <- gene_coloc_partners |>
  inner_join(gene_density, by = "gene") |>
  mutate(
    density_decile = ntile(neighborhood_genes, 10),
    density_decile = factor(density_decile, levels = 1:10, labels = paste0("Decile ", 1:10))
  )

significant_cis_coloc_pairs <- joined_coloc_results %>%
  filter(h4 > 0.8 & (is.na(cis_trans_a) | cis_trans_a == 'cis') & (is.na(cis_trans_b) | cis_trans_b == 'cis') & min_p_a < 5e-8 & min_p_b < 5e-8)

most_gene_a_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype' & data_type_a != 'methylation') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait = trait_name_b)

most_gene_b_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype' & data_type_b != 'methylation') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait = trait_name_a)

most_gene_a_colocalising_trait_category <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype' & data_type_a != 'methylation') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait_category = trait_category_b)

most_gene_b_colocalising_trait_category <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype' & data_type_b != 'methylation') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait_category = trait_category_a)

most_genes_colocalising_with_trait_categories <- rbind(most_gene_a_colocalising_trait_category, most_gene_b_colocalising_trait_category)

gene_trait_category_counts <- most_genes_colocalising_with_trait_categories %>%
  group_by(gene) %>%
  summarise(num_trait_categories = n_distinct(trait_category), .groups = 'drop')

gene_trait_category_count_data <- gene_trait_category_counts %>%
  count(num_trait_categories, name = "count") %>%
  mutate(count_offset = count + 0.1)  # Add small offset to make count=1 visible


p_gene_trait_category <- ggplot(gene_trait_category_count_data, aes(x = num_trait_categories, y = count_offset)) +
  geom_bar(stat = "identity", fill = cb_blue, color = "black", alpha = 0.7) +
  scale_x_continuous(breaks = seq(1, max(gene_trait_category_count_data$num_trait_categories))) +
  labs(
    title = "Number of trait categories colocalising with a gene",
    x = "Number of trait categories per gene",
    y = "Frequency",
    caption = paste( "Mean:", round(mean(gene_trait_category_counts$num_trait_categories), 2),
      "| Median:", round(median(gene_trait_category_counts$num_trait_categories), 2))
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_gene_trait_category_file <- file.path(report_analysis_dir, "pleiotropy_a_num_trait_categories_per_gene.png")
ggsave(p_gene_trait_category, filename = p_gene_trait_category_file, width = 8, height = 6)
print(p_gene_trait_category)

# New: Distribution of trait categories per colocalization group
coloc_group_trait_counts <- coloc_group_data %>%
  filter(data_type == 'phenotype') %>%
  group_by(coloc_group_id) %>%
  summarise(num_trait_categories = n_distinct(trait_category, na.rm = TRUE), .groups = 'drop') %>%
  filter(num_trait_categories > 0)

coloc_group_trait_count_data <- coloc_group_trait_counts %>%
  count(num_trait_categories, name = "count") %>%
  mutate(count_offset = count + 0.1)

p_coloc_group_trait_category <- ggplot(coloc_group_trait_count_data, aes(x = num_trait_categories, y = count_offset)) +
  geom_bar(stat = "identity", fill = cb_blue, color = "black", alpha = 0.7) +
  scale_x_continuous(breaks = seq(1, max(coloc_group_trait_count_data$num_trait_categories))) +
  labs(
    title = "Number of trait categories per colocalization group",
    x = "Number of trait categories per group",
    y = "Frequency",
    caption = paste("Mean:", round(mean(coloc_group_trait_counts$num_trait_categories), 2),
      "| Median:", round(median(coloc_group_trait_counts$num_trait_categories), 2))
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_coloc_group_trait_category_file <- file.path(report_analysis_dir, "pleiotropy_num_trait_categories_per_coloc_group.png")
ggsave(p_coloc_group_trait_category, filename = p_coloc_group_trait_category_file, width = 8, height = 6)
print(p_coloc_group_trait_category)

# Compute and plot histogram of the size of coloc groups (number of distinct studies per coloc group)
coloc_groups_sizes <- DBI::dbGetQuery(studies_conn, "
  SELECT coloc_group_id, COUNT(DISTINCT study_id) AS num_studies
  FROM coloc_groups
  GROUP BY coloc_group_id
")

coloc_group_count_data <- coloc_groups_sizes %>%
  count(num_studies, name = "count") %>%
  mutate(count_offset = count + 0.2)

p_coloc_group_size <- ggplot(coloc_group_count_data, aes(x = num_studies, y = count_offset)) +
  geom_bar(stat = "identity", fill = cb_blue, color = cb_blue, alpha = 0.7) +
  scale_x_continuous() +
  scale_y_log10() +
  labs(
    title = "Number of traits in colocilization group",
    x = "Number of traits in colocilization group",
    y = "Frequency (log scale)",
    caption = paste("Mean:", round(mean(coloc_groups_sizes$num_studies), 2),
                    "| Median:", round(median(coloc_groups_sizes$num_studies), 2))
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

coloc_group_size_file <- file.path(report_analysis_dir, "pleiotropy_coloc_group_size_histogram.png")
ggsave(p_coloc_group_size, filename = coloc_group_size_file, width = 8, height = 6)
print(p_coloc_group_size)



all_phenotype_studies <- DBI::dbGetQuery(studies_conn, "SELECT studies.*, traits.trait_name FROM studies
  JOIN traits ON studies.trait_id = traits.id
  WHERE studies.data_type = 'phenotype' AND studies.variant_type = 'common'
")

number_of_genes <- lapply(seq_along(all_phenotype_studies$id), function(i) {
  study <- all_phenotype_studies[i, , drop = FALSE]
  study_id <- study$id
  study_name <- study$study_name
  all_studies_in_coloc_groups_for_study <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT COUNT(DISTINCT studies.gene_id) AS num_genes
    FROM coloc_groups
    JOIN studies ON coloc_groups.study_id = studies.id
    JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
    WHERE study_extractions.min_p < 5e-8
    AND studies.data_type != 'methylation'
    AND coloc_groups.coloc_group_id IN (SELECT DISTINCT coloc_group_id
      FROM coloc_groups
      JOIN studies on coloc_groups.study_id = studies.id
      JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
      WHERE studies.id = {study_id}
      AND study_extractions.min_p < 5e-8
      AND (study_extractions.cis_trans = 'cis' OR study_extractions.cis_trans IS NULL)
    )
  "))
  all_studies_in_coloc_groups_for_study$study_id <- study_id
  all_studies_in_coloc_groups_for_study$trait_name <- study$trait_name
  all_studies_in_coloc_groups_for_study$sample_size <- study$sample_size
  return(all_studies_in_coloc_groups_for_study)
})
number_of_genes <- do.call(rbind, number_of_genes)

# Create a summary with counts and add small offset for visibility
count_data <- number_of_genes %>%
  count(num_genes, name = "count") %>%
  mutate(count_offset = count + 0.3)  # Add small offset to make count=1 visible

p <- ggplot(count_data, aes(x = num_genes, y = count_offset)) +
geom_bar(stat = "identity", fill = cb_blue, alpha = 1) +
  scale_x_continuous() +
  scale_y_log10() +
  labs(
    title = "Number of genes colocalising per trait",
    x = "Frequency (log scale)",
    y = "Number of genes colocalising per trait",
    caption = paste("Mean:", round(mean(number_of_genes$num_genes), 2),
                   "| Median:", round(median(number_of_genes$num_genes), 2))
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_num_genes_colocalising_file <- file.path(report_analysis_dir, "pleiotropy_d_num_genes_colocalising.png")
ggsave(p, filename = p_num_genes_colocalising_file, width = 8, height = 6)
print(p)

number_of_genes_summary <- number_of_genes %>%
  arrange(desc(num_genes)) %>%
  head(10)
knitr::kable(number_of_genes_summary, caption = "Top 10 studies by number of genes colocalising")

median_sample_size <- median(number_of_genes$sample_size, na.rm = TRUE)
median_num_genes <- median(number_of_genes$num_genes, na.rm = TRUE)

special_ld_block_ids <- c(60, 1058, 1059)

for (special_ld_block_id in special_ld_block_ids) {
valid_colocs <- dplyr::filter(joined_coloc_results, ld_block_id == special_ld_block_id & h4 >= 0.8 & min_p_a < 5e-8 & min_p_b < 5e-8 & !is.na(gene_a) & !is.na(gene_b) & gene_a != gene_b)
count_gene_pairs <- valid_colocs %>%
  group_by(gene_a, gene_b) %>%
  summarise(count = n(), .groups = 'drop') %>%
  arrange(desc(count))

ld_block_id <- special_ld_block_id
all_coloc_groups_in_ld_block <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT coloc_groups_wide.coloc_group_id, coloc_groups_wide.gene, coloc_groups_wide.data_type, gene_annotations.gene_biotype FROM coloc_groups_wide
  LEFT JOIN gene_annotations ON coloc_groups_wide.gene_id = gene_annotations.id
  WHERE ld_block_id = {ld_block_id}
  AND min_p <= 5e-8
  AND gene_annotations.gene_biotype = 'protein_coding'
"))

ld_block_info <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT * FROM ld_blocks WHERE id = {ld_block_id}"))

# Create gene pairs within each coloc group
gene_pairs <- all_coloc_groups_in_ld_block %>%
  filter(!is.na(gene) & data_type != 'methylation') %>%  # Filter out NA genes
  group_by(coloc_group_id) %>%
  do({
    genes_in_group <- unique(.$gene)
    genes_in_group <- genes_in_group[!is.na(genes_in_group)]  # Additional safety check
    if(length(genes_in_group) > 1) {
      # Create all pairwise combinations of genes in this coloc group
      gene_combinations <- combn(genes_in_group, 2, simplify = FALSE)
      data.frame(
        gene_a = sapply(gene_combinations, function(x) x[1]),
        gene_b = sapply(gene_combinations, function(x) x[2]),
        stringsAsFactors = FALSE
      )
    } else {
      data.frame(gene_a = character(), gene_b = character(), stringsAsFactors = FALSE)
    }
  }) %>%
  ungroup() %>%
  # Filter out any remaining NA pairs before counting
  filter(!is.na(gene_a) & !is.na(gene_b)) %>%
  # Count how many times each gene pair appears across coloc groups
  group_by(gene_a, gene_b) %>%
  summarise(
    weight = n(),  # Number of coloc groups where this gene pair appears
    .groups = 'drop'
  )

# Create igraph object and include isolated protein-coding genes within this LD block (no coloc edges)
library(igraph)
gene_graph <- graph_from_data_frame(gene_pairs, directed = FALSE)

block_genes <- protein_coding_genes |>
  filter(
    chr == ld_block_info$chr,
    start >= ld_block_info$start,
    stop  <= ld_block_info$stop
  ) |>
  pull(gene) |>
  unique()

missing_block_genes <- setdiff(block_genes, V(gene_graph)$name)
if (length(missing_block_genes) > 0) {
  gene_graph <- add_vertices(gene_graph, nv = length(missing_block_genes), name = missing_block_genes)
}

# Create visualization
library(ggraph)
library(tidygraph)

# Convert to tidygraph for easier plotting and add node attributes
# Map neighborhood gene density onto nodes (default 0 when missing)
gene_density_lookup <- gene_density |>
  select(gene, neighborhood_genes)

gene_tidy_graph <- as_tbl_graph(gene_graph) %>%
  activate(nodes) %>%
  mutate(
    connections = centrality_degree(),
    betweenness = centrality_betweenness(),
    closeness = centrality_closeness(),
    neighborhood_genes = gene_density_lookup$neighborhood_genes[match(name, gene_density_lookup$gene)],
    neighborhood_genes = tidyr::replace_na(neighborhood_genes, 0)
  )

num_isolated_genes <- sum(degree(gene_graph) == 0)
gene_components <- components(gene_graph)
component_sizes <- gene_components$csize
num_nontrivial_clusters <- sum(component_sizes >= 2)

# Plot the network
network_plot <- ggraph(gene_tidy_graph, layout = "fr", niter = 1000) +
  geom_edge_link(aes(alpha = weight), color = "#444444") +
  geom_node_point(aes(size = connections, fill = neighborhood_genes), shape = 21, color = "black", stroke = 0.4, alpha = 0.8) +
  geom_node_text(aes(label = name), size = 2.2, colour = "grey40", repel = TRUE) +
  scale_edge_alpha_continuous(range = c(0.1, 0.8)) +
  scale_size_continuous(range = c(2, 8)) +
  scale_fill_gradient(low = cb_light_blue, high = cb_blue, name = "# of genes ±1Mb") +
  labs(
    title = "Gene-Gene Colocalization Network within an LD Block",
    subtitle = paste("LD Block", ld_block_info$ld_block),
    caption = paste(
      "Total Genes:", vcount(gene_graph),
      # "| Links in coloc groups:", ecount(gene_graph),
      "| Isolated genes:", num_isolated_genes,
      "| Gene clusters (≥2 genes):", num_nontrivial_clusters
    )
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )

p_gene_gene_network_file <- file.path(report_analysis_dir, glue::glue("pleiotropy_gene_gene_network_{special_ld_block_id}.png"))
ggsave(network_plot, filename = p_gene_gene_network_file, width = 8, height = 6)
print(network_plot)
}
```

```{r relationship_between_gene_density_and_coloc_partners, echo=F}
# Get all LD blocks that have coloc groups with protein-coding genes (cis only)
ld_blocks_with_coloc <- DBI::dbGetQuery(studies_conn, "
  SELECT DISTINCT ld_block_id, ld_blocks.chr, ld_blocks.start, ld_blocks.stop, ld_blocks.ld_block
  FROM coloc_groups_wide
  JOIN ld_blocks ON coloc_groups_wide.ld_block_id = ld_blocks.id
  WHERE coloc_groups_wide.min_p < 5e-8
  AND coloc_groups_wide.gene IS NOT NULL
  AND coloc_groups_wide.cis_trans = 'cis'
  ORDER BY ld_block_id
")
protein_coding_genes <- DBI::dbGetQuery(studies_conn, "
  SELECT gene, chr, start, stop, gene_biotype
  FROM gene_annotations
  WHERE gene_biotype = 'protein_coding'
")

genes_dt <- as.data.table(protein_coding_genes)
# Define ±1Mb window around each gene; ensure ordering for foverlaps
genes_dt[, region_start := start - 1e6]
genes_dt[, region_end   := stop  + 1e6]
setkey(genes_dt, chr, start, stop)

# Count neighbouring protein-coding genes within ±1Mb (exclude self)
overlaps <- foverlaps(
  genes_dt[, .(gene, chr, region_start, region_end)],
  genes_dt,
  by.x = c("chr", "region_start", "region_end"),
  by.y = c("chr", "start", "stop"),
  type = "any",
  nomatch = 0
)

gene_density <- overlaps[
  ,
  .(neighborhood_genes = .N - 1L),  # subtract self
  by = gene
]

# For each LD block, calculate coloc partners limited to genes within that same block
ld_block_coloc_analysis <- lapply(seq_len(nrow(ld_blocks_with_coloc)), function(i) {
  block_info <- ld_blocks_with_coloc[i, ]
  block_id <- block_info$ld_block_id
  
  # Get all protein-coding genes within this LD block
  block_genes <- protein_coding_genes |>
    filter(
      chr == block_info$chr,
      start >= block_info$start,
      stop <= block_info$stop
    ) |>
    pull(gene) |>
    unique()
  
  if (length(block_genes) == 0) return(NULL)
  
  # Get coloc groups in this LD block (cis only, no trans)
  coloc_in_block <- DBI::dbGetQuery(studies_conn, glue::glue("
    SELECT coloc_group_id, gene
    FROM coloc_groups_wide
    WHERE ld_block_id = {block_id}
    AND min_p < 5e-8
    AND gene IS NOT NULL
    AND cis_trans = 'cis'
    AND data_type != 'methylation'
  ")) |>
    filter(gene %in% block_genes)
  
  if (nrow(coloc_in_block) == 0) {
    # No coloc groups, but still include genes with zero partners
    return(data.frame(
      ld_block_id = block_id,
      ld_block = block_info$ld_block,
      gene = block_genes,
      num_coloc_partner_genes_in_block = 0,
      stringsAsFactors = FALSE
    ))
  }
  
  # Create gene partner pairs within coloc groups, but only for genes in this block
  gene_partner_pairs_block <- coloc_in_block |>
    filter(gene %in% block_genes) |>
    group_by(coloc_group_id) |>
    summarise(genes = list(unique(gene)), .groups = "drop") |>
    filter(lengths(genes) > 1) |>
    mutate(pairs = map(genes, function(g) {
      if (length(g) < 2) {
        return(tibble::tibble(gene = character(), partner = character()))
      }
      cmb <- combn(g, 2)
      tibble::tibble(
        gene = c(cmb[1, ], cmb[2, ]),
        partner = c(cmb[2, ], cmb[1, ])
      )
    })) |>
    select(pairs) |>
    tidyr::unnest(pairs)
  
  # Only count partners that are also in this block (check if partner column exists)
  if (nrow(gene_partner_pairs_block) > 0 && "partner" %in% names(gene_partner_pairs_block)) {
    gene_partner_pairs_block <- gene_partner_pairs_block |>
      filter(partner %in% block_genes)
  } else {
    # If no pairs or no partner column, create empty result with correct structure
    gene_partner_pairs_block <- tibble::tibble(gene = character(), partner = character())
  }
  
  # Count coloc partners per gene (limited to this block)
  if (nrow(gene_partner_pairs_block) > 0 && "partner" %in% names(gene_partner_pairs_block)) {
    gene_coloc_partners_block <- gene_partner_pairs_block |>
      group_by(gene) |>
      summarise(num_coloc_partner_genes_in_block = n_distinct(partner), .groups = "drop")
  } else {
    gene_coloc_partners_block <- tibble::tibble(gene = character(), num_coloc_partner_genes_in_block = integer())
  }
  
  # Include all block genes, setting zero for those without coloc partners
  all_block_genes_df <- data.frame(
    gene = block_genes,
    stringsAsFactors = FALSE
  ) |>
    left_join(gene_coloc_partners_block, by = "gene") |>
    mutate(
      num_coloc_partner_genes_in_block = tidyr::replace_na(num_coloc_partner_genes_in_block, 0),
      ld_block_id = block_id,
      ld_block = block_info$ld_block
    )
  
  return(all_block_genes_df)
}) |>
  bind_rows()

# Join with gene density
ld_block_coloc_density <- ld_block_coloc_analysis |>
  left_join(gene_density, by = "gene") |>
  mutate(neighborhood_genes = tidyr::replace_na(neighborhood_genes, 0))

# Fit linear model
lm_density_coloc <- lm(num_coloc_partner_genes_in_block ~ neighborhood_genes, data = ld_block_coloc_density)
lm_density_coloc_summary <- summary(lm_density_coloc)
density_coloc_r2 <- lm_density_coloc_summary$r.squared
density_coloc_p <- pf(lm_density_coloc_summary$fstatistic[1], 
                      lm_density_coloc_summary$fstatistic[2], 
                      lm_density_coloc_summary$fstatistic[3], 
                      lower.tail = FALSE)
density_coloc_slope <- coef(lm_density_coloc)[2]

# Create annotation text
density_coloc_stats_text <- paste0(
  "Slope = ", round(density_coloc_slope, 3), "\n",
  "R² = ", round(density_coloc_r2, 3), "\n",
  "p = ", formatC(density_coloc_p, format = "e", digits = 2)
)

# Create scatter plot
gene_density_coloc_scatter <- ggplot(ld_block_coloc_density, 
  aes(x = neighborhood_genes, y = num_coloc_partner_genes_in_block)) +
  geom_point(alpha = 0.6, color = cb_blue, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.5) +
  annotate("label", x = -Inf, y = Inf, label = density_coloc_stats_text, 
           hjust = -0.1, vjust = 1.5, size = 3.5, family = "mono",
           fill = "white", alpha = 0.8) +
  labs(
    title = "Gene density vs colocalizing partners within LD blocks",
    x = "Number of protein-coding genes in ±1Mb window",
    y = "Number of colocalizing protein-coding genes",
    caption = paste("Each point is a gene. Coloc partners counted only within the same LD block")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

gene_density_coloc_file <- file.path(report_analysis_dir, "pleiotropy_b_gene_density_vs_coloc_without_methylation.png")
ggsave(gene_density_coloc_scatter, filename = gene_density_coloc_file, width = 8, height = 6)
print(gene_density_coloc_scatter)
```

## Genes that colocalise with the most traits (or trait categories?).  
Limited to cis-coloc pairs with h4 > 0.8 and p-values < 5e-8.

```{r genes_colocalising_with_most_traits, echo=F}
significant_cis_coloc_pairs <- joined_coloc_results %>%
  filter(h4 > 0.8 & (is.na(cis_trans_a) | cis_trans_a == 'cis') & (is.na(cis_trans_b) | cis_trans_b == 'cis') & min_p_a < 5e-8 & min_p_b < 5e-8)

most_gene_a_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype' & data_type_a != 'methylation') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait = trait_name_b)

most_gene_b_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype' & data_type_b != 'methylation') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait = trait_name_a)

most_gene_a_colocalising_trait_category <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype' & data_type_a != 'methylation') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait_category = trait_category_b)

most_gene_b_colocalising_trait_category <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype' & data_type_b != 'methylation') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait_category = trait_category_a)

most_genes_colocalising_with_traits <- rbind(most_gene_a_colocalising, most_gene_b_colocalising)
most_genes_colocalising_with_trait_categories <- rbind(most_gene_a_colocalising_trait_category, most_gene_b_colocalising_trait_category)

most_genes_colocalising_summary <- most_genes_colocalising_with_traits %>%
  group_by(gene) %>%
  summarise(num_traits = n_distinct(trait)) %>%
  arrange(desc(num_traits)) %>%
  head(50)

knitr::kable(most_genes_colocalising_summary, caption = "Genes that colocalise with the most traits")

top_gene <- most_genes_colocalising_summary$gene[1]
all_phenotypes_for_top_gene <- most_genes_colocalising_with_traits %>%
  filter(gene == top_gene) %>%
  select(trait) %>%
  distinct()
```

## Apparent pleiotropy due to LD

Association vs. colocalization: How often would we see a genetic association with a trait that doesn't have a coloc result?

For any causal variant:

* How many traits does it coloc with (significant h4)
* How many traits does it associate with (significant h3)
* Rate of incorrectness due to LD = significant h4 / significant h3
* How many significant h4 traits are spurious? (Spurious: h4 > 0.8 & failed the clustering step)

### Proximal pairs (within 1 kb)

Restricting to colocalization pairs where the two study extractions are on the same chromosome and within 1 kilobase (≤ 1,000 bp) of each other. This removes any confounding from long-range LD and focuses on pairs whose causal signals are essentially co-localised.

```{r h4_h3_ratio_proximal, echo=F}
hemoglobin_study_id <- 'ebi-a-GCST90002384'

# Filter to pairs on the same chromosome and within 1 kb
proximal_pairs <- joined_coloc_results %>%
  dplyr::filter(
    chr_a == chr_b,
    abs(bp_a - bp_b) <= 1000
  )

cat(glue::glue("Total coloc pairs within 1 kb: {scales::comma(nrow(proximal_pairs))} ",
               "out of {scales::comma(nrow(joined_coloc_results))} total ",
               "({round(nrow(proximal_pairs) / nrow(joined_coloc_results) * 100, 1)}%)\n\n"))

prox_num_h4 <- nrow(proximal_pairs[proximal_pairs$h4 > 0.8, ])
prox_num_h4_spurious <- nrow(proximal_pairs[proximal_pairs$h4 > 0.8 & proximal_pairs$false_positive, ])
prox_num_h3 <- nrow(proximal_pairs[proximal_pairs$h3 > 0.8, ])
prox_num_h3_spurious <- nrow(proximal_pairs[proximal_pairs$h3 > 0.8 & proximal_pairs$false_positive, ])

# Same hemoglobin example, restricted to proximal pairs
prox_hemoglobin <- proximal_pairs[proximal_pairs$study_name_a == hemoglobin_study_id | proximal_pairs$study_name_b == hemoglobin_study_id, ]
prox_hemo_h4 <- nrow(prox_hemoglobin[prox_hemoglobin$h4 > 0.8, ])
prox_hemo_h3 <- nrow(prox_hemoglobin[prox_hemoglobin$h3 > 0.8, ])
prox_hemo_h4_spurious <- nrow(prox_hemoglobin[prox_hemoglobin$h4 > 0.8 & prox_hemoglobin$false_positive, ])

prox_summary <- tibble::tribble(
  ~trait, ~num_significant_h4, ~num_spurious_h4, ~num_significant_h3, ~h3_h4_ratio,
  'All (≤1 kb)', prox_num_h4, prox_num_h4_spurious, prox_num_h3, ifelse(prox_num_h4 > 0, prox_num_h3 / prox_num_h4, NA),
  'Hemoglobin (≤1 kb)', prox_hemo_h4, prox_hemo_h4_spurious, prox_hemo_h3, ifelse(prox_hemo_h4 > 0, prox_hemo_h3 / prox_hemo_h4, NA)
)

knitr::kable(prox_summary, caption = "H3 vs H4 for proximal pairs (within 1 kb)")
```

### H3/H4 matrix by data type — proximal pairs only
This is a big matrix, skipping for now.
```{r h4_h3_ratio_proximal_matrix, echo=F, eval=F}
prox_coloc_matrix <- dplyr::filter(proximal_pairs, h4 > 0.8 | h3 > 0.8)

prox_h4_matrix <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(prox_h4_matrix) <- unique_data_types
colnames(prox_h4_matrix) <- unique_data_types

prox_h3_matrix <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(prox_h3_matrix) <- unique_data_types
colnames(prox_h3_matrix) <- unique_data_types

if (nrow(prox_coloc_matrix) > 0) {
  for (i in 1:nrow(prox_coloc_matrix)) {
    type_a <- prox_coloc_matrix$data_type_a[i]
    type_b <- prox_coloc_matrix$data_type_b[i]

    if (prox_coloc_matrix$h4[i] > 0.8) {
      prox_h4_matrix[type_a, type_b] <- prox_h4_matrix[type_a, type_b] + 1
      if (type_a != type_b) {
        prox_h4_matrix[type_b, type_a] <- prox_h4_matrix[type_b, type_a] + 1
      }
    } else if (prox_coloc_matrix$h3[i] > 0.8) {
      prox_h3_matrix[type_a, type_b] <- prox_h3_matrix[type_a, type_b] + 1
      if (type_a != type_b) {
        prox_h3_matrix[type_b, type_a] <- prox_h3_matrix[type_b, type_a] + 1
      }
    }
  }
}

prox_h4_matrix[lower.tri(prox_h4_matrix)] <- NA
prox_h4_display <- as.data.frame(prox_h4_matrix)
prox_h4_display[, unique_data_types] <- lapply(prox_h4_display[, unique_data_types], scales::comma)
knitr::kable(prox_h4_display, caption = "H4 > 0.8 between data types (≤1 kb pairs)")

prox_h3_matrix[lower.tri(prox_h3_matrix)] <- NA
prox_h3_display <- as.data.frame(prox_h3_matrix)
prox_h3_display[, unique_data_types] <- lapply(prox_h3_display[, unique_data_types], scales::comma)
knitr::kable(prox_h3_display, caption = "H3 > 0.8 between data types (≤1 kb pairs)")

prox_ratio <- as.data.frame(prox_h3_matrix) / as.data.frame(prox_h4_matrix)
knitr::kable(prox_ratio, caption = "Ratio of H3 > 0.8 to H4 > 0.8 between data types (≤1 kb pairs)")
```

### All pairs (unrestricted)

```{r h4_h3_ratio, echo=F}
num_significant_h4 <- nrow(joined_coloc_results[joined_coloc_results$h4 > 0.8, ])
num_spurious_h4 <- nrow(joined_coloc_results[joined_coloc_results$h4 > 0.8 & joined_coloc_results$false_positive, ])
num_significant_h3 <- nrow(joined_coloc_results[joined_coloc_results$h3 > 0.8, ])
num_spurious_h3 <- nrow(joined_coloc_results[joined_coloc_results$h3 > 0.8 & joined_coloc_results$false_positive, ])

hemoglobin_study_id <- 'ebi-a-GCST90002384'
hemoglobin_example <- joined_coloc_results[joined_coloc_results$study_name_a == hemoglobin_study_id | joined_coloc_results$study_name_b == hemoglobin_study_id, ]

hemoglobin_h4 <- nrow(hemoglobin_example[hemoglobin_example$h4 > 0.8, ])
hemoglobin_h3 <- nrow(hemoglobin_example[hemoglobin_example$h3 > 0.8, ])
hemoglobin_h4_spurious <- nrow(hemoglobin_example[hemoglobin_example$h4 > 0.8 & hemoglobin_example$false_positive, ])
hemoglobin_h3_spurious <- nrow(hemoglobin_example[hemoglobin_example$h3 > 0.8 & hemoglobin_example$false_positive, ])

h4_summary <- tibble::tribble(
  ~trait, ~num_significant_h4, ~num_spurious_h4, ~num_significant_h3, ~h3_h4_ratio,
  'All', num_significant_h4, num_spurious_h4, num_significant_h3, num_significant_h3 / num_significant_h4,
  'Hemoglobin', hemoglobin_h4, hemoglobin_h4_spurious, hemoglobin_h3, hemoglobin_h3 / hemoglobin_h4
)

knitr::kable(h4_summary)
```

#### Breakdown of ratio of h3 to h4 by data type

This is a big matrix, skipping for now.
```{r matrix_of_coloc_results, echo=F, eval=F}
coloc_matrix <- dplyr::filter(joined_coloc_results, h4 > 0.8 | h3 > 0.8)

# Create the n×n matrix
h4_coloc_matrix_by_type <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(h4_coloc_matrix_by_type) <- unique_data_types
colnames(h4_coloc_matrix_by_type) <- unique_data_types

h3_coloc_matrix_by_type <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(h3_coloc_matrix_by_type) <- unique_data_types
colnames(h3_coloc_matrix_by_type) <- unique_data_types

# Fill the matrix with coloc counts (combining symmetric entries)
for (i in 1:nrow(coloc_matrix)) {
  type_a <- coloc_matrix$data_type_a[i]
  type_b <- coloc_matrix$data_type_b[i]

  if (coloc_matrix$h4[i] > 0.8) {
    h4_coloc_matrix_by_type[type_a, type_b] <- h4_coloc_matrix_by_type[type_a, type_b] + 1
    if (type_a != type_b) {
      h4_coloc_matrix_by_type[type_b, type_a] <- h4_coloc_matrix_by_type[type_b, type_a] + 1
    }
  } else if (coloc_matrix$h3[i] > 0.8) {
    h3_coloc_matrix_by_type[type_a, type_b] <- h3_coloc_matrix_by_type[type_a, type_b] + 1
    if (type_a != type_b) {
      h3_coloc_matrix_by_type[type_b, type_a] <- h3_coloc_matrix_by_type[type_b, type_a] + 1
    }
  }
}

h4_coloc_matrix_by_type[lower.tri(h4_coloc_matrix_by_type)] <- NA
h4_coloc_matrix_by_type <- as.data.frame(h4_coloc_matrix_by_type)
h4_coloc_pretty <- h4_coloc_matrix_by_type
h4_coloc_pretty[, unique_data_types] <- lapply(h4_coloc_pretty[, unique_data_types], scales::comma)
knitr::kable(h4_coloc_pretty, caption = "Number of H4 > 0.8 between data types")

h3_coloc_matrix_by_type[lower.tri(h3_coloc_matrix_by_type)] <- NA
h3_coloc_matrix_by_type <- as.data.frame(h3_coloc_matrix_by_type)
h3_coloc_pretty <- h3_coloc_matrix_by_type
h3_coloc_pretty[, unique_data_types] <- lapply(h3_coloc_pretty[, unique_data_types], scales::comma)
knitr::kable(h3_coloc_pretty, caption = "Number of H3 > 0.8 between data types")

h3_h4_ratio <- h3_coloc_matrix_by_type / h4_coloc_matrix_by_type
knitr::kable(h3_h4_ratio, caption = "Ratio of H3 > 0.8 to H4 > 0.8 between data types")
```

# LD relationship between causal SNPs

### How often Causal SNPs are in LD with each other
```{r causal_snps_in_ld, echo=F}
# Find variant_ids that are in coloc_groups table under different coloc_group_ids
snp_with_multiple_coloc_groups <- DBI::dbGetQuery(
  studies_conn,
  "
  WITH coloc_groups_with_many_entries AS (
    SELECT coloc_group_id
    FROM coloc_groups
    GROUP BY coloc_group_id
    HAVING COUNT(*) > 5
  )
  SELECT variant_id, COUNT(DISTINCT coloc_group_id) AS num_coloc_groups
  FROM coloc_groups
  WHERE coloc_group_id IN (SELECT coloc_group_id FROM coloc_groups_with_many_entries)
  GROUP BY variant_id
  HAVING COUNT(DISTINCT coloc_group_id) > 1
  "
)

# Print number of such SNPs and preview data
num_snps_multiple_groups <- nrow(snp_with_multiple_coloc_groups)
cat('Number of variant_ids found in multiple coloc_group_ids:', num_snps_multiple_groups, '\n')

all_causal_snps <- DBI::dbGetQuery(studies_conn, "SELECT DISTINCT variant_id FROM coloc_groups")

dont_print <- DBI::dbExecute(ld_conn, glue::glue("ATTACH '{studies_db_file}' AS studies"))

formatted_variant_ids <- paste0(all_causal_snps$variant_id, collapse = ",")
snp_pair_ld_scores <- DBI::dbGetQuery(ld_conn, glue::glue("SELECT ld.*, studies.ld_blocks.ld_block AS ld_block FROM ld
  JOIN studies.ld_blocks ON ld.ld_block_id = studies.ld_blocks.id
  WHERE lead_variant_id IN ({formatted_variant_ids})
  AND proxy_variant_id IN ({formatted_variant_ids})
  "
))

snp_pair_ld_scores$r2 <- snp_pair_ld_scores$r ^ 2

r2_thresholds <- c(0.8, 0.9, 0.95, 0.99)

ld_results <- lapply(r2_thresholds, function(r2_threshold) {
  snp_pairs_in_ld <- snp_pair_ld_scores[snp_pair_ld_scores$r2 > r2_threshold, ]
  unique_causal_snps_in_ld <- length(unique(snp_pairs_in_ld$lead_variant_id))
  return(data.frame(
    r2_threshold = r2_threshold,
    unique_causal_snps_in_ld = unique_causal_snps_in_ld,
    number_of_snp_pairs_in_ld = nrow(snp_pairs_in_ld),
    percentage_of_causal_snps_in_ld = unique_causal_snps_in_ld / length(all_causal_snps$variant_id)
  ))
})

ld_results <- do.call(rbind, ld_results)
knitr::kable(ld_results, caption = "Number of unique causal SNPs in LD by r2 threshold")

snp_pairs_in_ld <- snp_pair_ld_scores[snp_pair_ld_scores$r2 > 0.8, ]
ld_block_summary <- snp_pairs_in_ld %>%
  group_by(ld_block_id, ld_block) %>%
  summarise(
    num_snp_pairs = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(num_snp_pairs)) %>%
  head(10)

knitr::kable(ld_block_summary, 
  caption = "Top 10 LD blocks by number of causal SNP pairs in LD",
  col.names = c("LD Block ID", "LD Block", "Number of SNP Pairs"))


```

## How often does a coloc group have multiple genes?  Multiple phenotypes? (cis only)

```{r multiple_genes_in_coloc_group, echo=F}
coloc_group_data <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*,
  studies.gene, studies.sample_size, studies.data_type, studies.tissue, studies.study_name,
  gene_annotations.chr, gene_annotations.start, gene_annotations.stop, gene_annotations.gene_biotype,
  study_extractions.cis_trans, traits.trait_category, traits.trait_name
  FROM coloc_groups
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN study_extractions ON coloc_groups.study_extraction_id = studies.study_extractions.id
  JOIN traits ON studies.trait_id = traits.id
  LEFT JOIN gene_annotations ON studies.gene_id = gene_annotations.id
  WHERE study_extractions.min_p < 5e-8
  AND (study_extractions.cis_trans = 'cis' OR study_extractions.cis_trans IS NULL)
")

# Filter to only keep coloc groups with multiple entries
coloc_group_data <- coloc_group_data |>
  group_by(coloc_group_id) |>
  filter(n() > 1) |>
  ungroup()

gene_phenotype_counts_per_coloc_group <- coloc_group_data |>
  group_by(coloc_group_id) |>
  summarise(
    unique_genes = n_distinct(gene, na.rm = TRUE),
    num_phenotypes = sum(data_type == "phenotype", na.rm = TRUE),
    .groups = 'drop'
  )


gene_phenotype_bins <- gene_phenotype_counts_per_coloc_group |>
  mutate(
    genes_bin = case_when(
      unique_genes == 0 ~ "No genes",
      unique_genes <= 5 ~ paste0(unique_genes, " genes"),
      TRUE ~ ">5 genes"
    ),
    phenotypes_bin = case_when(
      num_phenotypes == 0 ~ "No phenotypes",
      num_phenotypes <= 5 ~ paste0(num_phenotypes, " phenotypes"),
      TRUE ~ ">5 phenotypes"
    )
  ) |>
  mutate(
    genes_bin = factor(genes_bin, levels = c("No genes", paste0(1:5, " genes"), ">5 genes")),
    phenotypes_bin = factor(phenotypes_bin, levels = c("No phenotypes", paste0(1:5, " phenotypes"), ">5 phenotypes"))
  )

genes_phenotypes_matrix <- table(
  gene_phenotype_bins$genes_bin,
  gene_phenotype_bins$phenotypes_bin
)

genes_phenotypes_df <- as.data.frame(genes_phenotypes_matrix)
colnames(genes_phenotypes_df) <- c("genes_bin", "phenotypes_bin", "count")

genes_phenotypes_matrix_df <- genes_phenotypes_df |>
  tidyr::pivot_wider(
    names_from = phenotypes_bin,
    values_from = count,
    values_fill = 0
  )

knitr::kable(genes_phenotypes_matrix_df, caption = "Number of colocalization groups by unique genes and phenotypes (methQTL are neither genes nor phenotypes)")

percentage_of_coloc_groups <- genes_phenotypes_matrix_df |>
  mutate(across(-genes_bin, ~ .x / nrow(gene_phenotype_bins))) |>
  mutate(across(-genes_bin, ~ scales::percent(.x, accuracy = 0.01)))

knitr::kable(percentage_of_coloc_groups, caption = "Percentage of colocalization groups by unique genes and phenotypes (methQTL are neither genes nor phenotypes)")
```

### Distinct phenotype categories in the same colocalization group

How can we quantify the pleiotropy of phenotypes, when different phenotypes can be so similar?
How else can we quantify biological pleiotropy (vs. LD)?


```{r distinct_phenotype_categories_in_same_coloc_group, echo=F}
distinct_phenotype_categories_in_same_coloc_group <- coloc_group_data |>
  filter(data_type == 'phenotype') |>
  group_by(coloc_group_id) |>
  summarise(distinct_phenotype_categories = n_distinct(trait_category)) |>
  filter(distinct_phenotype_categories > 0)


trait_category_counts_per_coloc_group <- distinct_phenotype_categories_in_same_coloc_group |>
  mutate(
    trait_category_bin = case_when(
      distinct_phenotype_categories == 0 ~ "0 trait categories",
      TRUE ~ paste0(distinct_phenotype_categories, " trait categories")
    )
  ) |>
  group_by(trait_category_bin) |>
  summarise(count = n(), .groups = 'drop')

knitr::kable(trait_category_counts_per_coloc_group, caption = "Distinct trait categories in the same colocalization group")

```


# What fraction of known GWAS hits are no known molecular trait associated with it?

```{r gwas_hits_no_molecular_trait, echo=F}
all_phenotype_study_extractions <- DBI::dbGetQuery(studies_conn, "SELECT study_extractions.*,
  variant_annotations.feature_type, variant_annotations.consequence, variant_annotations.impact, variant_annotations.cdna_position,
  variant_annotations.cds_position, variant_annotations.protein_position, variant_annotations.amino_acids, variant_annotations.codons,
  variant_annotations.gene as snp_gene, traits.trait_name as trait_name, traits.trait_category as trait_category
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  JOIN variant_annotations ON study_extractions.variant_id = variant_annotations.id
  WHERE studies.data_type = 'phenotype' and studies.variant_type = 'common' and study_extractions.min_p < 5e-8
")

all_coloc_groups <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*, studies.data_type, study_extractions.min_p,
  variant_annotations.feature_type, variant_annotations.consequence, variant_annotations.impact, variant_annotations.cdna_position,
  variant_annotations.cds_position, variant_annotations.protein_position, variant_annotations.amino_acids, variant_annotations.codons,
  variant_annotations.gene as snp_gene, traits.trait_category as trait_category, traits.trait_name as trait_name
  FROM coloc_groups
  JOIN variant_annotations ON coloc_groups.variant_id = variant_annotations.id
  JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  WHERE study_extractions.min_p < 5e-8
")


all_study_extractions_in_coloc_groups <- unique(all_coloc_groups$study_extraction_id)
all_phenotype_study_extractions_without_coloc <- all_phenotype_study_extractions %>%
  filter(!id %in% all_study_extractions_in_coloc_groups)
study_extractions_with_phenotype_only_coloc <- all_coloc_groups %>%
  group_by(coloc_group_id) %>%
  filter(all(data_type == "phenotype")) %>%
  ungroup()
all_significant_non_molecular_coloc <- dplyr::bind_rows(all_phenotype_study_extractions_without_coloc, study_extractions_with_phenotype_only_coloc)

gwas_hits_and_molecular_traits_by_trait_category <- function(tc) {
  all_study_extractions_in_coloc_groups <- unique(all_coloc_groups$study_extraction_id)
  if (is.na(tc)) {
    all_phenotype_study_extractions_for_tc <- all_phenotype_study_extractions
  } else {
    all_phenotype_study_extractions_for_tc <- all_phenotype_study_extractions[all_phenotype_study_extractions$trait_category == tc, ]
  }

  all_phenotype_study_extractions_without_coloc <- all_phenotype_study_extractions_for_tc %>%
    filter(!id %in% all_study_extractions_in_coloc_groups)

  if (is.na(tc)) {
    study_extractions_with_phenotype_only_coloc <- all_coloc_groups %>%
      group_by(coloc_group_id) %>%
      filter(all(data_type == "phenotype")) %>%
      ungroup()
  } else {
    study_extractions_with_phenotype_only_coloc <- all_coloc_groups %>%
      group_by(coloc_group_id) %>%
      filter(any(trait_category == tc)) %>%
      filter(all(data_type == "phenotype")) %>%
      ungroup() %>%
      filter(trait_category == tc)
  }

  all_significant_non_molecular_coloc <- nrow(all_phenotype_study_extractions_without_coloc) + nrow(study_extractions_with_phenotype_only_coloc)

  percent_non_molecular_coloc <- (nrow(study_extractions_with_phenotype_only_coloc) + nrow(all_phenotype_study_extractions_without_coloc)) / nrow(all_phenotype_study_extractions_for_tc)

  if (is.na(tc)) tc <- "All"
  return(data.frame(
    trait_category = tc,
    all_study_extractions = nrow(all_phenotype_study_extractions_for_tc),
    all_study_extractions_without_coloc = nrow(all_phenotype_study_extractions_without_coloc),
    phenotype_only_coloc = nrow(study_extractions_with_phenotype_only_coloc),
    all_non_molecular_coloc = all_significant_non_molecular_coloc,
    percent_no_coloc = scales::percent(nrow(all_phenotype_study_extractions_without_coloc) / nrow(all_phenotype_study_extractions_for_tc), accuracy = 0.01),
    percent_non_molecular_coloc = scales::percent(percent_non_molecular_coloc, accuracy = 0.01)
  ))
}

cat("Breakdown of phenotype info by trait category")
gwas_hits_and_molecular_traits <- lapply(unique(all_coloc_groups$trait_category), function(tc) {
  gwas_hits_and_molecular_traits_by_trait_category(tc)
}) |> dplyr::bind_rows()

gwas_hits_and_molecular_traits <- gwas_hits_and_molecular_traits %>%
  arrange(trait_category != "All", trait_category)


knitr::kable(gwas_hits_and_molecular_traits, caption = "Breakdown of phenotype info by trait category")
```
