---
title: "Pleiotropy Investigation"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE, echo=F} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(ggplot2)
library(dplyr)
library(tidyr)
library(data.table)
library(purrr)
source('../../pipeline_steps/constants.R')

studies_db_file <- file.path(latest_results_dir, 'studies.db')
associations_db_file <- file.path(latest_results_dir, 'associations.db')
coloc_pairs_full_db_file <- file.path(latest_results_dir, 'coloc_pairs_full.db')
ld_db_file <- file.path(latest_results_dir, 'ld.db')

studies_conn <- duckdb::dbConnect(duckdb::duckdb(), studies_db_file, read_only = TRUE)
associations_conn <- duckdb::dbConnect(duckdb::duckdb(), associations_db_file, read_only = TRUE)
coloc_pairs_conn <- duckdb::dbConnect(duckdb::duckdb(), coloc_pairs_full_db_file, read_only = TRUE)
ld_conn <- duckdb::dbConnect(duckdb::duckdb(), ld_db_file, read_only = TRUE)

dont_print <- DBI::dbExecute(coloc_pairs_conn, glue::glue("ATTACH '{studies_db_file}' AS studies"))

cb_blue <- "#0072B2"
cb_light_blue <- "#80B9D9"
cb_orange <- "#E69F00"
cb_light_orange <- "#F3CF80"
```


### Questions to attempt to answer:

1. Association vs. colocalization: how often do we see a genetic association with a trait that doesn't have a coloc result?
   * Prevelance of H3 > 0.8 vs. H4 > 0.8
2. How many coloc groups have only qtls?  QTL and phenotype, QTL and multiple phenotypes
3. How many of these chosen causal SNPs are in LD with each other?
4. Modularity of pleiotropy
   * How can we quantify modularity?
   * How many genes are associated with a trait on average?
   * Does it change when including trans-QTLs?
   * What about gene-gene modularity? Could we have a graph where we count the number of associations between genes tagged?
   * Genes that colocalise with the most traits (or trait categories?).  
5. What fraction of known GWAS hits are no known molecular trait associated with it?
6. Is there consistency between of molecule-disease associations between common cis, rare cis and trans instruments?
7. Tissue specificity: can we break down coloc groups by tissue?
8. Can we calculate a 'pleiotropy score' for each SNP?  Would it be useful?


```{r load-data, echo=F}
joined_coloc_results <- DBI::dbGetQuery(coloc_pairs_conn, 'SELECT
  coloc_pairs.*,
  -- Study extraction A details
  study_extraction_a.unique_study_id AS study_extraction_a_unique_study_id,
  study_extraction_a.min_p AS min_p_a,
  study_extraction_a.cis_trans AS cis_trans_a,
  -- Study A details
  study_a.study_name AS study_name_a,
  study_a.gene AS gene_a,
  study_a.tissue AS tissue_a,
  study_a.sample_size AS sample_size_a,
  study_a.data_type AS data_type_a,
  trait_a.trait_category AS trait_category_a,
  trait_a.trait_name AS trait_name_a,
  -- Study extraction B details
  study_extraction_b.unique_study_id AS study_extraction_b_unique_study_id,
  study_extraction_b.min_p AS min_p_b,
  study_extraction_b.cis_trans AS cis_trans_b,
  -- Study B details
  study_b.study_name AS study_name_b,
  study_b.gene AS gene_b,
  study_b.tissue AS tissue_b,
  study_b.sample_size AS sample_size_b,
  study_b.data_type AS data_type_b,
  trait_b.trait_category AS trait_category_b,
  trait_b.trait_name AS trait_name_b
  FROM coloc_pairs
  -- Join for study extraction A
  JOIN studies.study_extractions AS study_extraction_a ON coloc_pairs.study_extraction_a_id = study_extraction_a.id
  JOIN studies.studies AS study_a ON study_extraction_a.study_id = study_a.id
  JOIN studies.traits AS trait_a ON study_a.trait_id = trait_a.id
  -- Join for study extraction B
  JOIN studies.study_extractions AS study_extraction_b ON coloc_pairs.study_extraction_b_id = study_extraction_b.id
  JOIN studies.studies AS study_b ON study_extraction_b.study_id = study_b.id
  JOIN studies.traits AS trait_b ON study_b.trait_id = trait_b.id
  WHERE coloc_pairs.h4 > 0.6 OR coloc_pairs.h3 > 0.6
')
unique_data_types <- DBI::dbGetQuery(studies_conn, "SELECT DISTINCT data_type FROM studies")$data_type
```

## Gene pleiotropy scores

```{r pleiotropy_scores, echo=F}
gene_pleiotropy_scores <- DBI::dbGetQuery(studies_conn, "SELECT * FROM gene_pleiotropy")

# Fit linear model for gene pleiotropy
lm_gene <- lm(distinct_protein_coding_genes ~ distinct_trait_categories, data = gene_pleiotropy_scores)
lm_gene_summary <- summary(lm_gene)
gene_r2 <- lm_gene_summary$r.squared
gene_p <- pf(lm_gene_summary$fstatistic[1], lm_gene_summary$fstatistic[2], lm_gene_summary$fstatistic[3], lower.tail = FALSE)
gene_slope <- coef(lm_gene)[2]

# Create annotation text
gene_stats_text <- paste0(
  "Slope = ", round(gene_slope, 3), "\n",
  "R² = ", round(gene_r2, 3), "\n",
  "p = ", formatC(gene_p, format = "e", digits = 2)
)

p_gene_pleiotropy <- ggplot(gene_pleiotropy_scores, aes(x = distinct_trait_categories, y = distinct_protein_coding_genes)) +
  geom_point(alpha = 0.6, size = 2, color = "#006400") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  annotate("label", x = -Inf, y = Inf, label = gene_stats_text, 
           hjust = -0.1, vjust = 1.5, size = 3.5, family = "mono",
           fill = "white", alpha = 0.8) +
  labs(
    title = "(E) Relationship between Gene Pleiotropy Metrics",
    x = "Number of Trait Categories Colocalising with Gene",
    y = "Number of Protein Coding Genes Colocalising with Gene"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(p_gene_pleiotropy, filename = file.path(results_analysis_dir, "pleiotropy_e_gene_pleiotropy_plot.png"), width = 8, height = 6)

snp_pleiotropy_scores <- DBI::dbGetQuery(studies_conn, "SELECT * FROM snp_pleiotropy")

# Fit linear model for SNP pleiotropy
lm_snp <- lm(distinct_protein_coding_genes ~ distinct_trait_categories, data = snp_pleiotropy_scores)
lm_snp_summary <- summary(lm_snp)
snp_r2 <- lm_snp_summary$r.squared
snp_p <- pf(lm_snp_summary$fstatistic[1], lm_snp_summary$fstatistic[2], lm_snp_summary$fstatistic[3], lower.tail = FALSE)
snp_slope <- coef(lm_snp)[2]

# Create annotation text
snp_stats_text <- paste0(
  "Slope = ", round(snp_slope, 3), "\n",
  "R² = ", round(snp_r2, 3), "\n",
  "p = ", formatC(snp_p, format = "e", digits = 2)
)

p_snp_pleiotropy <- ggplot(snp_pleiotropy_scores, aes(x = distinct_trait_categories, y = distinct_protein_coding_genes)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = cb_blue) +
  annotate("label", x = Inf, y = Inf, label = snp_stats_text, 
           hjust = 1.1, vjust = 1.5, size = 3.5, family = "mono",
           fill = "white", alpha = 0.8) +
  labs(
    title = "Relationship between SNP Pleiotropy Metrics",
    subtitle = "Trait Categories vs. Number of Protein Coding Genes Colocalising with SNP",
    x = "Number of Trait Categories",
    y = "Number of Protein Coding Genes Colocalising with SNP"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(p_snp_pleiotropy, filename = file.path(results_analysis_dir, "snp_pleiotropy_plot.png"), width = 8, height = 6)
```

```{r gene_pleiotropy_vs_methQTL, echo=F}
# Get coloc group data with data_type and gene information
coloc_group_meth_analysis <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*,
  studies.gene, studies.data_type, studies.tissue, studies.study_name,
  gene_annotations.gene_biotype,
  study_extractions.cis_trans
  FROM coloc_groups
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN study_extractions ON coloc_groups.study_extraction_id = studies.study_extractions.id
  LEFT JOIN gene_annotations ON studies.gene_id = gene_annotations.id
  WHERE study_extractions.min_p < 5e-8
")

# For each coloc group, calculate:
# 1. Whether it contains methylation data_type
# 2. Number of protein-coding genes that are NOT from methylation data_type
coloc_group_meth_summary <- coloc_group_meth_analysis |>
  group_by(coloc_group_id) |>
  summarise(
    has_methylation = any(data_type == "methylation", na.rm = TRUE),
    num_protein_coding_non_meth = sum(
      gene_biotype == "protein_coding" & 
      data_type != "methylation" & 
      !is.na(gene), 
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  # Convert to binary for logistic regression
  mutate(has_methylation = as.integer(has_methylation))

# Fit logistic regression model
glm_meth <- glm(has_methylation ~ num_protein_coding_non_meth, 
                data = coloc_group_meth_summary, 
                family = binomial(link = "logit"))
glm_summary <- summary(glm_meth)
glm_p <- coef(glm_summary)[2, "Pr(>|z|)"]
glm_coef <- coef(glm_meth)[2]

# Create annotation text
meth_stats_text <- paste0(
  "Coefficient = ", round(glm_coef, 3), "\n",
  "p = ", formatC(glm_p, format = "e", digits = 2)
)

# Create plot with logistic regression curve
meth_plot <- ggplot(coloc_group_meth_summary, 
  aes(x = num_protein_coding_non_meth, y = has_methylation)) +
  geom_jitter(height = 0.05, width = 0.1, alpha = 0.3, color = "#006400", size = 0.8) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              se = TRUE, color = "black", linewidth = 0.8) +
  annotate("label", x = -Inf, y = Inf, label = meth_stats_text, 
           hjust = -0.1, vjust = 1.5, size = 3.5, family = "mono",
           fill = "white", alpha = 0.8) +
  labs(
    title = "Relationship between Methylation Presence and Protein-Coding Genes",
    subtitle = "Logistic regression: Probability of methylation in coloc group vs. number of non-methylation protein-coding genes",
    x = "Number of protein-coding genes (excluding methylation)",
    y = "Has methylation in coloc group"
  ) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No", "Yes")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 9))

meth_plot_file <- file.path(results_analysis_dir, "pleiotropy_gene_vs_methQTL.png")
ggsave(meth_plot, filename = meth_plot_file, width = 10, height = 6)
print(meth_plot)
```

```{r l_shaped_pleiotropy, echo=F}
num_coloc_groups <- DBI::dbGetQuery(studies_conn, "SELECT COUNT(DISTINCT coloc_group_id) FROM coloc_groups")
coloc_group_data <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*,
  studies.gene, studies.sample_size, studies.data_type, studies.tissue, studies.study_name,
  gene_annotations.chr, gene_annotations.start, gene_annotations.stop, gene_annotations.gene_biotype,
  study_extractions.cis_trans
  FROM coloc_groups
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN studies.study_extractions ON coloc_groups.study_extraction_id = studies.study_extractions.id
  LEFT JOIN gene_annotations ON studies.gene_id = gene_annotations.id
  WHERE study_extractions.min_p < 5e-8
")
multiple_genes_in_coloc_group <- function(only_protein_coding = FALSE, include_trans = TRUE) {
  # Start with all coloc groups, but apply filters for counting
  coloc_group_analysis <- coloc_group_data %>%
    # Apply gene type filter if specified
    {if (only_protein_coding) {
      filter(., gene_biotype == "protein_coding")
    } else {
      .
    }} %>%
    # Apply cis/trans filter if specified
    {if (!include_trans) {
      filter(., cis_trans != "trans")
    } else {
      .
    }} %>%
    filter(data_type != "methylation") %>%
    group_by(coloc_group_id) %>%
    summarise(
      num_genes = n(),
      unique_genes = n_distinct(gene, na.rm = TRUE),
      genes = paste(unique(gene[!is.na(gene)]), collapse = ", "),
      .groups = 'drop'
    ) %>%
    mutate(
      genes = ifelse(genes == "", NA, genes),
      unique_genes = ifelse(genes == "" | is.na(genes), 0, unique_genes)
    )
  
  # Now we need to ensure we include ALL coloc groups, even those that might have been filtered out
  # Get all unique coloc_group_ids from the original data
  all_coloc_groups <- coloc_group_data %>%
    select(coloc_group_id) %>%
    distinct()
  
  # Left join to ensure all coloc groups are represented
  coloc_group_analysis <- all_coloc_groups %>%
    left_join(coloc_group_analysis, by = "coloc_group_id") %>%
    mutate(
      num_genes = ifelse(is.na(num_genes), 0, num_genes),
      unique_genes = ifelse(is.na(unique_genes), 0, unique_genes),
      genes = ifelse(is.na(genes), NA, genes)
    )

  coloc_group_analysis$unique_gene_bin <- case_when(
    coloc_group_analysis$unique_genes == 0 ~ "0",
    coloc_group_analysis$unique_genes <= 20 ~ as.character(coloc_group_analysis$unique_genes),
    TRUE ~ paste0(">20")
  )
  
  # Convert to factor with proper ordering
  coloc_group_analysis$unique_gene_bin <- factor(coloc_group_analysis$unique_gene_bin, 
                                                levels = c(as.character(0:20), ">20"))
  
  # Count by bins
  unique_gene_bin_counts <- coloc_group_analysis %>%
    group_by(unique_gene_bin) %>%
    summarise(count = n(), .groups = 'drop') %>%
    filter(!is.na(unique_gene_bin))
  
  # Filter out 0 gene groups for summary statistics to match the plot
  coloc_group_analysis_filtered <- coloc_group_analysis %>%
    filter(unique_genes > 0)
  
  p_gene_dist <- ggplot(unique_gene_bin_counts, aes(x = unique_gene_bin, y = count)) +
    geom_bar(stat = "identity", fill = cb_blue, color = "black", alpha = 0.7) +
    geom_text(aes(label = count), vjust = -0.5, size = 3) +
    scale_x_discrete(limits = c(as.character(1:20), ">20")) +
    labs(
      title = glue::glue("(B) Number of unique {if (only_protein_coding) 'protein coding' else 'all'} genes per coloc group{if (include_trans) ', including trans' else ''}"),
      x = "Number of unique genes per coloc group",
      y = "Frequency",
      caption = paste("Total groups (with genes):", nrow(coloc_group_analysis_filtered),
        "| Mean:", round(mean(coloc_group_analysis_filtered$unique_genes), 2),
        "| Median:", round(median(coloc_group_analysis_filtered$unique_genes), 2)
      )
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
  p_gene_dist_file <- file.path(results_analysis_dir, glue::glue("pleiotropy_b_num_genes_per_coloc_group_{if (only_protein_coding) 'protein_coding' else 'all'}_{if (include_trans) 'including_trans' else 'excluding_trans'}.png"))
  ggsave(p_gene_dist, filename = p_gene_dist_file, width = 8, height = 6)
  return(p_gene_dist)
}

multiple_genes_in_coloc_group(only_protein_coding = TRUE, include_trans = FALSE)

### Gene colocalization vs regional gene density (within ±1Mb)
# Use protein-coding genes only to match graph (B) context
protein_coding_genes <- DBI::dbGetQuery(studies_conn, "
  SELECT gene, chr, start, stop, gene_biotype
  FROM gene_annotations
  WHERE gene_biotype = 'protein_coding'
")

genes_dt <- as.data.table(protein_coding_genes)
# Define ±1Mb window around each gene; ensure ordering for foverlaps
genes_dt[, region_start := start - 1e6]
genes_dt[, region_end   := stop  + 1e6]
setkey(genes_dt, chr, start, stop)

# Count neighbouring protein-coding genes within ±1Mb (exclude self)
overlaps <- foverlaps(
  genes_dt[, .(gene, chr, region_start, region_end)],
  genes_dt,
  by.x = c("chr", "region_start", "region_end"),
  by.y = c("chr", "start", "stop"),
  type = "any",
  nomatch = 0
)

gene_density <- overlaps[
  ,
  .(neighborhood_genes = .N - 1L),  # subtract self
  by = gene
]

# For each gene, count distinct protein-coding genes it colocalises with (cis, min_p < 5e-8)
coloc_protein <- coloc_group_data |>
  filter(gene_biotype == "protein_coding", !is.na(gene)) |>
  group_by(coloc_group_id) |>
  summarise(genes = list(unique(gene)), .groups = "drop") |>
  filter(lengths(genes) > 0)

gene_partner_pairs <- coloc_protein |>
  mutate(pairs = map(genes, function(g) {
    if (length(g) < 2) {
      return(tibble::tibble(gene = character(), partner = character()))
    }
    cmb <- combn(g, 2)
    tibble::tibble(
      gene    = c(cmb[1, ], cmb[2, ]),
      partner = c(cmb[2, ], cmb[1, ])
    )
  })) |>
  select(pairs) |>
  tidyr::unnest(pairs)

gene_coloc_partners <- gene_partner_pairs |>
  group_by(gene) |>
  summarise(num_coloc_partner_genes = n_distinct(partner), .groups = "drop")

gene_coloc_density <- gene_coloc_partners |>
  inner_join(gene_density, by = "gene") |>
  mutate(
    density_decile = ntile(neighborhood_genes, 10),
    density_decile = factor(density_decile, levels = 1:10, labels = paste0("Decile ", 1:10))
  )

significant_cis_coloc_pairs <- joined_coloc_results %>%
  filter(h4 > 0.8 & (is.na(cis_trans_a) | cis_trans_a == 'cis') & (is.na(cis_trans_b) | cis_trans_b == 'cis') & min_p_a < 5e-8 & min_p_b < 5e-8)

most_gene_a_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype' & data_type_a != 'methylation') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait = trait_name_b)

most_gene_b_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype' & data_type_b != 'methylation') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait = trait_name_a)

most_gene_a_colocalising_trait_category <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype' & data_type_a != 'methylation') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait_category = trait_category_b)

most_gene_b_colocalising_trait_category <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype' & data_type_b != 'methylation') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait_category = trait_category_a)

most_genes_colocalising_with_trait_categories <- rbind(most_gene_a_colocalising_trait_category, most_gene_b_colocalising_trait_category)

gene_trait_category_counts <- most_genes_colocalising_with_trait_categories %>%
  group_by(gene) %>%
  summarise(num_trait_categories = n_distinct(trait_category), .groups = 'drop')

gene_trait_category_count_data <- gene_trait_category_counts %>%
  count(num_trait_categories, name = "count") %>%
  mutate(count_offset = count + 0.1)  # Add small offset to make count=1 visible


p_gene_trait_category <- ggplot(gene_trait_category_count_data, aes(x = num_trait_categories, y = count_offset)) +
  geom_bar(stat = "identity", fill = cb_blue, color = "black", alpha = 0.7) +
  scale_x_continuous() +
  labs(
    title = "(A) Number of trait categories colocalising with a gene",
    x = "Number of trait categories per gene",
    y = "Frequency (log scale)",
    caption = paste( "Mean:", round(mean(gene_trait_category_counts$num_trait_categories), 2),
      "| Median:", round(median(gene_trait_category_counts$num_trait_categories), 2))
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_gene_trait_category_file <- file.path(results_analysis_dir, "pleiotropy_a_num_trait_categories_per_gene.png")
ggsave(p_gene_trait_category, filename = p_gene_trait_category_file, width = 8, height = 6)
print(p_gene_trait_category)


all_phenotype_studies <- DBI::dbGetQuery(studies_conn, "SELECT studies.*, traits.trait_name FROM studies
  JOIN traits ON studies.trait_id = traits.id
  WHERE studies.data_type = 'phenotype' AND studies.variant_type = 'common'
")

number_of_genes <- lapply(seq_along(all_phenotype_studies$id), function(i) {
  study <- all_phenotype_studies[i, , drop = FALSE]
  study_id <- study$id
  study_name <- study$study_name
  all_studies_in_coloc_groups_for_study <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT COUNT(DISTINCT studies.gene_id) AS num_genes
    FROM coloc_groups
    JOIN studies ON coloc_groups.study_id = studies.id
    JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
    WHERE study_extractions.min_p < 5e-8
    AND studies.data_type != 'methylation'
    AND coloc_groups.coloc_group_id IN (SELECT DISTINCT coloc_group_id
      FROM coloc_groups
      JOIN studies on coloc_groups.study_id = studies.id
      JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
      WHERE studies.id = {study_id}
      AND study_extractions.min_p < 5e-8
      AND (study_extractions.cis_trans = 'cis' OR study_extractions.cis_trans IS NULL)
    )
  "))
  all_studies_in_coloc_groups_for_study$study_id <- study_id
  all_studies_in_coloc_groups_for_study$trait_name <- study$trait_name
  all_studies_in_coloc_groups_for_study$sample_size <- study$sample_size
  return(all_studies_in_coloc_groups_for_study)
})
number_of_genes <- do.call(rbind, number_of_genes)

# Create a summary with counts and add small offset for visibility
count_data <- number_of_genes %>%
  count(num_genes, name = "count") %>%
  mutate(count_offset = count + 0.3)  # Add small offset to make count=1 visible

p <- ggplot(count_data, aes(x = num_genes, y = count_offset)) +
geom_bar(stat = "identity", fill = cb_blue, alpha = 1) +
  scale_x_continuous() +
  scale_y_log10() +
  labs(
    title = "(D) Number of genes colocalising per trait",
    x = "Frequency (log scale)",
    y = "Number of genes colocalising per trait",
    caption = paste("Mean:", round(mean(number_of_genes$num_genes), 2),
                   "| Median:", round(median(number_of_genes$num_genes), 2))
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_num_genes_colocalising_file <- file.path(results_analysis_dir, "pleiotropy_d_num_genes_colocalising.png")
ggsave(p, filename = p_num_genes_colocalising_file, width = 8, height = 6)
print(p)

number_of_genes_summary <- number_of_genes %>%
  arrange(desc(num_genes)) %>%
  head(10)
knitr::kable(number_of_genes_summary, caption = "Top 10 studies by number of genes colocalising")

median_sample_size <- median(number_of_genes$sample_size, na.rm = TRUE)
median_num_genes <- median(number_of_genes$num_genes, na.rm = TRUE)

valid_colocs <- dplyr::filter(joined_coloc_results, ld_block_id == 60 & h4 >= 0.8 & min_p_a < 5e-8 & min_p_b < 5e-8 & !is.na(gene_a) & !is.na(gene_b) & gene_a != gene_b)
count_gene_pairs <- valid_colocs %>%
  group_by(gene_a, gene_b) %>%
  summarise(count = n(), .groups = 'drop') %>%
  arrange(desc(count))

ld_block_id <- 60
all_coloc_groups_in_ld_block <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT coloc_group_id, gene, data_type FROM coloc_groups_wide
  WHERE ld_block_id = {ld_block_id}
  AND min_p <= 5e-8
"))

ld_block_info <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT * FROM ld_blocks WHERE id = {ld_block_id}"))

# Create gene pairs within each coloc group
gene_pairs <- all_coloc_groups_in_ld_block %>%
  filter(!is.na(gene) & data_type != 'methylation') %>%  # Filter out NA genes
  group_by(coloc_group_id) %>%
  do({
    genes_in_group <- unique(.$gene)
    genes_in_group <- genes_in_group[!is.na(genes_in_group)]  # Additional safety check
    if(length(genes_in_group) > 1) {
      # Create all pairwise combinations of genes in this coloc group
      gene_combinations <- combn(genes_in_group, 2, simplify = FALSE)
      data.frame(
        gene_a = sapply(gene_combinations, function(x) x[1]),
        gene_b = sapply(gene_combinations, function(x) x[2]),
        stringsAsFactors = FALSE
      )
    } else {
      data.frame(gene_a = character(), gene_b = character(), stringsAsFactors = FALSE)
    }
  }) %>%
  ungroup() %>%
  # Filter out any remaining NA pairs before counting
  filter(!is.na(gene_a) & !is.na(gene_b)) %>%
  # Count how many times each gene pair appears across coloc groups
  group_by(gene_a, gene_b) %>%
  summarise(
    weight = n(),  # Number of coloc groups where this gene pair appears
    .groups = 'drop'
  )

# Create igraph object and include isolated protein-coding genes within this LD block (no coloc edges)
library(igraph)
gene_graph <- graph_from_data_frame(gene_pairs, directed = FALSE)

block_genes <- protein_coding_genes |>
  filter(
    chr == ld_block_info$chr,
    start >= ld_block_info$start,
    stop  <= ld_block_info$stop
  ) |>
  pull(gene) |>
  unique()

missing_block_genes <- setdiff(block_genes, V(gene_graph)$name)
if (length(missing_block_genes) > 0) {
  gene_graph <- add_vertices(gene_graph, nv = length(missing_block_genes), name = missing_block_genes)
}

# Create visualization
library(ggraph)
library(tidygraph)

# Convert to tidygraph for easier plotting and add node attributes
# Map neighborhood gene density onto nodes (default 0 when missing)
gene_density_lookup <- gene_density |>
  select(gene, neighborhood_genes)

gene_tidy_graph <- as_tbl_graph(gene_graph) %>%
  activate(nodes) %>%
  mutate(
    connections = centrality_degree(),
    betweenness = centrality_betweenness(),
    closeness = centrality_closeness(),
    neighborhood_genes = gene_density_lookup$neighborhood_genes[match(name, gene_density_lookup$gene)],
    neighborhood_genes = tidyr::replace_na(neighborhood_genes, 0)
  )

# Plot the network
network_plot <- ggraph(gene_tidy_graph, layout = "fr", niter = 1000) +
  geom_edge_link(aes(alpha = weight), color = "#444444") +
  geom_node_point(aes(size = connections, fill = neighborhood_genes), shape = 21, color = "black", stroke = 0.4, alpha = 0.8) +
  geom_node_text(aes(label = name), size = 2.2, colour = "grey40", repel = TRUE) +
  scale_edge_alpha_continuous(range = c(0.1, 0.8)) +
  scale_size_continuous(range = c(2, 8)) +
  scale_fill_gradient(low = cb_light_blue, high = cb_blue, name = "# of genes ±1Mb") +
  labs(
    title = "(C) Gene-Gene Colocalization Network within an LD Block",
    subtitle = paste("LD Block", ld_block_info$ld_block),
    caption = paste(
      "Total Genes:", vcount(gene_graph),
      "| Links in coloc groups:", ecount(gene_graph)
    )
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )

p_gene_gene_network_file <- file.path(results_analysis_dir, "pleiotropy_c_gene_gene_network.png")
ggsave(network_plot, filename = p_gene_gene_network_file, width = 8, height = 6)
print(network_plot)
```

```{r relationship_between_gene_density_and_coloc_partners, echo=F}
# Get all LD blocks that have coloc groups with protein-coding genes (cis only)
ld_blocks_with_coloc <- DBI::dbGetQuery(studies_conn, "
  SELECT DISTINCT ld_block_id, ld_blocks.chr, ld_blocks.start, ld_blocks.stop, ld_blocks.ld_block
  FROM coloc_groups_wide
  JOIN ld_blocks ON coloc_groups_wide.ld_block_id = ld_blocks.id
  WHERE coloc_groups_wide.min_p < 5e-8
  AND coloc_groups_wide.gene IS NOT NULL
  AND coloc_groups_wide.cis_trans = 'cis'
  ORDER BY ld_block_id
")
protein_coding_genes <- DBI::dbGetQuery(studies_conn, "
  SELECT gene, chr, start, stop, gene_biotype
  FROM gene_annotations
  WHERE gene_biotype = 'protein_coding'
")

genes_dt <- as.data.table(protein_coding_genes)
# Define ±1Mb window around each gene; ensure ordering for foverlaps
genes_dt[, region_start := start - 1e6]
genes_dt[, region_end   := stop  + 1e6]
setkey(genes_dt, chr, start, stop)

# Count neighbouring protein-coding genes within ±1Mb (exclude self)
overlaps <- foverlaps(
  genes_dt[, .(gene, chr, region_start, region_end)],
  genes_dt,
  by.x = c("chr", "region_start", "region_end"),
  by.y = c("chr", "start", "stop"),
  type = "any",
  nomatch = 0
)

gene_density <- overlaps[
  ,
  .(neighborhood_genes = .N - 1L),  # subtract self
  by = gene
]

# For each LD block, calculate coloc partners limited to genes within that same block
ld_block_coloc_analysis <- lapply(seq_len(nrow(ld_blocks_with_coloc)), function(i) {
  block_info <- ld_blocks_with_coloc[i, ]
  block_id <- block_info$ld_block_id
  
  # Get all protein-coding genes within this LD block
  block_genes <- protein_coding_genes |>
    filter(
      chr == block_info$chr,
      start >= block_info$start,
      stop <= block_info$stop
    ) |>
    pull(gene) |>
    unique()
  
  if (length(block_genes) == 0) return(NULL)
  
  # Get coloc groups in this LD block (cis only, no trans)
  coloc_in_block <- DBI::dbGetQuery(studies_conn, glue::glue("
    SELECT coloc_group_id, gene
    FROM coloc_groups_wide
    WHERE ld_block_id = {block_id}
    AND min_p < 5e-8
    AND gene IS NOT NULL
    AND cis_trans = 'cis'
    AND data_type != 'methylation'
  ")) |>
    filter(gene %in% block_genes)
  
  if (nrow(coloc_in_block) == 0) {
    # No coloc groups, but still include genes with zero partners
    return(data.frame(
      ld_block_id = block_id,
      ld_block = block_info$ld_block,
      gene = block_genes,
      num_coloc_partner_genes_in_block = 0,
      stringsAsFactors = FALSE
    ))
  }
  
  # Create gene partner pairs within coloc groups, but only for genes in this block
  gene_partner_pairs_block <- coloc_in_block |>
    filter(gene %in% block_genes) |>
    group_by(coloc_group_id) |>
    summarise(genes = list(unique(gene)), .groups = "drop") |>
    filter(lengths(genes) > 1) |>
    mutate(pairs = map(genes, function(g) {
      if (length(g) < 2) {
        return(tibble::tibble(gene = character(), partner = character()))
      }
      cmb <- combn(g, 2)
      tibble::tibble(
        gene = c(cmb[1, ], cmb[2, ]),
        partner = c(cmb[2, ], cmb[1, ])
      )
    })) |>
    select(pairs) |>
    tidyr::unnest(pairs)
  
  # Only count partners that are also in this block (check if partner column exists)
  if (nrow(gene_partner_pairs_block) > 0 && "partner" %in% names(gene_partner_pairs_block)) {
    gene_partner_pairs_block <- gene_partner_pairs_block |>
      filter(partner %in% block_genes)
  } else {
    # If no pairs or no partner column, create empty result with correct structure
    gene_partner_pairs_block <- tibble::tibble(gene = character(), partner = character())
  }
  
  # Count coloc partners per gene (limited to this block)
  if (nrow(gene_partner_pairs_block) > 0 && "partner" %in% names(gene_partner_pairs_block)) {
    gene_coloc_partners_block <- gene_partner_pairs_block |>
      group_by(gene) |>
      summarise(num_coloc_partner_genes_in_block = n_distinct(partner), .groups = "drop")
  } else {
    gene_coloc_partners_block <- tibble::tibble(gene = character(), num_coloc_partner_genes_in_block = integer())
  }
  
  # Include all block genes, setting zero for those without coloc partners
  all_block_genes_df <- data.frame(
    gene = block_genes,
    stringsAsFactors = FALSE
  ) |>
    left_join(gene_coloc_partners_block, by = "gene") |>
    mutate(
      num_coloc_partner_genes_in_block = tidyr::replace_na(num_coloc_partner_genes_in_block, 0),
      ld_block_id = block_id,
      ld_block = block_info$ld_block
    )
  
  return(all_block_genes_df)
}) |>
  bind_rows()

# Join with gene density
ld_block_coloc_density <- ld_block_coloc_analysis |>
  left_join(gene_density, by = "gene") |>
  mutate(neighborhood_genes = tidyr::replace_na(neighborhood_genes, 0))

# Fit linear model
lm_density_coloc <- lm(num_coloc_partner_genes_in_block ~ neighborhood_genes, data = ld_block_coloc_density)
lm_density_coloc_summary <- summary(lm_density_coloc)
density_coloc_r2 <- lm_density_coloc_summary$r.squared
density_coloc_p <- pf(lm_density_coloc_summary$fstatistic[1], 
                      lm_density_coloc_summary$fstatistic[2], 
                      lm_density_coloc_summary$fstatistic[3], 
                      lower.tail = FALSE)
density_coloc_slope <- coef(lm_density_coloc)[2]
overall_corr <- cor(ld_block_coloc_density$neighborhood_genes, 
                    ld_block_coloc_density$num_coloc_partner_genes_in_block, 
                    use = "complete.obs")

# Create annotation text
density_coloc_stats_text <- paste0(
  "Slope = ", round(density_coloc_slope, 3), "\n",
  "R² = ", round(density_coloc_r2, 3), "\n",
  "p = ", formatC(density_coloc_p, format = "e", digits = 2)
)

# Create scatter plot
gene_density_coloc_scatter <- ggplot(ld_block_coloc_density, 
  aes(x = neighborhood_genes, y = num_coloc_partner_genes_in_block)) +
  geom_point(alpha = 0.6, color = cb_blue, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.5) +
  annotate("label", x = -Inf, y = Inf, label = density_coloc_stats_text, 
           hjust = -0.1, vjust = 1.5, size = 3.5, family = "mono",
           fill = "white", alpha = 0.8) +
  labs(
    title = "(B b) Gene density vs colocalizing partners within LD blocks",
    subtitle = paste("Overall correlation (limited to partners within same LD block): r =", round(overall_corr, 3)),
    x = "Number of protein-coding genes in ±1Mb window",
    y = "Number of colocalizing protein-coding genes (same LD block only)",
    caption = paste("Each point is a gene. Coloc partners counted only within the same LD block")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10)
  )

gene_density_coloc_file <- file.path(results_analysis_dir, "pleiotropy_b_gene_density_vs_coloc_without_methylation.png")
ggsave(gene_density_coloc_scatter, filename = gene_density_coloc_file, width = 10, height = 8)
print(gene_density_coloc_scatter)
```

## Genes that colocalise with the most traits (or trait categories?).  
Limited to cis-coloc pairs with h4 > 0.8 and p-values < 5e-8.

```{r genes_colocalising_with_most_traits, echo=F}
significant_cis_coloc_pairs <- joined_coloc_results %>%
  filter(h4 > 0.8 & (is.na(cis_trans_a) | cis_trans_a == 'cis') & (is.na(cis_trans_b) | cis_trans_b == 'cis') & min_p_a < 5e-8 & min_p_b < 5e-8)

most_gene_a_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype' & data_type_a != 'methylation') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait = trait_name_b)

most_gene_b_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype' & data_type_b != 'methylation') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait = trait_name_a)

most_gene_a_colocalising_trait_category <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype' & data_type_a != 'methylation') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait_category = trait_category_b)

most_gene_b_colocalising_trait_category <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype' & data_type_b != 'methylation') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait_category = trait_category_a)

most_genes_colocalising_with_traits <- rbind(most_gene_a_colocalising, most_gene_b_colocalising)
most_genes_colocalising_with_trait_categories <- rbind(most_gene_a_colocalising_trait_category, most_gene_b_colocalising_trait_category)

most_genes_colocalising_summary <- most_genes_colocalising_with_traits %>%
  group_by(gene) %>%
  summarise(num_traits = n_distinct(trait)) %>%
  arrange(desc(num_traits)) %>%
  head(50)

knitr::kable(most_genes_colocalising_summary, caption = "Genes that colocalise with the most traits")

top_gene <- most_genes_colocalising_summary$gene[1]
all_phenotypes_for_top_gene <- most_genes_colocalising_with_traits %>%
  filter(gene == top_gene) %>%
  select(trait) %>%
  distinct()
```

## Apparent pleiotropy due to LD

Association vs. colocalization: How often would we see a genetic association with a trait that doesn't have a coloc result?

For any causal variant:

* How many traits does it coloc with (significant h4)
* How many traits does it associate with (significant h3)
* Rate of incorrectness due to LD = significant h4 / significant h3
* How many significant h4 traits are spurious? (Spurious: h4 > 0.8 & failed the clustering step)

```{r h4_h3_ratio, echo=F}
num_significant_h4 <- nrow(joined_coloc_results[joined_coloc_results$h4 > 0.8, ])
num_spurious_h4 <- nrow(joined_coloc_results[joined_coloc_results$h4 > 0.8 & joined_coloc_results$false_positive, ])
num_significant_h3 <- nrow(joined_coloc_results[joined_coloc_results$h3 > 0.8, ])
num_spurious_h3 <- nrow(joined_coloc_results[joined_coloc_results$h3 > 0.8 & joined_coloc_results$false_positive, ])

hemoglobin_study_id <- 'ebi-a-GCST90002384'
hemoglobin_example <- joined_coloc_results[joined_coloc_results$study_name_a == hemoglobin_study_id | joined_coloc_results$study_name_b == hemoglobin_study_id, ]

hemoglobin_h4 <- nrow(hemoglobin_example[hemoglobin_example$h4 > 0.8, ])
hemoglobin_h3 <- nrow(hemoglobin_example[hemoglobin_example$h3 > 0.8, ])
hemoglobin_h4_spurious <- nrow(hemoglobin_example[hemoglobin_example$h4 > 0.8 & hemoglobin_example$false_positive, ])
hemoglobin_h3_spurious <- nrow(hemoglobin_example[hemoglobin_example$h3 > 0.8 & hemoglobin_example$false_positive, ])

h4_summary <- tibble::tribble(
  ~trait, ~num_significant_h4, ~num_spurious_h4, ~num_significant_h3, ~h3_h4_ratio,
  'All', num_significant_h4, num_spurious_h4, num_significant_h3, num_significant_h3 / num_significant_h4,
  'Hemoglobin', hemoglobin_h4, hemoglobin_h4_spurious, hemoglobin_h3, hemoglobin_h3 / hemoglobin_h4
)

knitr::kable(h4_summary)
```

#### Breakdown of ratio of h3 to h4 by data type

```{r matrix_of_coloc_results, echo=F}
coloc_matrix <- dplyr::filter(joined_coloc_results, h4 > 0.8 | h3 > 0.8)

# Create the n×n matrix
h4_coloc_matrix_by_type <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(h4_coloc_matrix_by_type) <- unique_data_types
colnames(h4_coloc_matrix_by_type) <- unique_data_types

h3_coloc_matrix_by_type <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(h3_coloc_matrix_by_type) <- unique_data_types
colnames(h3_coloc_matrix_by_type) <- unique_data_types

# Fill the matrix with coloc counts (combining symmetric entries)
for (i in 1:nrow(coloc_matrix)) {
  type_a <- coloc_matrix$data_type_a[i]
  type_b <- coloc_matrix$data_type_b[i]

  if (coloc_matrix$h4[i] > 0.8) {
    h4_coloc_matrix_by_type[type_a, type_b] <- h4_coloc_matrix_by_type[type_a, type_b] + 1
    if (type_a != type_b) {
      h4_coloc_matrix_by_type[type_b, type_a] <- h4_coloc_matrix_by_type[type_b, type_a] + 1
    }
  } else if (coloc_matrix$h3[i] > 0.8) {
    h3_coloc_matrix_by_type[type_a, type_b] <- h3_coloc_matrix_by_type[type_a, type_b] + 1
    if (type_a != type_b) {
      h3_coloc_matrix_by_type[type_b, type_a] <- h3_coloc_matrix_by_type[type_b, type_a] + 1
    }
  }
}

h4_coloc_matrix_by_type[lower.tri(h4_coloc_matrix_by_type)] <- NA
h4_coloc_matrix_by_type <- as.data.frame(h4_coloc_matrix_by_type)
h4_coloc_pretty <- h4_coloc_matrix_by_type
h4_coloc_pretty[, unique_data_types] <- lapply(h4_coloc_pretty[, unique_data_types], scales::comma)
knitr::kable(h4_coloc_pretty, caption = "Number of H4 > 0.8 between data types")

h3_coloc_matrix_by_type[lower.tri(h3_coloc_matrix_by_type)] <- NA
h3_coloc_matrix_by_type <- as.data.frame(h3_coloc_matrix_by_type)
h3_coloc_pretty <- h3_coloc_matrix_by_type
h3_coloc_pretty[, unique_data_types] <- lapply(h3_coloc_pretty[, unique_data_types], scales::comma)
knitr::kable(h3_coloc_pretty, caption = "Number of H3 > 0.8 between data types")

h3_h4_ratio <- h3_coloc_matrix_by_type / h4_coloc_matrix_by_type
knitr::kable(h3_h4_ratio, caption = "Ratio of H3 > 0.8 to H4 > 0.8 between data types")
```

# LD relationship between causal SNPs

### How often Causal SNPs are in LD with each other
```{r causal_snps_in_ld, echo=F}

all_causal_snps <- DBI::dbGetQuery(studies_conn, "SELECT DISTINCT snp_id FROM coloc_groups")

dont_print <- DBI::dbExecute(ld_conn, glue::glue("ATTACH '{studies_db_file}' AS studies"))

formatted_snp_ids <- paste0(all_causal_snps$snp_id, collapse = ",")
snp_pair_ld_scores <- DBI::dbGetQuery(ld_conn, glue::glue("SELECT ld.*, studies.ld_blocks.ld_block AS ld_block FROM ld
  JOIN studies.ld_blocks ON ld.ld_block_id = studies.ld_blocks.id
  WHERE lead_snp_id IN ({formatted_snp_ids})
  AND variant_snp_id IN ({formatted_snp_ids})
  "
))

snp_pair_ld_scores$r2 <- snp_pair_ld_scores$r ^ 2

r2_thresholds <- c(0.8, 0.9, 0.95, 0.99)

ld_results <- lapply(r2_thresholds, function(r2_threshold) {
  snp_pairs_in_ld <- snp_pair_ld_scores[snp_pair_ld_scores$r2 > r2_threshold, ]
  unique_causal_snps_in_ld <- length(unique(snp_pairs_in_ld$lead_snp_id))
  return(data.frame(
    r2_threshold = r2_threshold,
    unique_causal_snps_in_ld = unique_causal_snps_in_ld,
    number_of_snp_pairs_in_ld = nrow(snp_pairs_in_ld),
    percentage_of_causal_snps_in_ld = unique_causal_snps_in_ld / length(all_causal_snps$snp_id)
  ))
})

ld_results <- do.call(rbind, ld_results)
knitr::kable(ld_results, caption = "Number of unique causal SNPs in LD by r2 threshold")

snp_pairs_in_ld <- snp_pair_ld_scores[snp_pair_ld_scores$r2 > 0.8, ]
ld_block_summary <- snp_pairs_in_ld %>%
  group_by(ld_block_id, ld_block) %>%
  summarise(
    num_snp_pairs = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(num_snp_pairs)) %>%
  head(10)

knitr::kable(ld_block_summary, 
  caption = "Top 10 LD blocks by number of causal SNP pairs in LD",
  col.names = c("LD Block ID", "LD Block", "Number of SNP Pairs"))


```

## How often does a coloc group have multiple genes?  Multiple phenotypes? (cis only)

```{r multiple_genes_in_coloc_group, echo=F}
coloc_group_data <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*,
  studies.gene, studies.sample_size, studies.data_type, studies.tissue, studies.study_name,
  gene_annotations.chr, gene_annotations.start, gene_annotations.stop, gene_annotations.gene_biotype,
  study_extractions.cis_trans, traits.trait_category, traits.trait_name
  FROM coloc_groups
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN study_extractions ON coloc_groups.study_extraction_id = studies.study_extractions.id
  JOIN traits ON studies.trait_id = traits.id
  LEFT JOIN gene_annotations ON studies.gene_id = gene_annotations.id
  WHERE study_extractions.min_p < 5e-8
  AND (study_extractions.cis_trans = 'cis' OR study_extractions.cis_trans IS NULL)
")

# Filter to only keep coloc groups with multiple entries
coloc_group_data <- coloc_group_data |>
  group_by(coloc_group_id) |>
  filter(n() > 1) |>
  ungroup()

gene_phenotype_counts_per_coloc_group <- coloc_group_data |>
  group_by(coloc_group_id) |>
  summarise(
    unique_genes = n_distinct(gene, na.rm = TRUE),
    num_phenotypes = sum(data_type == "phenotype", na.rm = TRUE),
    .groups = 'drop'
  )


gene_phenotype_bins <- gene_phenotype_counts_per_coloc_group |>
  mutate(
    genes_bin = case_when(
      unique_genes == 0 ~ "No genes",
      unique_genes <= 5 ~ paste0(unique_genes, " genes"),
      TRUE ~ ">5 genes"
    ),
    phenotypes_bin = case_when(
      num_phenotypes == 0 ~ "No phenotypes",
      num_phenotypes <= 5 ~ paste0(num_phenotypes, " phenotypes"),
      TRUE ~ ">5 phenotypes"
    )
  ) |>
  mutate(
    genes_bin = factor(genes_bin, levels = c("No genes", paste0(1:5, " genes"), ">5 genes")),
    phenotypes_bin = factor(phenotypes_bin, levels = c("No phenotypes", paste0(1:5, " phenotypes"), ">5 phenotypes"))
  )

genes_phenotypes_matrix <- table(
  gene_phenotype_bins$genes_bin,
  gene_phenotype_bins$phenotypes_bin
)

genes_phenotypes_df <- as.data.frame(genes_phenotypes_matrix)
colnames(genes_phenotypes_df) <- c("genes_bin", "phenotypes_bin", "count")

genes_phenotypes_matrix_df <- genes_phenotypes_df |>
  tidyr::pivot_wider(
    names_from = phenotypes_bin,
    values_from = count,
    values_fill = 0
  )

knitr::kable(genes_phenotypes_matrix_df, caption = "Number of coloc groups by unique genes and phenotypes (methQTL are neither genes nor phenotypes)")

percentage_of_coloc_groups <- genes_phenotypes_matrix_df |>
  mutate(across(-genes_bin, ~ .x / nrow(gene_phenotype_bins))) |>
  mutate(across(-genes_bin, ~ scales::percent(.x, accuracy = 0.01)))

knitr::kable(percentage_of_coloc_groups, caption = "Percentage of coloc groups by unique genes and phenotypes (methQTL are neither genes nor phenotypes)")
```

```{r official_pleiotropy_scores, echo=F}
all_genes <- DBI::dbGetQuery(studies_conn, "SELECT DISTINCT coloc_groups_wide.gene_id, gene_annotations.gene FROM coloc_groups_wide
  JOIN gene_annotations ON coloc_groups_wide.gene_id = gene_annotations.id
  WHERE gene_annotations.gene_biotype = 'protein_coding'
")

gene_pleiotropy_scores <- lapply(seq_along(all_genes$gene_id), function(i) {
  this_gene_id <- all_genes$gene_id[i]
  this_gene <- all_genes$gene[i]
  all_studies_in_coloc_groups_for_gene <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT coloc_groups_wide.*, gene_annotations.gene_biotype, gene_annotations.ensembl_id FROM coloc_groups_wide
    LEFT JOIN gene_annotations ON coloc_groups_wide.gene_id = gene_annotations.id
    WHERE coloc_group_id IN (SELECT DISTINCT coloc_group_id
      FROM coloc_groups_wide
      WHERE gene_id = {this_gene_id}
      AND (cis_trans = 'cis' OR cis_trans IS NULL)
    )
  "))
  distinct_trait_categories <- all_studies_in_coloc_groups_for_gene |>
    distinct(trait_category) |>
    nrow()
  distinct_protein_coding_genes <- all_studies_in_coloc_groups_for_gene |>
    filter(gene_biotype == 'protein_coding') |>
    distinct(gene_id) |>
    filter(gene_id != this_gene_id) |>
    nrow()
  return(data.frame(
    gene_id = this_gene_id,
    gene = this_gene,
    distinct_trait_categories = distinct_trait_categories,
    distinct_protein_coding_genes = distinct_protein_coding_genes
  ))
})
joined_gene_pleiotropy_scores <- dplyr::bind_rows(gene_pleiotropy_scores)
vroom::vroom_write(joined_gene_pleiotropy_scores, file.path(results_analysis_dir, "gene_pleiotropy_scores.tsv"))
```

### Distinct phenotype categories in the same coloc group

How can we quantify the pleiotropy of phenotypes, when different phenotypes can be so similar?
How else can we quantify biological pleiotropy (vs. LD)?


```{r distinct_phenotype_categories_in_same_coloc_group, echo=F}
distinct_phenotype_categories_in_same_coloc_group <- coloc_group_data |>
  filter(data_type == 'phenotype') |>
  group_by(coloc_group_id) |>
  summarise(distinct_phenotype_categories = n_distinct(trait_category)) |>
  filter(distinct_phenotype_categories > 0)


trait_category_counts_per_coloc_group <- distinct_phenotype_categories_in_same_coloc_group |>
  mutate(
    trait_category_bin = case_when(
      distinct_phenotype_categories == 0 ~ "0 trait categories",
      TRUE ~ paste0(distinct_phenotype_categories, " trait categories")
    )
  ) |>
  group_by(trait_category_bin) |>
  summarise(count = n(), .groups = 'drop')

knitr::kable(trait_category_counts_per_coloc_group, caption = "Distinct trait categories in the same coloc group")

```


# What fraction of known GWAS hits are no known molecular trait associated with it?

```{r gwas_hits_no_molecular_trait, echo=F}
all_phenotype_study_extractions <- DBI::dbGetQuery(studies_conn, "SELECT study_extractions.*,
  snp_annotations.feature_type, snp_annotations.consequence, snp_annotations.impact, snp_annotations.cdna_position,
  snp_annotations.cds_position, snp_annotations.protein_position, snp_annotations.amino_acids, snp_annotations.codons,
  snp_annotations.gene as snp_gene, traits.trait_name as trait_name, traits.trait_category as trait_category
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  JOIN snp_annotations ON study_extractions.snp_id = snp_annotations.id
  WHERE studies.data_type = 'phenotype' and studies.variant_type = 'common' and study_extractions.min_p < 5e-8
")

all_coloc_groups <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*, studies.data_type, study_extractions.min_p,
  snp_annotations.feature_type, snp_annotations.consequence, snp_annotations.impact, snp_annotations.cdna_position,
  snp_annotations.cds_position, snp_annotations.protein_position, snp_annotations.amino_acids, snp_annotations.codons,
  snp_annotations.gene as snp_gene, traits.trait_category as trait_category, traits.trait_name as trait_name
  FROM coloc_groups
  JOIN snp_annotations ON coloc_groups.snp_id = snp_annotations.id
  JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  WHERE study_extractions.min_p < 5e-8
")


all_study_extractions_in_coloc_groups <- unique(all_coloc_groups$study_extraction_id)
all_phenotype_study_extractions_without_coloc <- all_phenotype_study_extractions %>%
  filter(!id %in% all_study_extractions_in_coloc_groups)
study_extractions_with_phenotype_only_coloc <- all_coloc_groups %>%
  group_by(coloc_group_id) %>%
  filter(all(data_type == "phenotype")) %>%
  ungroup()
all_significant_non_molecular_coloc <- dplyr::bind_rows(all_phenotype_study_extractions_without_coloc, study_extractions_with_phenotype_only_coloc)

gwas_hits_and_molecular_traits_by_trait_category <- function(tc) {
  all_study_extractions_in_coloc_groups <- unique(all_coloc_groups$study_extraction_id)
  if (is.na(tc)) {
    all_phenotype_study_extractions_for_tc <- all_phenotype_study_extractions
  } else {
    all_phenotype_study_extractions_for_tc <- all_phenotype_study_extractions[all_phenotype_study_extractions$trait_category == tc, ]
  }

  all_phenotype_study_extractions_without_coloc <- all_phenotype_study_extractions_for_tc %>%
    filter(!id %in% all_study_extractions_in_coloc_groups)

  if (is.na(tc)) {
    study_extractions_with_phenotype_only_coloc <- all_coloc_groups %>%
      group_by(coloc_group_id) %>%
      filter(all(data_type == "phenotype")) %>%
      ungroup()
  } else {
    study_extractions_with_phenotype_only_coloc <- all_coloc_groups %>%
      group_by(coloc_group_id) %>%
      filter(any(trait_category == tc)) %>%
      filter(all(data_type == "phenotype")) %>%
      ungroup() %>%
      filter(trait_category == tc)
  }

  all_significant_non_molecular_coloc <- nrow(all_phenotype_study_extractions_without_coloc) + nrow(study_extractions_with_phenotype_only_coloc)

  percent_non_molecular_coloc <- (nrow(study_extractions_with_phenotype_only_coloc) + nrow(all_phenotype_study_extractions_without_coloc)) / nrow(all_phenotype_study_extractions_for_tc)

  if (is.na(tc)) tc <- "All"
  return(data.frame(
    trait_category = tc,
    all_study_extractions = nrow(all_phenotype_study_extractions_for_tc),
    all_study_extractions_without_coloc = nrow(all_phenotype_study_extractions_without_coloc),
    phenotype_only_coloc = nrow(study_extractions_with_phenotype_only_coloc),
    all_non_molecular_coloc = all_significant_non_molecular_coloc,
    percent_no_coloc = scales::percent(nrow(all_phenotype_study_extractions_without_coloc) / nrow(all_phenotype_study_extractions_for_tc), accuracy = 0.01),
    percent_non_molecular_coloc = scales::percent(percent_non_molecular_coloc, accuracy = 0.01)
  ))
}

cat("Breakdown of phenotype info by trait category")
gwas_hits_and_molecular_traits <- lapply(unique(all_coloc_groups$trait_category), function(tc) {
  gwas_hits_and_molecular_traits_by_trait_category(tc)
}) |> dplyr::bind_rows()

gwas_hits_and_molecular_traits <- gwas_hits_and_molecular_traits %>%
  arrange(trait_category != "All", trait_category)


knitr::kable(gwas_hits_and_molecular_traits, caption = "Breakdown of phenotype info by trait category")
```


```{r non_molecular_coloc_vep_annotations, echo=F}
# Deduplicate all_coloc_groups by snp_id and create histograms for VEP annotations
all_coloc_groups_dedup <- all_coloc_groups %>%
  group_by(snp_id) %>%
  slice(1) %>%  # Keep first occurrence of each snp_id
  ungroup()

all_significant_non_molecular_coloc_dedup <- all_significant_non_molecular_coloc %>%
  group_by(snp_id) %>%
  slice(1) %>%  # Keep first occurrence of each snp_id
  ungroup()


# percent_non_molecular_coloc <- (nrow(significant_phenotype_study_extractions_with_phenotype_only_coloc) + nrow(all_significant_phenotype_study_extractions_without_coloc)) / nrow(all_significant_phenotype_study_extractions)

# cat("All significant phenotype study extractions:", nrow(all_significant_phenotype_study_extractions), "\n")
# cat("All significant phenotype study extractions without coloc:", nrow(all_significant_phenotype_study_extractions_without_coloc), "\n")
# cat("Significant phenotype study extractions in a phenotype only coloc:", nrow(significant_phenotype_study_extractions_with_phenotype_only_coloc), "\n")
# cat("All significant non-molecular coloc:", nrow(all_significant_non_molecular_coloc), "\n")
# cat("Percentage of all significant phenotype study extractions that are not in a molecular coloc:", percent_non_molecular_coloc, "\n")
# Create histograms for consequence, impact, and cdna_position


all_consequence_counts <- all_coloc_groups_dedup %>%
  filter(!is.na(consequence)) %>%
  mutate(consequence_split = strsplit(consequence, ",")) %>%
  tidyr::unnest(consequence_split) %>%
  mutate(consequence_split = trimws(consequence_split)) %>%  # Remove whitespace
  count(consequence_split, sort = TRUE)

all_non_molecular_coloc_consequence_counts <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(consequence)) %>%
  mutate(consequence_split = strsplit(consequence, ",")) %>%
  tidyr::unnest(consequence_split) %>%
  mutate(consequence_split = trimws(consequence_split)) %>%  # Remove whitespace
  count(consequence_split, sort = TRUE)



# Calculate percent difference between all_coloc_groups and non_molecular_coloc consequences
consequence_comparison <- all_consequence_counts %>%
  full_join(all_non_molecular_coloc_consequence_counts, 
            by = "consequence_split", 
            suffix = c("_all", "_non_molecular")) %>%
  mutate(
    n_all = replace_na(n_all, 0),
    n_non_molecular = replace_na(n_non_molecular, 0),
    percent_diff = ((n_non_molecular - n_all) / n_all) * 100,
    proportion = (n_non_molecular / n_all) * 100
  ) %>%
  filter(n_all > 0)  # Only include consequences that appear in the main dataset

consequence_hist <- ggplot(consequence_comparison, aes(x = reorder(consequence_split, proportion), y = proportion)) +
  geom_bar(stat = "identity", fill = ifelse(consequence_comparison$proportion> 0, "red", "blue"), alpha = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid", size = 0.5) +
  labs(
    title = "Percent Difference in Consequence Prevalence",
    subtitle = "Non-molecular coloc vs All coloc groups (Positive = more common in non-molecular)",
    x = "VEP Consequence",
    y = "Proportion Difference (Non-molecular / All) * 100",
    caption = paste("Consequences with data in both datasets:", nrow(consequence_comparison))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 10)
  ) +
  coord_flip()  # Flip to make consequence names readable


# 2. Impact comparison
all_impact_counts <- all_coloc_groups_dedup %>%
  filter(!is.na(impact)) %>%
  count(impact, sort = TRUE)

all_non_molecular_coloc_impact_counts <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(impact)) %>%
  count(impact, sort = TRUE)

impact_comparison <- all_impact_counts %>%
  full_join(all_non_molecular_coloc_impact_counts, 
            by = "impact", 
            suffix = c("_all", "_non_molecular")) %>%
  mutate(
    n_all = replace_na(n_all, 0),
    n_non_molecular = replace_na(n_non_molecular, 0),
    percent_diff = ((n_non_molecular - n_all) / n_all) * 100,
    proportion = (n_non_molecular / n_all) * 100
  ) %>%
  filter(n_all > 0)

impact_hist <- ggplot(impact_comparison, aes(x = reorder(impact, proportion), y = proportion)) +
  geom_bar(stat = "identity", fill = ifelse(impact_comparison$proportion > 0, "red", "blue"), alpha = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid", size = 0.5) +
  labs(
    title = "Proportion Difference in Impact Prevalence",
    subtitle = "Non-molecular coloc vs All coloc groups (Positive = more common in non-molecular)",
    x = "VEP Impact",
    y = "Proportion Difference (Non-molecular / All) * 100",
    caption = paste("Impact levels with data in both datasets:", nrow(impact_comparison))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 10)
  ) +
  coord_flip()

# 3. Feature type comparison
all_feature_type_counts <- all_coloc_groups_dedup %>%
  filter(!is.na(feature_type)) %>%
  count(feature_type, sort = TRUE)

all_non_molecular_coloc_feature_type_counts <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(feature_type)) %>%
  count(feature_type, sort = TRUE)

feature_type_comparison <- all_feature_type_counts %>%
  full_join(all_non_molecular_coloc_feature_type_counts, 
            by = "feature_type", 
            suffix = c("_all", "_non_molecular")) %>%
  mutate(
    n_all = replace_na(n_all, 0),
    n_non_molecular = replace_na(n_non_molecular, 0),
    percent_diff = ((n_non_molecular - n_all) / n_all) * 100,
    proportion = (n_non_molecular / n_all) * 100
  ) %>%
  filter(n_all > 0)

feature_type_hist <- ggplot(feature_type_comparison, aes(x = reorder(feature_type, proportion), y = proportion)) +
  geom_bar(stat = "identity", fill = ifelse(feature_type_comparison$proportion > 0, "red", "blue"), alpha = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid", size = 0.5) +
  labs(
    title = "Proportion Difference in Feature Type Prevalence",
    subtitle = "Non-molecular coloc vs All coloc groups (Positive = more common in non-molecular)",
    x = "Feature Type",
    y = "Proportion Difference (Non-molecular / All) * 100",
    caption = paste("Feature types with data in both datasets:", nrow(feature_type_comparison))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 10)
  ) +
  coord_flip()


# 4. cDNA position comparison (histogram since it's continuous)
all_cdna_positions <- all_coloc_groups_dedup %>%
  filter(!is.na(cdna_position)) %>%
  mutate(cdna_position = as.numeric(cdna_position)) %>%
  filter(!is.na(cdna_position))

all_non_molecular_coloc_cdna_positions <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(cdna_position)) %>%
  mutate(cdna_position = as.numeric(cdna_position)) %>%
  filter(!is.na(cdna_position))

cdna_hist <- ggplot() +
  geom_histogram(data = all_cdna_positions, aes(x = cdna_position, y = ..density..), 
                 bins = 30, fill = "blue", alpha = 0.5, color = "black") +
  geom_histogram(data = all_non_molecular_coloc_cdna_positions, aes(x = cdna_position, y = ..density..), 
                 bins = 30, fill = "red", alpha = 0.5, color = "black") +
  labs(
    title = "Distribution of cDNA Positions",
    subtitle = "Blue = All coloc groups, Red = Non-molecular coloc groups",
    x = "cDNA Position",
    y = "Density",
    caption = paste("All coloc SNPs:", nrow(all_cdna_positions), 
                   "Non-molecular SNPs:", nrow(all_non_molecular_coloc_cdna_positions))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  )


# 5. Protein position comparison (histogram since it's continuous)
all_protein_positions <- all_coloc_groups_dedup %>%
  filter(!is.na(protein_position)) %>%
  mutate(protein_position = as.numeric(protein_position)) %>%
  filter(!is.na(protein_position))

all_non_molecular_coloc_protein_positions <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(protein_position)) %>%
  mutate(protein_position = as.numeric(protein_position)) %>%
  filter(!is.na(protein_position))

protein_hist <- ggplot() +
  geom_histogram(data = all_protein_positions, aes(x = protein_position, y = ..density..), 
                 bins = 30, fill = "blue", alpha = 0.5, color = "black") +
  geom_histogram(data = all_non_molecular_coloc_protein_positions, aes(x = protein_position, y = ..density..), 
                 bins = 30, fill = "red", alpha = 0.5, color = "black") +
  labs(
    title = "Distribution of Protein Positions",
    subtitle = "Blue = All coloc groups, Red = Non-molecular coloc groups",
    x = "Protein Position",
    y = "Density",
    caption = paste("All coloc SNPs:", nrow(all_protein_positions), 
                   "Non-molecular SNPs:", nrow(all_non_molecular_coloc_protein_positions))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  )


ensembl_gene_map <- DBI::dbGetQuery(studies_conn, "SELECT gene as gene_name, ensembl_id FROM gene_annotations")
all_gene_counts <- all_coloc_groups_dedup %>%
  left_join(ensembl_gene_map, by = c("snp_gene" = "ensembl_id")) %>%
  count(gene_name, sort = TRUE)

all_non_molecular_coloc_gene_counts <- all_significant_non_molecular_coloc_dedup %>%
  left_join(ensembl_gene_map, by = c("snp_gene" = "ensembl_id")) %>%
  count(gene_name, sort = TRUE)

na_gene_count <- all_non_molecular_coloc_gene_counts$n[is.na(all_non_molecular_coloc_gene_counts$gene_name)]

all_non_molecular_coloc_gene_counts <- all_non_molecular_coloc_gene_counts %>%
  filter(!is.na(gene_name))


# Show most common genes in non-molecular coloc groups
gene_hist <- ggplot(all_non_molecular_coloc_gene_counts %>% head(50), aes(x = reorder(gene_name, n), y = n)) +
  geom_bar(stat = "identity", fill = "red", alpha = 0.7) +
  labs(
    title = "Most Common Genes without molecular coloc results (Top 50)",
    subtitle = paste("Number of SNPs without known gene name:", na_gene_count),
    x = "Gene",
    y = "Count",
    caption = paste("Total unique genes in non-molecular coloc:", nrow(all_non_molecular_coloc_gene_counts))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 10)
  ) +
  coord_flip()

# Display the plots
print(consequence_hist)
print(impact_hist)
print(feature_type_hist)
print(cdna_hist)
print(protein_hist)
print(gene_hist)
```


# Can we calculate a 'pleiotropy score' for each SNP?

### Colocalisation proximity to genes: how can we quantify 'artefactual' pleiotropy?
