---
title: "Pleiotropy Investigation"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE, echo=F} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(ggplot2)
library(dplyr)
library(tidyr)
source('../../pipeline_steps/constants.R')

studies_db_file <- file.path(latest_results_dir, 'studies.db')
associations_db_file <- file.path(latest_results_dir, 'associations.db')
coloc_pairs_full_db_file <- file.path(latest_results_dir, 'coloc_pairs_full.db')
ld_db_file <- file.path(latest_results_dir, 'ld.db')

studies_conn <- duckdb::dbConnect(duckdb::duckdb(), studies_db_file, read_only = TRUE)
associations_conn <- duckdb::dbConnect(duckdb::duckdb(), associations_db_file, read_only = TRUE)
coloc_pairs_conn <- duckdb::dbConnect(duckdb::duckdb(), coloc_pairs_full_db_file, read_only = TRUE)
ld_conn <- duckdb::dbConnect(duckdb::duckdb(), ld_db_file, read_only = TRUE)

dont_print <- DBI::dbExecute(coloc_pairs_conn, glue::glue("ATTACH '{studies_db_file}' AS studies"))
```


### Questions to attempt to answer:

1. Association vs. colocalization: how often do we see a genetic association with a trait that doesn't have a coloc result?
   * Prevelance of H3 > 0.8 vs. H4 > 0.8
2. How many coloc groups have only qtls?  QTL and phenotype, QTL and multiple phenotypes
3. How many of these chosen causal SNPs are in LD with each other?
4. Modularity of pleiotropy
   * How can we quantify modularity?
   * How many genes are associated with a trait on average?
   * Does it change when including trans-QTLs?
   * What about gene-gene modularity? Could we have a graph where we count the number of associations between genes tagged?
   * Genes that colocalise with the most traits (or trait categories?).  
5. What fraction of known GWAS hits are no known molecular trait associated with it?
6. Is there consistency between of molecule-disease associations between common cis, rare cis and trans instruments?
7. Can we calculate a 'pleiotropy score' for each SNP?  Would it be useful?


```{r load-data, echo=F}
joined_coloc_results <- DBI::dbGetQuery(coloc_pairs_conn, 'SELECT
  coloc_pairs.*,
  -- Study extraction A details
  study_extraction_a.unique_study_id AS study_extraction_a_unique_study_id,
  study_extraction_a.min_p AS min_p_a,
  study_extraction_a.cis_trans AS cis_trans_a,
  -- Study A details
  study_a.study_name AS study_name_a,
  study_a.gene AS gene_a,
  study_a.tissue AS tissue_a,
  study_a.data_type AS data_type_a,
  trait_a.trait_category AS trait_category_a,
  trait_a.trait_name AS trait_name_a,
  -- Study extraction B details
  study_extraction_b.unique_study_id AS study_extraction_b_unique_study_id,
  study_extraction_b.min_p AS min_p_b,
  study_extraction_b.cis_trans AS cis_trans_b,
  -- Study B details
  study_b.study_name AS study_name_b,
  study_b.gene AS gene_b,
  study_b.tissue AS tissue_b,
  study_b.data_type AS data_type_b,
  trait_b.trait_category AS trait_category_b,
  trait_b.trait_name AS trait_name_b
  FROM coloc_pairs
  -- Join for study extraction A
  JOIN studies.study_extractions AS study_extraction_a ON coloc_pairs.study_extraction_a_id = study_extraction_a.id
  JOIN studies.studies AS study_a ON study_extraction_a.study_id = study_a.id
  JOIN studies.traits AS trait_a ON study_a.trait_id = trait_a.id
  -- Join for study extraction B
  JOIN studies.study_extractions AS study_extraction_b ON coloc_pairs.study_extraction_b_id = study_extraction_b.id
  JOIN studies.studies AS study_b ON study_extraction_b.study_id = study_b.id
  JOIN studies.traits AS trait_b ON study_b.trait_id = trait_b.id
  WHERE coloc_pairs.h4 > 0.6 OR coloc_pairs.h3 > 0.6
')
unique_data_types <- DBI::dbGetQuery(studies_conn, "SELECT DISTINCT data_type FROM studies")$data_type
```

## Apparent pleiotropy due to LD

Association vs. colocalization: How often would we see a genetic association with a trait that doesn't have a coloc result?

For any causal variant:

* How many traits does it coloc with (significant h4)
* How many traits does it associate with (significant h3)
* Rate of incorrectness due to LD = significant h4 / significant h3
* How many significant h4 traits are spurious? (Spurious: h4 > 0.8 & failed the clustering step)

```{r h4_h3_ratio, echo=F}
num_significant_h4 <- nrow(joined_coloc_results[joined_coloc_results$h4 > 0.8, ])
num_spurious_h4 <- nrow(joined_coloc_results[joined_coloc_results$h4 > 0.8 & joined_coloc_results$spurious, ])
num_significant_h3 <- nrow(joined_coloc_results[joined_coloc_results$h3 > 0.8, ])
num_spurious_h3 <- nrow(joined_coloc_results[joined_coloc_results$h3 > 0.8 & joined_coloc_results$spurious, ])

hemoglobin_study_id <- 'ebi-a-GCST90002384'
hemoglobin_example <- joined_coloc_results[joined_coloc_results$study_name_a == hemoglobin_study_id | joined_coloc_results$study_name_b == hemoglobin_study_id, ]

hemoglobin_h4 <- nrow(hemoglobin_example[hemoglobin_example$h4 > 0.8, ])
hemoglobin_h3 <- nrow(hemoglobin_example[hemoglobin_example$h3 > 0.8, ])
hemoglobin_h4_spurious <- nrow(hemoglobin_example[hemoglobin_example$h4 > 0.8 & hemoglobin_example$spurious, ])
hemoglobin_h3_spurious <- nrow(hemoglobin_example[hemoglobin_example$h3 > 0.8 & hemoglobin_example$spurious, ])

h4_summary <- tibble::tribble(
  ~trait, ~num_significant_h4, ~num_spurious_h4, ~num_significant_h3, ~h3_h4_ratio,
  'All', num_significant_h4, num_spurious_h4, num_significant_h3, num_significant_h3 / num_significant_h4,
  'Hemoglobin', hemoglobin_h4, hemoglobin_h4_spurious, hemoglobin_h3, hemoglobin_h3 / hemoglobin_h4
)

knitr::kable(h4_summary)
```

#### Breakdown of ratio of h3 to h4 by data type

```{r matrix_of_coloc_results, echo=F}
coloc_matrix <- dplyr::filter(joined_coloc_results, h4 > 0.8 | h3 > 0.8)

# Create the n×n matrix
h4_coloc_matrix_by_type <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(h4_coloc_matrix_by_type) <- unique_data_types
colnames(h4_coloc_matrix_by_type) <- unique_data_types

h3_coloc_matrix_by_type <- matrix(0, nrow = length(unique_data_types), ncol = length(unique_data_types))
rownames(h3_coloc_matrix_by_type) <- unique_data_types
colnames(h3_coloc_matrix_by_type) <- unique_data_types

# Fill the matrix with coloc counts (combining symmetric entries)
for (i in 1:nrow(coloc_matrix)) {
  type_a <- coloc_matrix$data_type_a[i]
  type_b <- coloc_matrix$data_type_b[i]

  if (coloc_matrix$h4[i] > 0.8) {
    h4_coloc_matrix_by_type[type_a, type_b] <- h4_coloc_matrix_by_type[type_a, type_b] + 1
    if (type_a != type_b) {
      h4_coloc_matrix_by_type[type_b, type_a] <- h4_coloc_matrix_by_type[type_b, type_a] + 1
    }
  } else if (coloc_matrix$h3[i] > 0.8) {
    h3_coloc_matrix_by_type[type_a, type_b] <- h3_coloc_matrix_by_type[type_a, type_b] + 1
    if (type_a != type_b) {
      h3_coloc_matrix_by_type[type_b, type_a] <- h3_coloc_matrix_by_type[type_b, type_a] + 1
    }
  }
}

h4_coloc_matrix_by_type[lower.tri(h4_coloc_matrix_by_type)] <- NA
h4_coloc_matrix_by_type <- as.data.frame(h4_coloc_matrix_by_type)
h4_coloc_pretty <- h4_coloc_matrix_by_type
h4_coloc_pretty[, unique_data_types] <- lapply(h4_coloc_pretty[, unique_data_types], scales::comma)
knitr::kable(h4_coloc_pretty, caption = "Number of H4 > 0.8 between data types")

h3_coloc_matrix_by_type[lower.tri(h3_coloc_matrix_by_type)] <- NA
h3_coloc_matrix_by_type <- as.data.frame(h3_coloc_matrix_by_type)
h3_coloc_pretty <- h3_coloc_matrix_by_type
h3_coloc_pretty[, unique_data_types] <- lapply(h3_coloc_pretty[, unique_data_types], scales::comma)
knitr::kable(h3_coloc_pretty, caption = "Number of H3 > 0.8 between data types")

h3_h4_ratio <- h3_coloc_matrix_by_type / h4_coloc_matrix_by_type
knitr::kable(h3_h4_ratio, caption = "Ratio of H3 > 0.8 to H4 > 0.8 between data types")
```


## Unpacking Spurious colocalizations

### Spurious colocalization rate by h4 score

Is there a relationship between the h4 score and the spuriousness of the colocalization?  If h4 > 0.8 too lenient?

```{r spurious_colocalization_rate_by_h4, echo=F}
spurious_results <- joined_coloc_results[joined_coloc_results$h4 > 0.8, ]

spurious_results_by_h4 <- spurious_results %>%
  mutate(h4 = round(h4, 2)) %>%
  group_by(h4) %>%
  summarise(
    total_results = n(),
    spurious_results = sum(spurious),
    spurious_percentage = (sum(spurious) / n()) * 100,
    .groups = 'drop'
  )

spurious_colocalization_rate_by_h4 <- ggplot(spurious_results_by_h4, aes(x = h4, y = spurious_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  labs(title = "Spurious colocalization rate by h4 score", x = "h4 score", y = "Percentage of spurious colocalizations (%)")

spurious_colocalization_rate_by_h4
ggsave("spurious_colocalization_rate_by_h4.png", spurious_colocalization_rate_by_h4, width = 10, height = 6)

spurious_results_by_h4 <- spurious_results_by_h4 %>%
  mutate(
    predicted_fpr = (1 - h4) * 100,
    calibration_error = spurious_percentage - predicted_fpr
  )

# Create calibration plot
calibration_plot <- ggplot(spurious_results_by_h4, aes(x = predicted_fpr, y = spurious_percentage)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_line(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red", alpha = 0.7) +
  labs(
    title = "Calibration Plot: Predicted vs Observed False Positive Rate",
    x = "Predicted False Positive Rate (1 - H4) (%)",
    y = "Observed Spurious Rate (%)",
    caption = "Red dashed line represents perfect calibration (predicted = observed)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

calibration_plot
ggsave("h4_calibration_plot.png", calibration_plot, width = 10, height = 6)

```


### Spurious colocalization rate by data type

```{r spurious_colocalization_rate, echo=F}
studies_by_data_type <- DBI::dbGetQuery(studies_conn, 'SELECT studies.data_type, COUNT(*) AS num_studies FROM study_extractions 
  JOIN studies ON study_extractions.study_id = studies.id
  GROUP BY studies.data_type
')

spurious_results <- joined_coloc_results[joined_coloc_results$spurious, ]
spurious_results_by_data_type <- spurious_results %>%
  select(data_type_a, data_type_b) %>%
  tidyr::pivot_longer(cols = c(data_type_a, data_type_b), 
              names_to = "position", 
              values_to = "data_type") %>%
  group_by(data_type) %>%
  summarise(n = n(), .groups = 'drop')

# Join with study counts to calculate rates
spurious_analysis <- spurious_results_by_data_type %>%
  left_join(studies_by_data_type, by = "data_type") %>%
  mutate(
    spurious_rate = n / num_studies,
    spurious_rate_percent = spurious_rate * 100,
    label_text = sprintf("%.1f%%\n(%d/%d)", spurious_rate_percent, n, num_studies)
  )

# Create histogram of spurious results by data type (adjusted for study count)
ggplot(spurious_analysis, aes(x = data_type, y = spurious_rate_percent)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = label_text), vjust = -0.5, size = 3) +
  labs(
    title = "Spurious colocalization rate by data type",
    x = "Data type",
    y = "Spurious rate (%)",
    caption = "Percentage of studies with spurious colocalizations (spurious/total studies)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )


```


```{r h3_h4_ratio_08, echo=F, eval=F}

h4_ratio_histogram <- function(threshold, filter_spurious = TRUE) {
  hemoglobin_example$hemoglobin_min_p <- ifelse(
    hemoglobin_example$study_name_a == hemoglobin_study_id,
    hemoglobin_example$min_p_a,
    hemoglobin_example$min_p_b
  )
  hemoglobin_example$min_p_bucket <- cut(
    hemoglobin_example$hemoglobin_min_p,
    breaks = c(0, 1e-20, 1e-8, 1e-7, 1e-6, 1e-5, 1e-4),
    labels = c("≤1e-20", "1e-20 to 1e-8", "1e-8 to 1e-7", "1e-7 to 1e-6", "1e-6 to 1e-5", "1e-5 to 1e-4"),
    include.lowest = TRUE
  )
  
  coloc_summary <- hemoglobin_example %>%
    group_by(min_p_bucket) %>%
    summarise(
      h3_gt_threshold = sum(h3 > threshold, na.rm = TRUE),
      h4_gt_threshold = sum(h4 > threshold, na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    mutate(
      h3_h4_ratio = ifelse(h4_gt_threshold > 0, h3_gt_threshold / h4_gt_threshold, NA),
      bucket_label = paste0(min_p_bucket, "\n(n=", h4_gt_threshold, ")")
    ) %>%
    mutate(
      min_p_bucket = factor(min_p_bucket, 
                           levels = c("≤1e-20", "1e-20 to 1e-8", "1e-8 to 1e-7", "1e-7 to 1e-6", "1e-6 to 1e-5", "1e-5 to 1e-4"),
                           ordered = TRUE),
      bucket_label = factor(bucket_label, 
                           levels = paste0(c("≤1e-20", "1e-20 to 1e-8", "1e-8 to 1e-7", "1e-7 to 1e-6", "1e-6 to 1e-5", "1e-5 to 1e-4"), "\n(h4 n=", h4_gt_threshold, ")"),
                           ordered = TRUE)
    )
  
  p <- ggplot(coloc_summary, aes(x = bucket_label, y = h3_h4_ratio)) +
    geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
    geom_text(aes(label = sprintf("%.2f", h3_h4_ratio)), 
              vjust = -0.5, size = 3) +
    labs(
      title = glue::glue("Ratio of H3 > {threshold} to H4 > {threshold} by min_p buckets"),
      x = glue::glue("min_p bucket (with H4 > {threshold} count)"),
      y = glue::glue("H3 > {threshold} / H4 > {threshold} ratio"),
      caption = "H4: shared causal variant, H3: distinct causal variants"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5)
    )
  return(p)
}

h4_ratio_histogram(0.8)
h4_ratio_histogram(0.6)
```

## How often does a coloc group have multiple genes?  Multiple phenotypes? (cis only)

```{r multiple_genes_in_coloc_group, echo=F}
coloc_group_data <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*,
  studies.gene, studies.sample_size, studies.data_type, studies.tissue, studies.study_name,
  gene_annotations.chr, gene_annotations.start, gene_annotations.stop, gene_annotations.gene_biotype,
  study_extractions.cis_trans
  FROM coloc_groups
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN studies.study_extractions ON coloc_groups.study_extraction_id = studies.study_extractions.id
  LEFT JOIN gene_annotations ON studies.gene_id = gene_annotations.id
  WHERE study_extractions.min_p < 1e-8
  AND (study_extractions.cis_trans = 'cis' OR study_extractions.cis_trans IS NULL)
")

# Filter to only keep coloc groups with multiple entries
coloc_group_data <- coloc_group_data |>
  group_by(coloc_group_id) |>
  filter(n() > 1) |>
  ungroup()



gene_phenotype_counts_per_coloc_group <- coloc_group_data |>
  filter(group_threshold == "strong") |>
  group_by(coloc_group_id) |>
  summarise(
    unique_genes = n_distinct(gene, na.rm = TRUE),
    num_phenotypes = sum(data_type == "phenotype", na.rm = TRUE),
    .groups = 'drop'
  )

gene_phenotype_bins <- gene_phenotype_counts_per_coloc_group |>
  mutate(
    genes_bin = case_when(
      unique_genes == 0 ~ "No genes",
      unique_genes <= 5 ~ paste0(unique_genes, " genes"),
      TRUE ~ ">5 genes"
    ),
    phenotypes_bin = case_when(
      num_phenotypes == 0 ~ "No phenotypes",
      num_phenotypes <= 5 ~ paste0(num_phenotypes, " phenotypes"),
      TRUE ~ ">5 phenotypes"
    )
  ) |>
  mutate(
    genes_bin = factor(genes_bin, levels = c("No genes", paste0(1:5, " genes"), ">5 genes")),
    phenotypes_bin = factor(phenotypes_bin, levels = c("No phenotypes", paste0(1:5, " phenotypes"), ">5 phenotypes"))
  )

genes_phenotypes_matrix <- table(
  gene_phenotype_bins$genes_bin,
  gene_phenotype_bins$phenotypes_bin
)

genes_phenotypes_df <- as.data.frame(genes_phenotypes_matrix)
colnames(genes_phenotypes_df) <- c("genes_bin", "phenotypes_bin", "count")

genes_phenotypes_matrix_df <- genes_phenotypes_df |>
  tidyr::pivot_wider(
    names_from = phenotypes_bin,
    values_from = count,
    values_fill = 0
  )

knitr::kable(genes_phenotypes_matrix_df, caption = "Number of coloc groups by unique genes and phenotypes (methQTL are neither genes nor phenotypes)")

percentage_of_coloc_groups <- genes_phenotypes_matrix_df |>
  mutate(across(-genes_bin, ~ .x / nrow(gene_phenotype_bins))) |>
  mutate(across(-genes_bin, ~ scales::percent(.x, accuracy = 0.01)))

knitr::kable(percentage_of_coloc_groups, caption = "Percentage of coloc groups by unique genes and phenotypes (methQTL are neither genes nor phenotypes)")



```

### Breakdown of coloc groups by number of genes 

```{r multiple_genes_in_coloc_group_protein_coding_trans, echo=F}
coloc_group_data <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*,
  studies.gene, studies.sample_size, studies.data_type, studies.tissue, studies.study_name,
  gene_annotations.chr, gene_annotations.start, gene_annotations.stop, gene_annotations.gene_biotype,
  study_extractions.cis_trans
  FROM coloc_groups
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN studies.study_extractions ON coloc_groups.study_extraction_id = studies.study_extractions.id
  LEFT JOIN gene_annotations ON studies.gene_id = gene_annotations.id
  WHERE study_extractions.min_p < 1e-8
")
multiple_genes_in_coloc_group <- function(threshold, only_protein_coding = FALSE, include_trans = TRUE) {
  if (only_protein_coding) {
    coloc_group_data <- coloc_group_data %>%
      filter(gene_biotype == "protein_coding")
  }

  if (!include_trans) {
    coloc_group_data <- coloc_group_data %>%
      filter(cis_trans != "trans")
  }

  coloc_group_analysis <- coloc_group_data %>%
    filter(group_threshold == threshold) %>%
    group_by(coloc_group_id) %>%
    summarise(
      num_genes = n(),
      unique_genes = n_distinct(gene, na.rm = TRUE),
      genes = paste(unique(gene[!is.na(gene)]), collapse = ", "),
      .groups = 'drop'
    ) %>%
    mutate(
      genes = ifelse(genes == "", NA, genes),
      unique_genes = ifelse(is.na(genes), 0, unique_genes)
    )

  coloc_group_analysis$unique_gene_bin <- case_when(
    coloc_group_analysis$unique_genes == 0 ~ "0",
    coloc_group_analysis$unique_genes <= 20 ~ as.character(coloc_group_analysis$unique_genes),
    TRUE ~ paste0(">20")
  )
  
  # Convert to factor with proper ordering
  coloc_group_analysis$unique_gene_bin <- factor(coloc_group_analysis$unique_gene_bin, 
                                                levels = c(as.character(0:20), ">20"))
  
  # Count by bins
  unique_gene_bin_counts <- coloc_group_analysis %>%
    group_by(unique_gene_bin) %>%
    summarise(count = n(), .groups = 'drop') %>%
    filter(!is.na(unique_gene_bin))
  
  p_gene_dist <- ggplot(unique_gene_bin_counts, aes(x = unique_gene_bin, y = count)) +
    geom_bar(stat = "identity", fill = "lightblue", color = "black", alpha = 0.7) +
    geom_text(aes(label = count), vjust = -0.5, size = 2) +
    labs(
      title = glue::glue("{if (only_protein_coding) 'protein coding' else 'all'} unique genes per coloc group, {if (include_trans) 'including trans' else 'excluding trans'}"),
      x = "Number of unique genes per coloc group",
      y = "Frequency (log scale)",
      caption = paste("Total groups:", nrow(coloc_group_analysis), 
                     "| Groups with 0 genes:", sum(coloc_group_analysis$unique_genes == 0),
                     "| Groups with >1 unique gene:", sum(coloc_group_analysis$unique_genes > 1))
    ) +
    theme_minimal() +
    scale_y_log10() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, size = 11)
    )
  return(p_gene_dist)
}

multiple_genes_in_coloc_group('strong', only_protein_coding = TRUE, include_trans = FALSE)
multiple_genes_in_coloc_group('strong', only_protein_coding = TRUE)
multiple_genes_in_coloc_group('strong')
```

## Genes that colocalise with the most traits (or trait categories?).  
Limited to cis-coloc pairs with h4 > 0.8 and p-values < 1e-8.

```{r genes_colocalising_with_most_traits, echo=F}
significant_cis_coloc_pairs <- joined_coloc_results %>%
  filter(h4 > 0.8 & (is.na(cis_trans_a) | cis_trans_a == 'cis') & (is.na(cis_trans_b) | cis_trans_b == 'cis') & min_p_a < 1e-8 & min_p_b < 1e-8)

most_gene_a_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_a) & data_type_b == 'phenotype') %>%
  group_by(gene_a) %>%
  rename(gene = gene_a, trait = trait_name_b)

most_gene_b_colocalising <- significant_cis_coloc_pairs %>%
  filter(!is.na(gene_b) & data_type_a == 'phenotype') %>%
  group_by(gene_b) %>%
  rename(gene = gene_b, trait = trait_name_a)

most_genes_colocalising_with_traits <- rbind(most_gene_a_colocalising, most_gene_b_colocalising)

most_genes_colocalising_summary <- most_genes_colocalising_with_traits %>%
  group_by(gene) %>%
  summarise(num_traits = n_distinct(trait)) %>%
  arrange(desc(num_traits)) %>%
  head(50)

knitr::kable(most_genes_colocalising_summary, caption = "Genes that colocalise with the most traits")

top_gene <- most_genes_colocalising_summary$gene[1]
all_phenotypes_for_top_gene <- most_genes_colocalising_with_traits %>%
  filter(gene == top_gene) %>%
  select(trait) %>%
  distinct()

knitr::kable(head(all_phenotypes_for_top_gene, 30), caption = glue::glue("Phenotypes for top gene: {top_gene}"))

```

# LD relationship between causal SNPs

### How often Causal SNPs are in LD with each other
```{r causal_snps_in_ld, echo=F}

all_causal_snps <- DBI::dbGetQuery(studies_conn, "SELECT DISTINCT snp_id FROM coloc_groups WHERE group_threshold = 'strong'")

dont_print <- DBI::dbExecute(ld_conn, glue::glue("ATTACH '{studies_db_file}' AS studies"))

formatted_snp_ids <- paste0(all_causal_snps$snp_id, collapse = ",")
snp_pair_ld_scores <- DBI::dbGetQuery(ld_conn, glue::glue("SELECT ld.*, studies.ld_blocks.ld_block AS ld_block FROM ld
  JOIN studies.ld_blocks ON ld.ld_block_id = studies.ld_blocks.id
  WHERE lead_snp_id IN ({formatted_snp_ids})
  AND variant_snp_id IN ({formatted_snp_ids})"
))

snp_pair_ld_scores$r2 <- snp_pair_ld_scores$r ^ 2

r2_thresholds <- c(0.8, 0.9, 0.95, 0.99)

ld_results <- lapply(r2_thresholds, function(r2_threshold) {
  snp_pairs_in_ld <- snp_pair_ld_scores[snp_pair_ld_scores$r2 > r2_threshold, ]
  unique_causal_snps_in_ld <- length(unique(snp_pairs_in_ld$lead_snp_id))
  return(data.frame(
    r2_threshold = r2_threshold,
    unique_causal_snps_in_ld = unique_causal_snps_in_ld,
    number_of_snp_pairs_in_ld = nrow(snp_pairs_in_ld),
    percentage_of_causal_snps_in_ld = unique_causal_snps_in_ld / length(all_causal_snps$snp_id)
  ))
})

ld_results <- do.call(rbind, ld_results)
knitr::kable(ld_results, caption = "Number of unique causal SNPs in LD by r2 threshold")

snp_pairs_in_ld <- snp_pair_ld_scores[snp_pair_ld_scores$r2 > 0.8, ]
ld_block_summary <- snp_pairs_in_ld %>%
  group_by(ld_block_id, ld_block) %>%
  summarise(
    num_snp_pairs = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(num_snp_pairs)) %>%
  head(10)

knitr::kable(ld_block_summary, 
  caption = "Top 10 LD blocks by number of causal SNP pairs in LD",
  col.names = c("LD Block ID", "LD Block", "Number of SNP Pairs"))


```

### How many SNPs are found in multiple coloc groups?

```{r snps_per_coloc_group, echo=F}
snps_per_coloc_group <- function(threshold) {
  snps_per_coloc_group_data <- DBI::dbGetQuery(studies_conn, glue::glue("WITH SnpGroupCounts AS (
    -- Step 1: Find the number of times each unique snp_id appears
    -- in each unique coloc_group_id for a specific threshold.
    SELECT
        snp_id,
        coloc_group_id,
        COUNT(*) AS num_snps_in_group
    FROM
        coloc_groups
    WHERE
        group_threshold = '{threshold}'
    GROUP BY
        snp_id,
        coloc_group_id
    )
    -- Step 2: Now, count how many different coloc_group_ids each snp_id has.
    SELECT
        snp_id,
        COUNT(coloc_group_id) AS number_of_coloc_groups
    FROM
        SnpGroupCounts
    GROUP BY
        snp_id
    HAVING
        COUNT(coloc_group_id) > 1;"))

  snp_coloc_group_distribution <- snps_per_coloc_group_data %>%
    group_by(number_of_coloc_groups) %>%
    summarise(
      number_of_snps = n(),
      .groups = 'drop'
    ) %>%
    arrange(number_of_coloc_groups)
  
  # Create histogram
  snp_coloc_group_histogram <- ggplot(snp_coloc_group_distribution, aes(x = number_of_coloc_groups, y = number_of_snps)) +
    geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.7) +
    scale_x_continuous(limits = c(1, NA)) +
    labs(
      title = glue::glue("Distribution of SNPs found in multiple coloc groups in {threshold} groups"),
      x = "Number of coloc groups per SNP",
      y = "Number of SNPs",
      caption = "SNPs that appear in multiple coloc groups"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  
  total_snps <- unique(DBI::dbGetQuery(studies_conn, glue::glue("SELECT COUNT(DISTINCT snp_id) FROM coloc_groups WHERE group_threshold = '{threshold}'")))
  total_snps <- total_snps$count

  return(list(histogram = snp_coloc_group_histogram, total_snps = total_snps, data = snp_coloc_group_distribution))
}

strong_snps_per_coloc_group <- snps_per_coloc_group('strong')
strong_snps_per_coloc_group$histogram
moderate_snps_per_coloc_group <- snps_per_coloc_group('moderate')

coloc_group_summary <- tibble::tribble(
  ~threshold, ~num_reused_snps, ~total_snps,
  'moderate', sum(moderate_snps_per_coloc_group$data$number_of_snps), moderate_snps_per_coloc_group$total_snps,
  'strong', sum(strong_snps_per_coloc_group$data$number_of_snps), strong_snps_per_coloc_group$total_snps
)

knitr::kable(coloc_group_summary)

# Create histogram of coloc group sizes
group_size_distribution <- DBI::dbGetQuery(studies_conn, "SELECT t.group_size, COUNT(*) AS number_of_groups 
  FROM 
    (SELECT coloc_group_id, COUNT(*) as group_size
    FROM coloc_groups WHERE group_threshold = 'strong'
    GROUP BY coloc_group_id)
  AS t
  GROUP BY t.group_size
  ORDER BY t.group_size")

ggplot(group_size_distribution, aes(x = group_size, y = number_of_groups)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(limits = c(1, NA)) +
  labs(
    title = glue::glue("Distribution of colocalization group sizes in strong groups"),
    x = "Size of coloc group",
    y = "Number of groups (log scale)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

# Universal vs. Modular Pleiotropy

### How many genes are associated with a trait on average?

```{r num_genes_colocalising, echo=F}
all_phenotype_studies <- DBI::dbGetQuery(studies_conn, "SELECT studies.*, traits.trait_name FROM studies
  JOIN traits ON studies.trait_id = traits.id
  WHERE studies.data_type = 'phenotype' AND studies.variant_type = 'common'
")

number_of_genes <- lapply(seq_along(all_phenotype_studies$id), function(i) {
  study <- all_phenotype_studies[i, , drop = FALSE]
  study_id <- study$id
  study_name <- study$study_name
  all_studies_in_coloc_groups_for_study <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT COUNT(DISTINCT studies.gene_id) AS num_genes
    FROM coloc_groups
    JOIN studies ON coloc_groups.study_id = studies.id
    JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
    WHERE study_extractions.min_p < 1e-8
    AND coloc_groups.coloc_group_id IN (SELECT DISTINCT coloc_group_id
      FROM coloc_groups
      JOIN studies on coloc_groups.study_id = studies.id
      JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
      WHERE coloc_groups.group_threshold = 'strong'
      AND studies.id = {study_id}
      AND study_extractions.min_p < 1e-8
      AND (study_extractions.cis_trans = 'cis' OR study_extractions.cis_trans IS NULL)
    )
  "))
  all_studies_in_coloc_groups_for_study$study_id <- study_id
  all_studies_in_coloc_groups_for_study$trait_name <- study$trait_name
  all_studies_in_coloc_groups_for_study$sample_size <- study$sample_size
  return(all_studies_in_coloc_groups_for_study)
})
number_of_genes <- do.call(rbind, number_of_genes)

ggplot(number_of_genes, aes(x = num_genes)) +
  geom_histogram(binwidth = 1, fill = "steelblue", alpha = 0.7) +
  scale_y_log10() +
  labs(
    title = "Distribution of number of genes colocalising per phenotype study",
    x = "Number of genes",
    y = "Count of studies (log scale)",
    caption = paste("Mean:", round(mean(number_of_genes$num_genes), 2),
                   "| Median:", round(median(number_of_genes$num_genes), 2))
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))


number_of_genes_summary <- number_of_genes %>%
  arrange(desc(num_genes)) %>%
  head(10)
knitr::kable(number_of_genes_summary, caption = "Top 10 studies by number of genes colocalising")

median_sample_size <- median(number_of_genes$sample_size, na.rm = TRUE)
median_num_genes <- median(number_of_genes$num_genes, na.rm = TRUE)

outlier_studies <- number_of_genes %>%
  filter(sample_size > median_sample_size & num_genes < median_num_genes) %>%
  arrange(desc(sample_size)) %>%
  mutate(
    sample_size = scales::comma(sample_size),
    num_genes = round(num_genes, 2)
  )

knitr::kable(head(outlier_studies, 10), 
             caption = paste0("Studies with large sample size (>", scales::comma(median_sample_size), 
                            ") but small number of genes (<", round(median_num_genes,1), ")"))


```

#### Does this change when we include trans-QTLs?
```{r num_genes_colocalising_trans, echo=F}
all_phenotype_studies <- DBI::dbGetQuery(studies_conn, "SELECT studies.*, traits.trait_name FROM studies
  JOIN traits ON studies.trait_id = traits.id
  WHERE studies.data_type = 'phenotype' AND studies.variant_type = 'common'
")

number_of_genes <- lapply(seq_along(all_phenotype_studies$id), function(i) {
  study <- all_phenotype_studies[i, , drop = FALSE]
  study_id <- study$id
  study_name <- study$study_name
  all_studies_in_coloc_groups_for_study <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT COUNT(DISTINCT studies.gene_id) AS num_genes
    FROM coloc_groups
    JOIN studies ON coloc_groups.study_id = studies.id
    JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
    WHERE study_extractions.min_p < 1e-8
    AND coloc_groups.coloc_group_id IN (SELECT DISTINCT coloc_group_id
      FROM coloc_groups
      JOIN studies on coloc_groups.study_id = studies.id
      JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
      WHERE coloc_groups.group_threshold = 'strong'
      AND studies.id = {study_id}
      AND study_extractions.min_p < 1e-8
    )
  "))
  all_studies_in_coloc_groups_for_study$study_id <- study_id
  all_studies_in_coloc_groups_for_study$trait_name <- study$trait_name
  all_studies_in_coloc_groups_for_study$sample_size <- study$sample_size
  return(all_studies_in_coloc_groups_for_study)
})
number_of_genes <- do.call(rbind, number_of_genes)

ggplot(number_of_genes, aes(x = num_genes)) +
  geom_histogram(binwidth = 1, fill = "steelblue", alpha = 0.7) +
  scale_y_log10() +
  labs(
    title = "Distribution of number of genes colocalising per phenotype study",
    x = "Number of genes",
    y = "Count of studies (log scale)",
    caption = paste("Mean:", round(mean(number_of_genes$num_genes), 2),
                   "| Median:", round(median(number_of_genes$num_genes), 2))
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))


number_of_genes_summary <- number_of_genes %>%
  arrange(desc(num_genes)) %>%
  head(10)
knitr::kable(number_of_genes_summary, caption = "Top 10 studies by number of genes colocalising")

median_sample_size <- median(number_of_genes$sample_size, na.rm = TRUE)
median_num_genes <- median(number_of_genes$num_genes, na.rm = TRUE)

outlier_studies <- number_of_genes %>%
  filter(sample_size > median_sample_size & num_genes < median_num_genes) %>%
  arrange(desc(sample_size)) %>%
  mutate(
    sample_size = scales::comma(sample_size),
    num_genes = round(num_genes, 2)
  )

knitr::kable(head(outlier_studies, 10), 
             caption = paste0("Studies with large sample size (>", scales::comma(median_sample_size), 
                            ") but small number of genes (<", round(median_num_genes,1), ")"))


```


### Relationship between number of genes colocalising for a study, and the sample size

```{r num_genes_colocalising_plot, echo=F}
lm_fit <- lm(num_genes~ sample_size, data = number_of_genes)
ggplot(number_of_genes, aes(x = sample_size, y = num_genes)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red", alpha = 0.3) +
  labs(
    title = "Relationship between number of genes colocalising for a study, and the sample size",
    x = "Sample size",
    y = "Number of genes colocalising",
    caption = paste("R² =", round(summary(lm_fit)$r.squared, 3), 
                   "| p-value =", format.pval(summary(lm_fit)$coefficients[2,4], digits = 3))
  )

```

### What about gene-gene modularity?
How can we look into this?  Some sort of recreation of epigraph db...

```{r gene_gene_modularity, echo=F}

ld_block_ids <- c(1, 60)
for (ld_block_id in ld_block_ids) {
  all_coloc_groups_in_ld_block <- DBI::dbGetQuery(studies_conn, glue::glue("SELECT coloc_groups.coloc_group_id, gene_annotations.gene FROM coloc_groups
    JOIN studies ON coloc_groups.study_id = studies.id
    JOIN gene_annotations ON studies.gene_id = gene_annotations.id
    WHERE coloc_groups.ld_block_id = {ld_block_id}
    AND coloc_groups.group_threshold = 'strong'
  "))

  # Create gene pairs within each coloc group
  gene_pairs <- all_coloc_groups_in_ld_block %>%
    group_by(coloc_group_id) %>%
    do({
      genes_in_group <- unique(.$gene)
      if(length(genes_in_group) > 1) {
        # Create all pairwise combinations of genes in this coloc group
        gene_combinations <- combn(genes_in_group, 2, simplify = FALSE)
        data.frame(
          gene_a = sapply(gene_combinations, function(x) x[1]),
          gene_b = sapply(gene_combinations, function(x) x[2]),
          stringsAsFactors = FALSE
        )
      } else {
        data.frame(gene_a = character(), gene_b = character(), stringsAsFactors = FALSE)
      }
    }) %>%
    ungroup() %>%
    # Count how many times each gene pair appears across coloc groups
    group_by(gene_a, gene_b) %>%
    summarise(
      weight = n(),  # Number of coloc groups where this gene pair appears
      .groups = 'drop'
    )

  # Create igraph object
  library(igraph)
  gene_graph <- graph_from_data_frame(gene_pairs, directed = FALSE)

  # Calculate network metrics
  network_metrics <- data.frame(
    metric = c("Number of nodes (genes)", "Number of edges", "Average degree", "Density", "Clustering coefficient"),
    value = c(
      vcount(gene_graph),
      ecount(gene_graph),
      mean(degree(gene_graph)),
      edge_density(gene_graph),
      transitivity(gene_graph, type = "global")
    )
  )

  knitr::kable(network_metrics, caption = "Gene-Gene Network Metrics")

  # Find genes with highest degree (most connections)
  gene_degree <- data.frame(
    gene = V(gene_graph)$name,
    degree = degree(gene_graph),
    betweenness = betweenness(gene_graph),
    closeness = closeness(gene_graph)
  ) %>%
    arrange(desc(degree)) %>%
    head(10)

  knitr::kable(gene_degree, caption = "Top 10 Genes by Degree (Most Connections)")

  # Create visualization
  library(ggraph)
  library(tidygraph)

  # Convert to tidygraph for easier plotting and add node attributes
  gene_tidy_graph <- as_tbl_graph(gene_graph) %>%
    activate(nodes) %>%
    mutate(
      degree = centrality_degree(),
      betweenness = centrality_betweenness(),
      closeness = centrality_closeness()
    )

  # Plot the network
  network_plot <- ggraph(gene_tidy_graph, layout = "fr") +
    geom_edge_link(aes(alpha = weight), color = "gray50") +
    geom_node_point(aes(size = degree), color = "steelblue", alpha = 0.7) +
    geom_node_text(aes(label = name), size = 3, repel = TRUE) +
    scale_edge_alpha_continuous(range = c(0.1, 0.8)) +
    scale_size_continuous(range = c(2, 8)) +
    labs(
      title = "Gene-Gene Network from Coloc Groups",
      subtitle = paste("LD Block", ld_block_id, "- All genes within coloc groups"),
      caption = paste("Nodes:", vcount(gene_graph), "Genes, Edges:", ecount(gene_graph), "Gene pairs across coloc groups")
    ) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5))

  print(network_plot)
}


```

# Random other ideas

### Colocalisation deserts in the genome
```{r colocalisation_deserts, echo=F}
all_snps_in_coloc_groups <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.coloc_group_id, study_extractions.gene_id, studies.data_type, coloc_groups.snp_id, snp_annotations.chr, snp_annotations.bp FROM coloc_groups
  JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
  JOIN studies ON coloc_groups.study_id = studies.id
  JOIN snp_annotations ON coloc_groups.snp_id = snp_annotations.id
  WHERE group_threshold = 'strong'
  AND study_extractions.min_p < 1e-8
  AND (study_extractions.cis_trans != 'trans' OR study_extractions.cis_trans IS NULL)
")

all_snps_in_coloc_groups_distinct <- all_snps_in_coloc_groups %>%
  group_by(coloc_group_id) %>%
  slice(1) %>%
  ungroup() %>%
  select(chr, bp) %>%
  distinct() %>%
  mutate(type = "All SNPs")

# Remove coloc groups where all gene_ids are NA
coloc_groups_with_genes <- all_snps_in_coloc_groups %>%
  group_by(coloc_group_id) %>%
  filter(!all(is.na(gene_id))) %>%
  ungroup() %>%
  select(chr, bp) %>%
  distinct() %>%
  mutate(type = "With Genes")

coloc_groups_without_genes <- all_snps_in_coloc_groups %>%
  group_by(coloc_group_id) %>%
  filter(all(is.na(gene_id))) %>%
  ungroup() %>%
  select(chr, bp) %>%
  distinct() %>%
  mutate(type = "Without Genes")

coloc_groups_with_only_phenotypes <- all_snps_in_coloc_groups %>%
  group_by(coloc_group_id) %>%
  filter(all(data_type == "phenotype")) %>%
  ungroup() %>%
  select(chr, bp) %>%
  distinct() %>%
  mutate(type = "Only Phenotypes")

coloc_groups_without_methylated_genes_with_genes <- all_snps_in_coloc_groups %>%
  group_by(coloc_group_id) %>%
  filter(!all(data_type == "methylated")) %>%
  ungroup() %>%
  select(chr, bp) %>%
  distinct() %>%
  mutate(type = "Without methQTLs")


all_coloc_groups <- bind_rows(
  coloc_groups_with_genes,
  coloc_groups_without_genes,
  coloc_groups_with_only_phenotypes,
  coloc_groups_without_methylated_genes_with_genes
)

all_coloc_groups$chr <- factor(all_coloc_groups$chr, 
                              levels = c(1:22, "X", "Y"))

# Create cumulative position for proper genome-wide plotting
chr_lengths <- c(
  "1" = 248956422, "2" = 242193529, "3" = 198295559, "4" = 190214555,
  "5" = 181538259, "6" = 170805979, "7" = 159345973, "8" = 145138636,
  "9" = 138394717, "10" = 133797422, "11" = 135086622, "12" = 133275309,
  "13" = 114364328, "14" = 107043718, "15" = 101991189, "16" = 90338345,
  "17" = 83257441, "18" = 80373285, "19" = 58617616, "20" = 64444167,
  "21" = 46709983, "22" = 50818468, "X" = 156040895, "Y" = 57227415
)

# Calculate cumulative positions
all_coloc_groups <- all_coloc_groups %>%
  mutate(
    chr_num = as.numeric(as.character(chr)),
    cumulative_bp = bp + cumsum(c(0, chr_lengths[1:(length(chr_lengths)-1)]))[chr_num]
  )

# Calculate chromosome boundaries for x-axis labels
chr_boundaries <- cumsum(c(0, chr_lengths))
chr_centers <- chr_boundaries[1:length(chr_boundaries)-1] + chr_lengths/2

# Option 1: Continuous genome-wide plot with jittered points
ggplot(all_coloc_groups, aes(x = cumulative_bp, y = type, color = type)) +
  geom_jitter(alpha = 0.4, size = 0.005, height = 0.2, width = 0) +
  geom_vline(xintercept = chr_boundaries, color = "gray80", alpha = 0.5) +
  labs(
    title = "Genome-wide Distribution of Coloc Groups by Type (Jittered)",
    x = "Genome Position",
    y = "Coloc Group Type",
    caption = paste("Total coloc groups:", nrow(all_coloc_groups))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 10),
    legend.position = "none",
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(
    breaks = chr_centers,
    labels = names(chr_lengths),
    expand = c(0, 0)
  ) +
  scale_color_manual(values = c("With Genes" = "darkblue", 
                               "Without Genes" = "red", 
                               "Only Phenotypes" = "orange", 
                               "Without methQTLs" = "green"))


# ggsave("coloc_groups_manhattan_jittered.png", p1, width = 10, height = 6)
```

# What fraction of known GWAS hits are no known molecular trait associated with it?

```{r gwas_hits_no_molecular_trait, echo=F}
all_phenotype_study_extractions <- DBI::dbGetQuery(studies_conn, "SELECT study_extractions.*,
  snp_annotations.feature_type, snp_annotations.consequence, snp_annotations.impact, snp_annotations.cdna_position,
  snp_annotations.cds_position, snp_annotations.protein_position, snp_annotations.amino_acids, snp_annotations.codons,
  snp_annotations.gene as snp_gene 
  FROM study_extractions
  JOIN studies ON study_extractions.study_id = studies.id
  JOIN traits ON studies.trait_id = traits.id
  JOIN snp_annotations ON study_extractions.snp_id = snp_annotations.id
  WHERE studies.data_type = 'phenotype' and studies.variant_type = 'common'
")

all_coloc_groups <- DBI::dbGetQuery(studies_conn, "SELECT coloc_groups.*, studies.data_type, study_extractions.min_p,
  snp_annotations.feature_type, snp_annotations.consequence, snp_annotations.impact, snp_annotations.cdna_position,
  snp_annotations.cds_position, snp_annotations.protein_position, snp_annotations.amino_acids, snp_annotations.codons,
  snp_annotations.gene as snp_gene
  FROM coloc_groups
  JOIN snp_annotations ON coloc_groups.snp_id = snp_annotations.id
  JOIN study_extractions ON coloc_groups.study_extraction_id = study_extractions.id
  JOIN studies ON coloc_groups.study_id = studies.id
  WHERE coloc_groups.group_threshold = 'strong'
")

ensembl_gene_map <- DBI::dbGetQuery(studies_conn, "SELECT gene as gene_name, ensembl_id FROM gene_annotations")

all_study_extractions_in_coloc_groups <- unique(all_coloc_groups$study_extraction_id)

all_phenotype_study_extractions_without_coloc <- all_phenotype_study_extractions %>%
  filter(!id %in% all_study_extractions_in_coloc_groups)

all_significant_phenotype_study_extractions <- all_phenotype_study_extractions %>%
  filter(min_p < 1e-8)

all_significant_phenotype_study_extractions_without_coloc <- all_phenotype_study_extractions_without_coloc %>%
  filter(min_p < 1e-8)

study_extractions_with_phenotype_only_coloc <- all_coloc_groups %>%
  group_by(coloc_group_id) %>%
  filter(all(data_type == "phenotype")) %>%
  ungroup()

significant_phenotype_study_extractions_with_phenotype_only_coloc <- study_extractions_with_phenotype_only_coloc %>%
  filter(min_p < 1e-8)

nrow(all_phenotype_study_extractions)
nrow(all_significant_phenotype_study_extractions)
nrow(all_significant_phenotype_study_extractions_without_coloc)
nrow(significant_phenotype_study_extractions_with_phenotype_only_coloc)

all_significant_non_molecular_coloc <- dplyr::bind_rows(all_significant_phenotype_study_extractions_without_coloc, significant_phenotype_study_extractions_with_phenotype_only_coloc)

(nrow(significant_phenotype_study_extractions_with_phenotype_only_coloc) + nrow(all_significant_phenotype_study_extractions_without_coloc)) / nrow(all_significant_phenotype_study_extractions)

#TODO: maybe figure this out, but deduplicate on if the study_extraction_id is from the same region, just finemapped

# Deduplicate all_coloc_groups by snp_id and create histograms for VEP annotations
all_coloc_groups_dedup <- all_coloc_groups %>%
  group_by(snp_id) %>%
  slice(1) %>%  # Keep first occurrence of each snp_id
  ungroup()

all_significant_non_molecular_coloc_dedup <- all_significant_non_molecular_coloc %>%
  group_by(snp_id) %>%
  slice(1) %>%  # Keep first occurrence of each snp_id
  ungroup()

cat("Original all_coloc_groups rows:", nrow(all_coloc_groups), "\n")
cat("Deduplicated rows:", nrow(all_coloc_groups_dedup), "\n")

# Create histograms for consequence, impact, and cdna_position


all_consequence_counts <- all_coloc_groups_dedup %>%
  filter(!is.na(consequence)) %>%
  mutate(consequence_split = strsplit(consequence, ",")) %>%
  tidyr::unnest(consequence_split) %>%
  mutate(consequence_split = trimws(consequence_split)) %>%  # Remove whitespace
  count(consequence_split, sort = TRUE)

all_non_molecular_coloc_consequence_counts <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(consequence)) %>%
  mutate(consequence_split = strsplit(consequence, ",")) %>%
  tidyr::unnest(consequence_split) %>%
  mutate(consequence_split = trimws(consequence_split)) %>%  # Remove whitespace
  count(consequence_split, sort = TRUE)



# Calculate percent difference between all_coloc_groups and non_molecular_coloc consequences
consequence_comparison <- all_consequence_counts %>%
  full_join(all_non_molecular_coloc_consequence_counts, 
            by = "consequence_split", 
            suffix = c("_all", "_non_molecular")) %>%
  mutate(
    n_all = replace_na(n_all, 0),
    n_non_molecular = replace_na(n_non_molecular, 0),
    percent_diff = ((n_non_molecular - n_all) / n_all) * 100,
    proportion = (n_non_molecular / n_all) * 100
  ) %>%
  filter(n_all > 0)  # Only include consequences that appear in the main dataset

consequence_hist <- ggplot(consequence_comparison, aes(x = reorder(consequence_split, proportion), y = proportion)) +
  geom_bar(stat = "identity", fill = ifelse(consequence_comparison$proportion> 0, "red", "blue"), alpha = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid", size = 0.5) +
  labs(
    title = "Percent Difference in Consequence Prevalence",
    subtitle = "Non-molecular coloc vs All coloc groups (Positive = more common in non-molecular)",
    x = "VEP Consequence",
    y = "Proportion Difference (Non-molecular / All) * 100",
    caption = paste("Consequences with data in both datasets:", nrow(consequence_comparison))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 10)
  ) +
  coord_flip()  # Flip to make consequence names readable

consequence_hist

# 2. Impact comparison
all_impact_counts <- all_coloc_groups_dedup %>%
  filter(!is.na(impact)) %>%
  count(impact, sort = TRUE)

all_non_molecular_coloc_impact_counts <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(impact)) %>%
  count(impact, sort = TRUE)

impact_comparison <- all_impact_counts %>%
  full_join(all_non_molecular_coloc_impact_counts, 
            by = "impact", 
            suffix = c("_all", "_non_molecular")) %>%
  mutate(
    n_all = replace_na(n_all, 0),
    n_non_molecular = replace_na(n_non_molecular, 0),
    percent_diff = ((n_non_molecular - n_all) / n_all) * 100,
    proportion = (n_non_molecular / n_all) * 100
  ) %>%
  filter(n_all > 0)

impact_hist <- ggplot(impact_comparison, aes(x = reorder(impact, proportion), y = proportion)) +
  geom_bar(stat = "identity", fill = ifelse(impact_comparison$proportion > 0, "red", "blue"), alpha = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid", size = 0.5) +
  labs(
    title = "Proportion Difference in Impact Prevalence",
    subtitle = "Non-molecular coloc vs All coloc groups (Positive = more common in non-molecular)",
    x = "VEP Impact",
    y = "Proportion Difference (Non-molecular / All) * 100",
    caption = paste("Impact levels with data in both datasets:", nrow(impact_comparison))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 10)
  ) +
  coord_flip()
impact_hist
# 3. Feature type comparison
all_feature_type_counts <- all_coloc_groups_dedup %>%
  filter(!is.na(feature_type)) %>%
  count(feature_type, sort = TRUE)

all_non_molecular_coloc_feature_type_counts <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(feature_type)) %>%
  count(feature_type, sort = TRUE)

feature_type_comparison <- all_feature_type_counts %>%
  full_join(all_non_molecular_coloc_feature_type_counts, 
            by = "feature_type", 
            suffix = c("_all", "_non_molecular")) %>%
  mutate(
    n_all = replace_na(n_all, 0),
    n_non_molecular = replace_na(n_non_molecular, 0),
    percent_diff = ((n_non_molecular - n_all) / n_all) * 100,
    proportion = (n_non_molecular / n_all) * 100
  ) %>%
  filter(n_all > 0)

feature_type_hist <- ggplot(feature_type_comparison, aes(x = reorder(feature_type, proportion), y = proportion)) +
  geom_bar(stat = "identity", fill = ifelse(feature_type_comparison$proportion > 0, "red", "blue"), alpha = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid", size = 0.5) +
  labs(
    title = "Proportion Difference in Feature Type Prevalence",
    subtitle = "Non-molecular coloc vs All coloc groups (Positive = more common in non-molecular)",
    x = "Feature Type",
    y = "Proportion Difference (Non-molecular / All) * 100",
    caption = paste("Feature types with data in both datasets:", nrow(feature_type_comparison))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 10)
  ) +
  coord_flip()

feature_type_hist

# 4. cDNA position comparison (histogram since it's continuous)
all_cdna_positions <- all_coloc_groups_dedup %>%
  filter(!is.na(cdna_position)) %>%
  mutate(cdna_position = as.numeric(cdna_position)) %>%
  filter(!is.na(cdna_position))

all_non_molecular_coloc_cdna_positions <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(cdna_position)) %>%
  mutate(cdna_position = as.numeric(cdna_position)) %>%
  filter(!is.na(cdna_position))

cdna_hist <- ggplot() +
  geom_histogram(data = all_cdna_positions, aes(x = cdna_position, y = ..density..), 
                 bins = 30, fill = "blue", alpha = 0.5, color = "black") +
  geom_histogram(data = all_non_molecular_coloc_cdna_positions, aes(x = cdna_position, y = ..density..), 
                 bins = 30, fill = "red", alpha = 0.5, color = "black") +
  labs(
    title = "Distribution of cDNA Positions",
    subtitle = "Blue = All coloc groups, Red = Non-molecular coloc groups",
    x = "cDNA Position",
    y = "Density",
    caption = paste("All coloc SNPs:", nrow(all_cdna_positions), 
                   "Non-molecular SNPs:", nrow(all_non_molecular_coloc_cdna_positions))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  )

cdna_hist

# 5. Protein position comparison (histogram since it's continuous)
all_protein_positions <- all_coloc_groups_dedup %>%
  filter(!is.na(protein_position)) %>%
  mutate(protein_position = as.numeric(protein_position)) %>%
  filter(!is.na(protein_position))

all_non_molecular_coloc_protein_positions <- all_significant_non_molecular_coloc_dedup %>%
  filter(!is.na(protein_position)) %>%
  mutate(protein_position = as.numeric(protein_position)) %>%
  filter(!is.na(protein_position))

protein_hist <- ggplot() +
  geom_histogram(data = all_protein_positions, aes(x = protein_position, y = ..density..), 
                 bins = 30, fill = "blue", alpha = 0.5, color = "black") +
  geom_histogram(data = all_non_molecular_coloc_protein_positions, aes(x = protein_position, y = ..density..), 
                 bins = 30, fill = "red", alpha = 0.5, color = "black") +
  labs(
    title = "Distribution of Protein Positions",
    subtitle = "Blue = All coloc groups, Red = Non-molecular coloc groups",
    x = "Protein Position",
    y = "Density",
    caption = paste("All coloc SNPs:", nrow(all_protein_positions), 
                   "Non-molecular SNPs:", nrow(all_non_molecular_coloc_protein_positions))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  )


all_gene_counts <- all_coloc_groups_dedup %>%
  left_join(ensembl_gene_map, by = c("snp_gene" = "ensembl_id")) %>%
  count(gene_name, sort = TRUE)

all_non_molecular_coloc_gene_counts <- all_significant_non_molecular_coloc_dedup %>%
  left_join(ensembl_gene_map, by = c("snp_gene" = "ensembl_id")) %>%
  count(gene_name, sort = TRUE)

na_gene_count <- all_non_molecular_coloc_gene_counts$n[is.na(all_non_molecular_coloc_gene_counts$gene_name)]

all_non_molecular_coloc_gene_counts <- all_non_molecular_coloc_gene_counts %>%
  filter(!is.na(gene_name))


# Show most common genes in non-molecular coloc groups
gene_hist <- ggplot(all_non_molecular_coloc_gene_counts %>% head(50), aes(x = reorder(gene_name, n), y = n)) +
  geom_bar(stat = "identity", fill = "red", alpha = 0.7) +
  labs(
    title = "Most Common Genes without molecular coloc results (Top 50)",
    subtitle = paste("Number of SNPs without known gene name:", na_gene_count),
    x = "Gene",
    y = "Count",
    caption = paste("Total unique genes in non-molecular coloc:", nrow(all_non_molecular_coloc_gene_counts))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 10)
  ) +
  coord_flip()

# Display the plots
print(consequence_hist)
print(impact_hist)
print(feature_type_hist)
print(cdna_hist)
print(protein_hist)
print(gene_hist)


#Other questions: for the 'credible set' / candidate snp for the non-colocing hits, are there patterns from VEP that we can learn about the type of variants that are not colocing?

```





# Can we calculate a 'pleiotropy score' for each SNP?

### Colocalisation proximity to genes: how can we quantify 'artefactual' pleiotropy?

```{r colocalisation_proximity, echo=F}
```